### 2015-03-16 swmpr_manu.tex
abstract

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

% for acronyms
\usepackage{acronym}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

% acronyms
\acrodef{NERRS}{National Estuarine Research Reserve System}
\acrodef{SWMP}{System Wide Monitoring Program}

%% END MACROS SECTION
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Marcus William Beck\textsuperscript{1, *}

\bigskip
\bf{1} ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
\\
\bigskip

* beck.marcus@epa.gov

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS} .  This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries across the United States.  \ac{SWMP} data have been used in a variety applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the large and increasing quantity of data provided by the monitoring network have prevented large-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  We describe a new open-source software package, SWMPr, developed in program R for use with \ac{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effectives approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.

\linenumbers

\section*{Introduction}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam Eq.~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\begin{equation}\label{eq:schemeP} 
D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
\end{equation}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Etiam eget sapien nibh.}

% For figure citations, please use "Fig." instead of "Figure".
Nulla mi mi, Fig.~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{figure}[h]
\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
\label{fig1}
\end{figure}

\begin{enumerate}
\item{react}
\item{diffuse free particles}
\item{increment time by dt and go to 1}
\end{enumerate}

% Results and Discussion can be combined.
\section*{Results}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}

Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat.

\subsection*{Sed ac quam id nisi malesuada congue.}

Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
\subsection*{Subsection 1}
Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\subsection*{Subsection 2}
\paragraph{3rd Level Heading.} Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\section*{Discussion}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}

CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec, auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Text}.

\section*{Supporting Information}

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{S1 Video}
\label{S1_Video}
{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Text}
\label{S1_Text}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Fig}
\label{S1_Fig}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S2 Fig}
\label{S2_Fig}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Table}
\label{S1_Table}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

\bibliographystyle{plost2015}
\bibliography{M:/docs/refs/refs}

\end{document}

### 2015-03-16 swmpr_manu.Rnw
started copying readme to manuscript template

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

% for acronyms
\usepackage{acronym}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

% acronyms
\acrodef{NERRS}{National Estuarine Research Reserve System}
\acrodef{SWMP}{System Wide Monitoring Program}

%% END MACROS SECTION

%knitr options
<<setup, cache = F, echo = F>>=
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Marcus William Beck\textsuperscript{1, *}

\bigskip
\bf{1} ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
\\
\bigskip

* beck.marcus@epa.gov

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS} .  This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries across the United States.  \ac{SWMP} data have been used in a variety applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the large and increasing quantity of data provided by the monitoring network have prevented large-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  We describe a new open-source software package, SWMPr, developed in program R for use with \ac{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effectives approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.

\linenumbers

\section*{Introduction}

Info about SWMP, etc.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

SWMPr is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).  SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.  SWMP data are maintained online by the Centralized Data Management Office (CDMO). This R package provides several functions to retrieve, organize, and analyze SWMP data from the CDMO.  Information on the CDMO web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.  Data can be downloaded directly from the CDMO using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.  Your computer's IP address must be registered by CDMO staff for direct downloads in R.  Detailed methods for importing SWMP data in R are described below.

All data obtained from the CDMO should be \href{http://cdmo.baruch.sc.edu/data/citation.cfm}{cited} using the format: \\

{\it National Estuarine Research Reserve System (NERRS). 2012. System-wide Monitoring Program. Data accessed from the NOAA NERRS Centralized Data Management Office website: http://cdmo.baruch.sc.edu/; accessed 12 October 2012.}

To cite this package: \\

{\it Beck MW. 2015. SWMPr: An R package for the National Estuarine Research Reserve System.  Version 1.6.0. https://github.com/fawda123/SWMPr}

\subsection*{Installing the package}

This package is currently under development and has not been extensively tested.  The development version can be installed from Github:

<<eval = F, message = F>>=
install.packages('devtools')
library(devtools)
install_github('fawda123/SWMPr')
library(SWMPr)
@
<<eval = T, echo = F, message = F, cache = F>>=
devtools::load_all('M:/docs/SWMPr')
@

Note that the current version of devtools (v1.6.1) was built under R version 3.1.1.  The SWMPr package may not install correctly with older versions of R.

\subsection*{Data retrieval}

SWMP data can be used in R after they are obtained directly from the CDMO through an online query or by using the retrieval functions provided in this package.  In the latter case, the IP address for the computer making the request must be registered with CDMO.  This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.  The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.  Metadata are included as a .csv file with data requested from the CDMO and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.  Again, these functions will only work if the computer's IP address is registered with CDMO.

<<eval = F>>=
# retrieve metadata for all sites
site_codes()

# retrieve metadata for a single site
site_codes_ind('apa')
@

Due to rate limitations on the server, the retrieval functions in this package return a limited number of records.  The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} or \texttt{while} loops) to obtain longer time series.  Data retrieval functions to access the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.  These are functions that call the existing methods on the CDMO web services.  \texttt{all\_params} returns the most recent 100 records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.

<<eval = F>>=
# all parameters for a station, most recent
all_params('hudscwq')

# get all parameters within a date range
all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))

# get single parameter within a date range
all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), param = 'do_mgl')

# single parameter for a station, most recent
single_param('hudscwq', 'do_mgl')
@

For larger requests, it's easier to obtain data outside of R using the CDMO query system and then importing within R using the \texttt{import\_local} function.  Data can be retrieved from the CDMO several ways.  The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO. The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested (expect bugs).  The zip downloads feature is an easy way to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to CDMO.  The \texttt{import\_local} function can be used after the folder is decompressed.

Occasionally, duplicate time stamps are present in the raw data.  The function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).  For water quality and nutrient data, duplicate time stamps are simply removed.  Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.  Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.  Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.

<<eval = F>>=
# import data for apaebmet that you downloaded

# this is an example path with the csv files, change as needed
path <- 'C:/my_path/'

# import, do not include file extension
import_local(path, 'apaebmet') 
@

Raw csv data have not been included in the package due to size limitations.  However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.  However, import time of the raw data may slow down use of the examples below and I have included binary data files (.RData) that are processed versions of the raw data.

% #The swmpr object class
% 
% All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.  The data include a datetimestamp column in the appropriate timezone for a station.  Note that the datetimestamp is standard time for each timezone and does not include daylight savings. Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.  The attributes for a swmpr object include `names` of the dataset, `class` (swmpr) `station name` (7 or 8 characters), `qaqc_cols` (logical), `date_rng` (POSIX vector), `timezone` (text string in country/city format), `stamp_class` (class of datetimestamp vector, POSIX or Date), and `parameters` (character vector).  Attributes of a swmpr object can be viewed as follows:
% 
% ```{r eval = T, cache = T}
% # import binary data
% data(apadbwq)
% dat <- apadbwq
% 
% # verify that dat is swmpr class
% class(dat)
% 
% # all attributes of dat
% names(attributes(dat))
% 
% # a single attribute of dat
% attr(dat, 'station')
% ```
% 
% The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.  A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.  Available methods for the swmpr class are described below and can also be viewed:
% 
% ```{r eval = T, cache = T}
% # available methods for swmpr class
% methods(class = 'swmpr')
% ```
% 
% #An overview of methods for swmpr objects
% 
% Three categories of functions are available: retrieve, organize, and analyze.  The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.  Methods defined for swmpr objects can be applied with the organize and analyze functions.  These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.  S3 methods are implemented in all cases.  
% 
% The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.  The `qaqc` function is a simple screen to retain values from the data with specified QAQC flags, described [here](http://cdmo.baruch.sc.edu/data/qaqc.cfm).  Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix 'f_'.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the '0' QAQC flag should be used, which is the default option for the `qaqc` function.  Data that do not satisfy QAQC criteria are converted to NA values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria. 
% 
% ```{r eval = F, cache = T}
% # qaqc screen for a swmpr object, retain only '0'
% qaqc(dat)
% 
% # retain all data regardless of flag
% qaqc(dat, qaqc_keep = NULL)
% 
% # retain only '0' and '-1' flags
% qaqc(dat, qaqc_keep = c(0, -1))
% 
% ```
% 
% Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with `qaqc`.  The `qaqcchk` function can be used to view this information.  Consult the [online documentation](http://cdmo.baruch.sc.edu/data/qaqc.cfm) for a description of each QAQC flag. 
% 
% ```{r eval = F, cache = T}
% # view the number observations in each QAQC flag
% qaqcchk(dat)
% 
% ```
% 
% Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The `rem_reps.swmpr` function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the `median`, `min`, or `max`.  The entire function call including treatment of NA values should be passed to the `FUN` argument (see the examples).  The function is meant to be used after `qaqc` processing, although it works with a warning if QAQC columns are present.
% 
% ```{r eval = F, cache = T}
% # get nutrient data
% data(apacpnut)
% swmp1 <- apacpnut
% swmp1 <- qaqc(swmp1)
% 
% # remove replicate nutrient data
% rem_reps(swmp1)
% 
% # use different function to aggregate replicates
% func <- function(x) max(x, na.rm = T)
% rem_reps(swmp1, FUN = func)
% ```
% 
% A subset method added to the existing `subset` function is available for swmpr objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator input as a character string passed to the argument, such as `>` or `<`.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., %Y-%m%-%d %H:%M in POSIX standards).  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. `2012-11-31 12:00`.  Finally, the function can be used to remove rows and columns that do not contain data. 
% 
% ```{r eval = F, cache = T}
% # select two parameters from dat
% subset(dat, select = c('rh', 'bp'))
% 
% # subset records greater than or equal to a date
% subset(dat, subset = '2013-01-01 0:00', operator = '>=')
% 
% # subset records within a date range
% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
% 
% # subset records within a date range, select two parameters
% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
%   select = c('atemp', 'totsorad'))
% 
% # remove rows/columns that do not contain data
% subset(dat, rem_rows = T, rem_cols = T)
% 
% ```
% 
% The `setstep` function formats a swmpr object to a continuous time series at a given time step.  This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.  The first argument of the function, `timestep`, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, `differ`, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.  Values for `differ` that are greater than one half the value of `timestep` are not allowed to prevent duplication of existing data.  Likewise, the default value for `differ` is one half the time step.  Rows that do not match any existing data within the limits of the `differ` argument are not discarded.  Output from the `setstep` function can be used with `subset` and to create a time series at a set interval with empty data removed.
% 
% ```{r eval = F, cache = T}
% # convert time series to two hour invervals
% # tolerance of +/- 30 minutes for matching existing data
% setstep(dat, timestep = 120, differ = 30)
% 
% # convert a nutrient time series to a continuous time series
% # then remove empty rows and columns
% data(apacpnut)
% dat_nut <- apacpnut
% dat_nut <- setstep(dat_nut, timestep = 60)
% subset(dat_nut, rem_rows = T, rem_cols = T)
% ```
% 
% The `comb` function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.  The `timestep` function is used internally such that `timestep` and `differ` are accepted arguments for `comb`.  The function requires one or more swmpr objects as input as separate, undefined arguments.  The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.  In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.  The time series for merging depends on the value passed to the `method` argument.  Passing `union` to `method` will create a time series that is continuous starting from the earliest date and the latest date for all input objects.  Passing `intersect` to `method` will create a time series that is continuous from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to `method` will merge all input objects based on a continuous time series for the given station.  The specified station must be present in the input data.  Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.  
% 
% ```{r eval = F, cache = T}
% # get nuts, wq, and met data as separate objects for the same station
% # note that most sites usually have one weather station
% data(apacpnut)
% data(apacpwq)
% data(apaebmet)
% swmp1 <- apacpnut
% swmp2 <- apacpwq
% swmp3 <- apaebmet
% 
% # combine nuts and wq data by union
% comb(swmp1, swmp2, method = 'union')
% 
% # combine nuts and wq data by intersect
% comb(swmp1, swmp3, method = 'intersect')
% 
% # combine nuts, wq, and met data by nuts time series, two hour time step
% comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
% 
% ```
% 
% The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.  The latter category includes a limited number of functions that were developed by myself or others.  The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include swmpr methods for `aggregate`, `filter`, and `approx` to deal with missing or noisy data and more general functions for exploratory data analaysis, such as `plot`, `summary`, and `hist` methods.  Decomposition functions (`decomp` and `decomp_cj`) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.    
% 
% The `aggregate` function aggregates parameter data for a swmpr object by set periods of observation.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.  A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the `round` method for IDate objects ([data.table](http://cran.r-project.org/web/packages/data.table/index.html) package).  This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.  Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).  Finally, the default argument for `na.action` is set to `na.pass` for swmpr objects to preserve the time series of the input data.
% 
% ```{r eval = F, cache = T}
% # combine, qaqc, remove empty columns
% dat <- comb(swmp1, swmp2, method = 'union')
% dat <- qaqc(dat)
% swmpr_in <- subset(dat, rem_cols = T)
% 
% # get mean DO by quarters
% aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
% 
% # get mean DO by quarters, remove NA when calculating means
% fun_in <- function(x) mean(x, na.rm = T)
% aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
% 
% ```
% 
% Time series can be smoothed to better characterize a signal independent of noise.  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The `smoother` function can be used to smooth parameters in a swmpr object using a specified window size.  This method is a simple wrapper to `filter`.  The `window` argument specifies the number of observations included in the moving average.  The `sides` argument specifies how the average is calculated for each observation (see the documentation for `filter`).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the `params` argument specifies which parameters to smooth.
% 
% ```{r eval = T, fig.height = 3, cache = T}
% # import data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # qaqc and subset imported data
% dat <- qaqc(swmp1)
% dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
% 
% # filter
% test <- smoother(dat, window = 50, params = 'do_mgl')
% 
% # plot to see the difference
% plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
% lines(test, select = 'do_mgl', col = 'red', lwd = 2)
% 
% ```
% 
% A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The `na.approx` function can be used to interpolate gaps in a swmpr object.  A required argument for the function is `maxgap` which defines the maximum gap size  for interpolation.
% 
% ```{r eval = T, fig.height = 6, cache = T}
% # get data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # qaqc and subset imported data
% dat <- qaqc(swmp1)
% dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
% 
% # interpolate, maxgap of 10 records
% test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
% 
% # interpolate maxgap of 30 records
% test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
% 
% # plot for comparison
% par(mfrow = c(3, 1))
% plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
% plot(do_mgl ~ datetimestamp, test, col = 'red', 
%   main = 'Interpolation - maximum gap of 10 records', type = 'l')
% lines(dat, select = 'do_mgl')
% plot(do_mgl ~ datetimestamp, test2, col = 'red', 
%   main = 'Interpolation - maximum gap of 30 records', type = 'l')
% lines(dat, select = 'do_mgl')
% ```
% 
% The `decomp` function is a simple wrapper to `decompose` that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.  The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.  By default, a moving average with a symmetric window is used to filter the seasonal component.  Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for `decompose`).  
% 
% The `decompose` function requires a ts object with a specified frequency as input.  The `decomp` function converts the input swmpr vector to a ts object prior to `decompose`.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of `'daily'` or `'seasonal'` can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the `setstep` function is also required to standardize the time step prior to decomposition.  
% 
% Note that the `decompose` function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
% 
% ```{r eval = T, fig.height = 6, cache = T}
% # get data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # subset for daily decomposition
% dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
% 
% # decomposition and plot
% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% plot(test)
% 
% ```
% 
% The next example illustrates how to handle missing values using the `decomp` function. The `decompose` function used internally within `decomp` currently cannot process time series with missing values.  A recommended approach is to use `na.approx` to interpolate the missing values prior to `decompose`.
% 
% ```{r eval = T, fig.height = 6, warning = F, cache = T}
% # get data
% dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
% 
% # this returns an error
% # test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% 
% # how many missing values?
% sum(is.na(dat$do_mgl))
% 
% # use na.approx to interpolate missing data
% dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
% 
% # decomposition and plot
% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% plot(test)
% 
% ```
% 
% An alternative approach to time series decomposition is provided by the `decomp_cj` function, which is a simple wrapper to the `decompTs` function in the wq package.  Theory describing this method is described in Cloern and Jassby (2010).  The function is similar to `decomp.swmpr` with a few key differences.  The `decomp.swmpr` function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.  The `decomp_cj` function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach is meant for nutrient data that are obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.  Accordingly, short-term variation less than one-month will be removed. Additional arguments passed to `decompTs` can be used with `decomp_cj`, such as `startyr`, `endyr`, and `type`.  Values passed to `type` are `mult` (default) or `add`, referring to multiplicative or additive decomposition.  See the documentation for `decompTs` for additional explanation and examples.   
% 
% ```{r eval = T, fig.height = 6, warning = F, cache = T}
% # get data
% data(apacpnut)
% dat <- apacpnut
% dat <- qaqc(dat, qaqc_keep = NULL)
% 
% # decomposition of chl, ggplot
% decomp_cj(dat, param = 'chla_n')
% ```
% 
% A reserve map with all stations can be obtained using the `map_reserve` function.  This function is a simple wrapper to functions in the ggmap package. The current function is limited to Google maps, which allows four map types that can be set with the `map_type` argument: terrain (default), satellite, roadmap, or hybrid.  The `zoom` argument may have to be chosen through trial and error depending on the spatial extent of the reserve.  See the help documentation for ggmap for more info on zoom.  Additionally, station locations are returned using the `site_codes_ind` function if the computer making the request has the IP address registered with CDMO. Otherwise, a local and possibly outdated file is used.  Use the contact at the CDMO [web services](http://cdmo.baruch.sc.edu/webservices.cfm) to register your IP.
% 
% ```{r fig.height = 5, message = F, cache = T}
% # plot the stations at Jacques Cousteau reserve
% map_reserve('jac')
% 
% # plot the stations at Padilla Bay reserve, satellite
% map_reserve('pdb', map_type = 'satellite', zoom = 12)
% ```
% 
% Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the `plot_summary` function.  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.  Monthly anomalies are in relation to the grand mean for the same month across all years.  All data are aggregated for quicker plotting.  Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.  Cumulative precipitation data are based on the daily maximum. The function returns a graphics object (Grob) of multiple ggplot objects.  An interactive Shiny widget that uses this function is available: [https://beckmw.shinyapps.io/swmp_summary/](https://beckmw.shinyapps.io/swmp_summary/).
% 
% ```{r fig.height = 7, fig.width = 13, message = F, cache = T}
% ## import data
% data(apacpnut)
% dat <- qaqc(apacpnut)
% 
% ## plot
% plot_summary(dat, param = 'chla_n')
% ```
% 
% Estimates of ecosystem metabolism provide a useful measure of overall system productivity.  These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface.  The diffusion-corrected DO flux estimates are averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of DO change. DO flux is averaged during night hours for respiration and averaged during day hours for net production. Respiration rates are assumed constant during day and night such that total daily rates are calculated as hourly respiration multiplied by 24. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days.  Respiration is subtracted from daily net production estimates to yield gross production.  
% 
% The `ecometab` function is used to implement an adaptation of the open-water method.  
% Several assumptions must be met for a valid interpretation of the results.  In general, the dissolved oxygen time series is assumed to represent the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Volumetric rates for gross production and total respiration are also based on total depth of the water column, which is assumed to be mixed.  Water column depth is based on mean value for the depth variable across the time series and is floored at 1 meter for very shallow stations.  Additionally, the volumetric reaeration coefficient requires an estimate of the anemometer height of the weather station, which is set as 10 meters by default.  The metadata should be consulted for exact height. Other assumptions may apply and the user should consult the relevant literature (see the references in the help file).  All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000). 
% 
% The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using `plot_metab`.
% 
% ```{r eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE}
% ## import water quality and weather data
% data(apadbwq)
% data(apaebmet)
% 
% ## qaqc, combine
% wq <- qaqc(apadbwq)
% met <- qaqc(apaebmet)
% dat <- comb(wq, met)
% 
% ## estimate metabolism
% res <- ecometab(dat, trace = FALSE)
% plot_metab(res)
% ```
% 
% #Function list
% 
% See help documentation for more details on each function (e.g., `?all_params`).
% 
% ##Retrieve
% 
% `all_params` Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to `exportAllParamsXMLNew` function on web services. 
% 
% `all_params_dtrng` Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to `exportAllParamsDateRangeXMLNew`.
% 
% `import_local` Import files from a local path.  The files must be in a specific format, specifically those returned from the CDMO using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
% 
% `single_param` Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to `exportSingleParamXMLNew` function on web services. 
% 
% ##Organize
% 
% `comb.swmpr` Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.
% 
% `qaqc.swmpr` Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.  
% 
% `qaqcchk.swmpr` View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.  The output is used to inform further processing but is not used explicitly. 
% 
% `rem_reps.swmpr` Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
% 
% `setstep.swmpr` Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in `comb.swmpr` and can also be used with individual stations.
% 
% `subset.swmpr` Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic `subset' function provided in the base package.
% 
% ##Analyze
% 
% `aggregate.swmpr` Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean. 
% 
% `aggregate_metab` Aggregate metabolism data from a swmpr object.  This is primarly used within `plot_metab` but may be useful for simple summaries of raw daily data.
% 
% `ecometab.swmpr` Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
% 
% `decomp.swmpr` Decompose a swmpr time series into trend, seasonal, and residual components.  This is a simple wrapper to `decompose`.  Decomposition of monthly or daily trends is possible.
% 
% `decomp_cj.swmpr` Decompose a swmpr time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to `decompTs` in the wq package.  Only monthly decomposition is possible.
% 
% `hist.swmpr` Plot a histogram for a swmpr object.
% 
% `lines.swmpr` Add lines to an existing swmpr plot.
% 
% `na.approx.swmpr` Linearly interpolate missing data (`NA` values) in a swmpr object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data. 
% 
% `plot.swmpr` Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
% 
% `plot_metab` Plot ecosystem metabolism estimates after running `ecometab` on a swmpr object.  
% 
% `plot_summary` Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
% 
% `smoother.swmpr` Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to `filter`.
% 
% ##Miscellaneous
% 
% `calcKL` Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the `ecometab` function.
% 
% `map_reserve` Create a map of all stations in a reserve using the ggmap package.
% 
% `metab_day` Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the `ecometab` function.
% 
% `param_names` Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).  Includes QAQC columns with 'f_' prefix. Used internally in other functions.
% 
% `parser` Parses html returned from CDMO web services, used internally in retrieval functions.
% 
% `site_codes` Metadata for all stations, wrapper to `exportStationCodesXMLNew` function on web services.
% 
% `site_codes_ind` Metadata for all stations at a single site, wrapper  to `NERRFilterStationCodesXMLNew` function on web services.
% 
% `swmpr` Creates object of swmpr class, used internally in retrieval functions.
% 
% `time_vec` Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.

\nameref{online_widget}

% Results and Discussion can be combined.
\section*{Results}

\section*{Discussion}

\section*{Supporting Information}

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{Online widget}
\label{online_widget}
{\bf Online widget.} Description of online widget.

\section*{Acknowledgments}

Text

\nolinenumbers

\bibliographystyle{plost2015}
\bibliography{M:/docs/refs/refs}

\end{document}

### 2015-03-16 swmpr_manu.tex
started copying readme to manuscript template

@@
 % Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
+\usepackage{nameref}
+\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 %% END MACROS SECTION
+
+%knitr options
+
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \section*{Introduction}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam Eq.~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
-\begin{equation}\label{eq:schemeP} 
-D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
-D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
-\end{equation}
+Info about SWMP, etc.
 
@@
 \section*{Materials and Methods}
-\subsection*{Etiam eget sapien nibh.}
 
-% For figure citations, please use "Fig." instead of "Figure".
-Nulla mi mi, Fig.~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
+SWMPr is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
+SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
+SWMP data are maintained online by the Centralized Data Management Office (CDMO).
+This R package provides several functions to retrieve, organize, and analyze SWMP data from the CDMO.
+Information on the CDMO web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+Data can be downloaded directly from the CDMO using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
+Your computer's IP address must be registered by CDMO staff for direct downloads in R.
+Detailed methods for importing SWMP data in R are described below.
 
-\begin{figure}[h]
-\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
-Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
+All data obtained from the CDMO should be \href{http://cdmo.baruch.sc.edu/data/citation.cfm}{cited} using the format: \\
 
-\begin{enumerate}
-\item{react}
-\item{diffuse free particles}
-\item{increment time by dt and go to 1}
-\end{enumerate}
+{\it National Estuarine Research Reserve System (NERRS).
+2012.
+System-wide Monitoring Program.
+Data accessed from the NOAA NERRS Centralized Data Management Office website: http://cdmo.baruch.sc.edu/; accessed 12 October 2012.}
+
+To cite this package: \\
+
+{\it Beck MW. 2015. SWMPr: An R package for the National Estuarine Research Reserve System.  Version 1.6.0. https://github.com/fawda123/SWMPr}
+
+\subsection*{Installing the package}
+
+This package is currently under development and has not been extensively tested.  The development version can be installed from Github:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
+\hlkwd{library}\hlstd{(devtools)}
+\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
+\hlkwd{library}\hlstd{(SWMPr)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Note that the current version of devtools (v1.6.1) was built under R version 3.1.1.  The SWMPr package may not install correctly with older versions of R.
+
+\subsection*{Data retrieval}
+
+SWMP data can be used in R after they are obtained directly from the CDMO through an online query or by using the retrieval functions provided in this package.
+In the latter case, the IP address for the computer making the request must be registered with CDMO.
+This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.
+Metadata are included as a .csv file with data requested from the CDMO and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
+Again, these functions will only work if the computer's IP address is registered with CDMO.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# retrieve metadata for all sites}
+\hlkwd{site_codes}\hlstd{()}
+
+\hlcom{# retrieve metadata for a single site}
+\hlkwd{site_codes_ind}\hlstd{(}\hlstr{'apa'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Due to rate limitations on the server, the retrieval functions in this package return a limited number of records.
+The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} or \texttt{while} loops) to obtain longer time series.
+Data retrieval functions to access the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
+These are functions that call the existing methods on the CDMO web services.
+\texttt{all\_params} returns the most recent 100 records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# all parameters for a station, most recent}
+\hlkwd{all_params}\hlstd{(}\hlstr{'hudscwq'}\hlstd{)}
+
+\hlcom{# get all parameters within a date range}
+\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{))}
+
+\hlcom{# get single parameter within a date range}
+\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+
+\hlcom{# single parameter for a station, most recent}
+\hlkwd{single_param}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlstr{'do_mgl'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+For larger requests, it's easier to obtain data outside of R using the CDMO query system and then importing within R using the \texttt{import\_local} function.
+Data can be retrieved from the CDMO several ways.
+The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO.
+The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested (expect bugs).
+The zip downloads feature is an easy way to obtain data from multiple stations in one request.
+The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
+It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to CDMO.
+The \texttt{import\_local} function can be used after the folder is decompressed.
+
+Occasionally, duplicate time stamps are present in the raw data.
+The function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
+For water quality and nutrient data, duplicate time stamps are simply removed.
+Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.
+Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
+Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.
+Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# import data for apaebmet that you downloaded}
+
+\hlcom{# this is an example path with the csv files, change as needed}
+\hlstd{path} \hlkwb{<-} \hlstr{'C:/my_path/'}
+
+\hlcom{# import, do not include file extension}
+\hlkwd{import_local}\hlstd{(path,} \hlstr{'apaebmet'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Raw csv data have not been included in the package due to size limitations.
+However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
+This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
+However, import time of the raw data may slow down use of the examples below and I have included binary data files (.RData) that are processed versions of the raw data.
+
+% #The swmpr object class
+% 
+% All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.
+The data include a datetimestamp column in the appropriate timezone for a station.
+Note that the datetimestamp is standard time for each timezone and does not include daylight savings.
+Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.
+The attributes for a swmpr object include `names` of the dataset, `class` (swmpr) `station name` (7 or 8 characters), `qaqc_cols` (logical), `date_rng` (POSIX vector), `timezone` (text string in country/city format), `stamp_class` (class of datetimestamp vector, POSIX or Date), and `parameters` (character vector).
+Attributes of a swmpr object can be viewed as follows:
+% 
+% ```{r eval = T, cache = T}
+% # import binary data
+% data(apadbwq)
+% dat <- apadbwq
+% 
+% # verify that dat is swmpr class
+% class(dat)
+% 
+% # all attributes of dat
+% names(attributes(dat))
+% 
+% # a single attribute of dat
+% attr(dat, 'station')
+% ```
+% 
+% The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.
+A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.
+Available methods for the swmpr class are described below and can also be viewed:
+% 
+% ```{r eval = T, cache = T}
+% # available methods for swmpr class
+% methods(class = 'swmpr')
+% ```
+% 
+% #An overview of methods for swmpr objects
+% 
+% Three categories of functions are available: retrieve, organize, and analyze.
+The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
+Methods defined for swmpr objects can be applied with the organize and analyze functions.
+These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.
+S3 methods are implemented in all cases.
+% 
+% The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
+The `qaqc` function is a simple screen to retain values from the data with specified QAQC flags, described [here](http://cdmo.baruch.sc.edu/data/qaqc.cfm).
+Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix 'f_'.
+Values in the QAQC column specify a flag from -5 to 5.
+Generally, only data with the '0' QAQC flag should be used, which is the default option for the `qaqc` function.
+Data that do not satisfy QAQC criteria are converted to NA values.
+Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
+Erroneous data entered as -99 are also removed.
+Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
+% 
+% ```{r eval = F, cache = T}
+% # qaqc screen for a swmpr object, retain only '0'
+% qaqc(dat)
+% 
+% # retain all data regardless of flag
+% qaqc(dat, qaqc_keep = NULL)
+% 
+% # retain only '0' and '-1' flags
+% qaqc(dat, qaqc_keep = c(0, -1))
+% 
+% ```
+% 
+% Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with `qaqc`.
+The `qaqcchk` function can be used to view this information.
+Consult the [online documentation](http://cdmo.baruch.sc.edu/data/qaqc.cfm) for a description of each QAQC flag.
+% 
+% ```{r eval = F, cache = T}
+% # view the number observations in each QAQC flag
+% qaqcchk(dat)
+% 
+% ```
+% 
+% Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
+The `rem_reps.swmpr` function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
+The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
+Other suggested functions include the `median`, `min`, or `max`.
+The entire function call including treatment of NA values should be passed to the `FUN` argument (see the examples).
+The function is meant to be used after `qaqc` processing, although it works with a warning if QAQC columns are present.
+% 
+% ```{r eval = F, cache = T}
+% # get nutrient data
+% data(apacpnut)
+% swmp1 <- apacpnut
+% swmp1 <- qaqc(swmp1)
+% 
+% # remove replicate nutrient data
+% rem_reps(swmp1)
+% 
+% # use different function to aggregate replicates
+% func <- function(x) max(x, na.rm = T)
+% rem_reps(swmp1, FUN = func)
+% ```
+% 
+% A subset method added to the existing `subset` function is available for swmpr objects.
+This function is used to subset the data by date and/or a selected parameter.
+The date can be a single value or as two dates to select records within the range.
+The former case requires a binary operator input as a character string passed to the argument, such as `>` or `<`.
+The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., %Y-%m%-%d %H:%M in POSIX standards).
+Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
+`2012-11-31 12:00`.
+Finally, the function can be used to remove rows and columns that do not contain data.
+% 
+% ```{r eval = F, cache = T}
+% # select two parameters from dat
+% subset(dat, select = c('rh', 'bp'))
+% 
+% # subset records greater than or equal to a date
+% subset(dat, subset = '2013-01-01 0:00', operator = '>=')
+% 
+% # subset records within a date range
+% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
+% 
+% # subset records within a date range, select two parameters
+% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+%   select = c('atemp', 'totsorad'))
+% 
+% # remove rows/columns that do not contain data
+% subset(dat, rem_rows = T, rem_cols = T)
+% 
+% ```
+% 
+% The `setstep` function formats a swmpr object to a continuous time series at a given time step.
+This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.
+The first argument of the function, `timestep`, specifies the desired time step in minutes starting from the nearest hour of the first observation.
+The second argument, `differ`, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
+Values for `differ` that are greater than one half the value of `timestep` are not allowed to prevent duplication of existing data.
+Likewise, the default value for `differ` is one half the time step.
+Rows that do not match any existing data within the limits of the `differ` argument are not discarded.
+Output from the `setstep` function can be used with `subset` and to create a time series at a set interval with empty data removed.
+% 
+% ```{r eval = F, cache = T}
+% # convert time series to two hour invervals
+% # tolerance of +/- 30 minutes for matching existing data
+% setstep(dat, timestep = 120, differ = 30)
+% 
+% # convert a nutrient time series to a continuous time series
+% # then remove empty rows and columns
+% data(apacpnut)
+% dat_nut <- apacpnut
+% dat_nut <- setstep(dat_nut, timestep = 60)
+% subset(dat_nut, rem_rows = T, rem_cols = T)
+% ```
+% 
+% The `comb` function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.
+The `timestep` function is used internally such that `timestep` and `differ` are accepted arguments for `comb`.
+The function requires one or more swmpr objects as input as separate, undefined arguments.
+The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.
+In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.
+The time series for merging depends on the value passed to the `method` argument.
+Passing `union` to `method` will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
+Passing `intersect` to `method` will create a time series that is continuous from the set of dates that are shared between all input objects.
+Finally, a seven or eight character station name passed to `method` will merge all input objects based on a continuous time series for the given station.
+The specified station must be present in the input data.
+Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.
+% 
+% ```{r eval = F, cache = T}
+% # get nuts, wq, and met data as separate objects for the same station
+% # note that most sites usually have one weather station
+% data(apacpnut)
+% data(apacpwq)
+% data(apaebmet)
+% swmp1 <- apacpnut
+% swmp2 <- apacpwq
+% swmp3 <- apaebmet
+% 
+% # combine nuts and wq data by union
+% comb(swmp1, swmp2, method = 'union')
+% 
+% # combine nuts and wq data by intersect
+% comb(swmp1, swmp3, method = 'intersect')
+% 
+% # combine nuts, wq, and met data by nuts time series, two hour time step
+% comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
+% 
+% ```
+% 
+% The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
+The latter category includes a limited number of functions that were developed by myself or others.
+The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.
+These functions include swmpr methods for `aggregate`, `filter`, and `approx` to deal with missing or noisy data and more general functions for exploratory data analaysis, such as `plot`, `summary`, and `hist` methods.
+Decomposition functions (`decomp` and `decomp_cj`) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.
+% 
+% The `aggregate` function aggregates parameter data for a swmpr object by set periods of observation.
+This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
+Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.
+A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the `round` method for IDate objects ([data.table](http://cran.r-project.org/web/packages/data.table/index.html) package).
+This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.
+Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).
+Finally, the default argument for `na.action` is set to `na.pass` for swmpr objects to preserve the time series of the input data.
+% 
+% ```{r eval = F, cache = T}
+% # combine, qaqc, remove empty columns
+% dat <- comb(swmp1, swmp2, method = 'union')
+% dat <- qaqc(dat)
+% swmpr_in <- subset(dat, rem_cols = T)
+% 
+% # get mean DO by quarters
+% aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
+% 
+% # get mean DO by quarters, remove NA when calculating means
+% fun_in <- function(x) mean(x, na.rm = T)
+% aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
+% 
+% ```
+% 
+% Time series can be smoothed to better characterize a signal independent of noise.
+Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
+The `smoother` function can be used to smooth parameters in a swmpr object using a specified window size.
+This method is a simple wrapper to `filter`.
+The `window` argument specifies the number of observations included in the moving average.
+The `sides` argument specifies how the average is calculated for each observation (see the documentation for `filter`).
+A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation.
+As before, the `params` argument specifies which parameters to smooth.
+% 
+% ```{r eval = T, fig.height = 3, cache = T}
+% # import data
+% data(apadbwq)
+% swmp1 <- apadbwq
+% 
+% # qaqc and subset imported data
+% dat <- qaqc(swmp1)
+% dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
+% 
+% # filter
+% test <- smoother(dat, window = 50, params = 'do_mgl')
+% 
+% # plot to see the difference
+% plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
+% lines(test, select = 'do_mgl', col = 'red', lwd = 2)
+% 
+% ```
+% 
+% A common issue with any statistical analysis is the treatment of missing values.
+Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
+In either case, an analyst should have a strong rationale for the chosen method.
+A common approach used to handle missing data in time series analysis is linear interpolation.
+A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
+A challenge with linear interpolation is an appropriate gap size for fitting missing observations.
+The ability of the interpolated data to approximate actual trends is a function of the gap size.
+Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.
+An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
+The `na.approx` function can be used to interpolate gaps in a swmpr object.
+A required argument for the function is `maxgap` which defines the maximum gap size  for interpolation.
+% 
+% ```{r eval = T, fig.height = 6, cache = T}
+% # get data
+% data(apadbwq)
+% swmp1 <- apadbwq
+% 
+% # qaqc and subset imported data
+% dat <- qaqc(swmp1)
+% dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
+% 
+% # interpolate, maxgap of 10 records
+% test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+% 
+% # interpolate maxgap of 30 records
+% test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
+% 
+% # plot for comparison
+% par(mfrow = c(3, 1))
+% plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
+% plot(do_mgl ~ datetimestamp, test, col = 'red', 
+%   main = 'Interpolation - maximum gap of 10 records', type = 'l')
+% lines(dat, select = 'do_mgl')
+% plot(do_mgl ~ datetimestamp, test2, col = 'red', 
+%   main = 'Interpolation - maximum gap of 30 records', type = 'l')
+% lines(dat, select = 'do_mgl')
+% ```
+% 
+% The `decomp` function is a simple wrapper to `decompose` that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
+The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.
+By default, a moving average with a symmetric window is used to filter the seasonal component.
+Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for `decompose`).
+% 
+% The `decompose` function requires a ts object with a specified frequency as input.
+The `decomp` function converts the input swmpr vector to a ts object prior to `decompose`.
+This requires an explicit input defining the frequency of the parameter in the time series.
+For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).
+The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
+For simplicity, character strings of `'daily'` or `'seasonal'` can be supplied in place of numeric values.
+A starting value of the time series must be supplied in the latter case.
+Use of the `setstep` function is also required to standardize the time step prior to decomposition.
+% 
+% Note that the `decompose` function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
+% 
+% ```{r eval = T, fig.height = 6, cache = T}
+% # get data
+% data(apadbwq)
+% swmp1 <- apadbwq
+% 
+% # subset for daily decomposition
+% dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
+% 
+% # decomposition and plot
+% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+% plot(test)
+% 
+% ```
+% 
+% The next example illustrates how to handle missing values using the `decomp` function.
+The `decompose` function used internally within `decomp` currently cannot process time series with missing values.
+A recommended approach is to use `na.approx` to interpolate the missing values prior to `decompose`.
+% 
+% ```{r eval = T, fig.height = 6, warning = F, cache = T}
+% # get data
+% dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
+% 
+% # this returns an error
+% # test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+% 
+% # how many missing values?
+% sum(is.na(dat$do_mgl))
+% 
+% # use na.approx to interpolate missing data
+% dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+% 
+% # decomposition and plot
+% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+% plot(test)
+% 
+% ```
+% 
+% An alternative approach to time series decomposition is provided by the `decomp_cj` function, which is a simple wrapper to the `decompTs` function in the wq package.
+Theory describing this method is described in Cloern and Jassby (2010).
+The function is similar to `decomp.swmpr` with a few key differences.
+The `decomp.swmpr` function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
+For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.
+The `decomp_cj` function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+This approach is meant for nutrient data that are obtained on a monthly cycle.
+The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.
+Accordingly, short-term variation less than one-month will be removed.
+Additional arguments passed to `decompTs` can be used with `decomp_cj`, such as `startyr`, `endyr`, and `type`.
+Values passed to `type` are `mult` (default) or `add`, referring to multiplicative or additive decomposition.
+See the documentation for `decompTs` for additional explanation and examples.
+% 
+% ```{r eval = T, fig.height = 6, warning = F, cache = T}
+% # get data
+% data(apacpnut)
+% dat <- apacpnut
+% dat <- qaqc(dat, qaqc_keep = NULL)
+% 
+% # decomposition of chl, ggplot
+% decomp_cj(dat, param = 'chla_n')
+% ```
+% 
+% A reserve map with all stations can be obtained using the `map_reserve` function.
+This function is a simple wrapper to functions in the ggmap package.
+The current function is limited to Google maps, which allows four map types that can be set with the `map_type` argument: terrain (default), satellite, roadmap, or hybrid.
+The `zoom` argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for ggmap for more info on zoom.
+Additionally, station locations are returned using the `site_codes_ind` function if the computer making the request has the IP address registered with CDMO.
+Otherwise, a local and possibly outdated file is used.
+Use the contact at the CDMO [web services](http://cdmo.baruch.sc.edu/webservices.cfm) to register your IP.
+% 
+% ```{r fig.height = 5, message = F, cache = T}
+% # plot the stations at Jacques Cousteau reserve
+% map_reserve('jac')
+% 
+% # plot the stations at Padilla Bay reserve, satellite
+% map_reserve('pdb', map_type = 'satellite', zoom = 12)
+% ```
+% 
+% Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the `plot_summary` function.
+The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
+Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.
+Monthly anomalies are in relation to the grand mean for the same month across all years.
+All data are aggregated for quicker plotting.
+Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.
+Cumulative precipitation data are based on the daily maximum.
+The function returns a graphics object (Grob) of multiple ggplot objects.
+An interactive Shiny widget that uses this function is available: [https://beckmw.shinyapps.io/swmp_summary/](https://beckmw.shinyapps.io/swmp_summary/).
+% 
+% ```{r fig.height = 7, fig.width = 13, message = F, cache = T}
+% ## import data
+% data(apacpnut)
+% dat <- qaqc(apacpnut)
+% 
+% ## plot
+% plot_summary(dat, param = 'chla_n')
+% ```
+% 
+% Estimates of ecosystem metabolism provide a useful measure of overall system productivity.
+These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
+The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface.
+The diffusion-corrected DO flux estimates are averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of DO change.
+DO flux is averaged during night hours for respiration and averaged during day hours for net production.
+Respiration rates are assumed constant during day and night such that total daily rates are calculated as hourly respiration multiplied by 24.
+The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days.
+Respiration is subtracted from daily net production estimates to yield gross production.
+% 
+% The `ecometab` function is used to implement an adaptation of the open-water method.  
+% Several assumptions must be met for a valid interpretation of the results.
+In general, the dissolved oxygen time series is assumed to represent the same water mass over time.
+Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.
+The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
+Volumetric rates for gross production and total respiration are also based on total depth of the water column, which is assumed to be mixed.
+Water column depth is based on mean value for the depth variable across the time series and is floored at 1 meter for very shallow stations.
+Additionally, the volumetric reaeration coefficient requires an estimate of the anemometer height of the weather station, which is set as 10 meters by default.
+The metadata should be consulted for exact height.
+Other assumptions may apply and the user should consult the relevant literature (see the references in the help file).
+All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000).
+% 
+% The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using `plot_metab`.
+% 
+% ```{r eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE}
+% ## import water quality and weather data
+% data(apadbwq)
+% data(apaebmet)
+% 
+% ## qaqc, combine
+% wq <- qaqc(apadbwq)
+% met <- qaqc(apaebmet)
+% dat <- comb(wq, met)
+% 
+% ## estimate metabolism
+% res <- ecometab(dat, trace = FALSE)
+% plot_metab(res)
+% ```
+% 
+% #Function list
+% 
+% See help documentation for more details on each function (e.g., `?all_params`).
+% 
+% ##Retrieve
+% 
+% `all_params` Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to `exportAllParamsXMLNew` function on web services. 
+% 
+% `all_params_dtrng` Retrieve records of all parameters within a given date range for a station.
+Optional argument for a single parameter.
+Maximum of 1000 records.
+Wrapper to `exportAllParamsDateRangeXMLNew`.
+% 
+% `import_local` Import files from a local path.
+The files must be in a specific format, specifically those returned from the CDMO using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
+% 
+% `single_param` Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to `exportSingleParamXMLNew` function on web services. 
+% 
+% ##Organize
+% 
+% `comb.swmpr` Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+Only different data types can be combined.
+% 
+% `qaqc.swmpr` Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.  
+% 
+% `qaqcchk.swmpr` View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.
+The output is used to inform further processing but is not used explicitly.
+% 
+% `rem_reps.swmpr` Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
+% 
+% `setstep.swmpr` Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in `comb.swmpr` and can also be used with individual stations.
+% 
+% `subset.swmpr` Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic `subset' function provided in the base package.
+% 
+% ##Analyze
+% 
+% `aggregate.swmpr` Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean. 
+% 
+% `aggregate_metab` Aggregate metabolism data from a swmpr object.  This is primarly used within `plot_metab` but may be useful for simple summaries of raw daily data.
+% 
+% `ecometab.swmpr` Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
+% 
+% `decomp.swmpr` Decompose a swmpr time series into trend, seasonal, and residual components.  This is a simple wrapper to `decompose`.  Decomposition of monthly or daily trends is possible.
+% 
+% `decomp_cj.swmpr` Decompose a swmpr time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to `decompTs` in the wq package.
+Only monthly decomposition is possible.
+% 
+% `hist.swmpr` Plot a histogram for a swmpr object.
+% 
+% `lines.swmpr` Add lines to an existing swmpr plot.
+% 
+% `na.approx.swmpr` Linearly interpolate missing data (`NA` values) in a swmpr object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data. 
+% 
+% `plot.swmpr` Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
+% 
+% `plot_metab` Plot ecosystem metabolism estimates after running `ecometab` on a swmpr object.  
+% 
+% `plot_summary` Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
+% 
+% `smoother.swmpr` Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to `filter`.
+% 
+% ##Miscellaneous
+% 
+% `calcKL` Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the `ecometab` function.
+% 
+% `map_reserve` Create a map of all stations in a reserve using the ggmap package.
+% 
+% `metab_day` Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the `ecometab` function.
+% 
+% `param_names` Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).  Includes QAQC columns with 'f_' prefix. Used internally in other functions.
+% 
+% `parser` Parses html returned from CDMO web services, used internally in retrieval functions.
+% 
+% `site_codes` Metadata for all stations, wrapper to `exportStationCodesXMLNew` function on web services.
+% 
+% `site_codes_ind` Metadata for all stations at a single site, wrapper  to `NERRFilterStationCodesXMLNew` function on web services.
+% 
+% `swmpr` Creates object of swmpr class, used internally in retrieval functions.
+% 
+% `time_vec` Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
+
+\nameref{online_widget}
 
@@
 \section*{Results}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\caption{
-{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-\begin{tabular}{|l|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-\end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{table}
-
-\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
-
-Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-
-\subsection*{Sed ac quam id nisi malesuada congue.}
-
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
-\subsection*{Subsection 1}
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\subsection*{Subsection 2}
-\paragraph{3rd Level Heading.} Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
 
 \section*{Discussion}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
-
-CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
-Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-Nam condimentum sem eget mollis euismod.
-Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
-Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
-Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
-For more information, see \nameref{S1_Text}.
 
@@
 % Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
-\subsection*{S1 Video}
-\label{S1_Video}
-{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S1 Text}
-\label{S1_Text}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S1 Fig}
-\label{S1_Fig}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S2 Fig}
-\label{S2_Fig}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S1 Table}
-\label{S1_Table}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
+\subsection*{Online widget}
+\label{online_widget}
+{\bf Online widget.} Description of online widget.
 
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+
+Text
 

### 2015-03-18 swmpr_manu.Rnw
finished formatting markdown as latex

@@
 <<setup, cache = F, echo = F>>=
+library(knitr)
 # set global chunk options
 opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
-options(replace.assign=TRUE,width=90,digits=1)
+options(replace.assign=TRUE,width=70,digits=1)
 @
@@
 
-% #The swmpr object class
-% 
-% All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.
+\subsection*{The swmpr object class}
+
+All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.
 The data include a datetimestamp column in the appropriate timezone for a station.
@@
 Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.
-The attributes for a swmpr object include `names` of the dataset, `class` (swmpr) `station name` (7 or 8 characters), `qaqc_cols` (logical), `date_rng` (POSIX vector), `timezone` (text string in country/city format), `stamp_class` (class of datetimestamp vector, POSIX or Date), and `parameters` (character vector).
+The attributes for a swmpr object include \texttt{names} of the dataset, \texttt{class} (swmpr) \texttt{station name} (7 or 8 characters), \texttt{qaqc\_cols} (logical), \texttt{date\_rng} (POSIX vector), \texttt{timezone} (text string in country/city format), \texttt{stamp\_class} (class of datetimestamp vector, POSIX or Date), and ), \texttt{parameters} (character vector).
 Attributes of a swmpr object can be viewed as follows:
-% 
-% ```{r eval = T, cache = T}
-% # import binary data
-% data(apadbwq)
-% dat <- apadbwq
-% 
-% # verify that dat is swmpr class
-% class(dat)
-% 
-% # all attributes of dat
-% names(attributes(dat))
-% 
-% # a single attribute of dat
-% attr(dat, 'station')
-% ```
-% 
-% The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.
+
+<<eval = T, cache = T>>=
+# import binary data
+data(apadbwq)
+dat <- apadbwq
+
+# verify that dat is swmpr class
+class(dat)
+
+# all attributes of dat
+names(attributes(dat))
+
+# a single attribute of dat
+attr(dat, 'station')
+@
+
+The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.
 A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.
 Available methods for the swmpr class are described below and can also be viewed:
-% 
-% ```{r eval = T, cache = T}
-% # available methods for swmpr class
-% methods(class = 'swmpr')
-% ```
-% 
-% #An overview of methods for swmpr objects
-% 
-% Three categories of functions are available: retrieve, organize, and analyze.
+
+<<eval = T, cache = T>>=
+# available methods for swmpr class
+methods(class = 'swmpr')
+@
+
+\subsection*{An overview of methods for swmpr objects}
+
+Three categories of functions are available: retrieve, organize, and analyze.
 The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
@@
 S3 methods are implemented in all cases.
-% 
-% The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
-The `qaqc` function is a simple screen to retain values from the data with specified QAQC flags, described [here](http://cdmo.baruch.sc.edu/data/qaqc.cfm).
-Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix 'f_'.
+
+The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
+The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.
+Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
 Values in the QAQC column specify a flag from -5 to 5.
-Generally, only data with the '0' QAQC flag should be used, which is the default option for the `qaqc` function.
+Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.
 Data that do not satisfy QAQC criteria are converted to NA values.
@@
 Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
-% 
-% ```{r eval = F, cache = T}
-% # qaqc screen for a swmpr object, retain only '0'
-% qaqc(dat)
-% 
-% # retain all data regardless of flag
-% qaqc(dat, qaqc_keep = NULL)
-% 
-% # retain only '0' and '-1' flags
-% qaqc(dat, qaqc_keep = c(0, -1))
-% 
-% ```
-% 
-% Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with `qaqc`.
-The `qaqcchk` function can be used to view this information.
-Consult the [online documentation](http://cdmo.baruch.sc.edu/data/qaqc.cfm) for a description of each QAQC flag.
-% 
-% ```{r eval = F, cache = T}
-% # view the number observations in each QAQC flag
-% qaqcchk(dat)
-% 
-% ```
-% 
-% Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
-The `rem_reps.swmpr` function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
+
+<<eval = F, cache = T>>=
+# qaqc screen for a swmpr object, retain only '0'
+qaqc(dat)
+
+# retain all data regardless of flag
+qaqc(dat, qaqc_keep = NULL)
+
+# retain only '0' and '-1' flags
+qaqc(dat, qaqc_keep = c(0, -1))
+@
+
+Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with ), \texttt{qaqc}.
+The \texttt{qaqcchk} function can be used to view this information.
+Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.
+
+<<eval = F, cache = T>>=
+# view the number observations in each QAQC flag
+qaqcchk(dat)
+@
+
+Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
+The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
 The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
-Other suggested functions include the `median`, `min`, or `max`.
-The entire function call including treatment of NA values should be passed to the `FUN` argument (see the examples).
-The function is meant to be used after `qaqc` processing, although it works with a warning if QAQC columns are present.
-% 
-% ```{r eval = F, cache = T}
-% # get nutrient data
-% data(apacpnut)
-% swmp1 <- apacpnut
-% swmp1 <- qaqc(swmp1)
-% 
-% # remove replicate nutrient data
-% rem_reps(swmp1)
-% 
-% # use different function to aggregate replicates
-% func <- function(x) max(x, na.rm = T)
-% rem_reps(swmp1, FUN = func)
-% ```
-% 
-% A subset method added to the existing `subset` function is available for swmpr objects.
+Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
+The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
+
+<<eval = F, cache = T>>=
+# get nutrient data
+data(apacpnut)
+swmp1 <- apacpnut
+swmp1 <- qaqc(swmp1)
+
+# remove replicate nutrient data
+rem_reps(swmp1)
+
+# use different function to aggregate replicates
+func <- function(x) max(x, na.rm = T)
+rem_reps(swmp1, FUN = func)
+@
+
+A subset method added to the existing \texttt{subset} function is available for swmpr objects.
 This function is used to subset the data by date and/or a selected parameter.
 The date can be a single value or as two dates to select records within the range.
-The former case requires a binary operator input as a character string passed to the argument, such as `>` or `<`.
-The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., %Y-%m%-%d %H:%M in POSIX standards).
+The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.
+The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).
 Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
-`2012-11-31 12:00`.
+\texttt{2012-11-31 12:00}.
 Finally, the function can be used to remove rows and columns that do not contain data.
-% 
-% ```{r eval = F, cache = T}
-% # select two parameters from dat
-% subset(dat, select = c('rh', 'bp'))
-% 
-% # subset records greater than or equal to a date
-% subset(dat, subset = '2013-01-01 0:00', operator = '>=')
-% 
-% # subset records within a date range
-% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
-% 
-% # subset records within a date range, select two parameters
-% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
-%   select = c('atemp', 'totsorad'))
-% 
-% # remove rows/columns that do not contain data
-% subset(dat, rem_rows = T, rem_cols = T)
-% 
-% ```
-% 
-% The `setstep` function formats a swmpr object to a continuous time series at a given time step.
+
+<<eval = F, cache = T>>=
+# select two parameters from dat
+subset(dat, select = c('rh', 'bp'))
+
+# subset records greater than or equal to a date
+subset(dat, subset = '2013-01-01 0:00', operator = '>=')
+
+# subset records within a date range
+subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
+
+# subset records within a date range, select two parameters
+subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+  select = c('atemp', 'totsorad'))
+
+# remove rows/columns that do not contain data
+subset(dat, rem_rows = T, rem_cols = T)
+@
+
+The \texttt{setstep} function formats a swmpr object to a continuous time series at a given time step.
 This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.
-The first argument of the function, `timestep`, specifies the desired time step in minutes starting from the nearest hour of the first observation.
-The second argument, `differ`, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
-Values for `differ` that are greater than one half the value of `timestep` are not allowed to prevent duplication of existing data.
-Likewise, the default value for `differ` is one half the time step.
-Rows that do not match any existing data within the limits of the `differ` argument are not discarded.
-Output from the `setstep` function can be used with `subset` and to create a time series at a set interval with empty data removed.
-% 
-% ```{r eval = F, cache = T}
-% # convert time series to two hour invervals
-% # tolerance of +/- 30 minutes for matching existing data
-% setstep(dat, timestep = 120, differ = 30)
-% 
-% # convert a nutrient time series to a continuous time series
-% # then remove empty rows and columns
-% data(apacpnut)
-% dat_nut <- apacpnut
-% dat_nut <- setstep(dat_nut, timestep = 60)
-% subset(dat_nut, rem_rows = T, rem_cols = T)
-% ```
-% 
-% The `comb` function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.
-The `timestep` function is used internally such that `timestep` and `differ` are accepted arguments for `comb`.
+The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
+The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
+Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.
+Likewise, the default value for \texttt{differ} is one half the time step.
+Rows that do not match any existing data within the limits of the \texttt{differ} argument are not discarded.
+Output from the \texttt{setstep} function can be used with \texttt{subset} and to create a time series at a set interval with empty data removed.
+
+<<eval = F, cache = T>>=
+# convert time series to two hour invervals
+# tolerance of +/- 30 minutes for matching existing data
+setstep(dat, timestep = 120, differ = 30)
+
+# convert a nutrient time series to a continuous time series
+# then remove empty rows and columns
+data(apacpnut)
+dat_nut <- apacpnut
+dat_nut <- setstep(dat_nut, timestep = 60)
+subset(dat_nut, rem_rows = T, rem_cols = T)
+@
+
+The \texttt{comb} function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.
+The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
 The function requires one or more swmpr objects as input as separate, undefined arguments.
@@
 In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.
-The time series for merging depends on the value passed to the `method` argument.
-Passing `union` to `method` will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
-Passing `intersect` to `method` will create a time series that is continuous from the set of dates that are shared between all input objects.
-Finally, a seven or eight character station name passed to `method` will merge all input objects based on a continuous time series for the given station.
+The time series for merging depends on the value passed to the \texttt{method} argument.
+Passing \texttt{union} to \texttt{method} will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
+Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
+Finally, a seven or eight character station name passed to \texttt{method} will merge all input objects based on a continuous time series for the given station.
 The specified station must be present in the input data.
 Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.
-% 
-% ```{r eval = F, cache = T}
-% # get nuts, wq, and met data as separate objects for the same station
-% # note that most sites usually have one weather station
-% data(apacpnut)
-% data(apacpwq)
-% data(apaebmet)
-% swmp1 <- apacpnut
-% swmp2 <- apacpwq
-% swmp3 <- apaebmet
-% 
-% # combine nuts and wq data by union
-% comb(swmp1, swmp2, method = 'union')
-% 
-% # combine nuts and wq data by intersect
-% comb(swmp1, swmp3, method = 'intersect')
-% 
-% # combine nuts, wq, and met data by nuts time series, two hour time step
-% comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
-% 
-% ```
-% 
-% The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
+
+<<eval = F, cache = T>>=
+# get nuts, wq, and met data as separate objects for the same station
+# note that most sites usually have one weather station
+data(apacpnut)
+data(apacpwq)
+data(apaebmet)
+swmp1 <- apacpnut
+swmp2 <- apacpwq
+swmp3 <- apaebmet
+
+# combine nuts and wq data by union
+comb(swmp1, swmp2, method = 'union')
+
+# combine nuts and wq data by intersect
+comb(swmp1, swmp3, method = 'intersect')
+
+# combine nuts, wq, and met data by nuts time series, two hour time step
+comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
+@
+
+The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
 The latter category includes a limited number of functions that were developed by myself or others.
 The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include swmpr methods for `aggregate`, `filter`, and `approx` to deal with missing or noisy data and more general functions for exploratory data analaysis, such as `plot`, `summary`, and `hist` methods.
-Decomposition functions (`decomp` and `decomp_cj`) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+These functions include swmpr methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
 The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.
-% 
-% The `aggregate` function aggregates parameter data for a swmpr object by set periods of observation.
+
+The \texttt{aggregate} function aggregates parameter data for a swmpr object by set periods of observation.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
@@
 A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the `round` method for IDate objects ([data.table](http://cran.r-project.org/web/packages/data.table/index.html) package).
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.
 This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.
 Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).
-Finally, the default argument for `na.action` is set to `na.pass` for swmpr objects to preserve the time series of the input data.
-% 
-% ```{r eval = F, cache = T}
-% # combine, qaqc, remove empty columns
-% dat <- comb(swmp1, swmp2, method = 'union')
-% dat <- qaqc(dat)
-% swmpr_in <- subset(dat, rem_cols = T)
-% 
-% # get mean DO by quarters
-% aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
-% 
-% # get mean DO by quarters, remove NA when calculating means
-% fun_in <- function(x) mean(x, na.rm = T)
-% aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
-% 
-% ```
-% 
-% Time series can be smoothed to better characterize a signal independent of noise.
+Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for swmpr objects to preserve the time series of the input data.
+
+<<eval = F, cache = T>>=
+# combine, qaqc, remove empty columns
+dat <- comb(swmp1, swmp2, method = 'union')
+dat <- qaqc(dat)
+swmpr_in <- subset(dat, rem_cols = T)
+
+# get mean DO by quarters
+aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
+
+# get mean DO by quarters, remove NA when calculating means
+fun_in <- function(x) mean(x, na.rm = T)
+aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
+@
+
+Time series can be smoothed to better characterize a signal independent of noise.
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The `smoother` function can be used to smooth parameters in a swmpr object using a specified window size.
-This method is a simple wrapper to `filter`.
-The `window` argument specifies the number of observations included in the moving average.
-The `sides` argument specifies how the average is calculated for each observation (see the documentation for `filter`).
+The \texttt{smoother} function can be used to smooth parameters in a swmpr object using a specified window size.
+This method is a simple wrapper to \texttt{filter}.
+The \texttt{window} argument specifies the number of observations included in the moving average.
+The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).
 A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation.
-As before, the `params` argument specifies which parameters to smooth.
-% 
-% ```{r eval = T, fig.height = 3, cache = T}
-% # import data
-% data(apadbwq)
-% swmp1 <- apadbwq
-% 
-% # qaqc and subset imported data
-% dat <- qaqc(swmp1)
-% dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
-% 
-% # filter
-% test <- smoother(dat, window = 50, params = 'do_mgl')
-% 
-% # plot to see the difference
-% plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
-% lines(test, select = 'do_mgl', col = 'red', lwd = 2)
-% 
-% ```
-% 
-% A common issue with any statistical analysis is the treatment of missing values.
+As before, the \texttt{params} argument specifies which parameters to smooth.
+
+<<eval = T, fig.height = 3, cache = T>>=
+# import data
+data(apadbwq)
+swmp1 <- apadbwq
+
+# qaqc and subset imported data
+dat <- qaqc(swmp1)
+dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
+
+# filter
+test <- smoother(dat, window = 50, params = 'do_mgl')
+
+# plot to see the difference
+plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
+lines(test, select = 'do_mgl', col = 'red', lwd = 2)
+@
+
+A common issue with any statistical analysis is the treatment of missing values.
 Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
@@
 An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
-The `na.approx` function can be used to interpolate gaps in a swmpr object.
-A required argument for the function is `maxgap` which defines the maximum gap size  for interpolation.
-% 
-% ```{r eval = T, fig.height = 6, cache = T}
-% # get data
-% data(apadbwq)
-% swmp1 <- apadbwq
-% 
-% # qaqc and subset imported data
-% dat <- qaqc(swmp1)
-% dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
-% 
-% # interpolate, maxgap of 10 records
-% test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-% 
-% # interpolate maxgap of 30 records
-% test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
-% 
-% # plot for comparison
-% par(mfrow = c(3, 1))
-% plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
-% plot(do_mgl ~ datetimestamp, test, col = 'red', 
-%   main = 'Interpolation - maximum gap of 10 records', type = 'l')
-% lines(dat, select = 'do_mgl')
-% plot(do_mgl ~ datetimestamp, test2, col = 'red', 
-%   main = 'Interpolation - maximum gap of 30 records', type = 'l')
-% lines(dat, select = 'do_mgl')
-% ```
-% 
-% The `decomp` function is a simple wrapper to `decompose` that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
+The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.
+A required argument for the function is \texttt{maxgap} which defines the maximum gap size  for interpolation.
+
+<<eval = T, fig.height = 6, cache = T>>=
+# get data
+data(apadbwq)
+swmp1 <- apadbwq
+
+# qaqc and subset imported data
+dat <- qaqc(swmp1)
+dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
+
+# interpolate, maxgap of 10 records
+test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+
+# interpolate maxgap of 30 records
+test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
+
+# plot for comparison
+par(mfrow = c(3, 1))
+plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
+plot(do_mgl ~ datetimestamp, test, col = 'red', 
+  main = 'Interpolation - maximum gap of 10 records', type = 'l')
+lines(dat, select = 'do_mgl')
+plot(do_mgl ~ datetimestamp, test2, col = 'red', 
+  main = 'Interpolation - maximum gap of 30 records', type = 'l')
+lines(dat, select = 'do_mgl')
+@
+
+The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
 The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.
 By default, a moving average with a symmetric window is used to filter the seasonal component.
-Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for `decompose`).
-% 
-% The `decompose` function requires a ts object with a specified frequency as input.
-The `decomp` function converts the input swmpr vector to a ts object prior to `decompose`.
+Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).
+
+The \texttt{decompose} function requires a ts object with a specified frequency as input.
+The \texttt{decomp} function converts the input swmpr vector to a ts object prior to \texttt{decompose}.
 This requires an explicit input defining the frequency of the parameter in the time series.
@@
 The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
-For simplicity, character strings of `'daily'` or `'seasonal'` can be supplied in place of numeric values.
+For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.
 A starting value of the time series must be supplied in the latter case.
-Use of the `setstep` function is also required to standardize the time step prior to decomposition.
-% 
-% Note that the `decompose` function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
-% 
-% ```{r eval = T, fig.height = 6, cache = T}
-% # get data
-% data(apadbwq)
-% swmp1 <- apadbwq
-% 
-% # subset for daily decomposition
-% dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
-% 
-% # decomposition and plot
-% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-% plot(test)
-% 
-% ```
-% 
-% The next example illustrates how to handle missing values using the `decomp` function.
-The `decompose` function used internally within `decomp` currently cannot process time series with missing values.
-A recommended approach is to use `na.approx` to interpolate the missing values prior to `decompose`.
-% 
-% ```{r eval = T, fig.height = 6, warning = F, cache = T}
-% # get data
-% dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
-% 
-% # this returns an error
-% # test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-% 
-% # how many missing values?
-% sum(is.na(dat$do_mgl))
-% 
-% # use na.approx to interpolate missing data
-% dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-% 
-% # decomposition and plot
-% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-% plot(test)
-% 
-% ```
-% 
-% An alternative approach to time series decomposition is provided by the `decomp_cj` function, which is a simple wrapper to the `decompTs` function in the wq package.
+Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+
+Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
+
+<<eval = T, fig.height = 6, cache = T>>=
+# get data
+data(apadbwq)
+swmp1 <- apadbwq
+
+# subset for daily decomposition
+dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
+
+# decomposition and plot
+test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+plot(test)
+@
+
+The next example illustrates how to handle missing values using the \texttt{decomp} function.
+The \texttt{decompose} function used internally within \texttt{decomp} currently cannot process time series with missing values.
+A recommended approach is to use \texttt{na.approx} to interpolate the missing values prior to \texttt{decompose}.
+
+<<eval = T, fig.height = 6, warning = F, cache = T>>=
+# get data
+dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
+
+# this returns an error
+# test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+
+# how many missing values?
+sum(is.na(dat$do_mgl))
+
+# use na.approx to interpolate missing data
+dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+
+# decomposition and plot
+test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+plot(test)
+@
+
+An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
 Theory describing this method is described in Cloern and Jassby (2010).
-The function is similar to `decomp.swmpr` with a few key differences.
-The `decomp.swmpr` function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
+The function is similar to \texttt{decomp.swmpr} with a few key differences.
+The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
 For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.
-The `decomp_cj` function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
 This approach is meant for nutrient data that are obtained on a monthly cycle.
@@
 Accordingly, short-term variation less than one-month will be removed.
-Additional arguments passed to `decompTs` can be used with `decomp_cj`, such as `startyr`, `endyr`, and `type`.
-Values passed to `type` are `mult` (default) or `add`, referring to multiplicative or additive decomposition.
-See the documentation for `decompTs` for additional explanation and examples.
-% 
-% ```{r eval = T, fig.height = 6, warning = F, cache = T}
-% # get data
-% data(apacpnut)
-% dat <- apacpnut
-% dat <- qaqc(dat, qaqc_keep = NULL)
-% 
-% # decomposition of chl, ggplot
-% decomp_cj(dat, param = 'chla_n')
-% ```
-% 
-% A reserve map with all stations can be obtained using the `map_reserve` function.
+Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
+Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
+See the documentation for \texttt{decompTs} for additional explanation and examples.
+
+<<eval = T, fig.height = 6, warning = F, cache = T>>=
+# get data
+data(apacpnut)
+dat <- apacpnut
+dat <- qaqc(dat, qaqc_keep = NULL)
+
+# decomposition of chl, ggplot
+decomp_cj(dat, param = 'chla_n')
+@
+
+A reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
 This function is a simple wrapper to functions in the ggmap package.
-The current function is limited to Google maps, which allows four map types that can be set with the `map_type` argument: terrain (default), satellite, roadmap, or hybrid.
-The `zoom` argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
+The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
 See the help documentation for ggmap for more info on zoom.
-Additionally, station locations are returned using the `site_codes_ind` function if the computer making the request has the IP address registered with CDMO.
+Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with CDMO.
 Otherwise, a local and possibly outdated file is used.
-Use the contact at the CDMO [web services](http://cdmo.baruch.sc.edu/webservices.cfm) to register your IP.
-% 
-% ```{r fig.height = 5, message = F, cache = T}
-% # plot the stations at Jacques Cousteau reserve
-% map_reserve('jac')
-% 
-% # plot the stations at Padilla Bay reserve, satellite
-% map_reserve('pdb', map_type = 'satellite', zoom = 12)
-% ```
-% 
-% Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the `plot_summary` function.
+Use the contact at the CDMO \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
+
+<<fig.height = 5, message = F, cache = T>>=
+# plot the stations at Jacques Cousteau reserve
+map_reserve('jac')
+
+# plot the stations at Padilla Bay reserve, satellite
+map_reserve('pdb', map_type = 'satellite', zoom = 12)
+@
+
+Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
@@
 The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny widget that uses this function is available: [https://beckmw.shinyapps.io/swmp_summary/](https://beckmw.shinyapps.io/swmp_summary/).
-% 
-% ```{r fig.height = 7, fig.width = 13, message = F, cache = T}
-% ## import data
-% data(apacpnut)
-% dat <- qaqc(apacpnut)
-% 
-% ## plot
-% plot_summary(dat, param = 'chla_n')
-% ```
-% 
-% Estimates of ecosystem metabolism provide a useful measure of overall system productivity.
+An interactive Shiny widget that uses this function is available: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}.
+
+<<fig.height = 7, fig.width = 13, message = F, cache = T>>=
+## import data
+data(apacpnut)
+dat <- qaqc(apacpnut)
+
+## plot
+plot_summary(dat, param = 'chla_n')
+@
+
+Estimates of ecosystem metabolism provide a useful measure of overall system productivity.
 These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
@@
 Respiration is subtracted from daily net production estimates to yield gross production.
-% 
-% The `ecometab` function is used to implement an adaptation of the open-water method.  
-% Several assumptions must be met for a valid interpretation of the results.
+
+The \texttt{ecometab} function is used to implement an adaptation of the open-water method.
+Several assumptions must be met for a valid interpretation of the results.
 In general, the dissolved oxygen time series is assumed to represent the same water mass over time.
@@
 All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000).
-% 
-% The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using `plot_metab`.
-% 
-% ```{r eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE}
-% ## import water quality and weather data
-% data(apadbwq)
-% data(apaebmet)
-% 
-% ## qaqc, combine
-% wq <- qaqc(apadbwq)
-% met <- qaqc(apaebmet)
-% dat <- comb(wq, met)
-% 
-% ## estimate metabolism
-% res <- ecometab(dat, trace = FALSE)
-% plot_metab(res)
-% ```
-% 
-% #Function list
-% 
-% See help documentation for more details on each function (e.g., `?all_params`).
-% 
-% ##Retrieve
-% 
-% `all_params` Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to `exportAllParamsXMLNew` function on web services. 
-% 
-% `all_params_dtrng` Retrieve records of all parameters within a given date range for a station.
+
+The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab}.
+
+<<eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE>>=
+## import water quality and weather data
+data(apadbwq)
+data(apaebmet)
+
+## qaqc, combine
+wq <- qaqc(apadbwq)
+met <- qaqc(apaebmet)
+dat <- comb(wq, met)
+
+## estimate metabolism
+res <- ecometab(dat, trace = FALSE)
+plot_metab(res)
+@
+
+\subsection*{Function list}
+
+See help documentation for more details on each function (e.g., \texttt{?all\_params}).
+
+\paragraph{Retrieve}
+
+\texttt{all\_params} Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.
+
+\texttt{all\_params\_dtrng} Retrieve records of all parameters within a given date range for a station.
 Optional argument for a single parameter.
 Maximum of 1000 records.
-Wrapper to `exportAllParamsDateRangeXMLNew`.
-% 
-% `import_local` Import files from a local path.
+Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.
+
+\texttt{import\_local} Import files from a local path.
 The files must be in a specific format, specifically those returned from the CDMO using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
-% 
-% `single_param` Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to `exportSingleParamXMLNew` function on web services. 
-% 
-% ##Organize
-% 
-% `comb.swmpr` Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+
+\texttt{single\_param} Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.
+
+\paragraph{Organize}
+
+\texttt{comb.swmpr} Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.
-% 
-% `qaqc.swmpr` Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.  
-% 
-% `qaqcchk.swmpr` View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.
+
+\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.
+
+\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.
 The output is used to inform further processing but is not used explicitly.
-% 
-% `rem_reps.swmpr` Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
-% 
-% `setstep.swmpr` Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in `comb.swmpr` and can also be used with individual stations.
-% 
-% `subset.swmpr` Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic `subset' function provided in the base package.
-% 
-% ##Analyze
-% 
-% `aggregate.swmpr` Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean. 
-% 
-% `aggregate_metab` Aggregate metabolism data from a swmpr object.  This is primarly used within `plot_metab` but may be useful for simple summaries of raw daily data.
-% 
-% `ecometab.swmpr` Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
-% 
-% `decomp.swmpr` Decompose a swmpr time series into trend, seasonal, and residual components.  This is a simple wrapper to `decompose`.  Decomposition of monthly or daily trends is possible.
-% 
-% `decomp_cj.swmpr` Decompose a swmpr time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to `decompTs` in the wq package.
+
+\texttt{rem\_reps.swmpr} Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
+
+\texttt{setstep.swmpr} Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
+
+\texttt{subset.swmpr} Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
+
+\paragraph{Analyze}
+
+\texttt{aggregate.swmpr} Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
+
+\texttt{aggregate\_metab} Aggregate metabolism data from a swmpr object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
+
+\texttt{ecometab.swmpr} Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
+
+\texttt{decomp.swmpr} Decompose a swmpr time series into trend, seasonal, and residual components.
+This is a simple wrapper to \texttt{decompose}.
+Decomposition of monthly or daily trends is possible.
+
+\texttt{decomp\_cj.swmpr} Decompose a swmpr time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \texttt{decompTs} in the wq package.
 Only monthly decomposition is possible.
-% 
-% `hist.swmpr` Plot a histogram for a swmpr object.
-% 
-% `lines.swmpr` Add lines to an existing swmpr plot.
-% 
-% `na.approx.swmpr` Linearly interpolate missing data (`NA` values) in a swmpr object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data. 
-% 
-% `plot.swmpr` Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
-% 
-% `plot_metab` Plot ecosystem metabolism estimates after running `ecometab` on a swmpr object.  
-% 
-% `plot_summary` Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
-% 
-% `smoother.swmpr` Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to `filter`.
-% 
-% ##Miscellaneous
-% 
-% `calcKL` Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the `ecometab` function.
-% 
-% `map_reserve` Create a map of all stations in a reserve using the ggmap package.
-% 
-% `metab_day` Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the `ecometab` function.
-% 
-% `param_names` Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).  Includes QAQC columns with 'f_' prefix. Used internally in other functions.
-% 
-% `parser` Parses html returned from CDMO web services, used internally in retrieval functions.
-% 
-% `site_codes` Metadata for all stations, wrapper to `exportStationCodesXMLNew` function on web services.
-% 
-% `site_codes_ind` Metadata for all stations at a single site, wrapper  to `NERRFilterStationCodesXMLNew` function on web services.
-% 
-% `swmpr` Creates object of swmpr class, used internally in retrieval functions.
-% 
-% `time_vec` Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
+
+\texttt{hist.swmpr} Plot a histogram for a swmpr object.
+
+\texttt{lines.swmpr} Add lines to an existing swmpr plot.
+
+\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a swmpr object.
+The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
+
+\texttt{plot.swmpr} Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
+
+\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a swmpr object.
+
+\texttt{plot\_summary} Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
+
+\texttt{smoother.swmpr} Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
+
+\paragraph{Miscellaneous}
+
+\texttt{calcKL} Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.
+
+\texttt{map\_reserve} Create a map of all stations in a reserve using the ggmap package.
+
+\texttt{metab\_day} Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.
+
+\texttt{param\_names} Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
+Includes QAQC columns with \texttt{f\_} prefix.
+Used internally in other functions.
+
+\texttt{parser} Parses html returned from CDMO web services, used internally in retrieval functions.
+
+\texttt{site\_codes} Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.
+
+\texttt{site\_codes\_ind} Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.
+
+\texttt{swmpr} Creates object of swmpr class, used internally in retrieval functions.
+
+\texttt{time\_vec} Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
 

### 2015-03-18 swmpr_manu.tex
finished formatting markdown as latex

--- 2015-03-16 swmpr_manu.tex
@@
 \section*{Materials and Methods}
-\subsection*{Etiam eget sapien nibh.}
 
-% For figure citations, please use "Fig." instead of "Figure".
-Nulla mi mi, Fig.~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
+SWMPr is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
+SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
+SWMP data are maintained online by the Centralized Data Management Office (CDMO).
+This R package provides several functions to retrieve, organize, and analyze SWMP data from the CDMO.
+Information on the CDMO web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+Data can be downloaded directly from the CDMO using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
+Your computer's IP address must be registered by CDMO staff for direct downloads in R.
+Detailed methods for importing SWMP data in R are described below.
 
-\begin{figure}[h]
-\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
-Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
+All data obtained from the CDMO should be \href{http://cdmo.baruch.sc.edu/data/citation.cfm}{cited} using the format: \\
 
-\begin{enumerate}
-\item{react}
-\item{diffuse free particles}
-\item{increment time by dt and go to 1}
-\end{enumerate}
+{\it National Estuarine Research Reserve System (NERRS).
+2012.
+System-wide Monitoring Program.
+Data accessed from the NOAA NERRS Centralized Data Management Office website: http://cdmo.baruch.sc.edu/; accessed 12 October 2012.}
+
+To cite this package: \\
+
+{\it Beck MW. 2015. SWMPr: An R package for the National Estuarine Research Reserve System.  Version 1.6.0. https://github.com/fawda123/SWMPr}
+
+\subsection*{Installing the package}
+
+This package is currently under development and has not been extensively tested.  The development version can be installed from Github:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
+\hlkwd{library}\hlstd{(devtools)}
+\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
+\hlkwd{library}\hlstd{(SWMPr)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Note that the current version of devtools (v1.6.1) was built under R version 3.1.1.  The SWMPr package may not install correctly with older versions of R.
+
+\subsection*{Data retrieval}
+
+SWMP data can be used in R after they are obtained directly from the CDMO through an online query or by using the retrieval functions provided in this package.
+In the latter case, the IP address for the computer making the request must be registered with CDMO.
+This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.
+Metadata are included as a .csv file with data requested from the CDMO and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
+Again, these functions will only work if the computer's IP address is registered with CDMO.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# retrieve metadata for all sites}
+\hlkwd{site_codes}\hlstd{()}
+
+\hlcom{# retrieve metadata for a single site}
+\hlkwd{site_codes_ind}\hlstd{(}\hlstr{'apa'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Due to rate limitations on the server, the retrieval functions in this package return a limited number of records.
+The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} or \texttt{while} loops) to obtain longer time series.
+Data retrieval functions to access the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
+These are functions that call the existing methods on the CDMO web services.
+\texttt{all\_params} returns the most recent 100 records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# all parameters for a station, most recent}
+\hlkwd{all_params}\hlstd{(}\hlstr{'hudscwq'}\hlstd{)}
+
+\hlcom{# get all parameters within a date range}
+\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{))}
+
+\hlcom{# get single parameter within a date range}
+\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+
+\hlcom{# single parameter for a station, most recent}
+\hlkwd{single_param}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlstr{'do_mgl'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+For larger requests, it's easier to obtain data outside of R using the CDMO query system and then importing within R using the \texttt{import\_local} function.
+Data can be retrieved from the CDMO several ways.
+The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO.
+The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested (expect bugs).
+The zip downloads feature is an easy way to obtain data from multiple stations in one request.
+The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
+It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to CDMO.
+The \texttt{import\_local} function can be used after the folder is decompressed.
+
+Occasionally, duplicate time stamps are present in the raw data.
+The function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
+For water quality and nutrient data, duplicate time stamps are simply removed.
+Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.
+Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
+Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.
+Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# import data for apaebmet that you downloaded}
+
+\hlcom{# this is an example path with the csv files, change as needed}
+\hlstd{path} \hlkwb{<-} \hlstr{'C:/my_path/'}
+
+\hlcom{# import, do not include file extension}
+\hlkwd{import_local}\hlstd{(path,} \hlstr{'apaebmet'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Raw csv data have not been included in the package due to size limitations.
+However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
+This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
+However, import time of the raw data may slow down use of the examples below and I have included binary data files (.RData) that are processed versions of the raw data.
+
+\subsection*{The swmpr object class}
+
+All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.
+The data include a datetimestamp column in the appropriate timezone for a station.
+Note that the datetimestamp is standard time for each timezone and does not include daylight savings.
+Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.
+The attributes for a swmpr object include \texttt{names} of the dataset, \texttt{class} (swmpr) \texttt{station name} (7 or 8 characters), \texttt{qaqc\_cols} (logical), \texttt{date\_rng} (POSIX vector), \texttt{timezone} (text string in country/city format), \texttt{stamp\_class} (class of datetimestamp vector, POSIX or Date), and ), \texttt{parameters} (character vector).
+Attributes of a swmpr object can be viewed as follows:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# import binary data}
+\hlkwd{data}\hlstd{(apadbwq)}
+\hlstd{dat} \hlkwb{<-} \hlstd{apadbwq}
+
+\hlcom{# verify that dat is swmpr class}
+\hlkwd{class}\hlstd{(dat)}
+\end{alltt}
+\begin{verbatim}
+## [1] "swmpr"      "data.frame"
+\end{verbatim}
+\begin{alltt}
+\hlcom{# all attributes of dat}
+\hlkwd{names}\hlstd{(}\hlkwd{attributes}\hlstd{(dat))}
+\end{alltt}
+\begin{verbatim}
+## [1] "names"       "row.names"   "class"       "station"    
+## [5] "parameters"  "qaqc_cols"   "date_rng"    "timezone"   
+## [9] "stamp_class"
+\end{verbatim}
+\begin{alltt}
+\hlcom{# a single attribute of dat}
+\hlkwd{attr}\hlstd{(dat,} \hlstr{'station'}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## [1] "apadbwq"
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.
+A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.
+Available methods for the swmpr class are described below and can also be viewed:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# available methods for swmpr class}
+\hlkwd{methods}\hlstd{(}\hlkwc{class} \hlstd{=} \hlstr{'swmpr'}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+##  [1] aggregate.swmpr       aggregate_metab.swmpr comb.swmpr           
+##  [4] decomp.swmpr          decomp_cj.swmpr       ecometab.swmpr       
+##  [7] hist.swmpr            lines.swmpr           na.approx.swmpr      
+## [10] plot.swmpr            plot_metab.swmpr      plot_summary.swmpr   
+## [13] qaqc.swmpr            qaqcchk.swmpr         rem_reps.swmpr       
+## [16] setstep.swmpr         smoother.swmpr        subset.swmpr
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{An overview of methods for swmpr objects}
+
+Three categories of functions are available: retrieve, organize, and analyze.
+The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
+Methods defined for swmpr objects can be applied with the organize and analyze functions.
+These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.
+S3 methods are implemented in all cases.
+
+The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
+The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.
+Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
+Values in the QAQC column specify a flag from -5 to 5.
+Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.
+Data that do not satisfy QAQC criteria are converted to NA values.
+Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
+Erroneous data entered as -99 are also removed.
+Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# qaqc screen for a swmpr object, retain only '0'}
+\hlkwd{qaqc}\hlstd{(dat)}
+
+\hlcom{# retain all data regardless of flag}
+\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwa{NULL}\hlstd{)}
+
+\hlcom{# retain only '0' and '-1' flags}
+\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with ), \texttt{qaqc}.
+The \texttt{qaqcchk} function can be used to view this information.
+Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# view the number observations in each QAQC flag}
+\hlkwd{qaqcchk}\hlstd{(dat)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
+The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
+The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
+Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
+The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get nutrient data}
+\hlkwd{data}\hlstd{(apacpnut)}
+\hlstd{swmp1} \hlkwb{<-} \hlstd{apacpnut}
+\hlstd{swmp1} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
+
+\hlcom{# remove replicate nutrient data}
+\hlkwd{rem_reps}\hlstd{(swmp1)}
+
+\hlcom{# use different function to aggregate replicates}
+\hlstd{func} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{max}\hlstd{(x,} \hlkwc{na.rm} \hlstd{= T)}
+\hlkwd{rem_reps}\hlstd{(swmp1,} \hlkwc{FUN} \hlstd{= func)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+A subset method added to the existing \texttt{subset} function is available for swmpr objects.
+This function is used to subset the data by date and/or a selected parameter.
+The date can be a single value or as two dates to select records within the range.
+The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.
+The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).
+Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
+\texttt{2012-11-31 12:00}.
+Finally, the function can be used to remove rows and columns that do not contain data.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# select two parameters from dat}
+\hlkwd{subset}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'rh'}\hlstd{,} \hlstr{'bp'}\hlstd{))}
+
+\hlcom{# subset records greater than or equal to a date}
+\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlstr{'2013-01-01 0:00'}\hlstd{,} \hlkwc{operator} \hlstd{=} \hlstr{'>='}\hlstd{)}
+
+\hlcom{# subset records within a date range}
+\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-01 6:00'}\hlstd{,} \hlstr{'2012-08-01 18:15'}\hlstd{))}
+
+\hlcom{# subset records within a date range, select two parameters}
+\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-01 6:00'}\hlstd{,} \hlstr{'2012-08-01 18:15'}\hlstd{),}
+  \hlkwc{select} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'atemp'}\hlstd{,} \hlstr{'totsorad'}\hlstd{))}
+
+\hlcom{# remove rows/columns that do not contain data}
+\hlkwd{subset}\hlstd{(dat,} \hlkwc{rem_rows} \hlstd{= T,} \hlkwc{rem_cols} \hlstd{= T)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+The \texttt{setstep} function formats a swmpr object to a continuous time series at a given time step.
+This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.
+The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
+The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
+Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.
+Likewise, the default value for \texttt{differ} is one half the time step.
+Rows that do not match any existing data within the limits of the \texttt{differ} argument are not discarded.
+Output from the \texttt{setstep} function can be used with \texttt{subset} and to create a time series at a set interval with empty data removed.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# convert time series to two hour invervals}
+\hlcom{# tolerance of +/- 30 minutes for matching existing data}
+\hlkwd{setstep}\hlstd{(dat,} \hlkwc{timestep} \hlstd{=} \hlnum{120}\hlstd{,} \hlkwc{differ} \hlstd{=} \hlnum{30}\hlstd{)}
+
+\hlcom{# convert a nutrient time series to a continuous time series}
+\hlcom{# then remove empty rows and columns}
+\hlkwd{data}\hlstd{(apacpnut)}
+\hlstd{dat_nut} \hlkwb{<-} \hlstd{apacpnut}
+\hlstd{dat_nut} \hlkwb{<-} \hlkwd{setstep}\hlstd{(dat_nut,} \hlkwc{timestep} \hlstd{=} \hlnum{60}\hlstd{)}
+\hlkwd{subset}\hlstd{(dat_nut,} \hlkwc{rem_rows} \hlstd{= T,} \hlkwc{rem_cols} \hlstd{= T)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+The \texttt{comb} function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.
+The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
+The function requires one or more swmpr objects as input as separate, undefined arguments.
+The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.
+In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.
+The time series for merging depends on the value passed to the \texttt{method} argument.
+Passing \texttt{union} to \texttt{method} will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
+Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
+Finally, a seven or eight character station name passed to \texttt{method} will merge all input objects based on a continuous time series for the given station.
+The specified station must be present in the input data.
+Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get nuts, wq, and met data as separate objects for the same station}
+\hlcom{# note that most sites usually have one weather station}
+\hlkwd{data}\hlstd{(apacpnut)}
+\hlkwd{data}\hlstd{(apacpwq)}
+\hlkwd{data}\hlstd{(apaebmet)}
+\hlstd{swmp1} \hlkwb{<-} \hlstd{apacpnut}
+\hlstd{swmp2} \hlkwb{<-} \hlstd{apacpwq}
+\hlstd{swmp3} \hlkwb{<-} \hlstd{apaebmet}
+
+\hlcom{# combine nuts and wq data by union}
+\hlkwd{comb}\hlstd{(swmp1, swmp2,} \hlkwc{method} \hlstd{=} \hlstr{'union'}\hlstd{)}
+
+\hlcom{# combine nuts and wq data by intersect}
+\hlkwd{comb}\hlstd{(swmp1, swmp3,} \hlkwc{method} \hlstd{=} \hlstr{'intersect'}\hlstd{)}
+
+\hlcom{# combine nuts, wq, and met data by nuts time series, two hour time step}
+\hlkwd{comb}\hlstd{(swmp1, swmp2, swmp3,} \hlkwc{timestep} \hlstd{=} \hlnum{120}\hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{'apacpnut'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
+The latter category includes a limited number of functions that were developed by myself or others.
+The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.
+These functions include swmpr methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.
+
+The \texttt{aggregate} function aggregates parameter data for a swmpr object by set periods of observation.
+This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
+Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.
+A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.
+This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.
+Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).
+Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for swmpr objects to preserve the time series of the input data.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# combine, qaqc, remove empty columns}
+\hlstd{dat} \hlkwb{<-} \hlkwd{comb}\hlstd{(swmp1, swmp2,} \hlkwc{method} \hlstd{=} \hlstr{'union'}\hlstd{)}
+\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(dat)}
+\hlstd{swmpr_in} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{rem_cols} \hlstd{= T)}
+
+\hlcom{# get mean DO by quarters}
+\hlkwd{aggregate}\hlstd{(swmpr_in,} \hlstr{'quarters'}\hlstd{,} \hlkwc{params} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'do_mgl'}\hlstd{))}
+
+\hlcom{# get mean DO by quarters, remove NA when calculating means}
+\hlstd{fun_in} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{mean}\hlstd{(x,} \hlkwc{na.rm} \hlstd{= T)}
+\hlkwd{aggregate}\hlstd{(swmpr_in,} \hlkwc{FUN} \hlstd{= fun_in,} \hlstr{'quarters'}\hlstd{,} \hlkwc{params} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'do_mgl'}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Time series can be smoothed to better characterize a signal independent of noise.
+Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
+The \texttt{smoother} function can be used to smooth parameters in a swmpr object using a specified window size.
+This method is a simple wrapper to \texttt{filter}.
+The \texttt{window} argument specifies the number of observations included in the moving average.
+The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).
+A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation.
+As before, the \texttt{params} argument specifies which parameters to smooth.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# import data}
+\hlkwd{data}\hlstd{(apadbwq)}
+\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
+
+\hlcom{# qaqc and subset imported data}
+\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
+\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-09 00:00'}\hlstd{,} \hlstr{'2012-07-24 00:00'}\hlstd{))}
+
+\hlcom{# filter}
+\hlstd{test} \hlkwb{<-} \hlkwd{smoother}\hlstd{(dat,} \hlkwc{window} \hlstd{=} \hlnum{50}\hlstd{,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+
+\hlcom{# plot to see the difference}
+\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp,} \hlkwc{data} \hlstd{= dat,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
+\hlkwd{lines}\hlstd{(test,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-15}
+
+}
+
+\end{knitrout}
+
+A common issue with any statistical analysis is the treatment of missing values.
+Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
+In either case, an analyst should have a strong rationale for the chosen method.
+A common approach used to handle missing data in time series analysis is linear interpolation.
+A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
+A challenge with linear interpolation is an appropriate gap size for fitting missing observations.
+The ability of the interpolated data to approximate actual trends is a function of the gap size.
+Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.
+An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
+The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.
+A required argument for the function is \texttt{maxgap} which defines the maximum gap size  for interpolation.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get data}
+\hlkwd{data}\hlstd{(apadbwq)}
+\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
+
+\hlcom{# qaqc and subset imported data}
+\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
+\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-01-22 00:00'}\hlstd{,} \hlstr{'2013-01-26 00:00'}\hlstd{))}
+
+\hlcom{# interpolate, maxgap of 10 records}
+\hlstd{test} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{10}\hlstd{)}
+
+\hlcom{# interpolate maxgap of 30 records}
+\hlstd{test2} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{30}\hlstd{)}
+
+\hlcom{# plot for comparison}
+\hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{1}\hlstd{))}
+\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, dat,} \hlkwc{main} \hlstd{=} \hlstr{'Raw'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
+\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
+  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 10 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
+\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test2,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
+  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 30 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
+\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-16}
+
+}
+
+\end{knitrout}
+
+The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
+The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.
+By default, a moving average with a symmetric window is used to filter the seasonal component.
+Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).
+
+The \texttt{decompose} function requires a ts object with a specified frequency as input.
+The \texttt{decomp} function converts the input swmpr vector to a ts object prior to \texttt{decompose}.
+This requires an explicit input defining the frequency of the parameter in the time series.
+For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).
+The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
+For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.
+A starting value of the time series must be supplied in the latter case.
+Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+
+Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get data}
+\hlkwd{data}\hlstd{(apadbwq)}
+\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
+
+\hlcom{# subset for daily decomposition}
+\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(swmp1,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-07-01 00:00'}\hlstd{,} \hlstr{'2013-07-31 00:00'}\hlstd{))}
+
+\hlcom{# decomposition and plot}
+\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
+\hlkwd{plot}\hlstd{(test)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-17}
+
+}
+
+\end{knitrout}
+
+The next example illustrates how to handle missing values using the \texttt{decomp} function.
+The \texttt{decompose} function used internally within \texttt{decomp} currently cannot process time series with missing values.
+A recommended approach is to use \texttt{na.approx} to interpolate the missing values prior to \texttt{decompose}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get data}
+\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(swmp1,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-06-01 00:00'}\hlstd{,} \hlstr{'2013-07-31 00:00'}\hlstd{))}
+
+\hlcom{# this returns an error}
+\hlcom{# test <- decomp(dat, param = 'do_mgl', frequency = 'daily')}
+
+\hlcom{# how many missing values?}
+\hlkwd{sum}\hlstd{(}\hlkwd{is.na}\hlstd{(dat}\hlopt{$}\hlstd{do_mgl))}
+\end{alltt}
+\begin{verbatim}
+## [1] 3
+\end{verbatim}
+\begin{alltt}
+\hlcom{# use na.approx to interpolate missing data}
+\hlstd{dat} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{10}\hlstd{)}
+
+\hlcom{# decomposition and plot}
+\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
+\hlkwd{plot}\hlstd{(test)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-18}
+
+}
+
+\end{knitrout}
+
+An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
+Theory describing this method is described in Cloern and Jassby (2010).
+The function is similar to \texttt{decomp.swmpr} with a few key differences.
+The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
+For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.
+The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+This approach is meant for nutrient data that are obtained on a monthly cycle.
+The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.
+Accordingly, short-term variation less than one-month will be removed.
+Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
+Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
+See the documentation for \texttt{decompTs} for additional explanation and examples.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# get data}
+\hlkwd{data}\hlstd{(apacpnut)}
+\hlstd{dat} \hlkwb{<-} \hlstd{apacpnut}
+\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwa{NULL}\hlstd{)}
+
+\hlcom{# decomposition of chl, ggplot}
+\hlkwd{decomp_cj}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-19}
+
+}
+
+\end{knitrout}
+
+A reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
+This function is a simple wrapper to functions in the ggmap package.
+The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for ggmap for more info on zoom.
+Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with CDMO.
+Otherwise, a local and possibly outdated file is used.
+Use the contact at the CDMO \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# plot the stations at Jacques Cousteau reserve}
+\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-201}
+
+}
+
+\begin{kframe}\begin{alltt}
+\hlcom{# plot the stations at Padilla Bay reserve, satellite}
+\hlkwd{map_reserve}\hlstd{(}\hlstr{'pdb'}\hlstd{,} \hlkwc{map_type} \hlstd{=} \hlstr{'satellite'}\hlstd{,} \hlkwc{zoom} \hlstd{=} \hlnum{12}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-202}
+
+}
+
+\end{knitrout}
+
+Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.
+The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
+Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.
+Monthly anomalies are in relation to the grand mean for the same month across all years.
+All data are aggregated for quicker plotting.
+Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.
+Cumulative precipitation data are based on the daily maximum.
+The function returns a graphics object (Grob) of multiple ggplot objects.
+An interactive Shiny widget that uses this function is available: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## import data}
+\hlkwd{data}\hlstd{(apacpnut)}
+\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apacpnut)}
+
+\hlcom{## plot}
+\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-21}
+
+}
+
+\end{knitrout}
+
+Estimates of ecosystem metabolism provide a useful measure of overall system productivity.
+These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
+The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface.
+The diffusion-corrected DO flux estimates are averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of DO change.
+DO flux is averaged during night hours for respiration and averaged during day hours for net production.
+Respiration rates are assumed constant during day and night such that total daily rates are calculated as hourly respiration multiplied by 24.
+The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days.
+Respiration is subtracted from daily net production estimates to yield gross production.
+
+The \texttt{ecometab} function is used to implement an adaptation of the open-water method.
+Several assumptions must be met for a valid interpretation of the results.
+In general, the dissolved oxygen time series is assumed to represent the same water mass over time.
+Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.
+The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
+Volumetric rates for gross production and total respiration are also based on total depth of the water column, which is assumed to be mixed.
+Water column depth is based on mean value for the depth variable across the time series and is floored at 1 meter for very shallow stations.
+Additionally, the volumetric reaeration coefficient requires an estimate of the anemometer height of the weather station, which is set as 10 meters by default.
+The metadata should be consulted for exact height.
+Other assumptions may apply and the user should consult the relevant literature (see the references in the help file).
+All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000).
+
+The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## import water quality and weather data}
+\hlkwd{data}\hlstd{(apadbwq)}
+\hlkwd{data}\hlstd{(apaebmet)}
+
+\hlcom{## qaqc, combine}
+\hlstd{wq} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apadbwq)}
+\hlstd{met} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apaebmet)}
+\hlstd{dat} \hlkwb{<-} \hlkwd{comb}\hlstd{(wq, met)}
+
+\hlcom{## estimate metabolism}
+\hlstd{res} \hlkwb{<-} \hlkwd{ecometab}\hlstd{(dat,} \hlkwc{trace} \hlstd{=} \hlnum{FALSE}\hlstd{)}
+\hlkwd{plot_metab}\hlstd{(res)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-22}
+
+}
+
+\end{knitrout}
+
+\subsection*{Function list}
+
+See help documentation for more details on each function (e.g., \texttt{?all\_params}).
+
+\paragraph{Retrieve}
+
+\texttt{all\_params} Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.
+
+\texttt{all\_params\_dtrng} Retrieve records of all parameters within a given date range for a station.
+Optional argument for a single parameter.
+Maximum of 1000 records.
+Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.
+
+\texttt{import\_local} Import files from a local path.
+The files must be in a specific format, specifically those returned from the CDMO using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
+
+\texttt{single\_param} Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.
+
+\paragraph{Organize}
+
+\texttt{comb.swmpr} Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+Only different data types can be combined.
+
+\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.
+
+\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.
+The output is used to inform further processing but is not used explicitly.
+
+\texttt{rem\_reps.swmpr} Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
+
+\texttt{setstep.swmpr} Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
+
+\texttt{subset.swmpr} Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
+
+\paragraph{Analyze}
+
+\texttt{aggregate.swmpr} Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
+
+\texttt{aggregate\_metab} Aggregate metabolism data from a swmpr object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
+
+\texttt{ecometab.swmpr} Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
+
+\texttt{decomp.swmpr} Decompose a swmpr time series into trend, seasonal, and residual components.
+This is a simple wrapper to \texttt{decompose}.
+Decomposition of monthly or daily trends is possible.
+
+\texttt{decomp\_cj.swmpr} Decompose a swmpr time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \texttt{decompTs} in the wq package.
+Only monthly decomposition is possible.
+
+\texttt{hist.swmpr} Plot a histogram for a swmpr object.
+
+\texttt{lines.swmpr} Add lines to an existing swmpr plot.
+
+\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a swmpr object.
+The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
+
+\texttt{plot.swmpr} Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
+
+\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a swmpr object.
+
+\texttt{plot\_summary} Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
+
+\texttt{smoother.swmpr} Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
+
+\paragraph{Miscellaneous}
+
+\texttt{calcKL} Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.
+
+\texttt{map\_reserve} Create a map of all stations in a reserve using the ggmap package.
+
+\texttt{metab\_day} Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.
+
+\texttt{param\_names} Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
+Includes QAQC columns with \texttt{f\_} prefix.
+Used internally in other functions.
+
+\texttt{parser} Parses html returned from CDMO web services, used internally in retrieval functions.
+
+\texttt{site\_codes} Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.
+
+\texttt{site\_codes\_ind} Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.
+
+\texttt{swmpr} Creates object of swmpr class, used internally in retrieval functions.
+
+\texttt{time\_vec} Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
+
+\nameref{online_widget}
 

### 2015-03-19 swmpr_manu.Rnw
started buffing up text

@@
 % acronyms
+\acrodef{CDMO}{Centralized Data Management Office}
 \acrodef{NERRS}{National Estuarine Research Reserve System}
+\acrodef{QAQC}{quality assurance/quality control}
 \acrodef{SWMP}{System Wide Monitoring Program}
@@
 Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.
-Overall, the package provides an effectives approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
+Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
 
@@
 
-Info about SWMP, etc.
-
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Materials and Methods}
+The development of automated sensors that have the ability to collect data at almost continuous intervals has facilitated the increase of standardized environmental monitoring programs.
+These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
 
@@
 SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
-SWMP data are maintained online by the Centralized Data Management Office (CDMO).
-This R package provides several functions to retrieve, organize, and analyze SWMP data from the CDMO.
-Information on the CDMO web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-Data can be downloaded directly from the CDMO using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
-Your computer's IP address must be registered by CDMO staff for direct downloads in R.
+SWMP data are maintained online by the \ac{CDMO}.
+This R package provides several functions to retrieve, organize, and analyze SWMP data from the \ac{CDMO}.
+Information on the \ac{CDMO} web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+Data can be downloaded directly from the \ac{CDMO} using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
+Your computer's IP address must be registered by \ac{CDMO} staff for direct downloads in R.
 Detailed methods for importing SWMP data in R are described below.
 
-All data obtained from the CDMO should be \href{http://cdmo.baruch.sc.edu/data/citation.cfm}{cited} using the format: \\
+\section*{Methods}
 
-{\it National Estuarine Research Reserve System (NERRS).
-2012.
-System-wide Monitoring Program.
-Data accessed from the NOAA NERRS Centralized Data Management Office website: http://cdmo.baruch.sc.edu/; accessed 12 October 2012.}
+\subsection*{The \ac{CDMO} and web portal for retrieving data}
 
-To cite this package: \\
-
-{\it Beck MW. 2015. SWMPr: An R package for the National Estuarine Research Reserve System.  Version 1.6.0. https://github.com/fawda123/SWMPr}
+The \acl{CDMO} is an online web portal for accessing monitoring data from the \ac{SWMP} network.
+\ac{CDMO} was established to address priority areas in support of \ac{SWMP} that generally focus on the continuation and advancement of data and information management.
+As such, the services provided by \ac{CDMO} allow access to over 35 milllion water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
+Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
+In most cases, an analyst will have a particular research question or management objective that he or she seeks to address by evaluating a particular dataset.
+The analysis will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
+Given the quantify of available information,
 
@@
 
-This package is currently under development and has not been extensively tested.  The development version can be installed from Github:
+The package can be installed from Github after installing the devtools library.
 
@@
 
-Note that the current version of devtools (v1.6.1) was built under R version 3.1.1.  The SWMPr package may not install correctly with older versions of R.
-
 \subsection*{Data retrieval}
 
-SWMP data can be used in R after they are obtained directly from the CDMO through an online query or by using the retrieval functions provided in this package.
-In the latter case, the IP address for the computer making the request must be registered with CDMO.
+SWMP data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.
 This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
 The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.
-Metadata are included as a .csv file with data requested from the CDMO and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
-Again, these functions will only work if the computer's IP address is registered with CDMO.
+Metadata are included as a .csv file with data requested from the \ac{CDMO} and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
+Again, these functions will only work if the computer's IP address is registered with \ac{CDMO}.
 
@@
 The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} or \texttt{while} loops) to obtain longer time series.
-Data retrieval functions to access the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
+Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
 These are functions that call the existing methods on the CDMO web services.
@@
 
-For larger requests, it's easier to obtain data outside of R using the CDMO query system and then importing within R using the \texttt{import\_local} function.
-Data can be retrieved from the CDMO several ways.
-The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO.
+For larger requests, it's easier to obtain data outside of R using the \ac{CDMO} query system and then importing within R using the \texttt{import\_local} function.
+Data can be retrieved from the \ac{CDMO} several ways.
+The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the \ac{CDMO}.
 The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested (expect bugs).
@@
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to CDMO.
+It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to \ac{CDMO}.
 The \texttt{import\_local} function can be used after the folder is decompressed.
@@
 However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
-This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
+This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
 However, import time of the raw data may slow down use of the examples below and I have included binary data files (.RData) that are processed versions of the raw data.
@@
 
-Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
+Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.
 The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
@@
 
-The next example illustrates how to handle missing values using the \texttt{decomp} function.
-The \texttt{decompose} function used internally within \texttt{decomp} currently cannot process time series with missing values.
-A recommended approach is to use \texttt{na.approx} to interpolate the missing values prior to \texttt{decompose}.
-
-<<eval = T, fig.height = 6, warning = F, cache = T>>=
-# get data
-dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
-
-# this returns an error
-# test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-
-# how many missing values?
-sum(is.na(dat$do_mgl))
-
-# use na.approx to interpolate missing data
-dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-
-# decomposition and plot
-test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-plot(test)
-@
-
 An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
-Theory describing this method is described in Cloern and Jassby (2010).
+Theory describing this method is described in \cite{Cloern10}.
 The function is similar to \texttt{decomp.swmpr} with a few key differences.
@@
 See the help documentation for ggmap for more info on zoom.
-Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with CDMO.
+Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with \ac{CDMO}.
 Otherwise, a local and possibly outdated file is used.
-Use the contact at the CDMO \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
+Use the contact at the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
 
@@
 map_reserve('jac')
-
-# plot the stations at Padilla Bay reserve, satellite
-map_reserve('pdb', map_type = 'satellite', zoom = 12)
 @
@@
 The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny widget that uses this function is available: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}.
+An interactive Shiny widget that uses this function is available (see the \nameref{supp_info}).
 
@@
 These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface.
+The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface \cite{Odum56}.
 The diffusion-corrected DO flux estimates are averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of DO change.
@@
 
-The \texttt{ecometab} function is used to implement an adaptation of the open-water method.
+The \texttt{ecometab} function is used to implement an adaptation of the open-water method \cite{Odum56,Caffrey14}.
 Several assumptions must be met for a valid interpretation of the results.
@@
 \texttt{import\_local} Import files from a local path.
-The files must be in a specific format, specifically those returned from the CDMO using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
+The files must be in a specific format, specifically those returned from the \ac{CDMO} using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
 
@@
 
-\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO.
+\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by \ac{CDMO}.
 The output is used to inform further processing but is not used explicitly.
@@
 
-\texttt{parser} Parses html returned from CDMO web services, used internally in retrieval functions.
+\texttt{parser} Parses html returned from \ac{CDMO} web services, used internally in retrieval functions.
 
@@
 
-\nameref{online_widget}
-
 % Results and Discussion can be combined.
@@
 \section*{Supporting Information}
+\label{supp_info}
 
 % Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
-\subsection*{Online widget}
-\label{online_widget}
-{\bf Online widget.} Description of online widget.
+\subsection*{Trends in SWMP parameters}
+\label{swmp_trends}
+\href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}
+\bigskip
+This widget is an interactive tool to evaluate trends in SWMP data within and between sites.
+Trends are described by an increase or decrease in values over time using a simple linear regression of summarized data.
+The regression for each station can be viewed by clicking on a map location.
+Trends at each station are plotted as circles that identify the direction and significance of the trend.
+The trend direction is blue for decreasing and red for increasing.
+The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.
+
+\subsection*{Montly and annual summary of SWMP parameters}
+\label{swmp_summary}
+\href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}
+
+This interactive widget provides graphical summaries of water quality, weather, and nutrient station data from SWMP.
+The drop down menus can be used to select the station, date range, and parameter for plotting.
+The raw data used for plotting include all SWMP records from the earliest date at each station after processing to remove QAQC flags.
+The data were downloaded from the \href{http://cdmo.baruch.sc.edu/}{CDMO web services} on November 25th, 2014 and include observations up to that date.
+Plots are based on daily averages for each parameter.
+Cumulative precipitation data are based on the daily maximum.
 
@@
 
-Text
+I acknowledge the significant efforts of \ac{NERRS} researchers and staff for providing access to high-quality monitoring data.
+Thanks particularly to Dwayne Porter and Melissa Ide from \ac{CDMO} for maintaining the online database.
+Thanks to Marie Bundy and Nikki Dix for providing me the opportunity to share this package with the broader \ac{NERRS} community.
+Thanks to Todd O'Brien for the inspiration for the online widgets in the supporting information.
+The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S.
+Environmental Protection Agency.
 
@@
 
-\bibliographystyle{plost2015}
-\bibliography{M:/docs/refs/refs}
+\bibliography{swmpr_refs}
 

### 2015-03-19 swmpr_manu.tex
started buffing up text

@@
 SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
-SWMP data are maintained online by the Centralized Data Management Office (CDMO).
-This R package provides several functions to retrieve, organize, and analyze SWMP data from the CDMO.
-Information on the CDMO web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-Data can be downloaded directly from the CDMO using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
-Your computer's IP address must be registered by CDMO staff for direct downloads in R.
+SWMP data are maintained online by the \ac{CDMO}.
+This R package provides several functions to retrieve, organize, and analyze SWMP data from the \ac{CDMO}.
+Information on the \ac{CDMO} web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
+Data can be downloaded directly from the \ac{CDMO} using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
+Your computer's IP address must be registered by \ac{CDMO} staff for direct downloads in R.
 Detailed methods for importing SWMP data in R are described below.
 
-All data obtained from the CDMO should be \href{http://cdmo.baruch.sc.edu/data/citation.cfm}{cited} using the format: \\
-
-{\it National Estuarine Research Reserve System (NERRS).
-2012.
-System-wide Monitoring Program.
-Data accessed from the NOAA NERRS Centralized Data Management Office website: http://cdmo.baruch.sc.edu/; accessed 12 October 2012.}
-
-To cite this package: \\
-
-{\it Beck MW. 2015. SWMPr: An R package for the National Estuarine Research Reserve System.  Version 1.6.0. https://github.com/fawda123/SWMPr}
+\section*{Methods}
+
+\subsection*{The \ac{CDMO} and web portal for retrieving data}
+
+The \acl{CDMO} is an online web portal for accessing monitoring data from the \ac{SWMP} network.
+\ac{CDMO} was established to address priority areas in support of \ac{SWMP} that generally focus on the continuation and advancement of data and information management.
+As such, the services provided by \ac{CDMO} allow access to over 35 milllion water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
+Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
+In most cases, an analyst will have a particular research question or management objective that he or she seeks to address by evaluating a particular dataset.
+The analysis will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
+Given the quantify of available information,
 
@@
 
-The next example illustrates how to handle missing values using the \texttt{decomp} function.
-The \texttt{decompose} function used internally within \texttt{decomp} currently cannot process time series with missing values.
-A recommended approach is to use \texttt{na.approx} to interpolate the missing values prior to \texttt{decompose}.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get data}
-\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(swmp1,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-06-01 00:00'}\hlstd{,} \hlstr{'2013-07-31 00:00'}\hlstd{))}
-
-\hlcom{# this returns an error}
-\hlcom{# test <- decomp(dat, param = 'do_mgl', frequency = 'daily')}
-
-\hlcom{# how many missing values?}
-\hlkwd{sum}\hlstd{(}\hlkwd{is.na}\hlstd{(dat}\hlopt{$}\hlstd{do_mgl))}
-\end{alltt}
-\begin{verbatim}
-## [1] 3
-\end{verbatim}
-\begin{alltt}
-\hlcom{# use na.approx to interpolate missing data}
-\hlstd{dat} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{10}\hlstd{)}
-
-\hlcom{# decomposition and plot}
-\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
-\hlkwd{plot}\hlstd{(test)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-18}
-
-}
-
-\end{knitrout}
-
 An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
-Theory describing this method is described in Cloern and Jassby (2010).
+Theory describing this method is described in \cite{Cloern10}.
 The function is similar to \texttt{decomp.swmpr} with a few key differences.
@@
 
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-19}
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-18}
 
@@
 
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-201}
-
-}
-
-\begin{kframe}\begin{alltt}
-\hlcom{# plot the stations at Padilla Bay reserve, satellite}
-\hlkwd{map_reserve}\hlstd{(}\hlstr{'pdb'}\hlstd{,} \hlkwc{map_type} \hlstd{=} \hlstr{'satellite'}\hlstd{,} \hlkwc{zoom} \hlstd{=} \hlnum{12}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-202}
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-19}
 
@@
 
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-21}
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-20}
 
@@
 
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-22}
+{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-21}
 
@@
 \texttt{time\_vec} Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
-
-\nameref{online_widget}
 

### 2015-03-19 swmpr_manu.Rnw
intro and methods beef up, figure captions, started tables, added lots of refs

@@
 \usepackage{nameref}
-\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
-\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 \usepackage{rotating}
+
+% packages I added
+\usepackage{cleveref}
+\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 
+% cleveref options
+\crefname{table}{Table}{Tables}
+\crefname{figure}{Fig.}{Figs.}
+\renewcommand{\figurename}{Fig.}
+
 % acronyms
@@
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS} .
+One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.
 This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries across the United States.
@@
 
-The development of automated sensors that have the ability to collect data at almost continuous intervals has facilitated the increase of standardized environmental monitoring programs.
+The development of low-cost, automated sensors that have the ability to collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.
 These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
+Applications from automated remote sensors are numerous with notable examples including prediction of harmful algal blooms and toxicants in aquatic systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and automated detection of airborne chemical warfare agents \cite{Sanders01}.
+Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.
+However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are more often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.
+Growing concerns over the use of adequate quality assurance and control methods, challenges for synthesis and interpretation, and increased emphasis on exploratory-based analytical techniques have characterized applications of data from automated monitoring programs \cite{Campbell13,Millie13}.
 
-SWMPr is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
-SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
-SWMP data are maintained online by the \ac{CDMO}.
-This R package provides several functions to retrieve, organize, and analyze SWMP data from the \ac{CDMO}.
-Information on the \ac{CDMO} web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-Data can be downloaded directly from the \ac{CDMO} using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
-Your computer's IP address must be registered by \ac{CDMO} staff for direct downloads in R.
-Detailed methods for importing SWMP data in R are described below.
+The \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}) is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.
+The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
+As part of this effort, the \acf{SWMP} program was implemented in 1995 at over 300 stations at each of the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The \ac{SWMP} network has provided a continous source of data collected at near real-time at each of the reserves with the intent to evaluate natural and anthrogogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
+These data have been useful for evaluating relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
+However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
+\ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity of data provided by \ac{SWMP} \cite{SWMP14}.
+
+This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
+SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
+Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
+Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.
+The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant addressing common challenges working with large datasets.
 
@@
 
-\subsection*{The \ac{CDMO} and web portal for retrieving data}
-
-The \acl{CDMO} is an online web portal for accessing monitoring data from the \ac{SWMP} network.
-\ac{CDMO} was established to address priority areas in support of \ac{SWMP} that generally focus on the continuation and advancement of data and information management.
-As such, the services provided by \ac{CDMO} allow access to over 35 milllion water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
-Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
-In most cases, an analyst will have a particular research question or management objective that he or she seeks to address by evaluating a particular dataset.
-The analysis will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
-Given the quantify of available information,
-
 \subsection*{Installing the package}
 
-The package can be installed from Github after installing the devtools library.
+The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}).
+The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.
+The package is loaded in the current workspace by using the \texttt{library} command.
 
@@
 
-\subsection*{Data retrieval}
+\subsection*{SWMP overview and data retrieval}
 
-SWMP data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
+The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
+Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
+Each reserve has no less than four water quality stations and one weather station at fixed locations.
+Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
+All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.
+The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).
+
+The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.
+As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
+Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
+In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
+All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.
+Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
+
+\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
+All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
+Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
+As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the function.
+
+\subsection*{Structure of the SWMPr package}
+
+The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.
+Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
+Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.
+An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.
+The following describes a general approach for using each category of functions based on a standard data workflow.
+
+\paragraph{Data retrieval}
+
+\ac{SWMP} data must be obtained from the \ac{CDMO} prior to organizing and analyzing the information.
+Two approaches can be used.
+First, the data can be obtained outside of R using the online query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
+Second, functions from the SWMPr package can be used to import the data directly from the online server (\Cref{tab:retrieve}).
 In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.
 This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.
-Metadata are included as a .csv file with data requested from the \ac{CDMO} and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
-Again, these functions will only work if the computer's IP address is registered with \ac{CDMO}.
+The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
+
+% table for retrieval funcitons
+<<results = 'asis', echo = FALSE>>=
+funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
+funcs <- paste0('\\texttt{', funcs, '}')
+
+descrips <- c(
+  'Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
+  'Retrieve records of all parameters within a given date range for a station.
+Optional argument for a single parameter.
+Maximum of 1000 records.
+Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.',
+  'Import files from a local path.  The files must be in a specific format, specifically those returned from the \\ac{CDMO} using the zip downloads option for a reserve.', 
+  'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
+  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
+  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
+  )
+
+to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+
+library(Hmisc)
+
+latex(
+  to_tab[, 'Description', drop = F],
+  file = '',
+  caption = 'Data retrieval functions available from the SWMPr package.',
+  rowlabel = 'Function',
+  colheads = 'Description',
+  rowname = to_tab$Functions,
+  caption.loc = 'top',
+  col.just=c("p{3.5in}"), 
+  label = 'tab:retrieve', 
+  table.env = FALSE
+  )
+@
 
@@
 
-Due to rate limitations on the server, the retrieval functions in this package return a limited number of records.
-The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} or \texttt{while} loops) to obtain longer time series.
+Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records.
+The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.
 Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
-These are functions that call the existing methods on the CDMO web services.
-\texttt{all\_params} returns the most recent 100 records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+These are functions that call the existing web protocol methods on the CDMO web services.
+\texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
 
@@
 # get single parameter within a date range
-all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), param = 'do_mgl')
+all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+  param = 'do_mgl')
 
@@
 
-For larger requests, it's easier to obtain data outside of R using the \ac{CDMO} query system and then importing within R using the \texttt{import\_local} function.
+For larger requests, data are easier to obtain outside of R using the \ac{CDMO} query system and then importing using the \texttt{import\_local} function.
 Data can be retrieved from the \ac{CDMO} several ways.
 The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the \ac{CDMO}.
-The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested (expect bugs).
+The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested.
 The zip downloads feature is an easy way to obtain data from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-It is recommended that all stations at a site and the complete date ranges are requested to avoid repeated requests to \ac{CDMO}.
 The \texttt{import\_local} function can be used after the folder is decompressed.
@@
 Occasionally, duplicate time stamps are present in the raw data.
-The function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
+The \texttt{import\_local} function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
 For water quality and nutrient data, duplicate time stamps are simply removed.
@@
 
-Raw csv data have not been included in the package due to size limitations.
-However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
-This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
-However, import time of the raw data may slow down use of the examples below and I have included binary data files (.RData) that are processed versions of the raw data.
-
-\subsection*{The swmpr object class}
-
-All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.
-The data include a datetimestamp column in the appropriate timezone for a station.
-Note that the datetimestamp is standard time for each timezone and does not include daylight savings.
-Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.
-The attributes for a swmpr object include \texttt{names} of the dataset, \texttt{class} (swmpr) \texttt{station name} (7 or 8 characters), \texttt{qaqc\_cols} (logical), \texttt{date\_rng} (POSIX vector), \texttt{timezone} (text string in country/city format), \texttt{stamp\_class} (class of datetimestamp vector, POSIX or Date), and ), \texttt{parameters} (character vector).
-Attributes of a swmpr object can be viewed as follows:
+All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.
+The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.
+Note that the \texttt{datetimestamp} is standard time for each timezone and does not include daylight savings.
+Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.
+The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}) \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
+Attributes of a \texttt{swmpr} object can be viewed as follows.
 
@@
 
-The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.
-A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.
-Available methods for the swmpr class are described below and can also be viewed:
+The \texttt{swmpr} object class was created for use with specific methods following the S3 object definition approach \cite{Wickham14}.
+A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.
+Available methods for the \texttt{swmpr} class are described below and can also be viewed:
 
@@
 
-\subsection*{An overview of methods for swmpr objects}
+Example data as raw, comma-separated files have not been included in the package due to size limitations.
+However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
+This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+Processed verions of the raw data are included with the package as binary data files (.RData) to decrease processing times with the examples.
+Information for each binary file can be viewed as follows.
 
-Three categories of functions are available: retrieve, organize, and analyze.
+<<eval = F>>=
+# view help files for complementary data
+# all files are samples from Apalachicola Bay
+
+# cat point station, nutrients
+?apacpnut
+
+# cat point station, water quality
+?apacpwq
+
+# dry bar station, water quality
+?apadbwq
+
+# east bay station, weater
+?apaebmet
+@
+
+\paragraph{Data organizing}
+
 The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
-Methods defined for swmpr objects can be applied with the organize and analyze functions.
-These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.
-S3 methods are implemented in all cases.
+The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
 
-The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
-The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.
-Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
+The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).
+Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
 Values in the QAQC column specify a flag from -5 to 5.
-Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.
-Data that do not satisfy QAQC criteria are converted to NA values.
+Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.
+Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
@@
 
-Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with ), \texttt{qaqc}.
+Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with ), \texttt{qaqc}.
 The \texttt{qaqcchk} function can be used to view this information.
@@
 <<eval = F, cache = T>>=
-# view the number observations in each QAQC flag
+# view the number of observations in each QAQC flag
 qaqcchk(dat)
@@
 The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
+The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
 Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
@@
 
-A subset method added to the existing \texttt{subset} function is available for swmpr objects.
+A subset method added to the existing \texttt{subset} function is available for \texttt{swmpr} objects.
 This function is used to subset the data by date and/or a selected parameter.
@@
 
-The \texttt{setstep} function formats a swmpr object to a continuous time series at a given time step.
+The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.
 This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.
@@
 
-The \texttt{comb} function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.
+The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.
 The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
-The function requires one or more swmpr objects as input as separate, undefined arguments.
+The function requires one or more \texttt{swmpr} objects as input as separate, undefined arguments.
 The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.
-In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.
+In general, the function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
 The time series for merging depends on the value passed to the \texttt{method} argument.
@@
 
+\paragraph{Data analysis}
+
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
 The latter category includes a limited number of functions that were developed by myself or others.
-The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include swmpr methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
+These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
 Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
-The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.
+The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
 
-The \texttt{aggregate} function aggregates parameter data for a swmpr object by set periods of observation.
+The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set periods of observation.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
 Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.
-A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.
+A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
 Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.
@@
 Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).
-Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for swmpr objects to preserve the time series of the input data.
+Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.
 
@@
 
-Time series can be smoothed to better characterize a signal independent of noise.
+Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a swmpr object using a specified window size.
+The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
 This method is a simple wrapper to \texttt{filter}.
@@
 As before, the \texttt{params} argument specifies which parameters to smooth.
+See \Cref{fig:smooth_ex} for the output from the code.
 
-<<eval = T, fig.height = 3, cache = T>>=
+<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\texttt{smoother} function.">>=
 # import data
@@
 An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
-The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.
+The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.
 A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.
+See \Cref{fig:interp_ex} for the output from the following code.
 
-<<eval = T, fig.height = 6, cache = T>>=
+<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
 # get data
@@
 
-The \texttt{decompose} function requires a ts object with a specified frequency as input.
-The \texttt{decomp} function converts the input swmpr vector to a ts object prior to \texttt{decompose}.
+The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.
+The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.
 This requires an explicit input defining the frequency of the parameter in the time series.
@@
 The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
-For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.
+For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.
 A starting value of the time series must be supplied in the latter case.
 Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
+\Cref{decomp_ex1} is an example of the \texttt{decomp} function.
 
-Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
-
-<<eval = T, fig.height = 6, cache = T>>=
+<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \\texttt{decomp} function.">>=
 # get data
@@
 An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
-Theory describing this method is described in \cite{Cloern10}.
+Theory describing this method is provided by Cloern and Jassby \cite{Cloern10}.
 The function is similar to \texttt{decomp.swmpr} with a few key differences.
 The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
-For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.
+For both functions, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.
 The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
-This approach is meant for nutrient data that are obtained on a monthly cycle.
-The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.
-Accordingly, short-term variation less than one-month will be removed.
+This approach works best for nutrient data that are typically obtained on a monthly cycle.
+The function will also work with continuous water quality or weather data but note that the data must first aggregated on the monthly scale before decomposition.
 Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
 Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-See the documentation for \texttt{decompTs} for additional explanation and examples.
+\Cref{decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
-<<eval = T, fig.height = 6, warning = F, cache = T>>=
+<<decomp_ex2, eval = T, fig.height = 6, warning = F, cache = T, fig.cap = "Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \\texttt{decomp\\_cj} function.">>=
 # get data
@@
 decomp_cj(dat, param = 'chla_n')
-@
-
-A reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
-This function is a simple wrapper to functions in the ggmap package.
-The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for ggmap for more info on zoom.
-Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with \ac{CDMO}.
-Otherwise, a local and possibly outdated file is used.
-Use the contact at the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
-
-<<fig.height = 5, message = F, cache = T>>=
-# plot the stations at Jacques Cousteau reserve
-map_reserve('jac')
 @
@@
 The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny widget that uses this function is available (see the \nameref{supp_info}).
+An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).
 
-<<fig.height = 7, fig.width = 13, message = F, cache = T>>=
+<<summary_ex, fig.height = 7, fig.width = 13, message = F, cache = T, fig.cap = "Summaries of a multi-year chlorophyll time series using the \\texttt{plot\\_summary} function.
+Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)">>=
 ## import data
@@
 
-The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab}.
+The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{metab_ex}).
 
-<<eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE>>=
+<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.">>=
 ## import water quality and weather data
@@
 
+\paragraph{Miscellaneous functions}
+
+Several additional functions are provided that cannot be described by the above categories.
+These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.
+
+Finally, a reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
+This function is a simple wrapper to functions in the ggmap package \cite{ggmap13}.
+The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for ggmap for more info on zoom.
+
+<<map_ex, fig.height = 5, message = F, cache = T, fig.cap = "Locations of all sites at the Jacques Cousteau reserve using the \\texttt{map\\_reserve} function.">>=
+# plot the stations at Jacques Cousteau reserve
+map_reserve('jac')
+@
+
 \subsection*{Function list}
-
-See help documentation for more details on each function (e.g., \texttt{?all\_params}).
-
-\paragraph{Retrieve}
-
-\texttt{all\_params} Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.
-
-\texttt{all\_params\_dtrng} Retrieve records of all parameters within a given date range for a station.
-Optional argument for a single parameter.
-Maximum of 1000 records.
-Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.
-
-\texttt{import\_local} Import files from a local path.
-The files must be in a specific format, specifically those returned from the \ac{CDMO} using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
-
-\texttt{single\_param} Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.
 
@@
 
-\texttt{comb.swmpr} Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+\texttt{comb.swmpr} Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.
 
-\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.
+\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present.
 
-\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by \ac{CDMO}.
+\texttt{qaqcchk.swmpr} View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different QAQC flags used by \ac{CDMO}.
 The output is used to inform further processing but is not used explicitly.
@@
 
-\texttt{setstep.swmpr} Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
+\texttt{setstep.swmpr} Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
+The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
 
-\texttt{subset.swmpr} Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
+\texttt{subset.swmpr} Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
 
@@
 
-\texttt{aggregate.swmpr} Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
+\texttt{aggregate.swmpr} Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
 
-\texttt{aggregate\_metab} Aggregate metabolism data from a swmpr object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
+\texttt{aggregate\_metab} Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
 
@@
 
-\texttt{decomp.swmpr} Decompose a swmpr time series into trend, seasonal, and residual components.
+\texttt{decomp.swmpr} Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
 This is a simple wrapper to \texttt{decompose}.
@@
 
-\texttt{decomp\_cj.swmpr} Decompose a swmpr time series into grandmean, annual, seasonal, and events components.
+\texttt{decomp\_cj.swmpr} Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
 This is a simple wrapper to \texttt{decompTs} in the wq package.
@@
 
-\texttt{hist.swmpr} Plot a histogram for a swmpr object.
+\texttt{hist.swmpr} Plot a histogram for a \texttt{swmpr} object.
 
-\texttt{lines.swmpr} Add lines to an existing swmpr plot.
+\texttt{lines.swmpr} Add lines to an existing \texttt{swmpr} plot.
 
-\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a swmpr object.
+\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
 The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
 
-\texttt{plot.swmpr} Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
+\texttt{plot.swmpr} Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.
 
-\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a swmpr object.
+\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.
 
@@
 
-\texttt{smoother.swmpr} Smooth swmpr objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
+\texttt{smoother.swmpr} Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
 
@@
 
-\texttt{site\_codes} Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.
-
-\texttt{site\_codes\_ind} Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.
-
-\texttt{swmpr} Creates object of swmpr class, used internally in retrieval functions.
+\texttt{swmpr} Creates object of \texttt{swmpr} class, used internally in retrieval functions.
 
@@
 
-\subsection*{Montly and annual summary of SWMP parameters}
+\subsection*{Monthly and annual summary of SWMP parameters}
 \label{swmp_summary}

### 2015-03-19 swmpr_manu.tex
intro and methods beef up, figure captions, started tables, added lots of refs

@@
 
-The development of automated sensors that have the ability to collect data at almost continuous intervals has facilitated the increase of standardized environmental monitoring programs.
+The development of low-cost, automated sensors that have the ability to collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.
 These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
-
-SWMPr is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
-SWMP was implemented by the National Estuarine Research Reserve System (\href{http://nerrs.noaa.gov/}{NERRS}) in 1995 to provide continuous monitoring data at over 300 stations in 28 estuaries across the United States.
-SWMP data are maintained online by the \ac{CDMO}.
-This R package provides several functions to retrieve, organize, and analyze SWMP data from the \ac{CDMO}.
-Information on the \ac{CDMO} web services are available \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-Data can be downloaded directly from the \ac{CDMO} using functions in this package, although it is easier to first download data outside of R and then use the \texttt{import\_local} function.
-Your computer's IP address must be registered by \ac{CDMO} staff for direct downloads in R.
-Detailed methods for importing SWMP data in R are described below.
+Applications from automated remote sensors are numerous with notable examples including prediction of harmful algal blooms and toxicants in aquatic systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and automated detection of airborne chemical warfare agents \cite{Sanders01}.
+Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.
+However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are more often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.
+Growing concerns over the use of adequate quality assurance and control methods, challenges for synthesis and interpretation, and increased emphasis on exploratory-based analytical techniques have characterized applications of data from automated monitoring programs \cite{Campbell13,Millie13}.
+
+The \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}) is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.
+The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
+As part of this effort, the \acf{SWMP} program was implemented in 1995 at over 300 stations at each of the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The \ac{SWMP} network has provided a continous source of data collected at near real-time at each of the reserves with the intent to evaluate natural and anthrogogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
+These data have been useful for evaluating relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
+However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
+\ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity of data provided by \ac{SWMP} \cite{SWMP14}.
+
+This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
+SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
+Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
+Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.
+The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant addressing common challenges working with large datasets.
 
@@
 
-\subsection*{Data retrieval}
-
-SWMP data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+\subsection*{SWMP overview and data retrieval}
+
+Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
+The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
+Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
+Each reserve has no less than four water quality stations and one weather station at fixed locations.
+Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
+All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.
+The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).
+
+The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.
+As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
+Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
+In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
+All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.
+Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
+
+\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
+All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
+Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
+As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the function.
+
+\subsection*{Structure of the SWMPr package}
+
+The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.
+Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
+Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.
+An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.
+The following describes a general approach for using each category of functions based on a standard data workflow.
+
+\paragraph{Data retrieval}
+
+\ac{SWMP} data must be obtained from the \ac{CDMO} prior to organizing and analyzing the information.
+Two approaches can be used.
+First, the data can be obtained outside of R using the online query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
+Second, functions from the SWMPr package can be used to import the data directly from the online server (\Cref{tab:retrieve}).
 In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.
 This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-The \href{http://cdmo.baruch.sc.edu/data/metadata.cfm}{metadata} should also be consulted for available data, including the parameters and date ranges for each monitoring station.
-Metadata are included as a .csv file with data requested from the \ac{CDMO} and can also be obtained using the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (individual site) functions.
-Again, these functions will only work if the computer's IP address is registered with \ac{CDMO}.
+The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
+
+% table for retrieval funcitons
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Data retrieval functions available from the SWMPr package.",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
+\begin{table}[!tbp]
+\caption{Data retrieval functions available from the SWMPr package.\label{tab:retrieve}} 
+\begin{center}
+\begin{tabular}{lp{3.5in}}
+\hline\hline
+\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
+\hline
+\texttt{all\_params}&Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
+\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
+Optional argument for a single parameter.
+Maximum of 1000 records.
+Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
+\texttt{import\_local}&Import files from a local path.
+The files must be in a specific format, specifically those returned from the \ac{CDMO} using the zip downloads option for a reserve.\tabularnewline
+\texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.
+Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
+\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
+\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
 
@@
 \hlcom{# get single parameter within a date range}
-\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
+\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),}
+  \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
 
@@
 
-\subsection*{An overview of methods for swmpr objects}
-
-Three categories of functions are available: retrieve, organize, and analyze.
+Example data as raw, comma-separated files have not been included in the package due to size limitations.
+However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
+This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+Processed verions of the raw data are included with the package as binary data files (.RData) to decrease processing times with the examples.
+Information for each binary file can be viewed as follows.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# view help files for complementary data}
+\hlcom{# all files are samples from Apalachicola Bay}
+
+\hlcom{# cat point station, nutrients}
+\hlopt{?}\hlstd{apacpnut}
+
+\hlcom{# cat point station, water quality}
+\hlopt{?}\hlstd{apacpwq}
+
+\hlcom{# dry bar station, water quality}
+\hlopt{?}\hlstd{apadbwq}
+
+\hlcom{# east bay station, weater}
+\hlopt{?}\hlstd{apaebmet}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\paragraph{Data organizing}
+
 The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
-Methods defined for swmpr objects can be applied with the organize and analyze functions.
-These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.
-S3 methods are implemented in all cases.
-
 The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
-The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.
-Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
+
+The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).
+Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
 Values in the QAQC column specify a flag from -5 to 5.
-Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.
-Data that do not satisfy QAQC criteria are converted to NA values.
+Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.
+Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
@@
 \begin{alltt}
-\hlcom{# view the number observations in each QAQC flag}
+\hlcom{# view the number of observations in each QAQC flag}
 \hlkwd{qaqcchk}\hlstd{(dat)}
@@
 
+\paragraph{Data analysis}
+
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
 The latter category includes a limited number of functions that were developed by myself or others.
-The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include swmpr methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
+These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
 Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
-The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.
-
-The \texttt{aggregate} function aggregates parameter data for a swmpr object by set periods of observation.
+The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
+
+The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set periods of observation.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
 Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.
-A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.
+A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
 Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.
@@
 As before, the \texttt{params} argument specifies which parameters to smooth.
+See \Cref{fig:smooth_ex} for the output from the code.
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-15}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/smooth_ex}
 
 }
+
+\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.\label{fig:smooth_ex}}
+\end{figure}
 
@@
 An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
-The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.
+The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.
 A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.
+See \Cref{fig:interp_ex} for the output from the following code.
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-16}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/interp_ex}
 
 }
+
+\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.\label{fig:interp_ex}}
+\end{figure}
 
@@
 The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
-For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.
+For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.
 A starting value of the time series must be supplied in the latter case.
 Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
-
 Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
+\Cref{decomp_ex1} is an example of the \texttt{decomp} function.
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-17}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex1}
 
@@
 
+\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.\label{fig:decomp_ex1}}
+\end{figure}
+
 \end{knitrout}
@@
 An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
-Theory describing this method is described in \cite{Cloern10}.
+Theory describing this method is provided by Cloern and Jassby \cite{Cloern10}.
 The function is similar to \texttt{decomp.swmpr} with a few key differences.
 The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
-For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.
+For both functions, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.
 The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
-This approach is meant for nutrient data that are obtained on a monthly cycle.
-The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.
-Accordingly, short-term variation less than one-month will be removed.
+This approach works best for nutrient data that are typically obtained on a monthly cycle.
+The function will also work with continuous water quality or weather data but note that the data must first aggregated on the monthly scale before decomposition.
 Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
 Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-See the documentation for \texttt{decompTs} for additional explanation and examples.
+\Cref{decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-18}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex2}
 
@@
 
-\end{knitrout}
-
-A reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
-This function is a simple wrapper to functions in the ggmap package.
-The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for ggmap for more info on zoom.
-Additionally, station locations are returned using the \texttt{site\_codes\_ind} function if the computer making the request has the IP address registered with \ac{CDMO}.
-Otherwise, a local and possibly outdated file is used.
-Use the contact at the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/webservices.cfm}{web services} to register your IP.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# plot the stations at Jacques Cousteau reserve}
-\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-19}
-
-}
+\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function.\label{fig:decomp_ex2}}
+\end{figure}
 
@@
 The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny widget that uses this function is available (see the \nameref{supp_info}).
+An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-20}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/summary_ex}
 
 }
+
+\caption[Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function]{Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function.
+Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)\label{fig:summary_ex}}
+\end{figure}
 
@@
 
-The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab}.
+The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{metab_ex}).
 
@@
 \end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-21}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/metab_ex}
 
@@
 
+\caption[Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida]{Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.\label{fig:metab_ex}}
+\end{figure}
+
+\end{knitrout}
+
+\paragraph{Miscellaneous functions}
+
+Several additional functions are provided that cannot be described by the above categories.
+These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.
+
+Finally, a reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
+This function is a simple wrapper to functions in the ggmap package \cite{ggmap13}.
+The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for ggmap for more info on zoom.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# plot the stations at Jacques Cousteau reserve}
+\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
+\end{alltt}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/map_ex}
+
+}
+
+\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.\label{fig:map_ex}}
+\end{figure}
+
 \end{knitrout}
@@
 
-See help documentation for more details on each function (e.g., \texttt{?all\_params}).
-
-\paragraph{Retrieve}
-
-\texttt{all\_params} Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.
-
-\texttt{all\_params\_dtrng} Retrieve records of all parameters within a given date range for a station.
-Optional argument for a single parameter.
-Maximum of 1000 records.
-Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.
-
-\texttt{import\_local} Import files from a local path.
-The files must be in a specific format, specifically those returned from the \ac{CDMO} using the [zip downloads](http://cdmo.baruch.sc.edu/aqs/zips.cfm) option for a reserve.
-
-\texttt{single\_param} Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.
-
 \paragraph{Organize}
 
-\texttt{comb.swmpr} Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+\texttt{comb.swmpr} Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.
 
-\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a swmpr object.  Only applies if QAQC columns are present.
-
-\texttt{qaqcchk.swmpr} View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by \ac{CDMO}.
+\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present.
+
+\texttt{qaqcchk.swmpr} View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different QAQC flags used by \ac{CDMO}.
 The output is used to inform further processing but is not used explicitly.
@@
 
-\texttt{setstep.swmpr} Format data from a swmpr object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
-
-\texttt{subset.swmpr} Subset by dates and/or columns for a swmpr object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
+\texttt{setstep.swmpr} Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
+The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
+
+\texttt{subset.swmpr} Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
 
@@
 
-\texttt{aggregate.swmpr} Aggregate swmpr objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
-
-\texttt{aggregate\_metab} Aggregate metabolism data from a swmpr object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
+\texttt{aggregate.swmpr} Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
+
+\texttt{aggregate\_metab} Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
 
@@
 
-\texttt{hist.swmpr} Plot a histogram for a swmpr object.
-
-\texttt{lines.swmpr} Add lines to an existing swmpr plot.
-
-\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a swmpr object.
+\texttt{hist.swmpr} Plot a histogram for a \texttt{swmpr} object.
+
+\texttt{lines.swmpr} Add lines to an existing \texttt{swmpr} plot.
+
+\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
 The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
 
-\texttt{plot.swmpr} Plot a univariate  time series for a swmpr object.  The parameter name must be specified.
-
-\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a swmpr object.
+\texttt{plot.swmpr} Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.
+
+\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.
 

### 2015-03-20 swmpr_manu.Rnw
started swmpr applications section, redo metab estimates for all data

@@
 \acrodef{CDMO}{Centralized Data Management Office}
+\acrodef{DO}{dissolved oxygen}
 \acrodef{NERRS}{National Estuarine Research Reserve System}
@@
 
-\section*{Methods}
+\section*{SWMP overview and data retrieval}
+
+Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
+The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
+Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
+Each reserve has no less than four water quality stations and one weather station at fixed locations.
+Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
+All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.
+The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).
+
+The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.
+As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
+Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
+In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from \ac{CDMO}.
+All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.
+Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
+
+\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the package's retrieval functions.
+Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
+All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
+Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
+As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the data retrieval functions in SWMPr.
+\href{http://cdmo.baruch.sc.edu/webservices.cfm}{Web services} are provided by \ac{CDMO} to provide direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
+The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.
+For example, the \texttt{all\_params} function in SWMPr uses the \texttt{exportAllParamsXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.
+The text below describes the data retrieval functions in more detail.
+
+\section*{Structure of the SWMPr package}
 
@@
 
-\subsection*{SWMP overview and data retrieval}
-
-Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
-The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
-Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
-Each reserve has no less than four water quality stations and one weather station at fixed locations.
-Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.
-The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).
-
-The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.
-As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
-Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
-In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
-All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
-For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.
-Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
-
-\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
-Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
-All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
-Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the function.
-
-\subsection*{Structure of the SWMPr package}
-
 The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.
@@
 
-\paragraph{Data retrieval}
+\subsection*{Data retrieval}
 
@@
 Two approaches can be used.
-First, the data can be obtained outside of R using the online query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
-Second, functions from the SWMPr package can be used to import the data directly from the online server (\Cref{tab:retrieve}).
+First, the data can be obtained outside of R using the graphical query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
+Second, functions from the SWMPr package can be used to import the data directly from the online server using \ac{CDMO} web services (\Cref{tab:retrieve}).
 In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.
@@
 
-% table for retrieval funcitons
+% table for retrieval functions
 <<results = 'asis', echo = FALSE>>=
@@
   file = '',
-  caption = 'Data retrieval functions available from the SWMPr package.',
+  caption = 'Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).',
   rowlabel = 'Function',
@@
 
-\paragraph{Data organizing}
+\subsection*{Data organizing}
 
 The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
-The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.
+The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types (\Cref{tab:organize}).
 
@@
 Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
+
+% table for organize functions
+<<results = 'asis', echo = FALSE>>=
+
+funcs <- c('comb.swmpr', 'qaqc.swmpr', 'qaqcchk.swmpr', 'rem\\_reps.swmpr', 'setstep.swmpr', 'subset.swmpr')
+funcs <- paste0('\\texttt{', funcs, '}')
+
+descrips <- c(
+  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+Only different data types can be combined.',
+  'Remove \\ac{QAQC} columns and remove data based on \\ac{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\ac{QAQC} columns are present. ', 
+  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\ac{QAQC} flags used by \\ac{CDMO}.
+The output is used to inform further processing but is not used explicitly.',
+  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
+  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb.swmpr} and can also be used with individual stations.',
+  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
+  )
+
+to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+
+library(Hmisc)
+
+latex(
+  to_tab[, 'Description', drop = F],
+  file = '',
+  caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",
+  rowlabel = 'Function',
+  colheads = 'Description',
+  rowname = to_tab$Functions,
+  caption.loc = 'top',
+  col.just=c("p{3.5in}"), 
+  label = 'tab:organize', 
+  table.env = FALSE
+  )
+@
 
@@
 
-\paragraph{Data analysis}
+\subsection*{Data analysis}
 
-The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
+The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
 The latter category includes a limited number of functions that were developed by myself or others.
@@
 Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.
+
+% table for analysis functions
+<<results = 'asis', echo = FALSE>>=
+
+funcs <- c('aggregate.swmpr', 'aggregate\\_metab.swmpr', 'ecometab.swmpr', 'decomp.swmpr', 'decomp\\_cj.swmpr', 'hist.swmpr', 'lines.swmpr', 'na.approx.swmpr', 'plot.swmpr', 'plot\\_metab.swmpr', 'plot\\_summary.swmpr', 'smoother.swmpr')
+funcs <- paste0('\\texttt{', funcs, '}')
+
+descrips <- c(
+  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.',
+  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw daily data.',
+  'Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.',
+  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose}.  Decomposition of monthly or daily trends is possible.',
+  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \\texttt{decompTs} in the wq package.
+Only monthly decomposition is possible.',
+  'Plot a histogram for a \\texttt{swmpr} object.',
+  'Add lines to an existing \\texttt{swmpr} plot.',
+  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data.', 
+  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
+  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
+  'Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.',
+  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \\texttt{filter}.'
+  )
+
+to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+
+library(Hmisc)
+
+latex(
+  to_tab[, 'Description', drop = F],
+  file = '',
+  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",
+  rowlabel = 'Function',
+  colheads = 'Description',
+  rowname = to_tab$Functions,
+  caption.loc = 'top',
+  col.just=c("p{3.5in}"), 
+  label = 'tab:analyze', 
+  table.env = FALSE
+  )
+@
 
@@
 Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
-\Cref{decomp_ex1} is an example of the \texttt{decomp} function.
+\Cref{fig:decomp_ex1} is an example of the \texttt{decomp} function.
 
@@
 Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-\Cref{decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
+\Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
@@
 ## plot
-plot_summary(dat, param = 'chla_n')
+plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
 @
@@
 These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration rates, corrected for air-sea gas diffusion at the surface \cite{Odum56}.
-The diffusion-corrected DO flux estimates are averaged during day and night for each 24 hour period in the time series, where flux is an hourly rate of DO change.
-DO flux is averaged during night hours for respiration and averaged during day hours for net production.
-Respiration rates are assumed constant during day and night such that total daily rates are calculated as hourly respiration multiplied by 24.
-The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days.
-Respiration is subtracted from daily net production estimates to yield gross production.
+The open-water method \cite{Odum56} is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model (see Ro and Hunt \cite{Ro06}, Thebault et al.
+\cite{Thebault08}).
+The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series.
+The nighttime average \ac{DO} flux is used to estimate respiration rates, while the daytime \ac{DO} flux is used to estimate net primary production.
+To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \ac{DO} flux rates are multiplied by 24.
+Similarly, the daytime \ac{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
+Respiration rates are subtracted from daily net production estimates to yield gross production rates.
+The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days
 
@@
 The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
-Volumetric rates for gross production and total respiration are also based on total depth of the water column, which is assumed to be mixed.
-Water column depth is based on mean value for the depth variable across the time series and is floored at 1 meter for very shallow stations.
-Additionally, the volumetric reaeration coefficient requires an estimate of the anemometer height of the weather station, which is set as 10 meters by default.
+Areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
+Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object.
+Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
+Additionally, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
 The metadata should be consulted for exact height.
-Other assumptions may apply and the user should consult the relevant literature (see the references in the help file).
-All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000).
+Other assumptions may apply and the user should consult the relevant literature.
+All calculations within the function are done using molar units (e.g., mmol O$^2$ m$^{-3}$).
+The output can be returned as mass units by changing the default argument.
+Input data must be in standard mass units for \ac{DO} (mg L-1)
 
-The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{metab_ex}).
+The following is an example that shows use of the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
 
@@
 
-\paragraph{Miscellaneous functions}
+\subsection*{Miscellaneous functions}
 
-Several additional functions are provided that cannot be described by the above categories.
+Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).
 These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.
 
-Finally, a reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
-This function is a simple wrapper to functions in the ggmap package \cite{ggmap13}.
-The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+% table for misc functions
+<<results = 'asis', echo = FALSE>>=
+
+funcs <- c('calckl', 'map\\_reserve', 'metab\\_day', 'param\\_names', 'parser', 'swmpr', 'time\\_vec')
+funcs <- paste0('\\texttt{', funcs, '}')
+
+descrips <- c(
+  'Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \\texttt{ecometab} function.',
+  'Create a map of all stations in a reserve using the ggmap package.',
+  'Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \\texttt{ecometab} function.',
+  'Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).  Includes \\ac{QAQC} columns with \\texttt{f\\_} prefix. Used internally in other functions.',
+  'Parses html returned from \\ac{CDMO} web services, used internally in retrieval functions.',
+  'Creates object of \\texttt{swmpr} class, used internally in retrieval functions.',
+  'Converts time vectors to \\texttt{POSIXct} objects with correct time zone for a site/station, used internally in retrieval functions.'
+  )
+
+to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+
+library(Hmisc)
+
+latex(
+  to_tab[, 'Description', drop = F],
+  file = '',
+  caption = "Miscellaneous functions available from the SWMPr package.
+Most are used within the main functions above but may be useful for more customized evaluationg of \\ac{SWMP} data.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
+  rowlabel = 'Function',
+  colheads = 'Description',
+  rowname = to_tab$Functions,
+  caption.loc = 'top',
+  col.just=c("p{3.5in}"), 
+  label = 'tab:misc', 
+  table.env = FALSE
+  )
+
+@
+
+For brevity, only the \texttt{reserve\_map} function is discussed.
+This function can be used to create a map with all stations at a reserve by passing arguments to functions in the ggmap package \cite{ggmap13}.
+The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
 The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for ggmap for more info on zoom.
+See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
 
@@
 
-\subsection*{Function list}
+\section*{Applications using the SWMPr package}
 
-\paragraph{Organize}
+The ability to evaluate environmental characteristics between estuaries within the \ac{NERRS} program has been greatly improved using functions in the SWMPr package.
+This section describes two examples of applications using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
+First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04}.
+A comprehensive evaluation of metabolic rates between estuaries has not been conducted using the most recent \ac{SWMP} data.
+The results are presented to facilitate additional research to evaluate factors that influence variation between sites.
 
-\texttt{comb.swmpr} Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
-Only different data types can be combined.
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} describe the theory and application of the open-water method to estimate ecosystem metabolism using two sites at each of the \ac{NERRS} reserve.
+Time series data included approximately five years of half-hour water quality observations at each site.
+Additionally, the air-sea gas exchange model, as implemented by the current \texttt{ecometab} function (see Ro and Hunt \cite{Ro06} and Thebault et al.
+\cite{Thebault08}), was not incorporated into the initial metabolism estimates such that a constant value for the reaeration coefficient was assumed.
+This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column  The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
+\cite{Caffrey14} for details).
 
-\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present.
+All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November of 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, saved again on a local hard drive as a binary .RData file, and uploaded.
+This resulted in a single \texttt{swmpr} object for each site.
+All files were then uploaded to a remote server.
+An R script was executed that retrieved and processed the \texttt{swmpr} objects for each site using the following functions in sequence:
 
-\texttt{qaqcchk.swmpr} View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different QAQC flags used by \ac{CDMO}.
-The output is used to inform further processing but is not used explicitly.
+The weather station with the most data was retained if more than one weather station was available for a single reserve.
 
-\texttt{rem\_reps.swmpr} Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
+<<eval = F>>=
+# packages to use
+library(SWMPr)
+library(httr)
+library(XML)
+library(foreach)
+library(doParallel)
 
-\texttt{setstep.swmpr} Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
-The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
+# names of files on server
+files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
+files_s3 <- rawToChar(files_s3)
+files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
+files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
 
-\texttt{subset.swmpr} Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
+# find only active water quality and weather sites
+# IP address must be registed with CDMO 
+meta <- site_codes()
+sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
+meta <- meta[sel, ]
 
-\paragraph{Analyze}
+# get wq sites
+wq_sites <- grep('wq$', meta$station_code, value = T)
 
-\texttt{aggregate.swmpr} Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
+##
+# process each site
 
-\texttt{aggregate\_metab} Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
+# setup parallel backend for processing
+cl <- makeCluster(8)
+registerDoParallel(cl)
+strt <- Sys.time()
 
-\texttt{ecometab.swmpr} Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
+# output
+metabs <- foreach(wq_site = wq_sites) %dopar% {
 
-\texttt{decomp.swmpr} Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \texttt{decompose}.
-Decomposition of monthly or daily trends is possible.
+  library(SWMPr)
 
-\texttt{decomp\_cj.swmpr} Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \texttt{decompTs} in the wq package.
-Only monthly decomposition is possible.
+  # progress
+  sink('log.txt')
+  cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
+  print(Sys.time() - strt)
+  sink()
 
-\texttt{hist.swmpr} Plot a histogram for a \texttt{swmpr} object.
+  # find corresponding wx station
+  met_site <- substr(wq_site, 1, 3)
+  met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
+  met_site <- gsub('\\.RData$', '', met_site)
 
-\texttt{lines.swmpr} Add lines to an existing \texttt{swmpr} plot.
+  # continue if wx data found
+  if(length(met_site) > 0){
 
-\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
-The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
+    ##
+    # load the files
 
-\texttt{plot.swmpr} Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.
+    # wq
+    raw_content <- paste0('https://s3.amazonaws.com/swmpalldata/', wq_site, '.RData')
+    raw_content <- httr::GET(raw_content)$content
+    connect <- rawConnection(raw_content)
+    load(connect)
+    wq <- get(wq_site)
+    rm(list = wq_site)
+    close(connect)
 
-\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.
+    # met
+    raw_content <- paste0('https://s3.amazonaws.com/swmpalldata/', met_site, '.RData')
+    raw_content <- httr::GET(raw_content)$content
+    connect <- rawConnection(raw_content)
+    load(connect)
+    met <- get(met_site)
+    rm(list = met_site)
+    close(connect)
 
-\texttt{plot\_summary} Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
+    ## 
+    # combine, subset up to 2013, estimate metabolism
+    dat <- SWMPr::comb(wq, met, method = met_site)
+    dat <- subset(dat, subset = '2014-01-01 0:0', operator = '<')
+    dat <- SWMPr::ecometab(dat)
 
-\texttt{smoother.swmpr} Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
+    dat
 
-\paragraph{Miscellaneous}
+  }
 
-\texttt{calcKL} Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.
+}
+names(metabs) <- wq_sites
+save(metabs, file = 'metabs.RData')
+@
 
-\texttt{map\_reserve} Create a map of all stations in a reserve using the ggmap package.
-
-\texttt{metab\_day} Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.
-
-\texttt{param\_names} Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
-Includes QAQC columns with \texttt{f\_} prefix.
-Used internally in other functions.
-
-\texttt{parser} Parses html returned from \ac{CDMO} web services, used internally in retrieval functions.
-
-\texttt{swmpr} Creates object of \texttt{swmpr} class, used internally in retrieval functions.
-
-\texttt{time\_vec} Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
-
-% Results and Discussion can be combined.
-\section*{Results}
-
-\section*{Discussion}
+\section*{Summary}
 
@@
 Thanks to Todd O'Brien for the inspiration for the online widgets in the supporting information.
+Thanks to Mike Murrell and Jim Hagy III for assistance with documentation and implementation of the ecosystem metabolism functions.
 The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S.

### 2015-03-20 swmpr_manu.tex
started swmpr applications section, redo metab estimates for all data

@@
 
-\section*{Methods}
-
-\subsection*{Installing the package}
-
-The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}).
-The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.
-The package is loaded in the current workspace by using the \texttt{library} command.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
-\hlkwd{library}\hlstd{(devtools)}
-\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
-\hlkwd{library}\hlstd{(SWMPr)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{SWMP overview and data retrieval}
+\section*{SWMP overview and data retrieval}
 
@@
 Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
-In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from the \ac{CDMO}.
+In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from \ac{CDMO}.
 All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
@@
 
-\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the retrieval functions provided in this package.
+\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the package's retrieval functions.
 Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
@@
 Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the function.
-
-\subsection*{Structure of the SWMPr package}
+As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the data retrieval functions in SWMPr.
+\href{http://cdmo.baruch.sc.edu/webservices.cfm}{Web services} are provided by \ac{CDMO} to provide direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
+The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.
+For example, the \texttt{all\_params} function in SWMPr uses the \texttt{exportAllParamsXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.
+The text below describes the data retrieval functions in more detail.
+
+\section*{Structure of the SWMPr package}
+
+\subsection*{Installing the package}
+
+The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}).
+The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.
+The package is loaded in the current workspace by using the \texttt{library} command.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
+\hlkwd{library}\hlstd{(devtools)}
+\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
+\hlkwd{library}\hlstd{(SWMPr)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 
-% table for retrieval funcitons
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Data retrieval functions available from the SWMPr package.",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
+% table for retrieval functions
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Data retrieval functions available from the SWMPr package.\label{tab:retrieve}} 
+\caption{Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} at the command line).\label{tab:retrieve}} 
 \begin{center}
@@
 Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
+
+% table for organize functions
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
+\begin{table}[!tbp]
+\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb.swmpr} at the command line).\label{tab:organize}} 
+\begin{center}
+\begin{tabular}{lp{3.5in}}
+\hline\hline
+\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
+\hline
+\texttt{comb.swmpr}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+Only different data types can be combined.\tabularnewline
+\texttt{qaqc.swmpr}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
+\texttt{qaqcchk.swmpr}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.
+The output is used to inform further processing but is not used explicitly.\tabularnewline
+\texttt{rem\_reps.swmpr}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
+\texttt{setstep.swmpr}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
+The function is used in \texttt{comb.swmpr} and can also be used with individual stations.\tabularnewline
+\texttt{subset.swmpr}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
 
@@
 
-\paragraph{Data analysis}
-
-The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.
+\subsection*{Data analysis}
+
+The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
 The latter category includes a limited number of functions that were developed by myself or others.
@@
 Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.
+
+% table for analysis functions
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
+\begin{table}[!tbp]
+\caption{Analysis functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \texttt{?aggregate.swmpr} at the command line).\label{tab:analyze}}
+\begin{center}
+\begin{tabular}{lp{3.5in}}
+\hline\hline
+\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
+\hline
+\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.
+Aggregation function is user-supplied but defaults to mean.\tabularnewline
+\texttt{aggregate\_metab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.
+This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.\tabularnewline
+\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.\tabularnewline
+\texttt{decomp.swmpr}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
+This is a simple wrapper to \texttt{decompose}.
+Decomposition of monthly or daily trends is possible.\tabularnewline
+\texttt{decomp\_cj.swmpr}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \texttt{decompTs} in the wq package.
+Only monthly decomposition is possible.\tabularnewline
+\texttt{hist.swmpr}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
+\texttt{lines.swmpr}&Add lines to an existing \texttt{swmpr} plot.\tabularnewline
+\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
+The maximum gap size that is interpolated is defined as a maximum number of records with missing data.\tabularnewline
+\texttt{plot.swmpr}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
+\texttt{plot\_metab.swmpr}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
+\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.\tabularnewline
+\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
 
@@
 \hlcom{## plot}
-\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
+\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{,} \hlkwc{years} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{2007}\hlstd{,} \hlnum{2013}\hlstd{))}
 \end{alltt}
@@
 The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
-Volumetric rates for gross production and total respiration are also based on total depth of the water column, which is assumed to be mixed.
-Water column depth is based on mean value for the depth variable across the time series and is floored at 1 meter for very shallow stations.
-Additionally, the volumetric reaeration coefficient requires an estimate of the anemometer height of the weather station, which is set as 10 meters by default.
+Areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
+Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object.
+Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
+Additionally, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
 The metadata should be consulted for exact height.
-Other assumptions may apply and the user should consult the relevant literature (see the references in the help file).
-All estimates are in mmol of oxygen but can be converted  to grams by changing the default arguments (i.e., 1mmol O2 = 32 mg O2, 1000 mg = 1g, multiply all estimates by 32/1000).
-
-The following is an example that shows how to use the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{metab_ex}).
+Other assumptions may apply and the user should consult the relevant literature.
+All calculations within the function are done using molar units (e.g., mmol O$^2$ m$^{-3}$).
+The output can be returned as mass units by changing the default argument.
+Input data must be in standard mass units for \ac{DO} (mg L-1)
+
+The following is an example that shows use of the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
 
@@
 
-\paragraph{Miscellaneous functions}
-
-Several additional functions are provided that cannot be described by the above categories.
+\subsection*{Miscellaneous functions}
+
+Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).
 These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.
 
-Finally, a reserve map with all stations can be obtained using the \texttt{map\_reserve} function.
-This function is a simple wrapper to functions in the ggmap package \cite{ggmap13}.
-The current function is limited to Google maps, which allows four map types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+% table for misc functions
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the SWMPr package.
+Most are used within the main functions above but may be useful for more customized evaluationg of \\ac{SWMP} data.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE)%
+\begin{table}[!tbp]
+\caption{Miscellaneous functions available from the SWMPr package.
+Most are used within the main functions above but may be useful for more customized evaluationg of \ac{SWMP} data.
+Full documentation for each function is in the help file (e.g., execute \texttt{?calckl} at the command line).\label{tab:misc}}
+\begin{center}
+\begin{tabular}{lp{3.5in}}
+\hline\hline
+\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
+\hline
+\texttt{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.\tabularnewline
+\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
+\texttt{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.\tabularnewline
+\texttt{param\_names}&Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
+Includes \ac{QAQC} columns with \texttt{f\_} prefix.
+Used internally in other functions.\tabularnewline
+\texttt{parser}&Parses html returned from \ac{CDMO} web services, used internally in retrieval functions.\tabularnewline
+\texttt{swmpr}&Creates object of \texttt{swmpr} class, used internally in retrieval functions.\tabularnewline
+\texttt{time\_vec}&Converts time vectors to \texttt{POSIXct} objects with correct time zone for a site/station, used internally in retrieval functions.\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\end{table}
+
+For brevity, only the \texttt{reserve\_map} function is discussed.
+This function can be used to create a map with all stations at a reserve by passing arguments to functions in the ggmap package \cite{ggmap13}.
+The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
 The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for ggmap for more info on zoom.
+See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
 
@@
 
-\subsection*{Function list}
-
-\paragraph{Organize}
-
-\texttt{comb.swmpr} Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
-Only different data types can be combined.
-
-\texttt{qaqc.swmpr} Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present.
-
-\texttt{qaqcchk.swmpr} View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different QAQC flags used by \ac{CDMO}.
-The output is used to inform further processing but is not used explicitly.
-
-\texttt{rem\_reps.swmpr} Remove replicate nutrient data that occur on the same day.  The default is to average replicates.
-
-\texttt{setstep.swmpr} Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
-The function is used in \texttt{comb.swmpr} and can also be used with individual stations.
-
-\texttt{subset.swmpr} Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.
-
-\paragraph{Analyze}
-
-\texttt{aggregate.swmpr} Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.
-
-\texttt{aggregate\_metab} Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.
-
-\texttt{ecometab.swmpr} Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.
-
-\texttt{decomp.swmpr} Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \texttt{decompose}.
-Decomposition of monthly or daily trends is possible.
-
-\texttt{decomp\_cj.swmpr} Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \texttt{decompTs} in the wq package.
-Only monthly decomposition is possible.
-
-\texttt{hist.swmpr} Plot a histogram for a \texttt{swmpr} object.
-
-\texttt{lines.swmpr} Add lines to an existing \texttt{swmpr} plot.
-
-\texttt{na.approx.swmpr} Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
-The maximum gap size that is interpolated is defined as a maximum number of records with missing data.
-
-\texttt{plot.swmpr} Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.
-
-\texttt{plot\_metab} Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.
-
-\texttt{plot\_summary} Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.
-
-\texttt{smoother.swmpr} Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.
-
-\paragraph{Miscellaneous}
-
-\texttt{calcKL} Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.
-
-\texttt{map\_reserve} Create a map of all stations in a reserve using the ggmap package.
-
-\texttt{metab\_day} Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.
-
-\texttt{param\_names} Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
-Includes QAQC columns with \texttt{f\_} prefix.
-Used internally in other functions.
-
-\texttt{parser} Parses html returned from \ac{CDMO} web services, used internally in retrieval functions.
-
-\texttt{swmpr} Creates object of \texttt{swmpr} class, used internally in retrieval functions.
-
-\texttt{time\_vec} Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.
-
-% Results and Discussion can be combined.
-\section*{Results}
-
-\section*{Discussion}
+\section*{Applications using the SWMPr package}
+
+The ability to evaluate environmental characteristics between estuaries within the \ac{NERRS} program has been greatly improved using functions in the SWMPr package.
+This section describes two examples of applications using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
+First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04}.
+A comprehensive evaluation of metabolic rates between estuaries has not been conducted using the most recent \ac{SWMP} data.
+The results are presented to facilitate additional research to evaluate factors that influence variation between sites.
+
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} describe the theory and application of the open-water method to estimate ecosystem metabolism using two sites at each of the \ac{NERRS} reserve.
+Time series data included approximately five years of half-hour water quality observations at each site.
+Additionally, the air-sea gas exchange model, as implemented by the current \texttt{ecometab} function (see Ro and Hunt \cite{Ro06} and Thebault et al.
+\cite{Thebault08}), was not incorporated into the initial metabolism estimates such that a constant value for the reaeration coefficient was assumed.
+This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column  The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
+\cite{Caffrey14} for details).
+
+All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November of 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function and then saved again on a local hard drive as a binary .RData file.
+This resulted in a single \texttt{swmpr} object for each site.
+All files were then uploaded to a remote server.
+An R script was executed that retrieved and processed the \texttt{swmpr} objects for each site using the following functions in sequence:
+
+\section*{Summary}
 

### 2015-03-23 swmpr_manu.Rnw
finished metab estimates, need to make plots and describe, need to wrap up applications section

@@
 
-\ac{SWMP} data can be used in R after they are obtained directly from the \ac{CDMO} through an online query or by using the package's retrieval functions.
-Prior to any data request, the site, parameter type, and date ranges need to be identified based on the analysis needs.
+Prior to any data request to the \ac{CDMO}, the site, parameter type, and date ranges need to be identified based on the analysis needs.
 All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
@@
 
-\ac{SWMP} data must be obtained from the \ac{CDMO} prior to organizing and analyzing the information.
-Two approaches can be used.
-First, the data can be obtained outside of R using the graphical query forms on the \ac{CDMO} website (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
-Second, functions from the SWMPr package can be used to import the data directly from the online server using \ac{CDMO} web services (\Cref{tab:retrieve}).
-In the latter case, the IP address for the computer making the request must be registered with \ac{CDMO}.
+Three approaches can be used for retrieving \ac{SWMP} data for import into R.
+First, functions from the SWMPr package can be used to import the data directly from the online server using \ac{CDMO} web services (\Cref{tab:retrieve}).
+To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.
 This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
@@
 <<results = 'asis', echo = FALSE>>=
-funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
+funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'import\\_remote', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   'Import files from a local path.  The files must be in a specific format, specifically those returned from the \\ac{CDMO} using the zip downloads option for a reserve.',
+  'Import site data from an independent server.
+These files have been downloaded from \\ac{CDMO} up to December 2014, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.',
   'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
@@
 
-For larger requests, data are easier to obtain outside of R using the \ac{CDMO} query system and then importing using the \texttt{import\_local} function.
+A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.
+This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in .RData binary format.
+This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.
+The files are hosted via third party web services and include almost all available \ac{SWMP} data for each site up to December 2014.
+Always use the \ac{CDMO} for the most recent data.
+The files can be accessed by copying the URL to a web browser (see the \Cref{sec:supp_info}) or by using the function as follows.
+
+<<eval = F>>=
+## import a file from an independent server
+import_remote('acebbwq')
+@
+
+A third approach is to use the \texttt{import\_local} function to import data into R that are locally available after downloading outside of R (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
+This approach may be best for larger, more customized data requests to \ac{CDMO}.
 Data can be retrieved from the \ac{CDMO} several ways.
@@
 The ability to evaluate environmental characteristics between estuaries within the \ac{NERRS} program has been greatly improved using functions in the SWMPr package.
-This section describes two examples of applications using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
+This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
+
 First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
 The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04}.
-A comprehensive evaluation of metabolic rates between estuaries has not been conducted using the most recent \ac{SWMP} data.
 The results are presented to facilitate additional research to evaluate factors that influence variation between sites.
-
 Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} describe the theory and application of the open-water method to estimate ecosystem metabolism using two sites at each of the \ac{NERRS} reserve.
@@
 
-All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November of 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
 After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, saved again on a local hard drive as a binary .RData file, and uploaded.
@@
 All files were then uploaded to a remote server.
-An R script was executed that retrieved and processed the \texttt{swmpr} objects for each site using the following functions in sequence:
-
-The weather station with the most data was retained if more than one weather station was available for a single reserve.
-
-<<eval = F>>=
-# packages to use
-library(SWMPr)
-library(httr)
-library(XML)
-library(foreach)
-library(doParallel)
-
-# names of files on server
-files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
-files_s3 <- rawToChar(files_s3)
-files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
-files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
-
-# find only active water quality and weather sites
-# IP address must be registed with CDMO 
-meta <- site_codes()
-sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
-meta <- meta[sel, ]
-
-# get wq sites
-wq_sites <- grep('wq$', meta$station_code, value = T)
-
-##
-# process each site
-
-# setup parallel backend for processing
-cl <- makeCluster(8)
-registerDoParallel(cl)
-strt <- Sys.time()
-
-# output
-metabs <- foreach(wq_site = wq_sites) %dopar% {
-
-  library(SWMPr)
-
-  # progress
-  sink('log.txt')
-  cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
-  print(Sys.time() - strt)
-  sink()
-
-  # find corresponding wx station
-  met_site <- substr(wq_site, 1, 3)
-  met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
-  met_site <- gsub('\\.RData$', '', met_site)
-
-  # continue if wx data found
-  if(length(met_site) > 0){
-
-    ##
-    # load the files
-
-    # wq
-    raw_content <- paste0('https://s3.amazonaws.com/swmpalldata/', wq_site, '.RData')
-    raw_content <- httr::GET(raw_content)$content
-    connect <- rawConnection(raw_content)
-    load(connect)
-    wq <- get(wq_site)
-    rm(list = wq_site)
-    close(connect)
-
-    # met
-    raw_content <- paste0('https://s3.amazonaws.com/swmpalldata/', met_site, '.RData')
-    raw_content <- httr::GET(raw_content)$content
-    connect <- rawConnection(raw_content)
-    load(connect)
-    met <- get(met_site)
-    rm(list = met_site)
-    close(connect)
-
-    ## 
-    # combine, subset up to 2013, estimate metabolism
-    dat <- SWMPr::comb(wq, met, method = met_site)
-    dat <- subset(dat, subset = '2014-01-01 0:0', operator = '<')
-    dat <- SWMPr::ecometab(dat)
-
-    dat
-
-  }
-
-}
-names(metabs) <- wq_sites
-save(metabs, file = 'metabs.RData')
-@
+An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined water quality and weather data for each site.
+
+<<eval = F, echo = F>>=
+# # packages to use
+# library(SWMPr)
+# library(httr)
+# library(XML)
+# library(foreach)
+# library(doParallel)
+# 
+# # names of files on server
+# files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
+# files_s3 <- rawToChar(files_s3)
+# files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
+# files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
+# files_s3 <- gsub('\\.RData$', '', files_s3) 
+# 
+# # find only active water quality and weather sites
+# # IP address must be registed with CDMO 
+# meta <- site_codes()
+# sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
+# meta <- meta[sel, ]
+# 
+# # get wq sites from meta, then filter by those on server
+# wq_sites <- grep('wq$', meta$station_code, value = T)
+# wq_sites <- files_s3[gsub('\\.RData$', '', files_s3) %in% wq_sites]
+# 
+# ##
+# # process each site
+# 
+# # setup parallel backend for processing
+# cl <- makeCluster(8)
+# registerDoParallel(cl)
+# strt <- Sys.time()
+# 
+# # output
+# metabs <- foreach(wq_site = wq_sites) %dopar% {
+#  
+#   library(SWMPr)
+#   
+#   # progress
+#   sink('log.txt')
+#   cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
+#   print(Sys.time() - strt)
+#   sink()
+#   
+#   # find corresponding wx station
+#   met_site <- substr(wq_site, 1, 3)
+#   met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
+#   met_site <- gsub('\\.RData$', '', met_site)
+#   
+#   # continue if wx data found
+#   if(length(met_site) > 0){
+#   
+#     ##
+#     # load the files
+#     
+#     # wq
+#     wq <- import_remote(wq_site)
+#     
+#     # met
+#     met <- import_remote(met_site)
+#     
+#     ## 
+#     # combine, subset up to 2013, estimate metabolism, reduce data volume
+#     dat <- comb(wq, met, method = met_site)
+#     dat <- subset(dat, subset = '2014-01-01 0:0', operator = '<')
+#     dat <- ecometab(dat)
+#     dat <- dat[1, ]
+#     
+#     dat
+#     
+#   }
+#   
+# }
+# names(metabs) <- wq_sites
+# save(metabs, file = 'metabs.RData')
+
+# make a plot.....
+
+@
+
+The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
+The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
+Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
+Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.
+Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.
+The trend direction is blue for decreasing and red for increasing.
+The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.
+The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations.
+The drop down menus can be used to select the station, date range, and parameter for plotting.
+Plots are based on daily averages for each parameter.
+The data used for each application are similar to those used to estimate ecosystem metabolism described above.
+
+% add figs to show each....
 
 \section*{Summary}
+
+Add some stuff....
 
@@
 % Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
-\subsection*{Trends in SWMP parameters}
+\subsection*{Trends in \ac{SWMP} parameters}
 \label{swmp_trends}
-\href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}
-\bigskip
-This widget is an interactive tool to evaluate trends in SWMP data within and between sites.
-Trends are described by an increase or decrease in values over time using a simple linear regression of summarized data.
-The regression for each station can be viewed by clicking on a map location.
-Trends at each station are plotted as circles that identify the direction and significance of the trend.
-The trend direction is blue for decreasing and red for increasing.
-The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.
-
-\subsection*{Monthly and annual summary of SWMP parameters}
+Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}
+
+\subsection*{Monthly and annual summary of \ac{SWMP} parameters}
 \label{swmp_summary}
-\href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}
-
-This interactive widget provides graphical summaries of water quality, weather, and nutrient station data from SWMP.
-The drop down menus can be used to select the station, date range, and parameter for plotting.
-The raw data used for plotting include all SWMP records from the earliest date at each station after processing to remove QAQC flags.
-The data were downloaded from the \href{http://cdmo.baruch.sc.edu/}{CDMO web services} on November 25th, 2014 and include observations up to that date.
-Plots are based on daily averages for each parameter.
-Cumulative precipitation data are based on the daily maximum.
+Summarize trends in \ac{SWMP} parameters for an individual site: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}
+
+\subsection*{\ac{SWMP} data as R binary files}
+
+Quickly access all \ac{SWMP} data: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}
+
+This is an online link for all \ac{SWMP} data from January 1994 to December 2014 as binary .RData files.
+Separate files are available for each site and parameter type, e.g., acebbnut.RData includes all available nutrient data for the Big Bay site at ACE Basin.
+Individual files can be downloaded by pasting the file name to the URL, e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, then using the \texttt{load} function to import the file in the current R workspace.
+Files can also be downloaded directly within R using \texttt{import\_remote} function.
 
@@
 Environmental Protection Agency.
+The use of trade names or products does not constitute endorsement by the US Government.
 

### 2015-03-23 swmpr_manu.tex
finished metab estimates, need to make plots and describe, need to wrap up applications section

@@
 The files must be in a specific format, specifically those returned from the \ac{CDMO} using the zip downloads option for a reserve.\tabularnewline
+\texttt{import\_remote}&Import site data from an independent server.
+These files have been downloaded from \ac{CDMO} up to December 2014, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.\tabularnewline
 \texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.
@@
 
-For larger requests, data are easier to obtain outside of R using the \ac{CDMO} query system and then importing using the \texttt{import\_local} function.
+A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.
+This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in .RData binary format.
+This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.
+The files are hosted via third party web services and include almost all available \ac{SWMP} data for each site up to December 2014.
+Always use the \ac{CDMO} for the most recent data.
+The files can be accessed by copying the URL to a web browser (see the \Cref{sec:supp_info}) or by using the function as follows.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## import a file from an independent server}
+\hlkwd{import_remote}\hlstd{(}\hlstr{'acebbwq'}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+A third approach is to use the \texttt{import\_local} function to import data into R that are locally available after downloading outside of R (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
+This approach may be best for larger, more customized data requests to \ac{CDMO}.
 Data can be retrieved from the \ac{CDMO} several ways.
@@
 
-All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November of 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function and then saved again on a local hard drive as a binary .RData file.
+All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, saved again on a local hard drive as a binary .RData file, and uploaded.
 This resulted in a single \texttt{swmpr} object for each site.
 All files were then uploaded to a remote server.
-An R script was executed that retrieved and processed the \texttt{swmpr} objects for each site using the following functions in sequence:
+An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined water quality and weather data for each site.
+
+The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
+The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
+Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
+Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.
+Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.
+The trend direction is blue for decreasing and red for increasing.
+The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.
+The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations.
+The drop down menus can be used to select the station, date range, and parameter for plotting.
+Plots are based on daily averages for each parameter.
+The data used for each application are similar to those used to estimate ecosystem metabolism described above.
+
+% add figs to show each....
 
 \section*{Summary}
+
+Add some stuff....
 

### 2015-03-24 swmpr_manu.Rnw
processed all new cdmo data for metabolism, added plot

@@
 A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.
-This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in .RData binary format.
+This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
 This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.
@@
 Always use the \ac{CDMO} for the most recent data.
-The files can be accessed by copying the URL to a web browser (see the \Cref{sec:supp_info}) or by using the function as follows.
+The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.
 
@@
 This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
-Processed verions of the raw data are included with the package as binary data files (.RData) to decrease processing times with the examples.
+Processed verions of the raw data are included with the package as binary data files (RData) to decrease processing times with the examples.
 Information for each binary file can be viewed as follows.
@@
 
-Estimates of ecosystem metabolism provide a useful measure of overall system productivity.
-These estimates are commonly used to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method \cite{Odum56} is a common approach to quantify net ecosystem metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model (see Ro and Hunt \cite{Ro06}, Thebault et al.
+Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model (see Ro and Hunt \cite{Ro06}, Thebault et al.
 \cite{Thebault08}).
@@
 
-The \texttt{ecometab} function is used to implement an adaptation of the open-water method \cite{Odum56,Caffrey14}.
+The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.
 Several assumptions must be met for a valid interpretation of the results.
@@
 The metadata should be consulted for exact height.
-Other assumptions may apply and the user should consult the relevant literature.
-All calculations within the function are done using molar units (e.g., mmol O$^2$ m$^{-3}$).
-The output can be returned as mass units by changing the default argument.
-Input data must be in standard mass units for \ac{DO} (mg L-1)
+Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).
 
@@
 First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04}.
-The results are presented to facilitate additional research to evaluate factors that influence variation between sites.
-Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} describe the theory and application of the open-water method to estimate ecosystem metabolism using two sites at each of the \ac{NERRS} reserve.
-Time series data included approximately five years of half-hour water quality observations at each site.
-Additionally, the air-sea gas exchange model, as implemented by the current \texttt{ecometab} function (see Ro and Hunt \cite{Ro06} and Thebault et al.
-\cite{Thebault08}), was not incorporated into the initial metabolism estimates such that a constant value for the reaeration coefficient was assumed.
-This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column  The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} using the most recent \ac{SWMP} data.
+Additionally, previous methods to evaluate metabolism at \ac{NERRS} reserves did not incorporate \textit{in situ} weather data to estimate the air-sea gas exchange model.
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \ac{NERRS} reserves.
+Additionally, the air-sea gas exchange model assumed a constant value for the reaeratio coefficientt.
+This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column.
+The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
 \cite{Caffrey14} for details).
 
-All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, saved again on a local hard drive as a binary .RData file, and uploaded.
-This resulted in a single \texttt{swmpr} object for each site.
+Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
+This resulted in a single \texttt{swmpr} object for each paramter at each site.
 All files were then uploaded to a remote server.
-An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined water quality and weather data for each site.
-
-<<eval = F, echo = F>>=
+An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site.
+
+<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\ac{NERRS} reserve.
+Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.
+Two sites were chosen from each reserve that had the longest available time series.', fig.height = 6, fig.width = 8>>=
 # # packages to use
@@
 # 
-# ##
-# # process each site
-# 
 # # setup parallel backend for processing
@@
 # 
-# # output
+# # process
 # metabs <- foreach(wq_site = wq_sites) %dopar% {
@@
 #     
+#     # if > 1 wx site, pick the one with more obs
+#     if(length(met_site) > 1){
+# 
+#       met_list <- vector('list', length(met_site))
+#       names(met_list) <- met_site
+#       for(met in met_site){
+#       
+#         met_tmp <- import_remote(met)
+#         met_list[[met]] <- met_tmp
+#       
+#       }
+#     
+#       # pick the one with most obs
+#       met_most <- which.max(unlist(lapply(met_list, nrow)))
+#       met <- met_list[met_most][[1]]
+# 
+#     # otherwise load only one
+#     } else {
+#     
+#       # met
+#       met <- import_remote(met_site)
+#     
+#     }
+#     
 #     ##
-#     # load the files
+#     # load the wq file
 #     
@@
 #     
-#     # met
-#     met <- import_remote(met_site)
-#     
 #     ## 
-#     # combine, subset up to 2013, estimate metabolism, reduce data volume
-#     dat <- comb(wq, met, method = met_site)
-#     dat <- subset(dat, subset = '2014-01-01 0:0', operator = '<')
+#     # combine, estimate metabolism, reduce data volume
+#     dat <- comb(wq, met, method = attr(met, 'station'))
 #     dat <- ecometab(dat)
@@
 
-# make a plot.....
+# # take all metab estimates and summarize by year
+# # done for two stations at each reserve with the longest time series
+# data(metabs)
+# 
+# metabs <- lapply(metabs, attr, 'metabolism')
+# metabs <- do.call('rbind', metabs)
+# metabs$site <- gsub('\\.[0-9]*$', '', as.character(row.names(metabs)))
+# row.names(metabs) <- 1:nrow(metabs)
+# 
+# library(dplyr)
+# library(tidyr)
+# metabs <- select(metabs, date, Pg, Rt, NEM, site) %>% 
+#   mutate(date = as.numeric(strftime(date, '%Y'))) %>% 
+#   gather('metab', 'value', 2:4) %>% 
+#   group_by(date, site, metab) %>% 
+#   summarize(value = mean(value, na.rm = T)) %>% 
+#   group_by(site, metab) %>% 
+#   na.omit %>% 
+#   mutate(reserve = substr(site, 1, 3)) %>% 
+#   spread(metab, value)
+# 
+# # select two sites from each reserve, those with longest records
+# metabs <- split(metabs, metabs$reserve)
+# metabs <- lapply(metabs, 
+#   function(x){
+#     
+#     num_yrs <- aggregate(date ~ site, x, length)
+#     num_yrs <- num_yrs[order(num_yrs$date, decreasing = T), ]
+#     
+#     if(nrow(num_yrs) >1){
+#       site_sel <- num_yrs$site[c(1, 2)]
+#     } else {
+#       site_sel <- num_yrs$site[1]
+#     }
+#     
+#     out <- x[x$site %in% site_sel, ]
+#     return(out)
+#     
+#   }
+#   
+# )
+# metabs <- do.call('rbind', metabs)
+# 
+# metabs_sum <- metabs
+# save(metabs_sum, file = 'data/metabs_sum.RData')
+
+data(metabs_sum)
+
+library(dplyr)
+library(tidyr)
+
+# summarize for annual means, get conf ints
+metabs_plo <- gather(metabs_sum, 'metab', 'value', 4:6) %>% 
+  group_by(site, metab) %>% 
+  summarize(
+    margs = qt(1 - 0.05/2, length(value) - 1) * sd(value)/sqrt(length(value)),
+    value = mean(value)
+    ) %>% 
+  mutate(
+    value = 0.032 * value,
+    margs = 0.032 * margs,
+    upper = value + margs, 
+    lower = value - margs
+    )
+
+# plot prep
+to_plo <- data.frame(metabs_plo)
+
+#reassign factors for ranking in plot
+sort_val <- order(to_plo[to_plo$metab == 'Pg', 'value'])
+to_plo$site <- factor(gsub('wq$', '', to_plo$site))
+to_plo$site <- factor(
+  to_plo$site,
+  levels = levels(to_plo$site)[sort_val],
+  labels = levels(to_plo$site)[sort_val]
+  )
+new_labs <- c('Mean net annual metabolism', 'Mean annual gross production', 'Mean annual total respiration')
+to_plo$metab <- factor(to_plo$metab, levels = c('NEM', 'Pg', 'Rt'), labels = new_labs)
+
+ylabs <- expression(paste('g ', O[2],' ', m^-2,' ', d^-1))
+p <- ggplot(to_plo, 
+    aes(x = site, y = value, group = metab, fill = metab)
+  ) + 
+  geom_bar(stat = 'identity', width = 1, colour = 'black') + 
+  scale_y_continuous(ylabs) + 
+  facet_wrap( ~ metab, ncol = 1, scales = 'free_y') + 
+  scale_x_discrete(name = element_blank()) + 
+  theme_classic() + 
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, 
+      size = 8),
+    legend.position = 'none'
+    ) +
+  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0)
+print(p)
+
+@
+
+% table summary of trends
+<<eval = F, cache = T, echo = F>>=
+
+data(metabs_sum)
+
+library(dplyr)
+library(tidyr)
+
+# summarize for annual means, get conf ints
+metabs_tab <- gather(metabs_sum, 'metab', 'value', 4:6)
+metabs_tab$yr_grp <- '1995 - 2004'
+metabs_tab$yr_grp[metabs_tab$year > 2004] <- '2005-2014'
+metabs_tab <- mutate(metabs_tab, value = 0.032 * value) %>% 
+  group_by(metabs_tab, site, metab, yr_grp)
+
+# summarize trends
 
@@
 
-% add figs to show each....
-
 \section*{Summary}
@@
 
-Quickly access all \ac{SWMP} data: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}
-
-This is an online link for all \ac{SWMP} data from January 1994 to December 2014 as binary .RData files.
-Separate files are available for each site and parameter type, e.g., acebbnut.RData includes all available nutrient data for the Big Bay site at ACE Basin.
-Individual files can be downloaded by pasting the file name to the URL, e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, then using the \texttt{load} function to import the file in the current R workspace.
-Files can also be downloaded directly within R using \texttt{import\_remote} function.
+Quickly access all \ac{SWMP} data from 1995 through 2014.
+See the help file for \texttt{import\_remote} for more information: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}
 

### 2015-03-24 swmpr_manu.tex
processed all new cdmo data for metabolism, added plot

@@
 First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04}.
-The results are presented to facilitate additional research to evaluate factors that influence variation between sites.
-Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} describe the theory and application of the open-water method to estimate ecosystem metabolism using two sites at each of the \ac{NERRS} reserve.
-Time series data included approximately five years of half-hour water quality observations at each site.
-Additionally, the air-sea gas exchange model, as implemented by the current \texttt{ecometab} function (see Ro and Hunt \cite{Ro06} and Thebault et al.
-\cite{Thebault08}), was not incorporated into the initial metabolism estimates such that a constant value for the reaeration coefficient was assumed.
-This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column  The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} using the most recent \ac{SWMP} data.
+Additionally, previous methods to evaluate metabolism at \ac{NERRS} reserves did not incorporate \textit{in situ} weather data to estimate the air-sea gas exchange model.
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \ac{NERRS} reserves.
+Additionally, the air-sea gas exchange model assumed a constant value for the reaeratio coefficientt.
+This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column.
+The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
 \cite{Caffrey14} for details).
 
-All water quality and weather observations for all \ac{NERRS} sites were obtained through a bulk data request in November 2014 using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, saved again on a local hard drive as a binary .RData file, and uploaded.
-This resulted in a single \texttt{swmpr} object for each site.
+Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
+After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
+This resulted in a single \texttt{swmpr} object for each paramter at each site.
 All files were then uploaded to a remote server.
-An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined water quality and weather data for each site.
+An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/metab_plo}
+
+}
+
+\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \ac{NERRS} reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \ac{NERRS} reserve.
+Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.
+Two sites were chosen from each reserve that had the longest available time series.\label{fig:metab_plo}}
+\end{figure}
+
+\end{knitrout}
+
+% table summary of trends
 

### 2015-03-25 swmpr_manu.Rnw
finished metabolism results and applications section

@@
 % Version 3.1 February 2015
-%
-% To compile to pdf, run:
-% latex plos.template
-% bibtex plos.template
-% latex plos.template
-% latex plos.template
-% dvipdf plos.template
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
 %
@@
 % For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
 %
@@
 
-The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).
+The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (see the \ac{CDMO} descriptions \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).
 Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
@@
 
-Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with ), \texttt{qaqc}.
+Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.
 The \texttt{qaqcchk} function can be used to view this information.
-Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.
 
@@
 
-A subset method added to the existing \texttt{subset} function is available for \texttt{swmpr} objects.
+A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.
 This function is used to subset the data by date and/or a selected parameter.
@@
 These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
-Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
 The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
@@
 Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
-This resulted in a single \texttt{swmpr} object for each paramter at each site.
+All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
+This resulted in a single \texttt{swmpr} object for each parameter at each site.
 All files were then uploaded to a remote server.
-An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site.
-
+An additional \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site to esimate metabolism.
+For brevity, only two sites at each reserve were evaluated.
+Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
+All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
+Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
+More importantly, the use of the data retreival, synthesis, and analysis functions to create the results shown in \Cref{fig:metab_plo,tab:metab_tab} illustrates the utility provided by the SWMPr package.
+
+% metabolism figure
 <<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\ac{NERRS} reserve.
 Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.
-Two sites were chosen from each reserve that had the longest available time series.', fig.height = 6, fig.width = 8>>=
+Two sites were chosen from each reserve that had the longest available time series.
+Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.', fig.height = 6, fig.width = 8, out.width = '\\textwidth'>>=
 # # packages to use
@@
 
+# get region
+data(stat_locs)
+names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
+metabs_plo$site <- gsub('wq$', '', metabs_plo$site)
+metabs_plo <- left_join(metabs_plo, stat_locs, by = 'site')
+region <- rep('Pacific', nrow(metabs_plo))
+region[metabs_plo$longitude > -110] <- 'Gulf'
+region[metabs_plo$latitude > 28 & metabs_plo$longitude > -84] <- 'Atlantic'
+
+# combine, select, sort
+metabs_plo <- select(metabs_plo, -c(latitude, longitude, gmt_off))
+metabs_plo$region <- region
+metabs_plo <- arrange(metabs_plo, region, site)
+
 # plot prep
@@
 #reassign factors for ranking in plot
-sort_val <- order(to_plo[to_plo$metab == 'Pg', 'value'])
-to_plo$site <- factor(gsub('wq$', '', to_plo$site))
+sort_val <- to_plo[to_plo$metab == 'Pg', c('region', 'value', 'site')]
+sort_val <- order(sort_val$region, sort_val$value)
+to_plo$site <- factor(to_plo$site)
 to_plo$site <- factor(
@@
   )
-new_labs <- c('Mean net annual metabolism', 'Mean annual gross production', 'Mean annual total respiration')
+new_labs <- c('Mean net metabolism', 'Mean gross production', 'Mean total respiration')
 to_plo$metab <- factor(to_plo$metab, levels = c('NEM', 'Pg', 'Rt'), labels = new_labs)
@@
 ylabs <- expression(paste('g ', O[2],' ', m^-2,' ', d^-1))
+library(ggplot2)
 p <- ggplot(to_plo, 
@@
   scale_y_continuous(ylabs) + 
-  facet_wrap( ~ metab, ncol = 1, scales = 'free_y') + 
+  facet_grid(metab ~ region, scales = 'free') + 
   scale_x_discrete(name = element_blank()) + 
-  theme_classic() + 
+  theme_bw() + 
   theme(
     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, 
-      size = 8),
+      size = 7),
     legend.position = 'none'
@@
 % table summary of trends
-<<eval = F, cache = T, echo = F>>=
+<<eval = T, results = 'asis', cache = T, echo = F>>=
 
@@
 
-# summarize for annual means, get conf ints
+# summarize by time periods
 metabs_tab <- gather(metabs_sum, 'metab', 'value', 4:6)
-metabs_tab$yr_grp <- '1995 - 2004'
-metabs_tab$yr_grp[metabs_tab$year > 2004] <- '2005-2014'
+metabs_tab$yr_grp <- '1995-2004'
+metabs_tab$yr_grp[metabs_tab$date > 2004] <- '2005-2014'
 metabs_tab <- mutate(metabs_tab, value = 0.032 * value) %>% 
-  group_by(metabs_tab, site, metab, yr_grp)
-
-# summarize trends
-
-@
-
-The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
+  group_by(site, metab, yr_grp) %>% 
+  arrange(yr_grp, date, metab) %>% 
+  summarize(ave = round(mean(value, na.rm = TRUE), 2)) %>% 
+  spread(yr_grp, ave) 
+trends <- apply(metabs_tab, 1, function(x){
+
+  val <- as.numeric(x[['2005-2014']])
+  trnd <- val > as.numeric(x[['1995-2004']])
+
+  if(is.na(trnd)) return(val)
+  if(trnd) return(paste0('{\\bf ', val, '}'))
+  if(!trnd) return(paste0('{\\it ', val, '}'))
+
+})
+metabs_tab[, '2005-2014'] <- trends
+metabs_tab <- gather(metabs_tab, 'period', 'value', 3:4) %>% 
+  unite('var', metab, period, sep = ' ') %>% 
+  spread(var, value) %>% 
+  mutate(site = gsub('wq$', '', site))
+
+# get region
+data(stat_locs)
+names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
+metabs_tab <- left_join(metabs_tab, stat_locs, by = 'site')
+region <- rep('Pacific', nrow(metabs_tab))
+region[metabs_tab$longitude > -110] <- 'Gulf'
+region[metabs_tab$latitude > 28 & metabs_tab$longitude > -84] <- 'Atlantic'
+
+# combine, select, sort
+metabs_tab <- select(metabs_tab, -c(latitude, longitude, gmt_off)) %>% 
+  mutate(region = region) %>% 
+  arrange(region, site)
+
+nrgroups <- as.numeric(table(metabs_tab$region))
+# table
+
+cap <- 'Trends in metabolism for two sites at each of the \\ac{NERRS} reserves.
+Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
+Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
+Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.'
+foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
+
+library(Hmisc)
+latex(
+  metabs_tab[, !names(metabs_tab) %in% c('site', 'region')], 
+  file = '', 
+  caption = cap, 
+  cgroup = c('NEM\\textsuperscript{{\\it a}}', 'Pg', 'Rt'), 
+  n.cgroup = c(2, 2, 2), 
+  colheads = rep(c('1995-2004', '2005-2014'), 3),
+  rowlabel = 'Site', 
+  rowname = metabs_tab$site,
+  digits = 2,
+  insert.bottom = foots,
+  rgroup = unique(metabs_tab$region),
+  n.rgroup = nrgroups,
+  label = 'tab:metab_tab',
+  size = 'footnotesize'
+  )
+
+@
+
+The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves that were created using the SWMPr package (see the \nameref{supp_info}).
 The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.

### 2015-03-25 swmpr_manu.tex
finished metabolism results and applications section

@@
 Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
-This resulted in a single \texttt{swmpr} object for each paramter at each site.
+All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
+This resulted in a single \texttt{swmpr} object for each parameter at each site.
 All files were then uploaded to a remote server.
-An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site.
-
+An additional \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site to esimate metabolism.
+For brevity, only two sites at each reserve were evaluated.
+Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
+All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
+Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
+More importantly, the use of the data retreival, synthesis, and analysis functions to create the results shown in \Cref{fig:metab_plo,tab:metab_tab} illustrates the utility provided by the SWMPr package.
+
+% metabolism figure
 \begin{knitrout}
@@
 
-{\centering \includegraphics[width=\maxwidth]{figure/metab_plo}
+{\centering \includegraphics[width=\textwidth]{figure/metab_plo}
 
@@
 Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.
-Two sites were chosen from each reserve that had the longest available time series.\label{fig:metab_plo}}
+Two sites were chosen from each reserve that had the longest available time series.
+Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.\label{fig:metab_plo}}
 \end{figure}
@@
 % table summary of trends
-
-The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
+%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{{\\it a}}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize")%
+\begin{table}[!tbp]
+{\footnotesize
+\caption{Trends in metabolism for two sites at each of the \ac{NERRS} reserves.
+Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
+Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
+Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.\label{tab:metab_tab}}
+\begin{center}
+\begin{tabular}{lllcllcll}
+\hline\hline
+\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{{\it a}}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline
+\cline{2-3} \cline{5-6} \cline{8-9}
+\multicolumn{1}{l}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}\tabularnewline
+\hline
+{\bfseries Atlantic}&&&&&&&&\tabularnewline
+~~acebb&-0.04&{\bf -0.03}&&5.11&{\it 3.81}&&-5.15&{\bf -3.82}\tabularnewline
+~~acesp&-0.26&{\bf -0.1}&&4.68&{\bf 4.93}&&-4.94&{\it -5.03}\tabularnewline
+~~cbmip&-0.14&{\it -0.17}&&1.02&{\bf 1.91}&&-1.16&{\it -2.08}\tabularnewline
+~~cbmmc&-0.43&{\bf -0.24}&&6.06&{\bf 10.13}&&-6.48&{\it -10.38}\tabularnewline
+~~cbvgi&0.13&{\it 0}&&6.56&{\it 4.47}&&-6.43&{\bf -4.46}\tabularnewline
+~~cbvtc&-0.15&{\bf -0.07}&&5.58&{\bf 6.07}&&-5.73&{\it -6.14}\tabularnewline
+~~delbl&-0.59&{\bf -0.16}&&6.68&{\it 6.51}&&-7.27&{\bf -6.67}\tabularnewline
+~~delsl&-0.37&{\bf -0.33}&&2.84&{\bf 3.22}&&-3.21&{\it -3.55}\tabularnewline
+~~grblr&&-0.02&&&1.28&&&-1.31\tabularnewline
+~~grbor&&-0.04&&&3.16&&&-3.2\tabularnewline
+~~gtmfm&-0.09&{\bf -0.04}&&3.36&{\bf 4.19}&&-3.45&{\it -4.23}\tabularnewline
+~~gtmpc&-0.48&{\bf -0.45}&&1.95&{\bf 2.91}&&-2.43&{\it -3.37}\tabularnewline
+~~hudsk&0.02&{\it 0.02}&&-0.36&{\it -0.39}&&0.38&{\bf 0.41}\tabularnewline
+~~hudtn&-0.03&{\it -0.03}&&3.75&{\it 3.43}&&-3.78&{\bf -3.46}\tabularnewline
+~~jacb6&0.05&{\it 0.03}&&3.41&{\bf 3.93}&&-3.36&{\it -3.89}\tabularnewline
+~~jacb9&0.01&{\it -0.02}&&1.8&{\bf 3.1}&&-1.79&{\it -3.12}\tabularnewline
+~~narnc&-0.64&{\bf -0.53}&&12.64&{\it 11.33}&&-13.25&{\bf -11.86}\tabularnewline
+~~narpc&-0.14&{\bf -0.03}&&7.36&{\it 6.3}&&-7.5&{\bf -6.32}\tabularnewline
+~~niwcb&-0.46&{\bf -0.28}&&5.5&{\bf 6.1}&&-5.95&{\it -6.38}\tabularnewline
+~~niwdc&-0.13&{\it -0.2}&&7.28&{\bf 8.02}&&-7.41&{\it -8.23}\tabularnewline
+~~nocec&-0.05&{\it -0.06}&&4.86&{\bf 4.88}&&-4.92&{\it -4.94}\tabularnewline
+~~nocrc&-0.19&{\it -0.19}&&5.93&{\it 5.31}&&-6.12&{\bf -5.5}\tabularnewline
+~~owcbr&-0.17&{\bf -0.09}&&1.45&{\bf 1.9}&&-1.62&{\it -1.99}\tabularnewline
+~~owcwm&-0.36&{\bf -0.19}&&7.02&{\it 5.36}&&-7.38&{\bf -5.54}\tabularnewline
+~~saphd&-1.28&{\bf -1}&&1.89&{\bf 3.77}&&-3.17&{\it -4.77}\tabularnewline
+~~sapld&-0.16&{\it -0.43}&&2.32&{\bf 3.56}&&-2.49&{\it -3.99}\tabularnewline
+~~welht&-0.07&{\bf -0.04}&&0.36&{\bf 0.63}&&-0.43&{\it -0.67}\tabularnewline
+~~welin&-1.87&{\it -2.49}&&3.61&{\it 2.68}&&-5.48&{\bf -5.18}\tabularnewline
+~~wqbcr&-0.01&{\bf 0.01}&&8.13&{\it 7.41}&&-8.14&{\bf -7.4}\tabularnewline
+~~wqbmh&0.02&{\it 0}&&2.82&{\it 2.76}&&-2.79&{\bf -2.75}\tabularnewline
+\hline
+{\bfseries Gulf}&&&&&&&&\tabularnewline
+~~apaeb&-0.35&{\it -0.35}&&4.19&{\bf 4.31}&&-4.54&{\it -4.67}\tabularnewline
+~~apaes&-0.71&{\bf -0.67}&&3.35&{\it 3.05}&&-4.06&{\bf -3.73}\tabularnewline
+~~gndbc&-1.02&{\bf -0.88}&&3.5&{\bf 4.3}&&-4.51&{\it -5.18}\tabularnewline
+~~gndbh&-1.81&{\it -2}&&2.19&{\it 1.77}&&-4&{\bf -3.77}\tabularnewline
+~~job09&-0.34&{\bf -0.25}&&5.32&{\bf 5.95}&&-5.66&{\it -6.2}\tabularnewline
+~~job10&-0.15&{\it -0.15}&&3.58&{\bf 4.03}&&-3.72&{\it -4.2}\tabularnewline
+~~marab&&-0.02&&&3.38&&&-3.4\tabularnewline
+~~marce&&-0.01&&&3.62&&&-3.63\tabularnewline
+~~rkbfb&-0.28&{\it -0.37}&&5.28&{\bf 6.25}&&-5.56&{\it -6.62}\tabularnewline
+~~rkbfu&-0.33&{\it -0.41}&&2.95&{\bf 3.24}&&-3.28&{\it -3.64}\tabularnewline
+~~wkbfr&-0.29&{\bf -0.14}&&6.27&{\bf 6.35}&&-6.57&{\bf -6.49}\tabularnewline
+~~wkbwb&-0.29&{\bf -0.16}&&6.81&{\it 6.54}&&-7.1&{\bf -6.69}\tabularnewline
+\hline
+{\bfseries Pacific}&&&&&&&&\tabularnewline
+~~elkap&0.03&{\bf 0.11}&&14.7&{\bf 19.84}&&-14.67&{\it -19.74}\tabularnewline
+~~elknm&-0.24&{\bf -0.08}&&11.95&{\bf 16.37}&&-12.18&{\it -16.45}\tabularnewline
+~~kachd&0.2&{\it -0.66}&&7.31&{\it 3.2}&&-7.12&{\bf -3.86}\tabularnewline
+~~kacsd&-0.06&{\it -0.26}&&5.15&{\it 4.2}&&-5.21&{\bf -4.45}\tabularnewline
+~~pdbbp&0.01&{\bf 0.08}&&6.29&{\bf 7.31}&&-6.27&{\it -7.23}\tabularnewline
+~~pdbby&-0.07&{\bf -0.01}&&8.39&{\it 5.92}&&-8.47&{\bf -5.93}\tabularnewline
+~~sfbcc&&-0.24&&&3.56&&&-3.79\tabularnewline
+~~sfbgc&&-0.55&&&7.32&&&-7.87\tabularnewline
+~~sosva&-0.2&{\bf -0.05}&&6.98&{\bf 7.83}&&-7.18&{\it -7.87}\tabularnewline
+~~soswi&-0.28&{\bf -0.08}&&4.41&{\bf 4.74}&&-4.69&{\it -4.82}\tabularnewline
+~~tjrbr&-1.33&{\bf -0.17}&&7.72&{\it 7.06}&&-9.05&{\bf -7.22}\tabularnewline
+~~tjros&-0.26&{\it -0.34}&&10.18&{\bf 11.61}&&-10.44&{\it -11.96}\tabularnewline
+\hline
+\end{tabular}\end{center}}
+
+\footnotesize{\textsuperscript{{\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}
+
+The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves that were created using the SWMPr package (see the \nameref{supp_info}).
 The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.

### 2015-03-25 swmpr_manu.Rnw
finished complete first draft

@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white')
 options(replace.assign=TRUE,width=70,digits=1)
@@
 However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
-\ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity of data provided by \ac{SWMP} \cite{SWMP14}.
+\ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.
 
@@
 Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.
-The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant addressing common challenges working with large datasets.
+The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.
 
@@
 \section*{Applications using the SWMPr package}
+\label{swmp_apps}
 
@@
 
-The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves that were created using the SWMPr package (see the \nameref{supp_info}).
+The second and third examples are two Shiny web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
 The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
@@
 
-Add some stuff....
+The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.
+Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessment of site-specific characteristics and more informed decisions to manage environmental resources.
+The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 300 stations within the 28 estuaries of the \acl{NERRS}.
+This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
+Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package to address common challenges working with large datasets.
+
+The SWMPr package provides several functions to retrieve, organize, and analyze \ac{SWMP} data to more effectively evaluate environmental trends.
+The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (\Cref{tab:retrieve}).
+Established \ac{QAQC} methods and data processing techniques are also enhanced with \ac{SWMP} by functions that filter observations for different \ac{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
+Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data, respectively.
+Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.
+In particular, the \texttt{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \cite{Odum56,Caffrey14}.
+The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the \ac{NERRS} estuaries using recent data to evaluate trends over time.
+Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
+Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \ac{SWMP} and \ac{NERRS}.
 
@@
 
-% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
 \subsection*{Trends in \ac{SWMP} parameters}

### 2015-03-25 swmpr_manu.tex
finished complete first draft

@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{alltt}
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[!ht]
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!ht]
 

### 2015-03-29 swmpr_manu.Rnw
proofing, finished up to analysis portion

@@
 %
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
-%
-% Please be sure to include all portions of an equation in the math environment.
-%
-% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
+% % % % % % % % % % % % % % % % % % % % % % %
 
@@
 \usepackage{cleveref}
-\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
+\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
+% \usepackage[hidelinks]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white')
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE)
 options(replace.assign=TRUE,width=70,digits=1)
@@
 \vspace*{0.35in}
+\urlstyle{same}
 
@@
 {\Large
-\textbf\newline{SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries}
+\textbf\newline{SWMPr: An R package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
 }
@@
 \end{flushleft}
-% Please keep the abstract below 300 words
+
 \section*{Abstract}
@@
 One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.
-This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries across the United States.
-\ac{SWMP} data have been used in a variety applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
-However, simple tools for processing and evaluating the large and increasing quantity of data provided by the monitoring network have prevented large-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
+This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries.
+\ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
+However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
 We describe a new open-source software package, SWMPr, developed in program R for use with \ac{SWMP} environmental data.
@@
 These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
-Applications from automated remote sensors are numerous with notable examples including prediction of harmful algal blooms and toxicants in aquatic systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and automated detection of airborne chemical warfare agents \cite{Sanders01}.
+Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \cite{NDBC15}.
 Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.
-However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are more often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.
+However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.
 Growing concerns over the use of adequate quality assurance and control methods, challenges for synthesis and interpretation, and increased emphasis on exploratory-based analytical techniques have characterized applications of data from automated monitoring programs \cite{Campbell13,Millie13}.
 
-The \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}) is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.
+The \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}} is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.
 The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acf{SWMP} program was implemented in 1995 at over 300 stations at each of the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
-The \ac{SWMP} network has provided a continous source of data collected at near real-time at each of the reserves with the intent to evaluate natural and anthrogogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
-These data have been useful for evaluating relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
+As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 300 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
+These data have been applied in evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
 However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
@@
 This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
-SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).
+SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acl{SWMP}.
 Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
-Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.
+Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.
 The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.
@@
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the \ac{CDMO} web portal, where multiple \ac{QAQC} measures are used to screen the information.
-The final data include all observations with relevant \ac{QAQC} flags indicating the appropriate qualifier (\href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{view codes}).
-
-The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data and information management.
+All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.
+The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.
+
+The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data management.
 As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
-Estuary data must be obtained from \ac{CDMO} prior to using most of the functions within the SWMPr package.
-In most cases, the analysis needs will typically define the location, date range, and parameters of interest that need to be obtained from \ac{CDMO}.
+Prior to any data request to the \ac{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs.
 All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
-For example, `apaebwq' is the water quality identifier for the East Bay station at the Apalachicola reserve.
+For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').
 Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
-
-Prior to any data request to the \ac{CDMO}, the site, parameter type, and date ranges need to be identified based on the analysis needs.
-All reserve names, stations, and date ranges for the water quality, weather, or nutrients data can be viewed on the \ac{CDMO} \href{http://cdmo.baruch.sc.edu/}{website}.
+All reserve names, stations, and date ranges for each parameter type can be viewed on the \ac{CDMO} website.
 Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered by \ac{CDMO} staff before using the data retrieval functions in SWMPr.
-\href{http://cdmo.baruch.sc.edu/webservices.cfm}{Web services} are provided by \ac{CDMO} to provide direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
+As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.
+Web services\footnote{\url{http://cdmo.baruch.sc.edu/webservices.cfm}} are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
 The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.
-For example, the \texttt{all\_params} function in SWMPr uses the \texttt{exportAllParamsXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.
-The text below describes the data retrieval functions in more detail.
+For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.
+The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.
 
@@
 
-The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}).
-The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.
-The package is loaded in the current workspace by using the \texttt{library} command.
+The SWMPr package was developed for use with the R ($\geq$v3.0.0) statistical programming language \cite{RDCT14}.
+The SWMPr package can be installed from GitHub\footnote{\url{https://github.com/fawda123/SWMPr}} by executing the following commands at the R console prompt.
+The package is loaded in the workspace using the \texttt{library} command.
 
@@
 
-The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.
-Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
-Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.
-An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.
-The following describes a general approach for using each category of functions based on a standard data workflow.
+The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.
+Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.
+Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.
+An additional group of `miscellaneous' functions are included as helpers for the main functions.
+The following describes a general approach for using each category of functions.
 
@@
 
-Three approaches can be used for retrieving \ac{SWMP} data for import into R.
-First, functions from the SWMPr package can be used to import the data directly from the online server using \ac{CDMO} web services (\Cref{tab:retrieve}).
-To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.
-This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.
-The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
+Three approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
+First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services.
+To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.\footnote{Follow instructions here: \url{http://cdmo.baruch.sc.edu/webservices.cfm}}  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
 
@@
 Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.',
-  'Import files from a local path.  The files must be in a specific format, specifically those returned from the \\ac{CDMO} using the zip downloads option for a reserve.',
+  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
   'Import site data from an independent server.
-These files have been downloaded from \\ac{CDMO} up to December 2014, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.',
+These files have been downloaded from \\ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.',
   'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
@@
 The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.
-Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
-These are functions that call the existing web protocol methods on the CDMO web services.
-\texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
 
@@
 
-A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.
-This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
-This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.
-The files are hosted via third party web services and include almost all available \ac{SWMP} data for each site up to December 2014.
-Always use the \ac{CDMO} for the most recent data.
+A second approach for importing \ac{SWMP} data into R uses the \texttt{import\_remote} function.
+This method imports site data from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
+This approach is advantageous because the files can be imported very quickly and an IP address does not need to be registered.
+The files include almost all available \ac{SWMP} data for each site up to December 2014.
+However, the files will not be regularly updated and \ac{CDMO} should always be used for the most recent and accurate data.
 The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.
@@
 <<eval = F>>=
-## import a file from an independent server
+# import a file from an independent server
 import_remote('acebbwq')
@@
 
-A third approach is to use the \texttt{import\_local} function to import data into R that are locally available after downloading outside of R (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
-This approach may be best for larger, more customized data requests to \ac{CDMO}.
-Data can be retrieved from the \ac{CDMO} several ways.
-The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the \ac{CDMO}.
-The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested.
+A third approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
+This approach is most appropriate for large, specific data requests.
+The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}.
+The function may also work using data from the data export system,\footnote{\url{http://cdmo.baruch.sc.edu/get/export.cfm}} although this feature has not been extensively tested.
 The zip downloads feature is an easy way to obtain data from multiple stations in one request.
@@
 For water quality and nutrient data, duplicate time stamps are simply removed.
-Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.
+Note that nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
 Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
-Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.
-Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.
+Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
+Only duplicate values at 15 minutes are averaged for weather data.
 
 <<eval = F>>=
-# import data for apaebmet that you downloaded
-
-# this is an example path with the csv files, change as needed
+# import local data for apaebmet
+
+# this is an example path with the decompressed csv files
 path <- 'C:/my_path/'
@@
 The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.
-Note that the \texttt{datetimestamp} is standard time for each timezone and does not include daylight savings.
 Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.
-The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}) \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
+The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
 Attributes of a \texttt{swmpr} object can be viewed as follows.
 
-<<eval = T, cache = T>>=
-# import binary data
-data(apadbwq)
-dat <- apadbwq
-
-# verify that dat is swmpr class
-class(dat)
-
-# all attributes of dat
-names(attributes(dat))
-
-# a single attribute of dat
+<<eval = F, cache = T>>=
+# import data
+dat <- import_remote('apadbwq')
+
+# view all attributes of dat
+attributes(dat)
+
+# view a single attribute of dat
 attr(dat, 'station')
@@
 
-<<eval = T, cache = T>>=
-# available methods for swmpr class
+<<eval = F, cache = T>>=
+# view available methods for swmpr class
 methods(class = 'swmpr')
@@
 
-Example data as raw, comma-separated files have not been included in the package due to size limitations.
-However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
-This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
-Processed verions of the raw data are included with the package as binary data files (RData) to decrease processing times with the examples.
-Information for each binary file can be viewed as follows.
-
-<<eval = F>>=
-# view help files for complementary data
-# all files are samples from Apalachicola Bay
-
-# cat point station, nutrients
-?apacpnut
-
-# cat point station, water quality
-?apacpwq
-
-# dry bar station, water quality
-?apadbwq
-
-# east bay station, weater
-?apaebmet
-@
+A sample dataset can be downloaded for use with the examples below.\footnote{\url{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}}  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+Processed verions of the raw data are also included with the package as binary data files (RData).
+These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
+Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).
 
@@
 
-The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.
-The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types (\Cref{tab:organize}).
-
-The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (see the \ac{CDMO} descriptions \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).
-Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.
-Values in the QAQC column specify a flag from -5 to 5.
-Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.
-Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.
+The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (\Cref{tab:organize}).
+
+The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags.\footnote{\ac{CDMO} flags: \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}} Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.
+The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
 Erroneous data entered as -99 are also removed.
-Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.
+The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
 
@@
   'Remove \\ac{QAQC} columns and remove data based on \\ac{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\ac{QAQC} columns are present. ', 
-  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\ac{QAQC} flags used by \\ac{CDMO}.
-The output is used to inform further processing but is not used explicitly.',
+  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\ac{QAQC} flags used by \\ac{CDMO}.  The output can be used to inform further processing.', 
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
@@
 The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
+The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
 Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
-The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).
-The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
+The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \ac{QAQC} columns are present.
 
@@
 data(apacpnut)
-swmp1 <- apacpnut
-swmp1 <- qaqc(swmp1)
+dat <- apacpnut
+dat <- qaqc(dat)
 
 # remove replicate nutrient data
-rem_reps(swmp1)
+rem_reps(dat)
 
@@
 func <- function(x) max(x, na.rm = T)
-rem_reps(swmp1, FUN = func)
+rem_reps(dat, FUN = func)
 @
@@
 The date can be a single value or as two dates to select records within the range.
-The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.
-The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).
+The former case requires a binary operator as a character string passed to the argument, such as \texttt{`>'} or \texttt{`<='}.
+The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').
 Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
-\texttt{2012-11-31 12:00}.
+\texttt{`2012-11-31 12:00'}.
 Finally, the function can be used to remove rows and columns that do not contain data.
@@
 <<eval = F, cache = T>>=
+# import data
+data(apaebmet)
+dat <- apaebmet
+
 # select two parameters from dat
@@
 The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
-The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
+The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.
 Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.
 Likewise, the default value for \texttt{differ} is one half the time step.
-Rows that do not match any existing data within the limits of the \texttt{differ} argument are not discarded.
-Output from the \texttt{setstep} function can be used with \texttt{subset} and to create a time series at a set interval with empty data removed.
+Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.
 
 <<eval = F, cache = T>>=
+# import, qaqc removal
+data(apadbwq)
+dat <- qaqc(apadbwq)
+
 # convert time series to two hour invervals
@@
 The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
-The function requires one or more \texttt{swmpr} objects as input as separate, undefined arguments.
-The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.
-In general, the function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
+All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input.
+The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
 The time series for merging depends on the value passed to the \texttt{method} argument.
-Passing \texttt{union} to \texttt{method} will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
+Passing \texttt{union} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
 Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
-Finally, a seven or eight character station name passed to \texttt{method} will merge all input objects based on a continuous time series for the given station.
-The specified station must be present in the input data.
-Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.
+Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
+Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).
 
 <<eval = F, cache = T>>=
-# get nuts, wq, and met data as separate objects for the same station
-# note that most sites usually have one weather station
+# get nuts, wq, and met data as separate objects for the same reserve
 data(apacpnut)
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
-The latter category includes a limited number of functions that were developed by myself or others.
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
 Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.
 The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

### 2015-03-29 swmpr_manu.tex
proofing, finished up to analysis portion

@@
 
-The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}).
-The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.
-The package is loaded in the current workspace by using the \texttt{library} command.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
-\hlkwd{library}\hlstd{(devtools)}
-\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
-\hlkwd{library}\hlstd{(SWMPr)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.
-Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
-Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.
-An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.
-The following describes a general approach for using each category of functions based on a standard data workflow.
+The SWMPr package was developed for use with the R ($\geq$v3.0.0) statistical programming language \cite{RDCT14}.
+The SWMPr package can be installed from GitHub\footnote{\url{https://github.com/fawda123/SWMPr}} by executing the following commands at the R console prompt.
+The package is loaded in the workspace using the \texttt{library} command.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> install.packages('devtools')
+> library(devtools)
+> install_github('fawda123/SWMPr')
+> library(SWMPr)
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.
+Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.
+Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.
+An additional group of `miscellaneous' functions are included as helpers for the main functions.
+The following describes a general approach for using each category of functions.
 
@@
 Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
-\texttt{import\_local}&Import files from a local path.
-The files must be in a specific format, specifically those returned from the \ac{CDMO} using the zip downloads option for a reserve.\tabularnewline
+\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \ac{CDMO} using the zip downloads option.\tabularnewline
 \texttt{import\_remote}&Import site data from an independent server.
-These files have been downloaded from \ac{CDMO} up to December 2014, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.\tabularnewline
+These files have been downloaded from \ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.\tabularnewline
 \texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# retrieve metadata for all sites}
-\hlkwd{site_codes}\hlstd{()}
-
-\hlcom{# retrieve metadata for a single site}
-\hlkwd{site_codes_ind}\hlstd{(}\hlstr{'apa'}\hlstd{)}
-\end{alltt}
+\begin{verbatim}
+> # retrieve metadata for all sites
+> site_codes()
+> 
+> # retrieve metadata for a single site
+> site_codes_ind('apa')
+\end{verbatim}
 \end{kframe}
@@
 The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.
-Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.
-These are functions that call the existing web protocol methods on the CDMO web services.
-\texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# all parameters for a station, most recent}
-\hlkwd{all_params}\hlstd{(}\hlstr{'hudscwq'}\hlstd{)}
-
-\hlcom{# get all parameters within a date range}
-\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{))}
-
-\hlcom{# get single parameter within a date range}
-\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),}
-  \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
-
-\hlcom{# single parameter for a station, most recent}
-\hlkwd{single_param}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlstr{'do_mgl'}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.
-This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
-This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.
-The files are hosted via third party web services and include almost all available \ac{SWMP} data for each site up to December 2014.
-Always use the \ac{CDMO} for the most recent data.
+Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # all parameters for a station, most recent
+> all_params('hudscwq')
+> 
+> # get all parameters within a date range
+> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
+> 
+> # get single parameter within a date range
+> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+>   param = 'do_mgl')
+> 
+> # single parameter for a station, most recent
+> single_param('hudscwq', 'do_mgl')
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+A second approach for importing \ac{SWMP} data into R uses the \texttt{import\_remote} function.
+This method imports site data from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
+This approach is advantageous because the files can be imported very quickly and an IP address does not need to be registered.
+The files include almost all available \ac{SWMP} data for each site up to December 2014.
+However, the files will not be regularly updated and \ac{CDMO} should always be used for the most recent and accurate data.
 The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## import a file from an independent server}
-\hlkwd{import_remote}\hlstd{(}\hlstr{'acebbwq'}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-A third approach is to use the \texttt{import\_local} function to import data into R that are locally available after downloading outside of R (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).
-This approach may be best for larger, more customized data requests to \ac{CDMO}.
-Data can be retrieved from the \ac{CDMO} several ways.
-The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the \ac{CDMO}.
-The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested.
+\begin{verbatim}
+> # import a file from an independent server
+> import_remote('acebbwq')
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+A third approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
+This approach is most appropriate for large, specific data requests.
+The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}.
+The function may also work using data from the data export system,\footnote{\url{http://cdmo.baruch.sc.edu/get/export.cfm}} although this feature has not been extensively tested.
 The zip downloads feature is an easy way to obtain data from multiple stations in one request.
@@
 For water quality and nutrient data, duplicate time stamps are simply removed.
-Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.
+Note that nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
 Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
-Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.
-Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# import data for apaebmet that you downloaded}
-
-\hlcom{# this is an example path with the csv files, change as needed}
-\hlstd{path} \hlkwb{<-} \hlstr{'C:/my_path/'}
-
-\hlcom{# import, do not include file extension}
-\hlkwd{import_local}\hlstd{(path,} \hlstr{'apaebmet'}\hlstd{)}
-\end{alltt}
+Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
+Only duplicate values at 15 minutes are averaged for weather data.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # import local data for apaebmet
+> 
+> # this is an example path with the decompressed csv files
+> path <- 'C:/my_path/'
+> 
+> # import, do not include file extension
+> import_local(path, 'apaebmet') 
+\end{verbatim}
 \end{kframe}
@@
 The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.
-Note that the \texttt{datetimestamp} is standard time for each timezone and does not include daylight savings.
 Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.
-The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}) \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
+The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
 Attributes of a \texttt{swmpr} object can be viewed as follows.
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# import binary data}
-\hlkwd{data}\hlstd{(apadbwq)}
-\hlstd{dat} \hlkwb{<-} \hlstd{apadbwq}
-
-\hlcom{# verify that dat is swmpr class}
-\hlkwd{class}\hlstd{(dat)}
-\end{alltt}
-\begin{verbatim}
-## [1] "swmpr"      "data.frame"
-\end{verbatim}
-\begin{alltt}
-\hlcom{# all attributes of dat}
-\hlkwd{names}\hlstd{(}\hlkwd{attributes}\hlstd{(dat))}
-\end{alltt}
-\begin{verbatim}
-## [1] "names"       "row.names"   "class"       "station"    
-## [5] "parameters"  "qaqc_cols"   "date_rng"    "timezone"   
-## [9] "stamp_class"
-\end{verbatim}
-\begin{alltt}
-\hlcom{# a single attribute of dat}
-\hlkwd{attr}\hlstd{(dat,} \hlstr{'station'}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## [1] "apadbwq"
+\begin{verbatim}
+> # import data
+> dat <- import_remote('apadbwq')
+>  
+> # view all attributes of dat
+> attributes(dat)
+>  
+> # view a single attribute of dat
+> attr(dat, 'station')
 \end{verbatim}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# available methods for swmpr class}
-\hlkwd{methods}\hlstd{(}\hlkwc{class} \hlstd{=} \hlstr{'swmpr'}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-##  [1] aggregate.swmpr       aggregate_metab.swmpr comb.swmpr           
-##  [4] decomp.swmpr          decomp_cj.swmpr       ecometab.swmpr       
-##  [7] hist.swmpr            lines.swmpr           na.approx.swmpr      
-## [10] plot.swmpr            plot_metab.swmpr      plot_summary.swmpr   
-## [13] qaqc.swmpr            qaqcchk.swmpr         rem_reps.swmpr       
-## [16] setstep.swmpr         smoother.swmpr        subset.swmpr
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Example data as raw, comma-separated files have not been included in the package due to size limitations.
-However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.
-This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
-Processed verions of the raw data are included with the package as binary data files (RData) to decrease processing times with the examples.
-Information for each binary file can be viewed as follows.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# view help files for complementary data}
-\hlcom{# all files are samples from Apalachicola Bay}
-
-\hlcom{# cat point station, nutrients}
-\hlopt{?}\hlstd{apacpnut}
-
-\hlcom{# cat point station, water quality}
-\hlopt{?}\hlstd{apacpwq}
-
-\hlcom{# dry bar station, water quality}
-\hlopt{?}\hlstd{apadbwq}
-
-\hlcom{# east bay station, weater}
-\hlopt{?}\hlstd{apaebmet}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+\begin{verbatim}
+> # view available methods for swmpr class
+> methods(class = 'swmpr')
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+A sample dataset can be downloaded for use with the examples below.\footnote{\url{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}}  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+Processed verions of the raw data are also included with the package as binary data files (RData).
+These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
+Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).
 
@@
 \texttt{qaqcchk.swmpr}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.
-The output is used to inform further processing but is not used explicitly.\tabularnewline
+The output can be used to inform further processing.\tabularnewline
 \texttt{rem\_reps.swmpr}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# qaqc screen for a swmpr object, retain only '0'}
-\hlkwd{qaqc}\hlstd{(dat)}
-
-\hlcom{# retain all data regardless of flag}
-\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwa{NULL}\hlstd{)}
-
-\hlcom{# retain only '0' and '-1' flags}
-\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{))}
-\end{alltt}
+\begin{verbatim}
+> # qaqc screen for a swmpr object, retain only '0'
+> qaqc(dat)
+> 
+> # retain all data regardless of flag
+> qaqc(dat, qaqc_keep = NULL)
+> 
+> # retain only '0' and '-1' flags
+> qaqc(dat, qaqc_keep = c(0, -1))
+\end{verbatim}
 \end{kframe}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# view the number of observations in each QAQC flag}
-\hlkwd{qaqcchk}\hlstd{(dat)}
-\end{alltt}
+\begin{verbatim}
+> # view the number of observations in each QAQC flag
+> qaqcchk(dat)
+\end{verbatim}
 \end{kframe}
@@
 The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.
+The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
 Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
-The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).
-The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get nutrient data}
-\hlkwd{data}\hlstd{(apacpnut)}
-\hlstd{swmp1} \hlkwb{<-} \hlstd{apacpnut}
-\hlstd{swmp1} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
-
-\hlcom{# remove replicate nutrient data}
-\hlkwd{rem_reps}\hlstd{(swmp1)}
-
-\hlcom{# use different function to aggregate replicates}
-\hlstd{func} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{max}\hlstd{(x,} \hlkwc{na.rm} \hlstd{= T)}
-\hlkwd{rem_reps}\hlstd{(swmp1,} \hlkwc{FUN} \hlstd{= func)}
-\end{alltt}
+The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \ac{QAQC} columns are present.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # get nutrient data
+> data(apacpnut)
+> dat <- apacpnut
+> dat <- qaqc(dat)
+> 
+> # remove replicate nutrient data
+> rem_reps(dat)
+> 
+> # use different function to aggregate replicates
+> func <- function(x) max(x, na.rm = T)
+> rem_reps(dat, FUN = func)
+\end{verbatim}
 \end{kframe}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# select two parameters from dat}
-\hlkwd{subset}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'rh'}\hlstd{,} \hlstr{'bp'}\hlstd{))}
-
-\hlcom{# subset records greater than or equal to a date}
-\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlstr{'2013-01-01 0:00'}\hlstd{,} \hlkwc{operator} \hlstd{=} \hlstr{'>='}\hlstd{)}
-
-\hlcom{# subset records within a date range}
-\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-01 6:00'}\hlstd{,} \hlstr{'2012-08-01 18:15'}\hlstd{))}
-
-\hlcom{# subset records within a date range, select two parameters}
-\hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-01 6:00'}\hlstd{,} \hlstr{'2012-08-01 18:15'}\hlstd{),}
-  \hlkwc{select} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'atemp'}\hlstd{,} \hlstr{'totsorad'}\hlstd{))}
-
-\hlcom{# remove rows/columns that do not contain data}
-\hlkwd{subset}\hlstd{(dat,} \hlkwc{rem_rows} \hlstd{= T,} \hlkwc{rem_cols} \hlstd{= T)}
-\end{alltt}
+\begin{verbatim}
+> # import data
+> data(apaebmet)
+> dat <- apaebmet
+> 
+> # select two parameters from dat
+> subset(dat, select = c('rh', 'bp'))
+> 
+> # subset records greater than or equal to a date
+> subset(dat, subset = '2013-01-01 0:00', operator = '>=')
+> 
+> # subset records within a date range
+> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
+> 
+> # subset records within a date range, select two parameters
+> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+>   select = c('atemp', 'totsorad'))
+> 
+> # remove rows/columns that do not contain data
+> subset(dat, rem_rows = T, rem_cols = T)
+\end{verbatim}
 \end{kframe}
@@
 The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
-The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.
+The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.
 Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.
 Likewise, the default value for \texttt{differ} is one half the time step.
-Rows that do not match any existing data within the limits of the \texttt{differ} argument are not discarded.
-Output from the \texttt{setstep} function can be used with \texttt{subset} and to create a time series at a set interval with empty data removed.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# convert time series to two hour invervals}
-\hlcom{# tolerance of +/- 30 minutes for matching existing data}
-\hlkwd{setstep}\hlstd{(dat,} \hlkwc{timestep} \hlstd{=} \hlnum{120}\hlstd{,} \hlkwc{differ} \hlstd{=} \hlnum{30}\hlstd{)}
-
-\hlcom{# convert a nutrient time series to a continuous time series}
-\hlcom{# then remove empty rows and columns}
-\hlkwd{data}\hlstd{(apacpnut)}
-\hlstd{dat_nut} \hlkwb{<-} \hlstd{apacpnut}
-\hlstd{dat_nut} \hlkwb{<-} \hlkwd{setstep}\hlstd{(dat_nut,} \hlkwc{timestep} \hlstd{=} \hlnum{60}\hlstd{)}
-\hlkwd{subset}\hlstd{(dat_nut,} \hlkwc{rem_rows} \hlstd{= T,} \hlkwc{rem_cols} \hlstd{= T)}
-\end{alltt}
+Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # import, qaqc removal
+> data(apadbwq)
+> dat <- qaqc(apadbwq)
+> 
+> # convert time series to two hour invervals
+> # tolerance of +/- 30 minutes for matching existing data
+> setstep(dat, timestep = 120, differ = 30)
+> 
+> # convert a nutrient time series to a continuous time series
+> # then remove empty rows and columns
+> data(apacpnut)
+> dat_nut <- apacpnut
+> dat_nut <- setstep(dat_nut, timestep = 60)
+> subset(dat_nut, rem_rows = T, rem_cols = T)
+\end{verbatim}
 \end{kframe}
@@
 The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
-The function requires one or more \texttt{swmpr} objects as input as separate, undefined arguments.
-The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.
-In general, the function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
+All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input.
+The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
 The time series for merging depends on the value passed to the \texttt{method} argument.
-Passing \texttt{union} to \texttt{method} will create a time series that is continuous starting from the earliest date and the latest date for all input objects.
+Passing \texttt{union} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
 Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
-Finally, a seven or eight character station name passed to \texttt{method} will merge all input objects based on a continuous time series for the given station.
-The specified station must be present in the input data.
-Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get nuts, wq, and met data as separate objects for the same station}
-\hlcom{# note that most sites usually have one weather station}
-\hlkwd{data}\hlstd{(apacpnut)}
-\hlkwd{data}\hlstd{(apacpwq)}
-\hlkwd{data}\hlstd{(apaebmet)}
-\hlstd{swmp1} \hlkwb{<-} \hlstd{apacpnut}
-\hlstd{swmp2} \hlkwb{<-} \hlstd{apacpwq}
-\hlstd{swmp3} \hlkwb{<-} \hlstd{apaebmet}
-
-\hlcom{# combine nuts and wq data by union}
-\hlkwd{comb}\hlstd{(swmp1, swmp2,} \hlkwc{method} \hlstd{=} \hlstr{'union'}\hlstd{)}
-
-\hlcom{# combine nuts and wq data by intersect}
-\hlkwd{comb}\hlstd{(swmp1, swmp3,} \hlkwc{method} \hlstd{=} \hlstr{'intersect'}\hlstd{)}
-
-\hlcom{# combine nuts, wq, and met data by nuts time series, two hour time step}
-\hlkwd{comb}\hlstd{(swmp1, swmp2, swmp3,} \hlkwc{timestep} \hlstd{=} \hlnum{120}\hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{'apacpnut'}\hlstd{)}
-\end{alltt}
+Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
+Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # get nuts, wq, and met data as separate objects for the same reserve
+> data(apacpnut)
+> data(apacpwq)
+> data(apaebmet)
+> swmp1 <- apacpnut
+> swmp2 <- apacpwq
+> swmp3 <- apaebmet
+> 
+> # combine nuts and wq data by union
+> comb(swmp1, swmp2, method = 'union')
+> 
+> # combine nuts and wq data by intersect
+> comb(swmp1, swmp3, method = 'intersect')
+> 
+> # combine nuts, wq, and met data by nuts time series, two hour time step
+> comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
+\end{verbatim}
 \end{kframe}
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
-The latter category includes a limited number of functions that were developed by myself or others.
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
@@
 Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.
 The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# combine, qaqc, remove empty columns}
-\hlstd{dat} \hlkwb{<-} \hlkwd{comb}\hlstd{(swmp1, swmp2,} \hlkwc{method} \hlstd{=} \hlstr{'union'}\hlstd{)}
-\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(dat)}
-\hlstd{swmpr_in} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{rem_cols} \hlstd{= T)}
-
-\hlcom{# get mean DO by quarters}
-\hlkwd{aggregate}\hlstd{(swmpr_in,} \hlstr{'quarters'}\hlstd{,} \hlkwc{params} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'do_mgl'}\hlstd{))}
-
-\hlcom{# get mean DO by quarters, remove NA when calculating means}
-\hlstd{fun_in} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{mean}\hlstd{(x,} \hlkwc{na.rm} \hlstd{= T)}
-\hlkwd{aggregate}\hlstd{(swmpr_in,} \hlkwc{FUN} \hlstd{= fun_in,} \hlstr{'quarters'}\hlstd{,} \hlkwc{params} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'do_mgl'}\hlstd{))}
-\end{alltt}
+\begin{verbatim}
+> # combine, qaqc, remove empty columns
+> dat <- comb(swmp1, swmp2, method = 'union')
+> dat <- qaqc(dat)
+> swmpr_in <- subset(dat, rem_cols = T)
+> 
+> # get mean DO by quarters
+> aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
+> 
+> # get mean DO by quarters, remove NA when calculating means
+> fun_in <- function(x) mean(x, na.rm = T)
+> aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
+\end{verbatim}
 \end{kframe}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# import data}
-\hlkwd{data}\hlstd{(apadbwq)}
-\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
-
-\hlcom{# qaqc and subset imported data}
-\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
-\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2012-07-09 00:00'}\hlstd{,} \hlstr{'2012-07-24 00:00'}\hlstd{))}
-
-\hlcom{# filter}
-\hlstd{test} \hlkwb{<-} \hlkwd{smoother}\hlstd{(dat,} \hlkwc{window} \hlstd{=} \hlnum{50}\hlstd{,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
-
-\hlcom{# plot to see the difference}
-\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp,} \hlkwc{data} \hlstd{= dat,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
-\hlkwd{lines}\hlstd{(test,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-\end{alltt}
+\begin{verbatim}
+> # import data
+> data(apadbwq)
+> swmp1 <- apadbwq
+> 
+> # qaqc and subset imported data
+> dat <- qaqc(swmp1)
+> dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
+> 
+> # filter
+> test <- smoother(dat, window = 50, params = 'do_mgl')
+> 
+> # plot to see the difference
+> plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
+> lines(test, select = 'do_mgl', col = 'red', lwd = 2)
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get data}
-\hlkwd{data}\hlstd{(apadbwq)}
-\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
-
-\hlcom{# qaqc and subset imported data}
-\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}
-\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(dat,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-01-22 00:00'}\hlstd{,} \hlstr{'2013-01-26 00:00'}\hlstd{))}
-
-\hlcom{# interpolate, maxgap of 10 records}
-\hlstd{test} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{10}\hlstd{)}
-
-\hlcom{# interpolate maxgap of 30 records}
-\hlstd{test2} \hlkwb{<-} \hlkwd{na.approx}\hlstd{(dat,} \hlkwc{params} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{maxgap} \hlstd{=} \hlnum{30}\hlstd{)}
-
-\hlcom{# plot for comparison}
-\hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{3}\hlstd{,} \hlnum{1}\hlstd{))}
-\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, dat,} \hlkwc{main} \hlstd{=} \hlstr{'Raw'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
-\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
-  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 10 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
-\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
-\hlkwd{plot}\hlstd{(do_mgl} \hlopt{~} \hlstd{datetimestamp, test2,} \hlkwc{col} \hlstd{=} \hlstr{'red'}\hlstd{,}
-  \hlkwc{main} \hlstd{=} \hlstr{'Interpolation - maximum gap of 30 records'}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{'l'}\hlstd{)}
-\hlkwd{lines}\hlstd{(dat,} \hlkwc{select} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}
-\end{alltt}
+\begin{verbatim}
+> # get data
+> data(apadbwq)
+> swmp1 <- apadbwq
+> 
+> # qaqc and subset imported data
+> dat <- qaqc(swmp1)
+> dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
+> 
+> # interpolate, maxgap of 10 records
+> test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+> 
+> # interpolate maxgap of 30 records
+> test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
+> 
+> # plot for comparison
+> par(mfrow = c(3, 1))
+> plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
+> plot(do_mgl ~ datetimestamp, test, col = 'red', 
++   main = 'Interpolation - maximum gap of 10 records', type = 'l')
+> lines(dat, select = 'do_mgl')
+> plot(do_mgl ~ datetimestamp, test2, col = 'red', 
++   main = 'Interpolation - maximum gap of 30 records', type = 'l')
+> lines(dat, select = 'do_mgl')
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get data}
-\hlkwd{data}\hlstd{(apadbwq)}
-\hlstd{swmp1} \hlkwb{<-} \hlstd{apadbwq}
-
-\hlcom{# subset for daily decomposition}
-\hlstd{dat} \hlkwb{<-} \hlkwd{subset}\hlstd{(swmp1,} \hlkwc{subset} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'2013-07-01 00:00'}\hlstd{,} \hlstr{'2013-07-31 00:00'}\hlstd{))}
-
-\hlcom{# decomposition and plot}
-\hlstd{test} \hlkwb{<-} \hlkwd{decomp}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{,} \hlkwc{frequency} \hlstd{=} \hlstr{'daily'}\hlstd{)}
-\hlkwd{plot}\hlstd{(test)}
-\end{alltt}
+\begin{verbatim}
+> # get data
+> data(apadbwq)
+> swmp1 <- apadbwq
+> 
+> # subset for daily decomposition
+> dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
+> 
+> # decomposition and plot
+> test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+> plot(test)
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# get data}
-\hlkwd{data}\hlstd{(apacpnut)}
-\hlstd{dat} \hlkwb{<-} \hlstd{apacpnut}
-\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwa{NULL}\hlstd{)}
-
-\hlcom{# decomposition of chl, ggplot}
-\hlkwd{decomp_cj}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{)}
-\end{alltt}
+\begin{verbatim}
+> # get data
+> data(apacpnut)
+> dat <- apacpnut
+> dat <- qaqc(dat, qaqc_keep = NULL)
+> 
+> # decomposition of chl, ggplot
+> decomp_cj(dat, param = 'chla_n')
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## import data}
-\hlkwd{data}\hlstd{(apacpnut)}
-\hlstd{dat} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apacpnut)}
-
-\hlcom{## plot}
-\hlkwd{plot_summary}\hlstd{(dat,} \hlkwc{param} \hlstd{=} \hlstr{'chla_n'}\hlstd{,} \hlkwc{years} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{2007}\hlstd{,} \hlnum{2013}\hlstd{))}
-\end{alltt}
+\begin{verbatim}
+> ## import data
+> data(apacpnut)
+> dat <- qaqc(apacpnut)
+> 
+> ## plot
+> plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## import water quality and weather data}
-\hlkwd{data}\hlstd{(apadbwq)}
-\hlkwd{data}\hlstd{(apaebmet)}
-
-\hlcom{## qaqc, combine}
-\hlstd{wq} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apadbwq)}
-\hlstd{met} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(apaebmet)}
-\hlstd{dat} \hlkwb{<-} \hlkwd{comb}\hlstd{(wq, met)}
-
-\hlcom{## estimate metabolism}
-\hlstd{res} \hlkwb{<-} \hlkwd{ecometab}\hlstd{(dat,} \hlkwc{trace} \hlstd{=} \hlnum{FALSE}\hlstd{)}
-\hlkwd{plot_metab}\hlstd{(res)}
-\end{alltt}
+\begin{verbatim}
+> ## import water quality and weather data
+> data(apadbwq)
+> data(apaebmet)
+> 
+> ## qaqc, combine
+> wq <- qaqc(apadbwq)
+> met <- qaqc(apaebmet)
+> dat <- comb(wq, met)
+> 
+> ## estimate metabolism
+> res <- ecometab(dat, trace = FALSE)
+> plot_metab(res)
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# plot the stations at Jacques Cousteau reserve}
-\hlkwd{map_reserve}\hlstd{(}\hlstr{'jac'}\hlstd{)}
-\end{alltt}
+\begin{verbatim}
+> # plot the stations at Jacques Cousteau reserve
+> map_reserve('jac')
+\end{verbatim}
 \end{kframe}\begin{figure}[!ht]

### 2015-03-30 swmpr_manu.Rnw
finished proofing

@@
 %
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
-% the final text of your manuscript.
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
 % -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
 %
@@
 %
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
 % % % % % % % % % % % % % % % % % % % % % % %
@@
 \usepackage[colorlinks=true,allcolors=Gray]{hyperref}
-% \usepackage[hidelinks]{hyperref}
 \usepackage[usenames,dvipsnames]{xcolor}
@@
 
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
 % cleveref options
@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE)
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small')
 options(replace.assign=TRUE,width=70,digits=1)
@@
 Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
-The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
+The SWMPr package is developed for the continuous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 An additional group of `miscellaneous' functions are included as helpers for the main functions.
-The following describes a general approach for using each category of functions.
+The following describes a general approach for working with \ac{SWMP} data using each category of functions.
 
@@
 <<eval = F, cache = T>>=
-# get nuts, wq, and met data as separate objects for the same reserve
+# get nut, wq, and met data as separate objects
 data(apacpnut)
@@
 
-# combine nuts and wq data by union
+# combine nut and wq data by union
 comb(swmp1, swmp2, method = 'union')
 
-# combine nuts and wq data by intersect
+# combine nut and wq data by intersect
 comb(swmp1, swmp3, method = 'intersect')
 
-# combine nuts, wq, and met data by nuts time series, two hour time step
+# combine nut, wq, and met data by nut time series, two hour time step
 comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
@@
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
+These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
 Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
@@
 
-The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set periods of observation.
-This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.
-Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.
+The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set units of time.
+This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
+Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.
 A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.
-This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.
-Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).
-Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.
+Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data.
 
@@
 descrips <- c(
-  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.',
-  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw daily data.',
-  'Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.',
-  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose}.  Decomposition of monthly or daily trends is possible.',
+  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
+  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw metabolism data.',
+  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
+  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.
+This is a simple wrapper to \\texttt{decompose} \\cite{Kendall83}.
+Decomposition of monthly or daily trends is possible.',
   'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \\texttt{decompTs} in the wq package.
+This is a simple wrapper to \\texttt{decompTs} in the wq package \\cite{Jassby14}.
 Only monthly decomposition is possible.',
   'Plot a histogram for a \\texttt{swmpr} object.',
-  'Add lines to an existing \\texttt{swmpr} plot.',
-  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data.', 
+  'Add lines to an existing plot created with \\texttt{plot.swmpr}.',
+  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
   'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
   'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
-  'Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.',
-  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \\texttt{filter}.'
+  'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
+  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\texttt{filter}.'
   )
@@
 <<eval = F, cache = T>>=
-# combine, qaqc, remove empty columns
-dat <- comb(swmp1, swmp2, method = 'union')
-dat <- qaqc(dat)
-swmpr_in <- subset(dat, rem_cols = T)
-
-# get mean DO by quarters
-aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
-
-# get mean DO by quarters, remove NA when calculating means
-fun_in <- function(x) mean(x, na.rm = T)
-aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
+# get data
+data(apacpwq)
+dat <- qaqc(apacpwq)
+
+# get mean DO by months
+aggregate(dat, 'months', params = c('do_mgl'))
 @
@@
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
-This method is a simple wrapper to \texttt{filter}.
-The \texttt{window} argument specifies the number of observations included in the moving average.
-The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).
-A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation.
+The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.
+The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
+The \texttt{sides} argument specifies how the average is calculated for each observation.
+Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation.
 As before, the \texttt{params} argument specifies which parameters to smooth.
-See \Cref{fig:smooth_ex} for the output from the code.
 
 <<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\texttt{smoother} function.">>=
-# import data
+# import data, qaqc and subset
 data(apadbwq)
-swmp1 <- apadbwq
-
-# qaqc and subset imported data
-dat <- qaqc(swmp1)
-dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
-
-# filter
-test <- smoother(dat, window = 50, params = 'do_mgl')
-
-# plot to see the difference
-plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
-lines(test, select = 'do_mgl', col = 'red', lwd = 2)
+dat <- qaqc(apadbwq)
+dat <- subset(dat, select = 'do_mgl', 
+  subset = c('2012-07-09 00:00', '2012-07-24 00:00')
+  )
+
+# smooth
+dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
+
+# plot raw and smoothed
+plot(dat, type = 'l')
+lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
 @
@@
 In either case, an analyst should have a strong rationale for the chosen method.
-A common approach used to handle missing data in time series analysis is linear interpolation.
+A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).
 A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
-A challenge with linear interpolation is an appropriate gap size for fitting missing observations.
-The ability of the interpolated data to approximate actual trends is a function of the gap size.
-Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.
-An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
-The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.
-A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.
-See \Cref{fig:interp_ex} for the output from the following code.
+However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.
+Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.
+An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
+The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
 
 <<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
-# get data
+# get data, qaqc and subset
 data(apadbwq)
-swmp1 <- apadbwq
-
-# qaqc and subset imported data
-dat <- qaqc(swmp1)
-dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
+dat <- qaqc(apadbwq)
+dat <- subset(dat, select = 'do_mgl', 
+  subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
 
 # interpolate, maxgap of 10 records
-test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)
 
 # interpolate maxgap of 30 records
-test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
+fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
 
@@
 par(mfrow = c(3, 1))
-plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
-plot(do_mgl ~ datetimestamp, test, col = 'red', 
-  main = 'Interpolation - maximum gap of 10 records', type = 'l')
-lines(dat, select = 'do_mgl')
-plot(do_mgl ~ datetimestamp, test2, col = 'red', 
-  main = 'Interpolation - maximum gap of 30 records', type = 'l')
-lines(dat, select = 'do_mgl')
-@
-
-The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
-The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.
-By default, a moving average with a symmetric window is used to filter the seasonal component.
-Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).
-
-The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.
-The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.
-This requires an explicit input defining the frequency of the parameter in the time series.
-For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).
-The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
-For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.
-A starting value of the time series must be supplied in the latter case.
-Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
-Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
-\Cref{fig:decomp_ex1} is an example of the \texttt{decomp} function.
+plot(dat, main = 'Raw')
+plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
+lines(dat)
+plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
+lines(dat)
+@
+
+The disaggregation of time series into additive of multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
+The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).
+An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
+The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
+For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (4 * 24).
+The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (4 * 24 * 365).
+For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
 
@@
 
-An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.
-Theory describing this method is provided by Cloern and Jassby \cite{Cloern10}.
-The function is similar to \texttt{decomp.swmpr} with a few key differences.
-The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.
-For both functions, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.
+An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.
 The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
 This approach works best for nutrient data that are typically obtained on a monthly cycle.
-The function will also work with continuous water quality or weather data but note that the data must first aggregated on the monthly scale before decomposition.
+The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.
+Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
+For both, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.
 Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
@@
 
-# decomposition of chl, ggplot
+# decomposition of chl
 decomp_cj(dat, param = 'chla_n')
@@
 
-Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.
+Detailed exploratory graphics are also useful for evaluating general trends in observed data.
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.
-Monthly anomalies are in relation to the grand mean for the same month across all years.
-All data are aggregated for quicker plotting.
-Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.
-Cumulative precipitation data are based on the daily maximum.
-The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).
+Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
+An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{supp_info}).
 
@@
 
-Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model (see Ro and Hunt \cite{Ro06}, Thebault et al.
-\cite{Thebault08}).
+Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}.
 The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series.
@@
 Several assumptions must be met for a valid interpretation of the results.
-In general, the dissolved oxygen time series is assumed to represent the same water mass over time.
+First, the \ac{DO} time series is assumed to be a sample of the same water mass over time.
 Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.
 The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
-Areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
+Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
 Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object.
 Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
-Additionally, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
+Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
 The metadata should be consulted for exact height.
@@
 
-The following is an example that shows use of the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
-
-<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.">>=
+The following is an example that shows use of the function from a combined water quality and weather data set.
+Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
+
+<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly means (95\\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.">>=
 ## import water quality and weather data
@@
 descrips <- c(
-  'Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \\texttt{ecometab} function.',
+  'Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \\texttt{ecometab} function.',
   'Create a map of all stations in a reserve using the ggmap package.',
-  'Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \\texttt{ecometab} function.',
-  'Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).  Includes \\ac{QAQC} columns with \\texttt{f\\_} prefix. Used internally in other functions.',
-  'Parses html returned from \\ac{CDMO} web services, used internally in retrieval functions.',
-  'Creates object of \\texttt{swmpr} class, used internally in retrieval functions.',
-  'Converts time vectors to \\texttt{POSIXct} objects with correct time zone for a site/station, used internally in retrieval functions.'
+  'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\texttt{ecometab} function.',
+  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes \\ac{QAQC} columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
+  'Parses HTML returned from \\ac{CDMO} data requests.  Used in the retrieval functions.',
+  'Creates a \\texttt{swmpr} object class.  Used in the data retrieval functions.',
+  'Converts time vectors to \\texttt{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.'
   )
@@
   caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for more customized evaluationg of \\ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
@@
 For brevity, only the \texttt{reserve\_map} function is discussed.
-This function can be used to create a map with all stations at a reserve by passing arguments to functions in the ggmap package \cite{ggmap13}.
+This function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{ggmap13}.
 The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
@@
 This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
-
-First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at the \ac{NERRS} reserve for all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update to results described in Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} using the most recent \ac{SWMP} data.
-Additionally, previous methods to evaluate metabolism at \ac{NERRS} reserves did not incorporate \textit{in situ} weather data to estimate the air-sea gas exchange model.
+First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each \ac{NERRS} reserve using all years of available data.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent \ac{SWMP} data.
 Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \ac{NERRS} reserves.
-Additionally, the air-sea gas exchange model assumed a constant value for the reaeratio coefficientt.
-This coefficient provides an estimate of the rate of air-sea gas exchange that varies as a function of wind speed, temperature, barometric pressure, salinity, and depth of the water column.
-The inclusion of weather data in the calculation allows for a more precise estimate of air-sea gas exchange and consequently more reliable estimates of ecosystem metabolism (see Caffrey et al.
+The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \texttt{ecometab} function provides a more accurate estimate by including weather data in the calculation (see Caffrey et al.
 \cite{Caffrey14} for details).
 
-Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.
-All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script).
-This resulted in a single \texttt{swmpr} object for each parameter at each site.
-All files were then uploaded to a remote server.
-An additional \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site to esimate metabolism.
-For brevity, only two sites at each reserve were evaluated.
+Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \ac{CDMO}.
+All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files.\footnote{Data processing script: \url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}} This resulted in a single \texttt{swmpr} object for each parameter at each site.
+All files were then uploaded to a remote server for online access.
+An additional R script\footnote{Metabolism processing script: \url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}} was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
+Two water quality sites with the longest time series at each reserve were used.
 Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
@@
 Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
-More importantly, the use of the data retreival, synthesis, and analysis functions to create the results shown in \Cref{fig:metab_plo,tab:metab_tab} illustrates the utility provided by the SWMPr package.
+More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.
 
@@
 Two sites were chosen from each reserve that had the longest available time series.
-Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.', fig.height = 6, fig.width = 8, out.width = '\\textwidth'>>=
+Sites were assigned to regions based on approximate geographic coordinates.', fig.height = 6, fig.width = 8, out.width = '\\textwidth'>>=
 # # packages to use
@@
 Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
-Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.'
+Sites were assigned to regions based on approximate geographic coordinates.'
 foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
@@
 
-The second and third examples are two Shiny web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).
-The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
+The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data (see the \nameref{supp_info}).
+The first web application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
-Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.
-Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.
-The trend direction is blue for decreasing and red for increasing.
-The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.
-The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations.
+Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.
+Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.
+The trend direction is blue for decreasing and red for increasing with time.
+The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function.
 The drop down menus can be used to select the station, date range, and parameter for plotting.
-Plots are based on daily averages for each parameter.
-The data used for each application are similar to those used to estimate ecosystem metabolism described above.
+The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.
 
@@
 The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.
-Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessment of site-specific characteristics and more informed decisions to manage environmental resources.
+Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.
 The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 300 stations within the 28 estuaries of the \acl{NERRS}.
 This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
-Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package to address common challenges working with large datasets.
-
-The SWMPr package provides several functions to retrieve, organize, and analyze \ac{SWMP} data to more effectively evaluate environmental trends.
+Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package.
+
+The SWMPr package provides several functions to retrieve, organize, and analyze \ac{SWMP} data to more effectively address common challenges working with large datasets.
 The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (\Cref{tab:retrieve}).
-Established \ac{QAQC} methods and data processing techniques are also enhanced with \ac{SWMP} by functions that filter observations for different \ac{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
-Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data, respectively.
+Established \ac{QAQC} methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different \ac{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
+Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data.
 Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.
@@
 
-I acknowledge the significant efforts of \ac{NERRS} researchers and staff for providing access to high-quality monitoring data.
-Thanks particularly to Dwayne Porter and Melissa Ide from \ac{CDMO} for maintaining the online database.
-Thanks to Marie Bundy and Nikki Dix for providing me the opportunity to share this package with the broader \ac{NERRS} community.
-Thanks to Todd O'Brien for the inspiration for the online widgets in the supporting information.
+I acknowledge the significant work of \ac{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
+Thanks to Dwayne Porter and Melissa Ide from \ac{CDMO} for maintaining the online database.
+Thanks to Marie Bundy and Nikki Dix for giving me the opportunity to share this package with the broader \ac{NERRS} community.
+Thanks to Todd O'Brien for the inspiration for the online widgets.
 Thanks to Mike Murrell and Jim Hagy III for assistance with documentation and implementation of the ecosystem metabolism functions.

### 2015-03-30 swmpr_manu.tex
finished proofing

@@
 
-\begin{knitrout}
+\begin{knitrout}\small
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get nuts, wq, and met data as separate objects for the same reserve
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # get nut, wq, and met data as separate objects
 > data(apacpnut)
@@
 > 
-> # combine nuts and wq data by union
+> # combine nut and wq data by union
 > comb(swmp1, swmp2, method = 'union')
 > 
-> # combine nuts and wq data by intersect
+> # combine nut and wq data by intersect
 > comb(swmp1, swmp3, method = 'intersect')
 > 
-> # combine nuts, wq, and met data by nuts time series, two hour time step
+> # combine nut, wq, and met data by nut time series, two hour time step
 > comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
@@
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.
+These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
 Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
@@
 \hline
-\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.
-Aggregation function is user-supplied but defaults to mean.\tabularnewline
+\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
 \texttt{aggregate\_metab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.
-This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.\tabularnewline
-\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.\tabularnewline
+This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
+\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
 \texttt{decomp.swmpr}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \texttt{decompose}.
+This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.
 Decomposition of monthly or daily trends is possible.\tabularnewline
 \texttt{decomp\_cj.swmpr}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \texttt{decompTs} in the wq package.
+This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.
 Only monthly decomposition is possible.\tabularnewline
 \texttt{hist.swmpr}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
-\texttt{lines.swmpr}&Add lines to an existing \texttt{swmpr} plot.\tabularnewline
-\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object.
-The maximum gap size that is interpolated is defined as a maximum number of records with missing data.\tabularnewline
+\texttt{lines.swmpr}&Add lines to an existing plot created with \texttt{plot.swmpr}.\tabularnewline
+\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
 \texttt{plot.swmpr}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
 \texttt{plot\_metab.swmpr}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
-\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.\tabularnewline
-\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.\tabularnewline
+\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
+\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
 \hline
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # combine, qaqc, remove empty columns
-> dat <- comb(swmp1, swmp2, method = 'union')
-> dat <- qaqc(dat)
-> swmpr_in <- subset(dat, rem_cols = T)
-> 
-> # get mean DO by quarters
-> aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
-> 
-> # get mean DO by quarters, remove NA when calculating means
-> fun_in <- function(x) mean(x, na.rm = T)
-> aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # get data
+> data(apacpwq)
+> dat <- qaqc(apacpwq)
+> 
+> # get mean DO by months
+> aggregate(dat, 'months', params = c('do_mgl'))
 \end{verbatim}
@@
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
-This method is a simple wrapper to \texttt{filter}.
-The \texttt{window} argument specifies the number of observations included in the moving average.
-The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).
-A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation.
+The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.
+The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
+The \texttt{sides} argument specifies how the average is calculated for each observation.
+Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation.
 As before, the \texttt{params} argument specifies which parameters to smooth.
-See \Cref{fig:smooth_ex} for the output from the code.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # import data
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # import data, qaqc and subset
 > data(apadbwq)
-> swmp1 <- apadbwq
-> 
-> # qaqc and subset imported data
-> dat <- qaqc(swmp1)
-> dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
-> 
-> # filter
-> test <- smoother(dat, window = 50, params = 'do_mgl')
-> 
-> # plot to see the difference
-> plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
-> lines(test, select = 'do_mgl', col = 'red', lwd = 2)
+> dat <- qaqc(apadbwq)
+> dat <- subset(dat, select = 'do_mgl', 
++   subset = c('2012-07-09 00:00', '2012-07-24 00:00')
++   )
+> 
+> # smooth
+> dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
+> 
+> # plot raw and smoothed
+> plot(dat, type = 'l')
+> lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
 \end{verbatim}
@@
 In either case, an analyst should have a strong rationale for the chosen method.
-A common approach used to handle missing data in time series analysis is linear interpolation.
+A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).
 A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
-A challenge with linear interpolation is an appropriate gap size for fitting missing observations.
-The ability of the interpolated data to approximate actual trends is a function of the gap size.
-Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.
-An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.
-The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.
-A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.
-See \Cref{fig:interp_ex} for the output from the following code.
-
-\begin{knitrout}
+However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.
+Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.
+An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
+The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # get data, qaqc and subset
+> data(apadbwq)
+> dat <- qaqc(apadbwq)
+> dat <- subset(dat, select = 'do_mgl', 
++   subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
+> 
+> # interpolate, maxgap of 10 records
+> fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)
+> 
+> # interpolate maxgap of 30 records
+> fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
+> 
+> # plot for comparison
+> par(mfrow = c(3, 1))
+> plot(dat, main = 'Raw')
+> plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
+> lines(dat)
+> plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
+> lines(dat)
+\end{verbatim}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/interp_ex}
+
+}
+
+\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.\label{fig:interp_ex}}
+\end{figure}
+
+\end{knitrout}
+
+The disaggregation of time series into additive of multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
+The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).
+An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
+The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
+For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (4 * 24).
+The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (4 * 24 * 365).
+For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+
+\begin{knitrout}\small
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
@@
 > 
-> # qaqc and subset imported data
-> dat <- qaqc(swmp1)
-> dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
-> 
-> # interpolate, maxgap of 10 records
-> test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-> 
-> # interpolate maxgap of 30 records
-> test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
-> 
-> # plot for comparison
-> par(mfrow = c(3, 1))
-> plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
-> plot(do_mgl ~ datetimestamp, test, col = 'red', 
-+   main = 'Interpolation - maximum gap of 10 records', type = 'l')
-> lines(dat, select = 'do_mgl')
-> plot(do_mgl ~ datetimestamp, test2, col = 'red', 
-+   main = 'Interpolation - maximum gap of 30 records', type = 'l')
-> lines(dat, select = 'do_mgl')
-\end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/interp_ex}
-
-}
-
-\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.\label{fig:interp_ex}}
-\end{figure}
-
-\end{knitrout}
-
-The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.
-The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.
-By default, a moving average with a symmetric window is used to filter the seasonal component.
-Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).
-
-The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.
-The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.
-This requires an explicit input defining the frequency of the parameter in the time series.
-For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).
-The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).
-For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.
-A starting value of the time series must be supplied in the latter case.
-Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
-Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
-\Cref{fig:decomp_ex1} is an example of the \texttt{decomp} function.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get data
-> data(apadbwq)
-> swmp1 <- apadbwq
-> 
 > # subset for daily decomposition
@@
 > 
-> # decomposition of chl, ggplot
+> # decomposition of chl
 > decomp_cj(dat, param = 'chla_n')
@@
 
-Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.
+Detailed exploratory graphics are also useful for evaluating general trends in observed data.
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.
-Monthly anomalies are in relation to the grand mean for the same month across all years.
-All data are aggregated for quicker plotting.
-Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.
-Cumulative precipitation data are based on the daily maximum.
-The function returns a graphics object (Grob) of multiple ggplot objects.
-An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).
-
-\begin{knitrout}
+Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
+An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{supp_info}).
+
+\begin{knitrout}\small
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
@@
 
-The following is an example that shows use of the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
-
-\begin{knitrout}
+The following is an example that shows use of the function from a combined water quality and weather data set.
+Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
+
+\begin{knitrout}\small
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
@@
 
-\caption[Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida]{Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.\label{fig:metab_ex}}
+\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.\label{fig:metab_ex}}
 \end{figure}
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for more customized evaluationg of \\ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE)%
@@
 \caption{Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for more customized evaluationg of \ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \ac{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \texttt{?calckl} at the command line).\label{tab:misc}}
@@
 \hline
-\texttt{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  This is only used within the \texttt{ecometab} function.\tabularnewline
+\texttt{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \texttt{ecometab} function.\tabularnewline
 \texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
-\texttt{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  This is only used within the \texttt{ecometab} function.\tabularnewline
-\texttt{param\_names}&Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality).
+\texttt{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \texttt{ecometab} function.\tabularnewline
+\texttt{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
 Includes \ac{QAQC} columns with \texttt{f\_} prefix.
-Used internally in other functions.\tabularnewline
-\texttt{parser}&Parses html returned from \ac{CDMO} web services, used internally in retrieval functions.\tabularnewline
-\texttt{swmpr}&Creates object of \texttt{swmpr} class, used internally in retrieval functions.\tabularnewline
-\texttt{time\_vec}&Converts time vectors to \texttt{POSIXct} objects with correct time zone for a site/station, used internally in retrieval functions.\tabularnewline
+Used in the data retrieval functions.\tabularnewline
+\texttt{parser}&Parses HTML returned from \ac{CDMO} data requests.  Used in the retrieval functions.\tabularnewline
+\texttt{swmpr}&Creates a \texttt{swmpr} object class.  Used in the data retrieval functions.\tabularnewline
+\texttt{time\_vec}&Converts time vectors to \texttt{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.\tabularnewline
 \hline
@@
 Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
-More importantly, the use of the data retreival, synthesis, and analysis functions to create the results shown in \Cref{fig:metab_plo,tab:metab_tab} illustrates the utility provided by the SWMPr package.
+More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.
 
 % metabolism figure
-\begin{knitrout}
+\begin{knitrout}\small
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!ht]
@@
 Two sites were chosen from each reserve that had the longest available time series.
-Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.\label{fig:metab_plo}}
+Sites were assigned to regions based on approximate geographic coordinates.\label{fig:metab_plo}}
 \end{figure}
@@
 Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
-Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.\label{tab:metab_tab}}
+Sites were assigned to regions based on approximate geographic coordinates.\label{tab:metab_tab}}
 \begin{center}

### 2015-03-30 swmpr_manu.Rnw
Fixed title case in title, some examples still had left over arguments from old plot methods

@@
 {\Large
-\textbf\newline{SWMPr: An R package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
+\textbf\newline{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
 }
@@
 # plot raw and smoothed
-plot(dat, type = 'l')
-lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
+plot(dat)
+lines(dat_smooth, col = 'red', lwd = 2)
 @

### 2015-03-30 swmpr_manu.tex
Fixed title case in title, some examples still had left over arguments from old plot methods

@@
 % See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
 %
@@
 > # plot raw and smoothed
-> plot(dat, type = 'l')
-> lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
+> plot(dat)
+> lines(dat_smooth, col = 'red', lwd = 2)
 \end{verbatim}

### 2015-04-01 swmpr_manu.Rnw
Kahle13 ref update
updated text for function name changes - aggregate to aggreswmp and aggregate_metab to aggremetab

@@
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
-These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
+These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
 Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
@@
 
-The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set units of time.
+The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
@@
 
-funcs <- c('aggregate.swmpr', 'aggregate\\_metab.swmpr', 'ecometab.swmpr', 'decomp.swmpr', 'decomp\\_cj.swmpr', 'hist.swmpr', 'lines.swmpr', 'na.approx.swmpr', 'plot.swmpr', 'plot\\_metab.swmpr', 'plot\\_summary.swmpr', 'smoother.swmpr')
+funcs <- c('aggreswmp.swmpr', 'aggremetab.swmpr', 'ecometab.swmpr', 'decomp.swmpr', 'decomp\\_cj.swmpr', 'hist.swmpr', 'lines.swmpr', 'na.approx.swmpr', 'plot.swmpr', 'plot\\_metab.swmpr', 'plot\\_summary.swmpr', 'smoother.swmpr')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   file = '',
-  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",
+  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp.swmpr} at the command line).",
   rowlabel = 'Function',
@@
 # get mean DO by months
-aggregate(dat, 'months', params = c('do_mgl'))
+aggreswmp(dat, 'months', params = c('do_mgl'))
 @
@@
 For brevity, only the \texttt{reserve\_map} function is discussed.
-This function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{ggmap13}.
+This function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
 The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.

### 2015-04-01 swmpr_manu.tex
Kahle13 ref update
updated text for function name changes - aggregate to aggreswmp and aggregate_metab to aggremetab

@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
 \begin{table}[!tbp]
 \caption{Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \texttt{?aggregate.swmpr} at the command line).\label{tab:analyze}}
+Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp.swmpr} at the command line).\label{tab:analyze}}
 \begin{center}
@@
 \hline
-\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
-\texttt{aggregate\_metab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.
+\texttt{aggreswmp.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
+\texttt{aggremetab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.
 This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
@@
 > # get mean DO by months
-> aggregate(dat, 'months', params = c('do_mgl'))
+> aggreswmp(dat, 'months', params = c('do_mgl'))
 \end{verbatim}

### 2015-04-30 internal_review_response.Rnw
started revisions from jwh

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}

%acronyms
% \acrodef{}{}

%knitr options
<<setup, cache = F, echo = F>>=
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\setlength{\parskip}{5mm}
\setlength{\parindent}{0in}

\begin{document}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to review from Dr. Jeff Hollister, ``SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries'', Beck, MW.}

Response to reviewer comments are provided in italics.  The original review is included as a pdf with markup comment. The original pages and line numbers are noted below.

Page 1: I describe SWMPr, an open-source R package, for use ...

{\it Changed.}

Page 2: growth

{\it Changed.}

Page 2: is limited

{\it Changed.}

Page 2: identifying the signal from the

{\it Changed.}

Page 2: This feels like you are questioning the use of this type of data? Also, try to transition to the next paragraph. Feels a bit abrupt.

{\it Modified as follows: `A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs [6, 7].  An important monitoring network for coastal environments in the United States is the National Estuarine Research Reserve System (NERRS). This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.'}

Page 2: both for

{\it Changed.}

Page 2: Why just abiotic? You should add a bit about this. Are the other data not available, just haven't done it yet? If the later, let me know, that'd be fun to work on!

{\it The description in the text is vague in this regard and I have modified for clarity: `Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.  '}

Page 3: space

{\it Added.}

Page 3: On CRAN now, correct? IF so update installation directions.

{\it Installation instructions were updated}

Page 3: Would bet a lot of readers won't have any idea what this means. Since the swmpr object is central to the package, maybe a section describing its structure and a line or two on S3 might be a good idea.

{\it Agreed.  This material was already in the text but not as a separate section.  I added a section describing the object structure, including the following text for clarity: `The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, [16]), such that specific methods for a generic function are developed for the object class.}

Page 4: Why only 100? web service limit?

{\it Added footnote to the table that describes the limitation.}

Page 4: Same as above. If there are service limitations should outline those. Just saw this explained below!

{\it Same as above.}

Page 4: This is going to be a tough one to keep relevant. What's the reason for including it?

Page 4: Couldn't you build this functionality into the function itself? Perhaps next version... No need to include prior to publishing this.

Page 5: Might be more informative to show the results for each request... dplyr style?

Page 5: If you built in the loop to get around the limit would this not be necessary? This seems like a bandaid that will be less and less relevant over time.

Page 5: Unless it is necessary, I wouldn't even mention this.

Page 5: Would be nice to have the decompression as part of the function.

Page 5: What is the source of the duplicates? Does it make sense to just remove them? Are the data duplicated? Could you just flag these in the output as opposed to removing? If you did this, you could then let the user decide when to remove or not remove with rem\_reps().

Page 6: Re: my comment above about the object... Might make this a separate section. Include a table instead of prose to describe the object. Could closely follow the output of str(my\_swmpr\_obj).

Page 7: setstep.swmpr??

Page 7: Not sure how best to refer to these S3 methods. You do it both ways (with .smpr in the table) and without in the text. Could keep it that way if you add a line to table caption. I bring it up, because I think new R users would be confused by this (I was for awhile...)

Page 9: Does PLOS allow Boxes? Like Box 3. Example of qaqc() funciton usage... That would be a good way to show these.

Page 9: quote

Page 9: quote

Page 10: A section on package dependencies might be a good addition.

Page 10: Include an example of this in the code examples below.

Page 10: rewrite. ``a signal from noisy data'' ??

Page 13: or

Page 14: May not be obvious that this is 60 min / 15 min... And character string options are a great idea!

Page 16: I'm just being picky, but the name on this one is tripping me up. I think I'd prefer to see eco\_metab. But, I realize that may not be an easy change, so up to you!

Page 17: Do all of these functions need to be exported? Could they simply be internal functions that the users don't see? If so, you could shorten or eliminate this section. Also you could create a Viz category of functions instead of this one. In that you could talk about the plot capabilites separate from the analysis section, and the include the reserve\_map function there.

Page 17: Should this be map\_reserve???

Page 19: Yank this. Goes without saying.

Page 20: Some of the figures have code that directly preceeds it. Other figures (like Fig 8) it is a bit more difficult to link code to figure. Would be nice to provide a link to code for all the figures. This way a reader could download a single script to reproduce the figures.

Page 20: Are these examples necessary? Feel kinda tacked on and why have links as supplementary info and not part of the text? If you are going to keep these, I would flesh this out a bit more, add in some screen captures as figures and links directly in the text.

Page 21: Same comment as above. Be nice to have ready access to code that reproduces this table.

\end{document}

### 2015-04-30 internal_review_response.tex
started revisions from jwh

--- 2015-04-30 internal_review_response.Rnw
@@
-\documentclass[letterpaper,12pt]{article}
+\documentclass[letterpaper,12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
@@
 %knitr options
-<<setup, cache = F, echo = F>>=
-# set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
-options(replace.assign=TRUE,width=90,digits=1)
-@
 
@@
 \setlength{\parindent}{0in}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}

### 2015-04-30 swmpr_manu.Rnw
started revisions from jwh

@@
 \acrodef{CDMO}{Centralized Data Management Office}
+\acrodef{CRAN}{Comprehensive R Archive Network}
 \acrodef{DO}{dissolved oxygen}
@@
 One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.
-This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries.
+This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
 \ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
 However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
-We describe a new open-source software package, SWMPr, developed in program R for use with \ac{SWMP} environmental data.
+I describe SWMPr, an open-source R package, for use with \ac{SWMP} environmental data.
 The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.
@@
 
-The development of low-cost, automated sensors that have the ability to collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.
+The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.
 These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
@@
 Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.
-However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.
-Growing concerns over the use of adequate quality assurance and control methods, challenges for synthesis and interpretation, and increased emphasis on exploratory-based analytical techniques have characterized applications of data from automated monitoring programs \cite{Campbell13,Millie13}.
-
-The \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}} is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.
-The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 300 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information.
+A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \cite{Campbell13,Millie13}.
+
+An important monitoring network for coastal environments in the United States is the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.
+This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
+As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
 The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
-These data have been applied in evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
+These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
 However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
@@
 Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
-The SWMPr package is developed for the continuous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.
+Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
+As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 
-The SWMPr package was developed for use with the R ($\geq$v3.0.0) statistical programming language \cite{RDCT14}.
-The SWMPr package can be installed from GitHub\footnote{\url{https://github.com/fawda123/SWMPr}} by executing the following commands at the R console prompt.
+The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}.
+The SWMPr package can be downloaded from the \ac{CRAN}\footnote{\url{http://cran.r-project.org/web/packages/SWMPr/index.html} for the \ac{CRAN} release, the development version can be installed  from \url{https://github.com/fawda123/SWMPr}.} by executing the following commands at the R console prompt.
 The package is loaded in the workspace using the \texttt{library} command.
@@
 <<eval = F, message = F>>=
-install.packages('devtools')
-library(devtools)
-install_github('fawda123/SWMPr')
+install.packages('SWMPr')
 library(SWMPr)
@@
 The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.
-Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.
+Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
 Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.
 An additional group of `miscellaneous' functions are included as helpers for the main functions.
-The following describes a general approach for working with \ac{SWMP} data using each category of functions.
+The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.
 
@@
 descrips <- c(
-  'Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
-  'Retrieve records of all parameters within a given date range for a station.
-Optional argument for a single parameter.
-Maximum of 1000 records.
-Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.',
+  'Retrieve up to 100 records\\textsuperscript{\\textit{a}} starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
+  'Retrieve up to 1000 records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
   'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
@@
 to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+foot <- '\\textsuperscript{\\textit{a}} \\footnotesize Server limits prevent retrieving more records in a single request.'
 
@@
   col.just=c("p{3.5in}"), 
+  insert.bottom = foot,
   label = 'tab:retrieve', 
@@
 
+\subsection*{The \texttt{swmpr} object class}
+
 All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.
-The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.
-Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.
+The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
+Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.
 The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
@@
 
-The \texttt{swmpr} object class was created for use with specific methods following the S3 object definition approach \cite{Wickham14}.
+The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.
+This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for a generic function are developed for the object class.
 A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.
@@
 Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.
-The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 300 stations within the 28 estuaries of the \acl{NERRS}.
+The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the \acl{NERRS}.
 This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.

### 2015-04-30 swmpr_manu.tex
started revisions from jwh

@@
 \begin{verbatim}
-> install.packages('devtools')
-> library(devtools)
-> install_github('fawda123/SWMPr')
+> install.packages('SWMPr')
 > library(SWMPr)
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), insert.bottom = foot,     label = "tab:retrieve", table.env = FALSE)%
 \begin{table}[!tbp]
@@
 \hline
-\texttt{all\_params}&Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
-\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
+\texttt{all\_params}&Retrieve up to 100 records\textsuperscript{\textit{a}} starting with the most recent at a given station, all parameters.
+Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
+\texttt{all\_params\_dtrng}&Retrieve up to 1000 records of all parameters within a given date range for a station.
 Optional argument for a single parameter.
-Maximum of 1000 records.
 Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
@@
 
-\end{table}
+\textsuperscript{\textit{a}} \footnotesize Server limits prevent retrieving more records in a single request.\end{table}
 
@@
 > all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
->   param = 'do_mgl')
++   param = 'do_mgl')
 > 

### 2015-05-03 internal_review_response.Rnw
Revisions for changes in CDMO import functions

@@
 
-{\it Added footnote to the table that describes the limitation.}
+{\it I've modified all functions that have limits on the number of records that can be downloaded, so this is no longer relevant.}
 
@@
 
+{\it Agreed, the description of this function was removed from the manuscript but it is retained in the package.
+I think it's useful to include but shouldn't be advertised as a primary means of importing data.}
+
 Page 4: Couldn't you build this functionality into the function itself? Perhaps next version... No need to include prior to publishing this.
+
+{\it The most recent version now includes this functionality.}
 
@@
 
+{\it Yes, I've removed this function from the manuscript as noted above.}
+
 Page 5: Unless it is necessary, I wouldn't even mention this.
+
+{\it Omitted.}
 

### 2015-05-03 swmpr_manu.Rnw
Revisions for changes in CDMO import functions

@@
 
-Three approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
+Two approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
 First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services.
@@
 <<results = 'asis', echo = FALSE>>=
-funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'import\\_remote', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
+funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
 descrips <- c(
-  'Retrieve up to 100 records\\textsuperscript{\\textit{a}} starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
-  'Retrieve up to 1000 records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
+  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
+  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
   'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
-  'Import site data from an independent server.
-These files have been downloaded from \\ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.',
-  'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
+  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
   'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
@@
 to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
-foot <- '\\textsuperscript{\\textit{a}} \\footnotesize Server limits prevent retrieving more records in a single request.'
 
@@
   col.just=c("p{3.5in}"), 
-  insert.bottom = foot,
   label = 'tab:retrieve', 
@@
 
-Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records.
-The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.
-Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Data retrieval functions to import data directly into R from the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records with each request.
+However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.
+Download time can be excessive for longer time series.
 
@@
 
-A second approach for importing \ac{SWMP} data into R uses the \texttt{import\_remote} function.
-This method imports site data from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
-This approach is advantageous because the files can be imported very quickly and an IP address does not need to be registered.
-The files include almost all available \ac{SWMP} data for each site up to December 2014.
-However, the files will not be regularly updated and \ac{CDMO} should always be used for the most recent and accurate data.
-The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.
-
-<<eval = F>>=
-# import a file from an independent server
-import_remote('acebbwq')
-@
-
-A third approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
+The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
 This approach is most appropriate for large, specific data requests.
 The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}.
-The function may also work using data from the data export system,\footnote{\url{http://cdmo.baruch.sc.edu/get/export.cfm}} although this feature has not been extensively tested.
-The zip downloads feature is an easy way to obtain data from multiple stations in one request.
+The zip downloads feature can be used to obtain data from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).

### 2015-05-03 swmpr_manu.tex
Revisions for changes in CDMO import functions

@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), insert.bottom = foot,     label = "tab:retrieve", table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
 \begin{table}[!tbp]
@@
 \hline
-\texttt{all\_params}&Retrieve up to 100 records\textsuperscript{\textit{a}} starting with the most recent at a given station, all parameters.
-Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
-\texttt{all\_params\_dtrng}&Retrieve up to 1000 records of all parameters within a given date range for a station.
+\texttt{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
+\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
 Optional argument for a single parameter.
@@
 \texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \ac{CDMO} using the zip downloads option.\tabularnewline
-\texttt{import\_remote}&Import site data from an independent server.
-These files have been downloaded from \ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.\tabularnewline
-\texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.
-Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
+\texttt{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
 \texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
@@
 
-\textsuperscript{\textit{a}} \footnotesize Server limits prevent retrieving more records in a single request.\end{table}
+\end{table}
 
@@
 
-A second approach for importing \ac{SWMP} data into R uses the \texttt{import\_remote} function.
-This method imports site data from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.
-This approach is advantageous because the files can be imported very quickly and an IP address does not need to be registered.
-The files include almost all available \ac{SWMP} data for each site up to December 2014.
-However, the files will not be regularly updated and \ac{CDMO} should always be used for the most recent and accurate data.
-The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # import a file from an independent server
-> import_remote('acebbwq')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-A third approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
+The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
 This approach is most appropriate for large, specific data requests.
@@
 > subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
->   select = c('atemp', 'totsorad'))
++   select = c('atemp', 'totsorad'))
 > 

### 2015-05-04 internal_review_response.Rnw
revisions up to new table addition for attributes

@@
 
+{\it I think this might create unecessary clutter if all the examples are shown.  I've added an example in the next section for the swmpr object class that shows the raw data.}
+
 Page 5: If you built in the loop to get around the limit would this not be necessary? This seems like a bandaid that will be less and less relevant over time.
@@
 
+{\it The function was modified to include this feature.
+The text was updated accordingly `The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed'}
+
 Page 5: What is the source of the duplicates?
@@
 
+{\it This section was revised for clarity, `Occasionally, non-unique observations are present in the raw data.
+These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.
+Erroneous duplicates with non-unique time stamps may also be present.'  I will consider adding functionality to handle duplicates but I don't think the issue of actual duplicates with non-unique time stamps is a huge issue.}
+
 Page 6: Re: my comment above about the object... Might make this a separate section. Include a table instead of prose to describe the object. Could closely follow the output of str(my\_swmpr\_obj).
+
+{\it A table was added to replace the paragraph description of the object attributes.  A separate section was also added in response to a previous comment}.
 

### 2015-05-04 internal_review_response.tex
revisions up to new table addition for attributes

@@
 
+{\it I think this might create unecessary clutter if all the examples are shown.  I've added an example in the next section for the swmpr object class that shows the raw data.}
+
 Page 5: If you built in the loop to get around the limit would this not be necessary? This seems like a bandaid that will be less and less relevant over time.
+
+{\it Yes, I've removed this function from the manuscript as noted above.}
 
@@
 
+{\it Omitted.}
+
 Page 5: Would be nice to have the decompression as part of the function.
+
+{\it The function was modified to include this feature.
+The text was updated accordingly `The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed'}
 

### 2015-05-04 swmpr_manu.Rnw
revisions up to new table addition for attributes

@@
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \texttt{import\_local} function can be used after the folder is decompressed.
-
-Occasionally, duplicate time stamps are present in the raw data.
-The \texttt{import\_local} function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
+The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed
+
+Occasionally, non-unique observations are present in the raw data.
+These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.
+Erroneous duplicates with non-unique time stamps may also be present.
+The \texttt{import\_local} function handles duplicate entries differently depending on the data type.
 For water quality and nutrient data, duplicate time stamps are simply removed.
-Note that nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
-Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
+Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
+Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
 Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
-Only duplicate values at 15 minutes are averaged for weather data.
+Only duplicate values at 15 minutes are averaged for weather data during import.
 
@@
 # import, do not include file extension
-import_local(path, 'apaebmet') 
+import_local(path, 'apadbwq') 
 @
@@
 Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.
-The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
-Attributes of a \texttt{swmpr} object can be viewed as follows.
+The following shows an example of the raw data imported using \texttt{all\_params}.
+
+<<>>=
+# import all paramaters for the station
+# three most recent records
+exdat <- all_params('apadbwq', Max = 3, trace = F)
+exdat
+@
+
+The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (\cref{tab:attributes}).
+These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
+The attributes are not visible with the raw data but can be viewed as follows.
 
 <<eval = F, cache = T>>=
-# import data
-dat <- import_remote('apadbwq')
+# import sample data from package
+data(apadbwq)
+dat <- apadbwq
 
@@
 attr(dat, 'station')
+@
+
+% attributes table
+<<results = 'asis', echo = FALSE>>=
+# create the attributes table
+data(apadbwq)
+dat <- apadbwq
+
+# view all attributes of dat
+atts <- paste0('\\texttt{', names(attributes(dat)), '}')
+atts <- gsub('_', '\\_', atts, fixed = T)
+atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
+atts_desc <- c(
+  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class',
+  'Row names of the data set, inherited from the \\texttt{data.frame} object class',
+  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
+  'Station identifier used by \\ac{NERRS} as a string with 7 or 8 characters',
+  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
+  'Indicates if \\ac{QAQC} columns are present in the raw data',
+  'Start and end dates for the raw data',
+  'Timezone of the station using the city/country format\\textsuperscript{\\textit{a}}',
+  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
+)
+
+to_tab <- data.frame(Attributes = atts, Class = atts_class, Description = atts_desc, stringsAsFactors = F)
+foot <- c('\\textsuperscript{\\textit{a}}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.
+For example, ``America/Jamaica" is the same as Eastern Standard Time.')
+
+library(Hmisc)
+
+latex(
+  to_tab[, -1, drop = F],
+  file = '',
+  caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.  Similar information can be obtained using the \\texttt{str} function.",
+  rowlabel = 'Attributes',
+  colheads = c('Class', 'Description'),
+  rowname = to_tab$Attributes,
+  caption.loc = 'top',
+  insert.bottom = foot,
+  col.just=c("p{0.75in}", "p{3.25in}"), 
+  label = 'tab:attributes', 
+  table.env = FALSE
+  )
 @

### 2015-05-04 swmpr_manu.tex
revisions up to new table addition for attributes

@@
 The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}.
-The function may also work using data from the data export system,\footnote{\url{http://cdmo.baruch.sc.edu/get/export.cfm}} although this feature has not been extensively tested.
-The zip downloads feature is an easy way to obtain data from multiple stations in one request.
+The zip downloads feature can be used to obtain data from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \texttt{import\_local} function can be used after the folder is decompressed.
-
-Occasionally, duplicate time stamps are present in the raw data.
-The \texttt{import\_local} function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).
+The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed
+
+Occasionally, non-unique observations are present in the raw data.
+These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.
+Erroneous duplicates with non-unique time stamps may also be present.
+The \texttt{import\_local} function handles duplicate entries differently depending on the data type.
 For water quality and nutrient data, duplicate time stamps are simply removed.
-Note that nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
-Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
+Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
+Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
 Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
-Only duplicate values at 15 minutes are averaged for weather data.
+Only duplicate values at 15 minutes are averaged for weather data during import.
 
@@
 > # import, do not include file extension
-> import_local(path, 'apaebmet') 
+> import_local(path, 'apadbwq') 
 \end{verbatim}
@@
 Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.
-The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).
-Attributes of a \texttt{swmpr} object can be viewed as follows.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # import data
-> dat <- import_remote('apadbwq')
+The following shows an example of the raw data imported using \texttt{all\_params}.
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # import all paramaters for the station
+> # three most recent records
+> exdat <- all_params('apadbwq', Max = 3, trace = F)
+> exdat
+##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
+## 1 2015-04-21 13:00:00   21      0   0.07        0   0     0     99
+## 2 2015-04-21 13:15:00   21      0   0.05        0   0     0     99
+## 3 2015-04-21 13:30:00   22      0   0.03        0   0     0     98
+##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
+## 1        0      9        0  0.02       0  8    0    2      0       NA
+## 2        0      9        0  0.02       0  8    0    2      0       NA
+## 3        0      8        0  0.02       0  8    0   16      0       NA
+##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
+## 1         -2    NA      -1  -0.01     NA        3         
+## 2         -2    NA      -1   0.00     NA        3         
+## 3         -2    NA      -1   0.00     NA        3
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (\cref{tab:attributes}).
+These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
+Attributes of a \texttt{swmpr} object can also be viewed in R as follows.
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # import sample data from package
+> data(apadbwq)
+> dat <- apadbwq
 > 
@@
 \end{knitrout}
+
+% attributes table
+%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.
+Similar information can be obtained using the \\texttt{str} function.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE)%
+\begin{table}[!tbp]
+\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.  Similar information can be obtained using the \texttt{str} function.\label{tab:attributes}} 
+\begin{center}
+\begin{tabular}{lp{0.75in}p{3.25in}}
+\hline\hline
+\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
+\hline
+\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class\tabularnewline
+\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class\tabularnewline
+\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}\tabularnewline
+\texttt{station}&character&Station identifier used by \ac{NERRS} as a string with 7 or 8 characters\tabularnewline
+\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}\tabularnewline
+\texttt{qaqc\_cols}&logical&Indicates if \ac{QAQC} columns are present in the raw data\tabularnewline
+\texttt{date\_rng}&POSIXct&Start and end dates for the raw data\tabularnewline
+\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{\textit{a}}\tabularnewline
+\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated\tabularnewline
+\hline
+\end{tabular}\end{center}
+
+\textsuperscript{\textit{a}}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.
+For example, ``America/Jamaica" is the same as Eastern Standard Time.\end{table}
 

### 2015-05-04 internal_review_response.Rnw
through the part about package dependencies

@@
 
+{\it See response to the next comment.}
+
 Page 7: Not sure how best to refer to these S3 methods.
@@
 
+{\it Agreed, I've removed the method identifier from each function in the tables to avoid confusion.}
+
 Page 9: Does PLOS allow Boxes? Like Box 3. Example of qaqc() funciton usage... That would be a good way to show these.
+
+{\it I've checked a few papers from PLOS One that describe software packages and most follow the format I've used, i.e., presentation of code preceded by the command prompt.}
 
@@
 
+{\it Added.}
+
 Page 9: quote
 
+{\it Added.}
+
 Page 10: A section on package dependencies might be a good addition.
+
+{\it I've added a footnote that directs readers to the package index page on CRAN where the dependencies and package imports can be viewed.
+I'm not sure it's worth describing each in detail since this is readily available in the source code.}
 

### 2015-05-04 swmpr_manu.Rnw
through the part about package dependencies

@@
 
-funcs <- c('comb.swmpr', 'qaqc.swmpr', 'qaqcchk.swmpr', 'rem\\_reps.swmpr', 'setstep.swmpr', 'subset.swmpr')
+funcs <- c('comb', 'qaqc', 'qaqcchk', 'rem\\_reps', 'setstep', 'subset')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
-  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb.swmpr} and can also be used with individual stations.',
+  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb} and can also be used with individual stations.',
   'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
@@
   file = '',
-  caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",
+  caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} at the command line).",
   rowlabel = 'Function',
@@
 Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.
-The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
+The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
 The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
@@
 The time series for merging depends on the value passed to the \texttt{method} argument.
-Passing \texttt{union} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
-Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
+Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
+Passing \texttt{`intersect`} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
 Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
-The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.
+The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages\footnote{See the listing on \ac{CRAN} for package imports and dependencies for SWMPr: \url{http://cran.r-project.org/web/packages/SWMPr/index.html}}.
 These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
@@
 
-funcs <- c('aggreswmp.swmpr', 'aggremetab.swmpr', 'ecometab.swmpr', 'decomp.swmpr', 'decomp\\_cj.swmpr', 'hist.swmpr', 'lines.swmpr', 'na.approx.swmpr', 'plot.swmpr', 'plot\\_metab.swmpr', 'plot\\_summary.swmpr', 'smoother.swmpr')
+funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   'Plot a histogram for a \\texttt{swmpr} object.',
-  'Add lines to an existing plot created with \\texttt{plot.swmpr}.',
+  'Add lines to an existing plot created with \\texttt{plot}.',
   'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
@@
   file = '',
-  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp.swmpr} at the command line).",
+  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} at the command line).",
   rowlabel = 'Function',

### 2015-05-04 swmpr_manu.tex
through the part about package dependencies

@@
 These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
-Attributes of a \texttt{swmpr} object can also be viewed in R as follows.
+The attributes are not visible with the raw data but can be viewed as follows.
 
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb.swmpr} at the command line).\label{tab:organize}} 
+\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb} at the command line).\label{tab:organize}} 
 \begin{center}
@@
 \hline
-\texttt{comb.swmpr}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+\texttt{comb}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.\tabularnewline
-\texttt{qaqc.swmpr}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
-\texttt{qaqcchk.swmpr}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.
+\texttt{qaqc}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
+\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.
 The output can be used to inform further processing.\tabularnewline
-\texttt{rem\_reps.swmpr}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
-\texttt{setstep.swmpr}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
-The function is used in \texttt{comb.swmpr} and can also be used with individual stations.\tabularnewline
-\texttt{subset.swmpr}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
+\texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
+\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
+The function is used in \texttt{comb} and can also be used with individual stations.\tabularnewline
+\texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
 \hline
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp.swmpr} at the command line).\label{tab:analyze}}
+\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} at the command line).\label{tab:analyze}} 
 \begin{center}
@@
 \hline
-\texttt{aggreswmp.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
-\texttt{aggremetab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.
+\texttt{aggreswmp}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
+\texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.
 This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
-\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
-\texttt{decomp.swmpr}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
+\texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
+\texttt{decomp}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
 This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.
 Decomposition of monthly or daily trends is possible.\tabularnewline
-\texttt{decomp\_cj.swmpr}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
+\texttt{decomp\_cj}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
 This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.
 Only monthly decomposition is possible.\tabularnewline
-\texttt{hist.swmpr}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
-\texttt{lines.swmpr}&Add lines to an existing plot created with \texttt{plot.swmpr}.\tabularnewline
-\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
-\texttt{plot.swmpr}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
-\texttt{plot\_metab.swmpr}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
-\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
-\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
+\texttt{hist}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
+\texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
+\texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
+\texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
+\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
+\texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
+\texttt{smoother}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
 \hline

### 2015-05-05 internal_review_response.Rnw
almost done with internal revisions, upload tangled script for reproducing figures/tables

@@
 
+{\it The example for \texttt{aggregswmp} was modified to show this issue.}
+
 Page 10: rewrite. ``a signal from noisy data'' ??
+
+{\it Changed.}
+
+Page 12, line 262: Omit.
+
+{\it The sentence was revised to remove the following: `...with arguments passed to the \texttt{filter} function from the R base installation'.}
 
@@
 
+{\it Changed.}
+
 Page 14: May not be obvious that this is 60 min / 15 min... And character string options are a great idea!
 
+{\it Text was modified for clarity: `...96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
+The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).'}
+
 Page 16: I'm just being picky, but the name on this one is tripping me up. I think I'd prefer to see eco\_metab. But, I realize that may not be an easy change, so up to you!
+
+{\it I'm trying to reduce the number of underscores in the function names so I've kept \texttt{ecometab}.}
+
+Page 14, line 310: Omit `respectively'.
+
+{\it Omitted.}
 
@@
 
+{\it I'm hesitant to change the Namespace of the package too much because I don't want to break any functions that depend on the exported functions in the miscellaneous category.
+I've also exported a lot of these functions because some users might be interested in seeing the source code, such as for the \texttt{calckl} function used by \texttt{ecometab}.
+That being said, I have shortened the section to emphasize these points and have moved the \texttt{map\_reserve} function to the previous section.
+I suppose looking at site locations could be considered `analysis' if you are using spatial location to interpret trends.}
+
 Page 17: Should this be map\_reserve???
 
+{\it Yes, changed.}
+
 Page 19: Yank this. Goes without saying.
+
+{\it Removed.}
 
@@
 
+{\it I've added an R script to the supporting information that reproduces the figures/tables in the manuscript.}
+
 Page 20: Are these examples necessary?
@@
 
+{\it }
+
 Page 21: Same comment as above. Be nice to have ready access to code that reproduces this table.
 
+{\it See response to above comment.}
+
 \end{document}

### 2015-05-05 swmpr_manu.Rnw
almost done with internal revisions, upload tangled script for reproducing figures/tables

@@
   file = '',
-  caption = 'Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).',
+  caption = "Retrieval functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 foot <- c('\\textsuperscript{\\textit{a}}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.
-For example, ``America/Jamaica" is the same as Eastern Standard Time.')
+For example, ``America/Jamaica" is used for Eastern Standard Time.')
 
@@
   file = '',
-  caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.  Similar information can be obtained using the \\texttt{str} function.",
+  caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",
   rowlabel = 'Attributes',
@@
   file = '',
-  caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} at the command line).",
+  caption = "Organizing functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
-Passing \texttt{`intersect`} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
+Passing \texttt{`intersect'} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
 Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
@@
 Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.
-Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data.
+Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
 
@@
 
-funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
+funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   'Add lines to an existing plot created with \\texttt{plot}.',
+  'Create a map of all stations in a reserve using the ggmap package.',
   'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
@@
   file = '',
-  caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} at the command line).",
+  caption = "Analysis functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 
-<<eval = F, cache = T>>=
-# get data
-data(apacpwq)
-dat <- qaqc(apacpwq)
-
-# get mean DO by months
-aggreswmp(dat, 'months', params = c('do_mgl'))
-@
-
-Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).
+<<eval = T, cache = T>>=
+# get data, keep all observations
+data(apacpnut)
+dat <- qaqc(apacpnut, qaqc_keep = NULL)
+
+# aggregate by quarters
+agg_dat <- aggreswmp(dat, by = 'quarters')
+nrow(agg_dat)
+
+# aggregate by quarters, remove rows with NA values
+# note the reduction in the number of rows
+agg_dat2 <- aggreswmp(dat, by = 'quarters', na.action = na.omit)
+nrow(agg_dat2)
+@
+
+Time series can be smoothed to better characterize a signal from noisy data (\Cref{fig:smooth_ex}).
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.
+The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
 The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
@@
 
-The disaggregation of time series into additive of multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
+The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
 The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).
@@
 The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
-For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (4 * 24).
-The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (4 * 24 * 365).
+For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
+The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
 For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
@@
 Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
-For both, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.
+For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
 Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
@@
 
+Finally, the \texttt{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
+This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
+The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
+
+<<map_ex, fig.height = 5, message = F, cache = T, fig.cap = "Locations of all sites at the Jacques Cousteau reserve using the \\texttt{map\\_reserve} function.">>=
+# plot the stations at Jacques Cousteau reserve
+map_reserve('jac')
+@
+
 \subsection*{Miscellaneous functions}
@@
 
-funcs <- c('calckl', 'map\\_reserve', 'metab\\_day', 'param\\_names', 'parser', 'swmpr', 'time\\_vec')
+funcs <- c('calckl', 'metab\\_day', 'param\\_names', 'parser', 'swmpr', 'time\\_vec')
 funcs <- paste0('\\texttt{', funcs, '}')
@@
   'Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \\texttt{ecometab} function.',
-  'Create a map of all stations in a reserve using the ggmap package.',
   'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\texttt{ecometab} function.',
@@
 
-For brevity, only the \texttt{reserve\_map} function is discussed.
-This function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
-The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
-
-<<map_ex, fig.height = 5, message = F, cache = T, fig.cap = "Locations of all sites at the Jacques Cousteau reserve using the \\texttt{map\\_reserve} function.">>=
-# plot the stations at Jacques Cousteau reserve
-map_reserve('jac')
-@
-
 \section*{Applications using the SWMPr package}
@@
 
-The ability to evaluate environmental characteristics between estuaries within the \ac{NERRS} program has been greatly improved using functions in the SWMPr package.
 This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
@@
 
+\subsection*{Source code for reproducing figures/tables}
+
+Source code as an R script for reproducing figures and tables in the manuscript: \href{}{}
+
 \section*{Acknowledgments}

### 2015-05-05 swmpr_manu.tex
almost done with internal revisions, upload tangled script for reproducing figures/tables

@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} at the command line).\label{tab:retrieve}} 
+\caption{Retrieval functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} for individual functions or \texttt{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}}
 \begin{center}
@@
 % attributes table
-%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.
-Similar information can be obtained using the \\texttt{str} function.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE)%
+%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.  Similar information can be obtained using the \texttt{str} function.\label{tab:attributes}} 
+\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.\label{tab:attributes}} 
 \begin{center}
@@
 \textsuperscript{\textit{a}}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.
-For example, ``America/Jamaica" is the same as Eastern Standard Time.\end{table}
+For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}
 
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb} at the command line).\label{tab:organize}} 
+\caption{Organizing functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \texttt{?comb} for individual functions or \texttt{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}}
 \begin{center}
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
 \begin{table}[!tbp]
-\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} at the command line).\label{tab:analyze}} 
+\caption{Analysis functions available from the SWMPr package.
+Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} for individual functions or \texttt{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}}
 \begin{center}
@@
 \texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
+\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
 \texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
@@
 \begin{verbatim}
-> # get data
-> data(apacpwq)
-> dat <- qaqc(apacpwq)
-> 
-> # get mean DO by months
-> aggreswmp(dat, 'months', params = c('do_mgl'))
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).
+> # get data, keep all observations
+> data(apacpnut)
+> dat <- qaqc(apacpnut, qaqc_keep = NULL)
+> 
+> # aggregate by quarters
+> agg_dat <- aggreswmp(dat, by = 'quarters')
+> nrow(agg_dat)
+## [1] 47
+> # aggregate by quarters, remove rows with NA values
+> # note the reduction in the number of rows
+> agg_dat2 <- aggreswmp(dat, by = 'quarters', na.action = na.omit)
+> nrow(agg_dat2)
+## [1] 16
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Time series can be smoothed to better characterize a signal from noisy data (\Cref{fig:smooth_ex}).
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.
+The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
 The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
@@
 \caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.\label{fig:metab_ex}}
+\end{figure}
+
+\end{knitrout}
+
+Finally, the \texttt{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
+This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
+The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \texttt{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> # plot the stations at Jacques Cousteau reserve
+> map_reserve('jac')
+\end{verbatim}
+\end{kframe}\begin{figure}[!ht]
+
+{\centering \includegraphics[width=\maxwidth]{figure/map_ex}
+
+}
+
+\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.\label{fig:map_ex}}
 \end{figure}
@@
 \texttt{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \texttt{ecometab} function.\tabularnewline
-\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
 \texttt{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \texttt{ecometab} function.\tabularnewline
@@
 
-For brevity, only the \texttt{reserve\_map} function is discussed.
-This function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
-The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument may have to be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # plot the stations at Jacques Cousteau reserve
-> map_reserve('jac')
-\end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/map_ex}
-
-}
-
-\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.\label{fig:map_ex}}
-\end{figure}
-
-\end{knitrout}
-
 \section*{Applications using the SWMPr package}

### 2015-05-05 swmpr_manu_code.R

@@ -1,0 +1,690 @@
+## ----setup, cache = F, echo = F------------------------------------------
+library(knitr)
+# set global chunk options
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small')
+options(replace.assign=TRUE,width=70,digits=1)
+
+## ----eval = F, message = F-----------------------------------------------
+## install.packages('SWMPr')
+## library(SWMPr)
+
+## ----eval = T, echo = F, message = F, cache = F--------------------------
+devtools::load_all('M:/docs/SWMPr')
+
+## ----results = 'asis', echo = FALSE--------------------------------------
+funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
+funcs <- paste0('\\texttt{', funcs, '}')
+
+descrips <- c(
+  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
+  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
+  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
+  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
+  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
+  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
+  )
+
+to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
+
+library(Hmisc)
+
+latex(
+  to_tab[, 'Description', drop = F],
+  file = '',
+  caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",
+  rowlabel = 'Function',
[…]

### 2015-05-05 internal_review_response.Rnw
internal review revisions done

@@
 
-{\it Response to review from Dr. Jeff Hollister, ``SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries'', Beck, MW.}
+{\it Response to review from Dr. Jeff Hollister, ``SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries'', Beck, MW. May 2015.}
 
-Response to reviewer comments are provided in italics.  The original review is included as a pdf with markup comment. The original pages and line numbers are noted below.
+Response to reviewer comments are provided in italics.  The original review is included as a pdf with markup comments. The original pages and line numbers are noted below.
 
@@
 {\it Modified as follows: `A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs [6, 7].
-An important monitoring network for coastal environments in the United States is the National Estuarine Research Reserve System (NERRS).
+An invaluable source of monitoring data for coastal environments in the United States is provided by the National Estuarine Research Reserve System (NERRS).
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.'}
@@
 
-{\it The description in the text is vague in this regard and I have modified for clarity: `Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
+{\it The description in the text is vague and I've modified for clarity: `Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
 As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.
@@
 I added a section describing the object structure, including the following text for clarity: `The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system, [16]), such that specific methods for a generic function are developed for the object class.}
+This object-oriented approach is standard for R (i.e., the S3 object  system, [16]), such that specific methods for generic functions are developed for the object class.}
 
@@
 
-{\it I've modified all functions that have limits on the number of records that can be downloaded, so this is no longer relevant.}
+{\it I've modified all functions that have limits on the number of records that can be downloaded.
+Text was amended to remove any description of the limits, with the exception of the following: `Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.
+However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.
+Download time can be excessive for longer time series.'}
 
@@
 
-{\it Agreed, the description of this function was removed from the manuscript but it is retained in the package.
+{\it Agreed, the description of this function was removed from the manuscript but it is retained in the package and briefly noted in the supporting information.
 I think it's useful to include but shouldn't be advertised as a primary means of importing data.}
@@
 
-{\it The most recent version now includes this functionality.}
+{\it I've added this functionality for the revision.}
 
@@
 {\it The function was modified to include this feature.
-The text was updated accordingly `The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed'}
+The text was updated accordingly `The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.'}
 
@@
 
-{\it I've checked a few papers from PLOS One that describe software packages and most follow the format I've used, i.e., presentation of code preceded by the command prompt.}
+{\it I've checked a few papers from PLOS One that describe software packages and most follow the format I've used, i.e., command prompt followed by code, no identifying box.}
 
@@
 {\it I've added a footnote that directs readers to the package index page on CRAN where the dependencies and package imports can be viewed.
-I'm not sure it's worth describing each in detail since this is readily available in the source code.}
+I'm not sure it's worth describing each in detail since this is readily available online and in the source code.}
 
@@
 
-{\it I've added an R script to the supporting information that reproduces the figures/tables in the manuscript.}
+{\it I've added an R script to the supporting information that reproduces the figures/tables in the manuscript, including a link to the data.}
 
@@
 
-{\it }
+{\it I've added links to each app as footnotes and also included a figure for the SWMP comparison map.
+The following text was added to the end of the paragraph: `Both of the apps provide previously unavailable utilities to interact with SWMP data using an intuitive online interface.
+This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
+The applications have been used extensively with over 300 hours noted in a single month.'}
 

### 2015-05-05 swmpr_manu.Rnw
internal review revisions done

@@
 
-An important monitoring network for coastal environments in the United States is the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.
+An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
 The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
@@
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed
+The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.
 
@@
 The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for a generic function are developed for the object class.
+This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.
 A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.
@@
 Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
-An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{supp_info}).
+An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{swmp_apps} section).
 
@@
 
-The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data (see the \nameref{supp_info}).
-The first web application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.
+The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data.
+The first web application\footnote{Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}} evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}).
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
@@
 The trend direction is blue for decreasing and red for increasing with time.
-The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function.
-The drop down menus can be used to select the station, date range, and parameter for plotting.
+The second application\footnote{Summarize trends in \ac{SWMP} parameters for an individual site: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}} provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function (output is identical to \cref{fig:summary_ex}).
+Drop down menus can be used to select the station, date range, and parameter for plotting.
 The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.
+Both of the apps provide previously unavailable utilities to interact with \ac{SWMP} data using an intuitive online interface.
+This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
+The applications have been used extensively with over 300 hours noted in a single month.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
+\caption{Online application for comparing trends in \ac{SWMP} data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
+\label{fig:swmp_comp}
+\end{center}
+\end{figure}
 
@@
 
-\subsection*{Trends in \ac{SWMP} parameters}
-\label{swmp_trends}
-Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}
-
-\subsection*{Monthly and annual summary of \ac{SWMP} parameters}
-\label{swmp_summary}
-Summarize trends in \ac{SWMP} parameters for an individual site: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}
-
 \subsection*{\ac{SWMP} data as R binary files}
 
-Quickly access all \ac{SWMP} data from 1995 through 2014.
+All \ac{SWMP} data from 1995 through 2014, one file per site.
 See the help file for \texttt{import\_remote} for more information: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}
@@
 
-Source code as an R script for reproducing figures and tables in the manuscript: \href{}{}
+Source code as an R script for reproducing figures and tables in the manuscript: \href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R}
+
+\subsection*{Data to reproduce figures/tables}
+
+Data used to create the figures and tables.  Files are in the .RData format: \href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data}
 

### 2015-05-08 swmpr_manu.Rnw
removed unused pdfs in figure folder, figures not shown in text

@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small')
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.00\\textwidth')
 options(replace.assign=TRUE,width=70,digits=1)
@@
   label = 'tab:retrieve', 
-  table.env = FALSE
+  table.env = FALSE, 
+  where = '!tbp'
   )
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # import sample data from package
@@
   'Start and end dates for the raw data',
-  'Timezone of the station using the city/country format\\textsuperscript{\\textit{a}}',
+'Timezone of the station using the city/country format\\textsuperscript{a}',
   'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
@@
 to_tab <- data.frame(Attributes = atts, Class = atts_class, Description = atts_desc, stringsAsFactors = F)
-foot <- c('\\textsuperscript{\\textit{a}}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.
+foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.
 For example, ``America/Jamaica" is used for Eastern Standard Time.')
@@
   label = 'tab:attributes', 
-  table.env = FALSE
+  table.env = FALSE, 
+  where = '!tbp'
   )
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # view available methods for swmpr class
@@
   label = 'tab:organize', 
-  table.env = FALSE
-  )
-@
-
-<<eval = F, cache = T>>=
+  table.env = FALSE, 
+  where = '!tbp'
+  )
+@
+
+<<eval = F, cache = F>>=
 # qaqc screen for a swmpr object, retain only '0'
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # view the number of observations in each QAQC flag
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # get nutrient data
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # import data
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # import, qaqc removal
@@
 
-<<eval = F, cache = T>>=
+<<eval = F, cache = F>>=
 # get nut, wq, and met data as separate objects
@@
   label = 'tab:analyze', 
-  table.env = FALSE
-  )
-@
-
-<<eval = T, cache = T>>=
+  table.env = FALSE, 
+  where = '!tbp'
+  )
+@
+
+<<eval = T, cache = F>>=
 # get data, keep all observations
@@
 
-<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly means (95\\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.">>=
+<<metab_ex, eval = TRUE, cache = T, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly means (95\\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.">>=
 ## import water quality and weather data
@@
   label = 'tab:misc', 
-  table.env = FALSE
+  table.env = FALSE, 
+  where = '!tbp'
   )
@@
 Two sites were chosen from each reserve that had the longest available time series.
-Sites were assigned to regions based on approximate geographic coordinates.', fig.height = 6, fig.width = 8, out.width = '\\textwidth'>>=
+Sites were assigned to regions based on approximate geographic coordinates.', fig.height = 6, fig.width = 8>>=
 # # packages to use
@@
   label = 'tab:metab_tab',
-  size = 'footnotesize'
+  size = 'footnotesize', 
+  where = '!tbp'
   )
@@
 \begin{center}
-\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
+\includegraphics[width = 0\textwidth]{figure/swmp_comp.pdf}
 \caption{Online application for comparing trends in \ac{SWMP} data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}

### 2015-05-08 swmpr_manu.tex
removed unused pdfs in figure folder, figures not shown in text

@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
@@
 % attributes table
-%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE)%
+%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
@@
 \texttt{date\_rng}&POSIXct&Start and end dates for the raw data\tabularnewline
-\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{\textit{a}}\tabularnewline
+\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}\tabularnewline
 \texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated\tabularnewline
@@
 
-\textsuperscript{\textit{a}}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.
+\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.
 For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/smooth_ex}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/smooth_ex-1}
 
@@
 
-\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.\label{fig:smooth_ex}}
+\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.}\label{fig:smooth_ex}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/interp_ex}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/interp_ex-1}
 
@@
 
-\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.\label{fig:interp_ex}}
+\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex1}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/decomp_ex1-1}
 
@@
 
-\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.\label{fig:decomp_ex1}}
+\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/decomp_ex2}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/decomp_ex2-1}
 
@@
 
-\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function.\label{fig:decomp_ex2}}
+\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function.}\label{fig:decomp_ex2}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/summary_ex}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/summary_ex-1}
 
@@
 \caption[Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function]{Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function.
-Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)\label{fig:summary_ex}}
+Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)}\label{fig:summary_ex}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/metab_ex}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/metab_ex-1}
 
@@
 
-\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.\label{fig:metab_ex}}
+\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.}\label{fig:metab_ex}
 \end{figure}
@@
 \end{verbatim}
-\end{kframe}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\maxwidth]{figure/map_ex}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/map_ex-1}
 
@@
 
-\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.\label{fig:map_ex}}
+\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.}\label{fig:map_ex}
 \end{figure}
@@
 Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE)%
+Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
@@
 \begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!ht]
-
-{\centering \includegraphics[width=\textwidth]{figure/metab_plo}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.00\textwidth]{figure/metab_plo-1}
 
@@
 Two sites were chosen from each reserve that had the longest available time series.
-Sites were assigned to regions based on approximate geographic coordinates.\label{fig:metab_plo}}
+Sites were assigned to regions based on approximate geographic coordinates.}\label{fig:metab_plo}
 \end{figure}
@@
 % table summary of trends
-%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{{\\it a}}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize")%
+%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{{\\it a}}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%
 \begin{table}[!tbp]

### 2015-05-10 swmpr_manu.Rnw
removed all footnotes

@@
 % Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
+\usepackage{setspace} 
+\doublespacing
 
@@
 
-An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.
+An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}).
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+As part of this effort, the \acf{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
 The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
@@
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.
+All data are made accessible through the \ac{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.
 The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.
@@
 As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.
-Web services\footnote{\url{http://cdmo.baruch.sc.edu/webservices.cfm}} are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
+Web services are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
 The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.
@@
 The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}.
-The SWMPr package can be downloaded from the \ac{CRAN}\footnote{\url{http://cran.r-project.org/web/packages/SWMPr/index.html} for the \ac{CRAN} release, the development version can be installed  from \url{https://github.com/fawda123/SWMPr}.} by executing the following commands at the R console prompt.
+The SWMPr package can be downloaded from the \ac{CRAN} by executing the following commands at the R console prompt.
 The package is loaded in the workspace using the \texttt{library} command.
@@
 First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services.
-To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.\footnote{Follow instructions here: \url{http://cdmo.baruch.sc.edu/webservices.cfm}}  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
+To do so, the IP address for the computer making the request must be registered by following instructions on the \ac{CDMO} website.
+The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
 
@@
 This approach is most appropriate for large, specific data requests.
-The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}.
+The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \ac{CDMO} website.
 The zip downloads feature can be used to obtain data from multiple stations in one request.
@@
 
-A sample dataset can be downloaded for use with the examples below.\footnote{\url{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}}  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
-Processed verions of the raw data are also included with the package as binary data files (RData).
+A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).
+This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.
 These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
@@
 
-The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags.\footnote{\ac{CDMO} flags: \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}} Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
+Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
 Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
-The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages\footnote{See the listing on \ac{CRAN} for package imports and dependencies for SWMPr: \url{http://cran.r-project.org/web/packages/SWMPr/index.html}}.
+The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
 These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
@@
 For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
-A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
+For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).
+Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
 
@@
 Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \ac{CDMO}.
-All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files.\footnote{Data processing script: \url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}} This resulted in a single \texttt{swmpr} object for each parameter at each site.
+All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}).
+This resulted in a single \texttt{swmpr} object for each parameter at each site.
 All files were then uploaded to a remote server for online access.
-An additional R script\footnote{Metabolism processing script: \url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}} was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
+An additional R script (see \nameref{met_scripts}) was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
 Two water quality sites with the longest time series at each reserve were used.
@@
 The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data.
-The first web application\footnote{Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}} evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}).
+The first web application evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
@@
 The trend direction is blue for decreasing and red for increasing with time.
-The second application\footnote{Summarize trends in \ac{SWMP} parameters for an individual site: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}} provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function (output is identical to \cref{fig:summary_ex}).
+The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
+The output is identical to \cref{fig:summary_ex}.
 Drop down menus can be used to select the station, date range, and parameter for plotting.
@@
 
+\subsection*{Sample data files}
+\label{supp_zip}
+
+\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \ac{CDMO}.
+These data can be retrieved for use in R with the \texttt{import\_local} function.
+The same data are included with the SWMPr package as binary RData files.
+
 \subsection*{\ac{SWMP} data as R binary files}
 
-All \ac{SWMP} data from 1995 through 2014, one file per site.
-See the help file for \texttt{import\_remote} for more information: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}
+\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \ac{SWMP} data from 1995 through 2014, one file per site.
+See the help file for \texttt{import\_remote} for more information.
+
+\subsection*{Metabolism scripts}
+\label{met_scripts}
+
+\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data processing script for all data requested from the CDMO using the zip downloads feature.
+\url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291} Metabolism script for all data from the previous script.
 
@@
 
-Source code as an R script for reproducing figures and tables in the manuscript: \href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R}
+\href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R} Source code as an R script for reproducing figures and tables in the manuscript.
 
@@
 
-Data used to create the figures and tables.  Files are in the .RData format: \href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data}
+\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the .RData format.
 

### 2015-05-10 swmpr_manu.Rnw
all acronyms defined using glossaries package, added glossary to beginning of text

@@
 % Version 3.1 February 2015
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig." instead of "Figure".
-% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - tabs/spacing/line breaks within cells to alter layout or alignment
-% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
-% - colors, shading, or graphic objects
-% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
 
@@
 \usepackage{rotating}
-
-% packages I added
-\usepackage{cleveref}
-\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
-\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 
+% packages I added
+\usepackage{cleveref}
+\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage[acronym]{glossaries}
+\makeglossaries
+\include{glossary}
+
 %% Include all macros below
@@
 
-% acronyms
-\acrodef{CDMO}{Centralized Data Management Office}
-\acrodef{CRAN}{Comprehensive R Archive Network}
-\acrodef{DO}{dissolved oxygen}
-\acrodef{NERRS}{National Estuarine Research Reserve System}
-\acrodef{QAQC}{quality assurance/quality control}
-\acrodef{SWMP}{System Wide Monitoring Program}
+% acronyms from glossaries package
+\newacronym{CDMO}{CDMO}{Centralized Data Management Office}
+\newacronym{CRAN}{CRAN}{Comprehensive R Archive Network}
+\newacronym{DO}{DO}{dissolved oxygen}
+\newacronym{NERRS}{NERRS}{National Estuarine Research Reserve System}
+\newacronym{QAQC}{QAQC}{quality assurance/quality control}
+\newacronym{SWMP}{SWMP}{System Wide Monitoring Program}
 
@@
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.
+One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \gls{NERRS}.
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-\ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
+\gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
 However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
-I describe SWMPr, an open-source R package, for use with \ac{SWMP} environmental data.
+I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.
 The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.
@@
 Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
-
 \linenumbers
+\glsaddall
+\printglossaries
 
@@
 
-An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}).
+An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acf{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
-The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
+As part of this effort, the \acrlong{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
 These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
-However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.
-\ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.
-
-This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
-SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acl{SWMP}.
+However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.
+\gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.
+
+This article describes a software package that was developed to address research needs of the \gls{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
+SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acrlong{SWMP}.
 Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
 Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.
-The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.
+The software is provided specifically for use with \gls{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.
 
@@
 
-Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
+Four core data elements are collected through the \gls{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
 Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
-As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.
+As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \gls{SWMP} data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the \ac{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.
-The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.
-
-The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data management.
-As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.
-Prior to any data request to the \ac{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs.
-All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+All data are made accessible through the \gls{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \gls{QAQC} measures are used to screen the information for accuracy and reliability.
+The final data include all timestamped observations including relevant \gls{QAQC} flags with the appropriate qualifier.
+
+The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.
+As such, \gls{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.
+Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs.
+All stations in the \gls{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
 For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').
 Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
-All reserve names, stations, and date ranges for each parameter type can be viewed on the \ac{CDMO} website.
+All reserve names, stations, and date ranges for each parameter type can be viewed on the \gls{CDMO} website.
 Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.
-Web services are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
-The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.
-For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.
+As noted below, the computer's IP address must be registered with \gls{CDMO} before using the data retrieval functions in SWMPr.
+Web services are provided by \gls{CDMO} for direct access to \gls{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
+The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \gls{CDMO} web services.
+For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \gls{SWMP} sites.
 The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.
@@
 The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}.
-The SWMPr package can be downloaded from the \ac{CRAN} by executing the following commands at the R console prompt.
+The SWMPr package can be downloaded from the \gls{CRAN} by executing the following commands at the R console prompt.
 The package is loaded in the workspace using the \texttt{library} command.
@@
 
-Two approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
-First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services.
-To do so, the IP address for the computer making the request must be registered by following instructions on the \ac{CDMO} website.
-The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.
+Two approaches can be used to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
+First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services.
+To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.
+The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.
 
@@
   'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
-  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
+  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\gls{CDMO} using the zip downloads option.',
   'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
@@
 
-Data retrieval functions to import data directly into R from the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
-Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records with each request.
+Data retrieval functions to import data directly into R from the \gls{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.
 However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.
@@
 
-The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.
+The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \gls{CDMO}.
 This approach is most appropriate for large, specific data requests.
-The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \ac{CDMO} website.
+The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \gls{CDMO} website.
 The zip downloads feature can be used to obtain data from multiple stations in one request.
@@
 The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
-Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.
+Correspondingg \gls{QAQC} columns for each parameter are also returned if provided by the initial data request.
 The following shows an example of the raw data imported using \texttt{all\_params}.
@@
   'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
-  'Station identifier used by \\ac{NERRS} as a string with 7 or 8 characters',
+  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters',
   "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
-  'Indicates if \\ac{QAQC} columns are present in the raw data',
+  'Indicates if \\gls{QAQC} columns are present in the raw data',
   'Start and end dates for the raw data',
@@
 A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).
-This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.
+This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.
 These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.
@@
 Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
-Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.
-The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).
+Values in the QAQC column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.
+The \gls{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
@@
 Only different data types can be combined.',
-  'Remove \\ac{QAQC} columns and remove data based on \\ac{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\ac{QAQC} columns are present. ', 
-  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\ac{QAQC} flags used by \\ac{CDMO}.  The output can be used to inform further processing.', 
+  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
+  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\gls{QAQC} flags used by \\gls{CDMO}.
+The output can be used to inform further processing.',
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
@@
 
-Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.
+Viewing the number of observations for each parameter that are assigned to a \gls{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.
 The \texttt{qaqcchk} function can be used to view this information.
@@
 
-Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.
+Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.
 The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
@@
 The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).
-The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \ac{QAQC} columns are present.
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.
 
@@
 Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
-An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{swmp_apps} section).
+An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \gls{SWMP} sites (see the \nameref{swmp_apps} section).
 
@@
 The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}.
-The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series.
-The nighttime average \ac{DO} flux is used to estimate respiration rates, while the daytime \ac{DO} flux is used to estimate net primary production.
-To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \ac{DO} flux rates are multiplied by 24.
-Similarly, the daytime \ac{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
+The diffusion-corrected \gls{DO} flux estimates are averaged separately over each day and night of the time series.
+The nighttime average \gls{DO} flux is used to estimate respiration rates, while the daytime \gls{DO} flux is used to estimate net primary production.
+To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \gls{DO} flux rates are multiplied by 24.
+Similarly, the daytime \gls{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
 Respiration rates are subtracted from daily net production estimates to yield gross production rates.
@@
 Several assumptions must be met for a valid interpretation of the results.
-First, the \ac{DO} time series is assumed to be a sample of the same water mass over time.
+First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.
 Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.
@@
 Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).
-These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.
+These functions are generally used within the main functions but may be useful for more customized evaluation of \gls{SWMP} data.
 
@@
   'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\texttt{ecometab} function.',
-  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes \\ac{QAQC} columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
-  'Parses HTML returned from \\ac{CDMO} data requests.  Used in the retrieval functions.',
+  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes \\gls{QAQC} columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
+  'Parses HTML returned from \\gls{CDMO} data requests.  Used in the retrieval functions.',
   'Creates a \\texttt{swmpr} object class.  Used in the data retrieval functions.',
@@
   caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \\gls{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
@@
 This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
-First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each \ac{NERRS} reserve using all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent \ac{SWMP} data.
-Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \ac{NERRS} reserves.
+First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each \gls{NERRS} reserve using all years of available data.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent \gls{SWMP} data.
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \gls{NERRS} reserves.
 The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \texttt{ecometab} function provides a more accurate estimate by including weather data in the calculation (see Caffrey et al.
@@
 
-Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \ac{CDMO}.
+Water quality and weather observations from January 1995 to December 2014 for all \gls{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \gls{CDMO}.
 All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}).
@@
 % metabolism figure
-<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\ac{NERRS} reserve.
+<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\gls{NERRS} reserve.
 Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.
@@
 
-cap <- 'Trends in metabolism for two sites at each of the \\ac{NERRS} reserves.
+cap <- 'Trends in metabolism for two sites at each of the \\gls{NERRS} reserves.
 Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
@@
 
-The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data.
-The first web application evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
+The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \gls{SWMP} data.
+The first web application evaluates trends in \gls{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
@@
 The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.
-Both of the apps provide previously unavailable utilities to interact with \ac{SWMP} data using an intuitive online interface.
+Both of the apps provide previously unavailable utilities to interact with \gls{SWMP} data using an intuitive online interface.
 This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
@@
 \includegraphics[width = 0\textwidth]{figure/swmp_comp.pdf}
-\caption{Online application for comparing trends in \ac{SWMP} data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
+\caption{Online application for comparing trends in \gls{SWMP} data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
 \label{fig:swmp_comp}
@@
 Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.
-The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the \acl{NERRS}.
+The \acrlong{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the \acrlong{NERRS}.
 This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
-Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package.
-
-The SWMPr package provides several functions to retrieve, organize, and analyze \ac{SWMP} data to more effectively address common challenges working with large datasets.
+Although the data provided by \gls{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \gls{NERRS} researchers and staff to more effectively evaluate \gls{SWMP} data could be greatly improved using the SWMPr package.
+
+The SWMPr package provides several functions to retrieve, organize, and analyze \gls{SWMP} data to more effectively address common challenges working with large datasets.
 The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (\Cref{tab:retrieve}).
-Established \ac{QAQC} methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different \ac{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
+Established \gls{QAQC} methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different \gls{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
 Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data.
@@
 In particular, the \texttt{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \cite{Odum56,Caffrey14}.
-The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the \ac{NERRS} estuaries using recent data to evaluate trends over time.
+The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the \gls{NERRS} estuaries using recent data to evaluate trends over time.
 Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
-Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \ac{SWMP} and \ac{NERRS}.
+Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \gls{SWMP} and \gls{NERRS}.
 
@@
 
-\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \ac{CDMO}.
+\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \gls{CDMO}.
 These data can be retrieved for use in R with the \texttt{import\_local} function.
@@
 
-\subsection*{\ac{SWMP} data as R binary files}
-
-\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \ac{SWMP} data from 1995 through 2014, one file per site.
+\subsection*{\gls{SWMP} data as R binary files}
+
+\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \gls{SWMP} data from 1995 through 2014, one file per site.
 See the help file for \texttt{import\_remote} for more information.
@@
 
-I acknowledge the significant work of \ac{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
-Thanks to Dwayne Porter and Melissa Ide from \ac{CDMO} for maintaining the online database.
-Thanks to Marie Bundy and Nikki Dix for giving me the opportunity to share this package with the broader \ac{NERRS} community.
+I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
+Thanks to Dwayne Porter and Melissa Ide from \gls{CDMO} for maintaining the online database.
+Thanks to Marie Bundy and Nikki Dix for giving me the opportunity to share this package with the broader \gls{NERRS} community.
 Thanks to Todd O'Brien for the inspiration for the online widgets.

### 2015-05-10 swmpr_manu.tex
all acronyms defined using glossaries package, added glossary to beginning of text

@@
 
-% packages I added
-\usepackage{cleveref}
-\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
-\usepackage[usenames,dvipsnames]{xcolor}
-
 % Remove comment for double spacing
@@
 Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
-\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \ac{CDMO} using the zip downloads option.\tabularnewline
+\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \gls{CDMO} using the zip downloads option.\tabularnewline
 \texttt{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
@@
 \texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}\tabularnewline
-\texttt{station}&character&Station identifier used by \ac{NERRS} as a string with 7 or 8 characters\tabularnewline
+\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters\tabularnewline
 \texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}\tabularnewline
-\texttt{qaqc\_cols}&logical&Indicates if \ac{QAQC} columns are present in the raw data\tabularnewline
+\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data\tabularnewline
 \texttt{date\_rng}&POSIXct&Start and end dates for the raw data\tabularnewline
@@
 Only different data types can be combined.\tabularnewline
-\texttt{qaqc}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
-\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.
+\texttt{qaqc}&Remove \gls{QAQC} columns and remove data based on \gls{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \gls{QAQC} columns are present. \tabularnewline
+\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \gls{QAQC} flags used by \gls{CDMO}.
 The output can be used to inform further processing.\tabularnewline
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \\gls{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
@@
 \caption{Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \ac{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of \gls{SWMP} data.
 Full documentation for each function is in the help file (e.g., execute \texttt{?calckl} at the command line).\label{tab:misc}}
@@
 \texttt{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
-Includes \ac{QAQC} columns with \texttt{f\_} prefix.
+Includes \gls{QAQC} columns with \texttt{f\_} prefix.
 Used in the data retrieval functions.\tabularnewline
-\texttt{parser}&Parses HTML returned from \ac{CDMO} data requests.  Used in the retrieval functions.\tabularnewline
+\texttt{parser}&Parses HTML returned from \gls{CDMO} data requests.  Used in the retrieval functions.\tabularnewline
 \texttt{swmpr}&Creates a \texttt{swmpr} object class.  Used in the data retrieval functions.\tabularnewline
@@
 
-\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \ac{NERRS} reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \ac{NERRS} reserve.
+\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \gls{NERRS} reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \gls{NERRS} reserve.
 Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.
@@
 {\footnotesize
-\caption{Trends in metabolism for two sites at each of the \ac{NERRS} reserves.
+\caption{Trends in metabolism for two sites at each of the \gls{NERRS} reserves.
 Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).

### 2015-05-10 cover_letter.tex
cover letter added

\documentclass[a4paper,12pt]{article} 
\usepackage{times}
\usepackage[top=1in,bottom=1in,left=1in,right=1in,nohead]{geometry}
\usepackage{graphics}

\begin{document}
\renewcommand{\rmdefault}{ptm}
\pagestyle{empty}

\setlength{\parindent}{0mm} 
\setlength{\parskip}{5mm}

\begin{flushright}
\today
\end{flushright}

\emph{To:}\newline
Dr. Damian Pattinson\newline
Editorial Director\newline
PLOS ONE

\emph{From:}\newline
Dr. Marcus W. Beck\newline
ORISE Research Participation Program\newline
US Environmental Protection Agency\newline
beck.marcus@epa.gov, 850-934-2480\vspace{0.1in}

Enclosed please find my manuscript, entitled `SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries', to be considered as a research article in PLOS ONE. A continuing challenge in the environmental sciences is the effective use of continuous monitoring data.  The increasing quantity of data from automated sensor networks suggests that more appropriate methods to synthesize and interpret this information are needed. This article describes the SWMPr (`swamper') software package that was developed to address research needs of the US National Estuarine Research Reserve System (NERRS) using the open-source statistical programming language R.  The System-Wide Monitoring Program (SWMP) was implemented by NERRS in 1995 to collect continuous water quality data at over 140 fixed stations at each of the estuaries in the reserve system.  The SWMPr package address many common issues working with large datasets from automated sensor networks, such as data pre-processing, combining data from different sources, and exploratory analyses to identify parameters of interest. Additionally, a cross-reserve comparison of water quality trends and ecosystem metabolism estimates is provided to illustrate the utility of the package.

The manuscript has not been previously submitted to PLOS ONE and I have had no prior interactions with PLOS. Please note that Dr. Jeffrey Hollister (US EPA) has already provided a review of an earlier draft. I greatly appreciate the opportunity to publish my work in PLOS ONE.  Below is a suggested list of appropriate editors to handle my manuscript.

Zaid Abdo,
Institution and Department: Agricultural Research Service,
UNITED STATES

Ruth H. Carmichael,
Dauphin Island Sea Lab; University of South Alabama,
UNITED STATES

Juan Carlos Molinero,
GEOMAR: Helmholtz Center for Ocean Research,
GERMANY

Tomoya Iwata,
University of Yamanashi,
JAPAN

\vspace{0.1in}

\hspace{4.5in}Respectfully,

\hspace{4.5in}Dr. Marcus W. Beck

\end{document}

### 2015-05-10 swmpr_manu.Rnw
cover letter added

@@
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \gls{NERRS}.
+One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \acrlong{NERRS}.
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
@@
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acrlong{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+As part of this effort, the \acrlong{SWMP} (\acrshort{SWMP}) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
 The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
@@
 
-\section*{Summary}
+\section*{Conclusions}
 
@@
 
+\section*{Acknowledgments}
+
+I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
+Thanks to Todd O'Brien for the inspiration for the online widgets.
+Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions.
+Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.
+The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S.
+Environmental Protection Agency.
+The use of trade names or products does not constitute endorsement by the US Government.
+
+\nolinenumbers
+
+\bibliography{swmpr_refs}
+
 \section*{Supporting Information}
@@
 
-\subsection*{Sample data files}
+\subsection*{S1 Sample data files}
 \label{supp_zip}
@@
 
-\subsection*{\gls{SWMP} data as R binary files}
+\subsection*{S2 \gls{SWMP} data as R binary files}
 
@@
 
-\subsection*{Metabolism scripts}
+\subsection*{S3 Metabolism scripts}
 \label{met_scripts}
@@
 
-\subsection*{Source code for reproducing figures/tables}
+\subsection*{S4 Source code for reproducing figures/tables}
 
@@
 
-\subsection*{Data to reproduce figures/tables}
+\subsection*{S5 Data to reproduce figures/tables}
 
@@
 
-\section*{Acknowledgments}
-
-I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
-Thanks to Dwayne Porter and Melissa Ide from \gls{CDMO} for maintaining the online database.
-Thanks to Marie Bundy and Nikki Dix for giving me the opportunity to share this package with the broader \gls{NERRS} community.
-Thanks to Todd O'Brien for the inspiration for the online widgets.
-Thanks to Mike Murrell and Jim Hagy III for assistance with documentation and implementation of the ecosystem metabolism functions.
-The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S.
-Environmental Protection Agency.
-The use of trade names or products does not constitute endorsement by the US Government.
-
-\nolinenumbers
-
-\bibliography{swmpr_refs}
-
 \end{document}

### 2015-05-11 swmpr_manu.Rnw
minor edits before submission

@@
 \renewcommand{\figurename}{Fig.}
+
+% bold/italic text
+\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
 
@@
 
-An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).
+An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}\glsunset{NERRS}).
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
@@
 The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.
-As such, \gls{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.
+As such, \gls{CDMO} provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.
 Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs.
@@
 
-The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.
+The SWMPr package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
 Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
@@
 
-Two approaches can be used to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
+Two basic approaches using SWMPr are available to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
 First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services.
@@
 Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.
-However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.
+However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.
 Download time can be excessive for longer time series.
@@
 Occasionally, non-unique observations are present in the raw data.
-These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.
-Erroneous duplicates with non-unique time stamps may also be present.
+These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.
 The \texttt{import\_local} function handles duplicate entries differently depending on the data type.
@@
 # this is an example path with the decompressed csv files
-path <- 'C:/my_path/'
+path <- 'C:/my_path'
+
+# this is an example path for zipped csv files
+path <- 'C:/my_path.zip'
 
@@
 atts_desc <- c(
-  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class',
-  'Row names of the data set, inherited from the \\texttt{data.frame} object class',
-  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
-  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters',
-  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
-  'Indicates if \\gls{QAQC} columns are present in the raw data',
-  'Start and end dates for the raw data',
-'Timezone of the station using the city/country format\\textsuperscript{a}',
-  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
+  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
+  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
+  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
+  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters.',
+  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
+  'Indicates if \\gls{QAQC} columns are present in the raw data.',
+  'Start and end dates for the data.',
+'Timezone of the station using the city/country format\\textsuperscript{a}.',
+  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
 )
@@
 This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.
-A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.
+A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.
 Available methods for the \texttt{swmpr} class are described below and can also be viewed:
@@
 This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.
-These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.
+These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.
 These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
@@
 
-The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
-Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
-Values in the QAQC column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.
-The \gls{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).
+The \texttt{qaqc} function is a simple screen to retain observations from the data with specified \gls{QAQC} flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
+Each parameter in the imported \texttt{swmpr} object will have a corresponding \gls{QAQC} column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+Values in the \gls{QAQC} column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.
+The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
 Erroneous data entered as -99 are also removed.
-The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
+The function returns the original data with the \gls{QAQC} columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
 
@@
   'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
-  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\gls{QAQC} flags used by \\gls{CDMO}.
-The output can be used to inform further processing.',
+  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each \\gls{QAQC} flag used by \\gls{CDMO}.  The output can be used to inform further processing.', 
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
-  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb} and can also be used with individual stations.',
+  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
   'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
@@
 Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
-The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).
+The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).
 The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.
@@
 <<eval = F, cache = F>>=
-# get nutrient data
+# get nutrient data, filter qaqc flags
 data(apacpnut)
@@
 The date can be a single value or as two dates to select records within the range.
-The former case requires a binary operator as a character string passed to the argument, such as \texttt{`>'} or \texttt{`<='}.
+The former case requires a binary operator as a character string passed to the \texttt{operator} argument, such as \texttt{`>'} or \texttt{`<='}.
 The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').
@@
 The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.
-This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.
+This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.
 The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
@@
 The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.
-The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
-All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input.
+The \texttt{setstep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
+All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input to \texttt{comb}.
 The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
@@
 Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
-Passing \texttt{`intersect'} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.
+Passing \texttt{`intersect'} to \texttt{method} will create a continuous time series from the set of dates that are shared between all input objects.
 Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
@@
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
-Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.
+Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.
 A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
@@
   'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
-  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw metabolism data.',
+  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
   'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
@@
   'Add lines to an existing plot created with \\texttt{plot}.',
-  'Create a map of all stations in a reserve using the ggmap package.',
+  'Create a map of all stations in a reserve using the ggmap package \\cite{Kahle13}.',
   'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
   'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
-  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
+  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a combined \\texttt{swmpr} object.',  
   'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
@@
 
-# smooth
+# smooth using a window of 50 observations
 dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
@@
 
-<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \\texttt{decomp} function.">>=
+<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \\texttt{decomp} function.">>=
 # get data
 data(apadbwq)
-swmp1 <- apadbwq
+dat <- apadbwq
 
 # subset for daily decomposition
-dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
-
-# decomposition and plot
-test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-plot(test)
+dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
+
+# daily decomposition of DO and plot
+dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+plot(dc_dat)
 @
@@
 The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.
-Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
+The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
 For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
@@
 First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.
-Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.
+Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.
 The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
@@
 
-The following is an example that shows use of the function from a combined water quality and weather data set.
+The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.
 Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
@@
 Two water quality sites with the longest time series at each reserve were used.
-Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
+Mean annual metabolism values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
 All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
@@
 Sites were assigned to regions based on approximate geographic coordinates.'
-foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
+foots <- '\\footnotesize{\\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
 
@@
   caption = cap, 
-  cgroup = c('NEM\\textsuperscript{{\\it a}}', 'Pg', 'Rt'), 
+  cgroup = c('NEM\\textsuperscript{a}', 'Pg', 'Rt'), 
   n.cgroup = c(2, 2, 2), 
@@
 Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.
-The trend direction is blue for decreasing and red for increasing with time.
+The trend over time is blue for decreasing and red for increasing.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
-The output is identical to \cref{fig:summary_ex}.
-Drop down menus can be used to select the station, date range, and parameter for plotting.
+The output is identical to \cref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
 The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.
@@
 Established \gls{QAQC} methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different \gls{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
-Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data.
+Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize and combine time series.
 Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.
@@
 Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
-Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \gls{SWMP} and \gls{NERRS}.
+Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \gls{CDMO}, \gls{SWMP}, and \gls{NERRS}.
 
@@
 Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.
-The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S.
+The views expressed in this article are those of the author and do not necessarily reflect the views or policies of the U.S.
 Environmental Protection Agency.
@@
 
-\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data processing script for all data requested from the CDMO using the zip downloads feature.
+\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data pre-processing script for all data requested from the CDMO using the zip downloads feature.
 \url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291} Metabolism script for all data from the previous script.
@@
 
-\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the .RData format.
+\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the RData format.
 

### 2015-05-11 swmpr_manu.tex
minor edits before submission

@@
 
+% bold/italic text
+\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
+
 % acronyms from glossaries package
@@
 > # this is an example path with the decompressed csv files
-> path <- 'C:/my_path/'
+> path <- 'C:/my_path'
+> 
+> # this is an example path for zipped csv files
+> path <- 'C:/my_path.zip'
 > 
@@
 \hline
-\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class\tabularnewline
-\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class\tabularnewline
-\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}\tabularnewline
-\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters\tabularnewline
-\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}\tabularnewline
-\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data\tabularnewline
-\texttt{date\_rng}&POSIXct&Start and end dates for the raw data\tabularnewline
-\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}\tabularnewline
-\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated\tabularnewline
+\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class.\tabularnewline
+\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class.\tabularnewline
+\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}.\tabularnewline
+\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters.\tabularnewline
+\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}, \texttt{`turb'}, etc.\tabularnewline
+\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data.\tabularnewline
+\texttt{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
+\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
+\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
 \hline
@@
 \texttt{qaqc}&Remove \gls{QAQC} columns and remove data based on \gls{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \gls{QAQC} columns are present. \tabularnewline
-\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \gls{QAQC} flags used by \gls{CDMO}.
+\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each \gls{QAQC} flag used by \gls{CDMO}.
 The output can be used to inform further processing.\tabularnewline
 \texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
-\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.
-The function is used in \texttt{comb} and can also be used with individual stations.\tabularnewline
+\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \texttt{comb}.\tabularnewline
 \texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
@@
 \begin{verbatim}
-> # get nutrient data
+> # get nutrient data, filter qaqc flags
 > data(apacpnut)
@@
 \texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.
-This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
+This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
 \texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
@@
 \texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
-\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
+\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package \cite{Kahle13}.\tabularnewline
 \texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
 \texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
-\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
+\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a combined \texttt{swmpr} object.\tabularnewline
 \texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
@@
 > 
-> # smooth
+> # smooth using a window of 50 observations
 > dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
@@
 > data(apadbwq)
-> swmp1 <- apadbwq
+> dat <- apadbwq
 > 
 > # subset for daily decomposition
-> dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
-> 
-> # decomposition and plot
-> test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-> plot(test)
+> dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
+> 
+> # daily decomposition of DO and plot
+> dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+> plot(dc_dat)
 \end{verbatim}
@@
 
-\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
+\caption[An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
 \end{figure}
@@
 % table summary of trends
-%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{{\\it a}}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%
+%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{a}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%
 \begin{table}[!tbp]
@@
 \hline\hline
-\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{{\it a}}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline
+\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline
 \cline{2-3} \cline{5-6} \cline{8-9}
@@
 
-\footnotesize{\textsuperscript{{\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}
+\footnotesize{\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}
 

### 2015-09-01 swmpr_manu.tex
created plos reject branch for future ref

@@
 \usepackage{setspace} 
-\doublespacing
+% \doublespacing
 
@@
 ##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
-## 1 2015-04-21 13:00:00   21      0   0.07        0   0     0     99
-## 2 2015-04-21 13:15:00   21      0   0.05        0   0     0     99
-## 3 2015-04-21 13:30:00   22      0   0.03        0   0     0     98
+## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
+## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
+## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
 ##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
-## 1        0      9        0  0.02       0  8    0    2      0       NA
-## 2        0      9        0  0.02       0  8    0    2      0       NA
-## 3        0      8        0  0.02       0  8    0   16      0       NA
+## 1        0      5        0     2       0  8    0   13      0       NA
+## 2        0      5        0     2       0  8    0    9      0       NA
+## 3        0      6        0     2       0  8    0   11      0       NA
 ##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
-## 1         -2    NA      -1  -0.01     NA        3         
-## 2         -2    NA      -1   0.00     NA        3         
-## 3         -2    NA      -1   0.00     NA        3
+## 1         -2    NA      -1     NA     NA       -1         
+## 2         -2    NA      -1     NA     NA       -1         
+## 3         -2    NA      -1     NA     NA       -1
 \end{verbatim}

### 2015-09-01 beck.Rnw
working template for R journal

% !TeX root = RJwrapper.tex
\title{Capitalized Title Here}
\author{by Author One, Author Two and Author Three}

\maketitle

\abstract{
An abstract of less than 150 words.
}

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

\section{Section title in sentence case}

This section may contain a figure such as Figure~\ref{figure:rlogo}.

\begin{figure}[htbp]
  \centering
  \includegraphics{Rlogo}
  \caption{The logo of R.}
  \label{figure:rlogo}
\end{figure}

\section{Another section}

There will likely be several sections, perhaps including code snippets, such as:

\begin{example}
  x <- 1:10
  result <- myFunction(x)
\end{example}

\section{Summary}

This file is only a basic article template. For full details of \emph{The R Journal} style and information on how to prepare your article for submission, see the \href{http://journal.r-project.org/share/author-guide.pdf}{Instructions for Authors}.

\bibliography{beck}

\address{Author One\\
  Affiliation\\
  Address\\
  Country\\}
\email{author1@work}

\address{Author Two\\
  Affiliation\\
  Address\\
  Country\\}
\email{author2@work}

\address{Author Three\\
  Affiliation\\
  Address\\
  Country\\}
\email{author3@work}

### 2015-09-01 swmpr_manu_code.R

@@ -1,690 +1,0 @@
-## ----setup, cache = F, echo = F------------------------------------------
-library(knitr)
-# set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small')
-options(replace.assign=TRUE,width=70,digits=1)
-
-## ----eval = F, message = F-----------------------------------------------
-## install.packages('SWMPr')
-## library(SWMPr)
-
-## ----eval = T, echo = F, message = F, cache = F--------------------------
-devtools::load_all('M:/docs/SWMPr')
-
-## ----results = 'asis', echo = FALSE--------------------------------------
-funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
-funcs <- paste0('\\texttt{', funcs, '}')
-
-descrips <- c(
-  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
-  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
-  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
-  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
-  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
-  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
-  )
-
-to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
-
-library(Hmisc)
-
-latex(
-  to_tab[, 'Description', drop = F],
-  file = '',
-  caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",
-  rowlabel = 'Function',
[…]

### 2015-09-02 swmpr_manu.tex
successful compile using R journal template

@@
-% Template for PLoS
-% Version 3.1 February 2015
-
-\documentclass[10pt,letterpaper]{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[a4paper]{report}\usepackage[]{graphicx}\usepackage{xcolor}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{alltt}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
 \usepackage[utf8]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% fixltx2e package for \textsubscript
-\usepackage{fixltx2e}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% rotating package for sideways tables
-\usepackage{rotating}
-
-% Remove comment for double spacing
-\usepackage{setspace} 
-% \doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-% for acronyms
-\usepackage{acronym}
-
-% packages I added
-\usepackage{cleveref}
-\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
-\usepackage[usenames,dvipsnames]{xcolor}
-\usepackage[acronym]{glossaries}
-\makeglossaries
-\include{glossary}
-
-%% Include all macros below
-
-% cleveref options
-\crefname{table}{Table}{Tables}
-\crefname{figure}{Fig.}{Figs.}
-\renewcommand{\figurename}{Fig.}
+\usepackage[T1]{fontenc}
+\usepackage{RJournal}
+\usepackage{amsmath,amssymb,array}
+\usepackage{booktabs}
+
+%% load any required packages
 
@@
 \newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
-
-% acronyms from glossaries package
-\newacronym{CDMO}{CDMO}{Centralized Data Management Office}
-\newacronym{CRAN}{CRAN}{Comprehensive R Archive Network}
-\newacronym{DO}{DO}{dissolved oxygen}
-\newacronym{NERRS}{NERRS}{National Estuarine Research Reserve System}
-\newacronym{QAQC}{QAQC}{quality assurance/quality control}
-\newacronym{SWMP}{SWMP}{System Wide Monitoring Program}
-
-%% END MACROS SECTION
-
-%knitr options
-
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
-\vspace*{0.35in}
-\urlstyle{same}
-
-% Title must be 250 characters or less.
-% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
-\begin{flushleft}
-{\Large
-\textbf\newline{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Marcus William Beck\textsuperscript{1, *}
-
-\bigskip
-\bf{1} ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
-\\
-\bigskip
-
-* beck.marcus@epa.gov
-
-\end{flushleft}
+
+%% do not edit, for illustration only
+\sectionhead{Contributed research article}
+\volume{XX}
+\volnumber{YY}
+\year{20ZZ}
+\month{AAAA}
+
+\begin{article}
+
+%knitr options
+
+\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
+\author{by Marcus W Beck}
+% ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
+
+\maketitle
 
@@
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \acrlong{NERRS}.
+One example in the United States is the System-Wide Monitoring Program (SWMP) that was implemented in 1995 by the federally-funded National Estuarine Research Reserve System.
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-\gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
+SWMP data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
 However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
-I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.
+I describe SWMPr, an open-source R package, for use with SWMP environmental data.
 The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.
@@
 Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
-\linenumbers
-\glsaddall
-\printglossaries
-
-\section*{Introduction}
+
+\section{Introduction}
 
@@
 
-An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}\glsunset{NERRS}).
+An invaluable source of monitoring data for coastal environments in the United States is provided by the National Estuarine Research Reserve System (NERRS, \url{http://www.nerrs.noaa.gov/}).
 This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the \acrlong{SWMP} (\acrshort{SWMP}) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
-The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
+As part of this effort, the System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The SWMP network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
 These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
-However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.
-\gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.
-
-This article describes a software package that was developed to address research needs of the \gls{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.
-SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acrlong{SWMP}.
+However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current SWMP data.
+NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \cite{SWMP14}.
+
+This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R \cite{RDCT14}.
+SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.
 Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
 Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.
-The software is provided specifically for use with \gls{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.
-
-\section*{SWMP overview and data retrieval}
-
-Four core data elements are collected through the \gls{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
+The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large datasets.
+
+\section{SWMP overview and data retrieval}
+
+Four core data elements are collected through the SWMP monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
 Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
-As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \gls{SWMP} data and, consequently, the most challenging to evaluate.
+As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the \gls{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \gls{QAQC} measures are used to screen the information for accuracy and reliability.
-The final data include all timestamped observations including relevant \gls{QAQC} flags with the appropriate qualifier.
-
-The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.
-As such, \gls{CDMO} provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.
-Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs.
-All stations in the \gls{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+All data are made accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.
+The final data include all timestamped observations including relevant QAQC flags with the appropriate qualifier.
+
+The CDMO web portal was established to support priority areas of SWMP that focus on the continuation and advancement of data management.
+As such, CDMO provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic QAQC procedures.
+Prior to any data request to the CDMO, the location, parameter type, and date ranges need to be identified based on the analysis needs.
+All stations in the SWMP network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
 For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').
 Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
-All reserve names, stations, and date ranges for each parameter type can be viewed on the \gls{CDMO} website.
+All reserve names, stations, and date ranges for each parameter type can be viewed on the CDMO website.
 Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered with \gls{CDMO} before using the data retrieval functions in SWMPr.
-Web services are provided by \gls{CDMO} for direct access to \gls{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.
-The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \gls{CDMO} web services.
-For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \gls{SWMP} sites.
+As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in SWMPr.
+Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.
+The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on CDMO web services.
+For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.
 The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.
 
-\section*{Structure of the SWMPr package}
-
-\subsection*{Installing the package}
+\section{Structure of the SWMPr package}
+
+\subsection{Installing the package}
 
 The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}.
-The SWMPr package can be downloaded from the \gls{CRAN} by executing the following commands at the R console prompt.
+The SWMPr package can be downloaded from CRAN by executing the following commands at the R console prompt.
 The package is loaded in the workspace using the \texttt{library} command.
@@
 
-\subsection*{Data retrieval}
-
-Two basic approaches using SWMPr are available to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}).
-First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services.
-To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.
-The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.
+\subsection{Data retrieval}
+
+Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
+First, functions from the package can be used to import the data directly from the online server using CDMO web services.
+To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.
+The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
 
@@
 Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
-\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \gls{CDMO} using the zip downloads option.\tabularnewline
+\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
 \texttt{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
@@
 
-Data retrieval functions to import data directly into R from the \gls{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
-Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.
+Data retrieval functions to import data directly into R from the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.
 However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.
@@
 
-The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \gls{CDMO}.
+The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from CDMO.
 This approach is most appropriate for large, specific data requests.
-The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \gls{CDMO} website.
+The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website.
 The zip downloads feature can be used to obtain data from multiple stations in one request.
@@
 
-\subsection*{The \texttt{swmpr} object class}
+\subsection{The \texttt{swmpr} object class}
 
@@
 The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
-Correspondingg \gls{QAQC} columns for each parameter are also returned if provided by the initial data request.
+Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.
 The following shows an example of the raw data imported using \texttt{all\_params}.
@@
 
-The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (\cref{tab:attributes}).
+The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).
 These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
@@
 \texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}.\tabularnewline
-\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters.\tabularnewline
+\texttt{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
 \texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}, \texttt{`turb'}, etc.\tabularnewline
-\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data.\tabularnewline
+\texttt{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
 \texttt{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
@@
 
-A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).
-This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.
+A sample dataset can be downloaded for use with the examples below.
+This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
 These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.
@@
 
-\subsection*{Data organizing}
-
-The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (\Cref{tab:organize}).
-
-The \texttt{qaqc} function is a simple screen to retain observations from the data with specified \gls{QAQC} flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
-Each parameter in the imported \texttt{swmpr} object will have a corresponding \gls{QAQC} column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
-Values in the \gls{QAQC} column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.
-The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).
+\subsection{Data organizing}
+
+The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).
+
+The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
+Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.
+The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
 Erroneous data entered as -99 are also removed.
-The function returns the original data with the \gls{QAQC} columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
+The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
 
@@
 Only different data types can be combined.\tabularnewline
-\texttt{qaqc}&Remove \gls{QAQC} columns and remove data based on \gls{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \gls{QAQC} columns are present. \tabularnewline
-\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each \gls{QAQC} flag used by \gls{CDMO}.
+\texttt{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present. \tabularnewline
+\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.
 The output can be used to inform further processing.\tabularnewline
@@
 
-Viewing the number of observations for each parameter that are assigned to a \gls{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.
+Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \texttt{qaqc}.
 The \texttt{qaqcchk} function can be used to view this information.
@@
 
-Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.
+Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
 The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
@@
 The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).
-The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.
+The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
 
@@
 
-\subsection*{Data analysis}
-
-The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).
+\subsection{Data analysis}
+
+The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).
 The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
@@
 
-Time series can be smoothed to better characterize a signal from noisy data (\Cref{fig:smooth_ex}).
+Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
@@
 In either case, an analyst should have a strong rationale for the chosen method.
-A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).
+A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (Figure~\ref{fig:interp_ex}).
 A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
@@
 The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
-The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).
+The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
 An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
@@
 Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-\Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
+Figure~\ref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
@@
 Detailed exploratory graphics are also useful for evaluating general trends in observed data.
-Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (Figure~\ref{fig:summary_ex}).
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
 Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
-An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \gls{SWMP} sites (see the \nameref{swmp_apps} section).
+An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
 
@@
 The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}.
-The diffusion-corrected \gls{DO} flux estimates are averaged separately over each day and night of the time series.
-The nighttime average \gls{DO} flux is used to estimate respiration rates, while the daytime \gls{DO} flux is used to estimate net primary production.
-To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \gls{DO} flux rates are multiplied by 24.
-Similarly, the daytime \gls{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
+The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series.
+The nighttime average DO flux is used to estimate respiration rates, while the daytime DO flux is used to estimate net primary production.
+To generate daily integrated rates, respiration rates are assumed constant such that hourly night time DO flux rates are multiplied by 24.
+Similarly, the daytime DO flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
 Respiration rates are subtracted from daily net production estimates to yield gross production rates.
@@
 Several assumptions must be met for a valid interpretation of the results.
-First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.
+First, the DO time series is assumed to be a sample of the same water mass over time.
 Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.
@@
 The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.
-Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
+Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (Figure~\ref{fig:metab_ex}).
 
@@
 
-\subsection*{Miscellaneous functions}
-
-Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).
-These functions are generally used within the main functions but may be useful for more customized evaluation of \gls{SWMP} data.
+\subsection{Miscellaneous functions}
+
+Several additional functions are provided that do not fit the above categories (Table~\ref{tab:misc}).
+These functions are generally used within the main functions but may be useful for more customized evaluation of SWMP data.
 
@@
 %latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \\gls{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
@@
 \caption{Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \gls{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
 Full documentation for each function is in the help file (e.g., execute \texttt{?calckl} at the command line).\label{tab:misc}}
@@
 \texttt{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
-Includes \gls{QAQC} columns with \texttt{f\_} prefix.
+Includes QAQC columns with \texttt{f\_} prefix.
 Used in the data retrieval functions.\tabularnewline
-\texttt{parser}&Parses HTML returned from \gls{CDMO} data requests.  Used in the retrieval functions.\tabularnewline
+\texttt{parser}&Parses HTML returned from CDMO data requests.  Used in the retrieval functions.\tabularnewline
 \texttt{swmpr}&Creates a \texttt{swmpr} object class.  Used in the data retrieval functions.\tabularnewline
@@
 
-\section*{Applications using the SWMPr package}
+\section{Applications using the SWMPr package}
 \label{swmp_apps}
@@
 This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
-First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each \gls{NERRS} reserve using all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent \gls{SWMP} data.
-Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the \gls{NERRS} reserves.
+First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each NERRS reserve using all years of available data.
+The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent SWMP data.
+Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.
 The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \texttt{ecometab} function provides a more accurate estimate by including weather data in the calculation (see Caffrey et al.
@@
 
-Water quality and weather observations from January 1995 to December 2014 for all \gls{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \gls{CDMO}.
-All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}).
+Water quality and weather observations from January 1995 to December 2014 for all NERRS sites were obtained through a bulk data request using the zip downloads feature of CDMO.
+All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files.
 This resulted in a single \texttt{swmpr} object for each parameter at each site.
 All files were then uploaded to a remote server for online access.
-An additional R script (see \nameref{met_scripts}) was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
+An additional R script was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
 Two water quality sites with the longest time series at each reserve were used.
-Mean annual metabolism values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.
+Mean annual metabolism values at each site, organized by region, are shown in Figure~\ref{fig:metab_plo}, whereas decadal comparisons are shown in Table~\ref{tab:metab_tab}.
 All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
@@
 
-\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \gls{NERRS} reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \gls{NERRS} reserve.
+\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve.
 Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.
@@
 {\footnotesize
-\caption{Trends in metabolism for two sites at each of the \gls{NERRS} reserves.
+\caption{Trends in metabolism for two sites at each of the NERRS reserves.
 Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
@@
 
-The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \gls{SWMP} data.
-The first web application evaluates trends in \gls{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
+The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of SWMP data.
+The first web application evaluates trends in SWMP data within and between sites using an interactive map (Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
@@
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
-The output is identical to \cref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
-The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.
-Both of the apps provide previously unavailable utilities to interact with \gls{SWMP} data using an intuitive online interface.
+The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
+Both of the apps provide previously unavailable utilities to interact with SWMP data using an intuitive online interface.
 This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
@@
 \includegraphics[width = 0\textwidth]{figure/swmp_comp.pdf}
-\caption{Online application for comparing trends in \gls{SWMP} data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
+\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
 \label{fig:swmp_comp}
@@
 
-\section*{Conclusions}
+\section{Conclusions}
 
@@
 Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.
-The \acrlong{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the \acrlong{NERRS}.
+The System-Wide Monitoring Program has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the National Estuarine Research Reserve System.
 This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
-Although the data provided by \gls{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \gls{NERRS} researchers and staff to more effectively evaluate \gls{SWMP} data could be greatly improved using the SWMPr package.
-
-The SWMPr package provides several functions to retrieve, organize, and analyze \gls{SWMP} data to more effectively address common challenges working with large datasets.
-The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (\Cref{tab:retrieve}).
-Established \gls{QAQC} methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different \gls{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
+Although the data provided by SWMP are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of NERRS researchers and staff to more effectively evaluate SWMP data could be greatly improved using the SWMPr package.
+
+The SWMPr package provides several functions to retrieve, organize, and analyze SWMP data to more effectively address common challenges working with large datasets.
+The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (Table~\ref{tab:retrieve}).
+Established QAQC methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different QAQC flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
 Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize and combine time series.
@@
 In particular, the \texttt{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \cite{Odum56,Caffrey14}.
-The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the \gls{NERRS} estuaries using recent data to evaluate trends over time.
+The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the NERRS estuaries using recent data to evaluate trends over time.
 Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
-Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \gls{CDMO}, \gls{SWMP}, and \gls{NERRS}.
-
-\section*{Acknowledgments}
-
-I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.
+Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by CDMO, SWMP, and NERRS.
+
+\section{Acknowledgments}
+
+I acknowledge the significant work of NERRS researchers and staff that has allowed access to high-quality monitoring data.
 Thanks to Todd O'Brien for the inspiration for the online widgets.
@@
 
-\nolinenumbers
-
-\bibliography{swmpr_refs}
-
-\section*{Supporting Information}
-\label{supp_info}
-
-\subsection*{S1 Sample data files}
-\label{supp_zip}
-
-\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \gls{CDMO}.
-These data can be retrieved for use in R with the \texttt{import\_local} function.
-The same data are included with the SWMPr package as binary RData files.
-
-\subsection*{S2 \gls{SWMP} data as R binary files}
-
-\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \gls{SWMP} data from 1995 through 2014, one file per site.
-See the help file for \texttt{import\_remote} for more information.
-
-\subsection*{S3 Metabolism scripts}
-\label{met_scripts}
-
-\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data pre-processing script for all data requested from the CDMO using the zip downloads feature.
-\url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291} Metabolism script for all data from the previous script.
-
-\subsection*{S4 Source code for reproducing figures/tables}
-
-\href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R} Source code as an R script for reproducing figures and tables in the manuscript.
-
-\subsection*{S5 Data to reproduce figures/tables}
-
-\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the RData format.
+\bibliography{beck}
+
+\end{article}
 

### 2015-09-02 beck.tex
working version following R journal wrapper file

--- 2015-09-02 swmpr_manu.tex
@@
-\documentclass[a4paper]{report}\usepackage[]{graphicx}\usepackage{xcolor}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{RJournal}
-\usepackage{amsmath,amssymb,array}
-\usepackage{booktabs}
-
-%% load any required packages
-
-% bold/italic text
-\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-
-%% do not edit, for illustration only
-\sectionhead{Contributed research article}
-\volume{XX}
-\volnumber{YY}
-\year{20ZZ}
-\month{AAAA}
-
-\begin{article}
-
-%knitr options
-
 \title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
 \author{by Marcus W Beck}
-% ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
 
@@
 
-\section*{Abstract}
+\abstract{
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
@@
 Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
+}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > install.packages('SWMPr')
 > library(SWMPr)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # retrieve metadata for all sites
@@
 > site_codes_ind('apa')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # all parameters for a station, most recent
@@
 > single_param('hudscwq', 'do_mgl')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import local data for apaebmet
@@
 > import_local(path, 'apadbwq') 
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import all paramaters for the station
@@
 ## 3         -2    NA      -1     NA     NA       -1
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import sample data from package
@@
 > attr(dat, 'station')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # view available methods for swmpr class
 > methods(class = 'swmpr')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # qaqc screen for a swmpr object, retain only '0'
@@
 > qaqc(dat, qaqc_keep = c(0, -1))
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # view the number of observations in each QAQC flag
 > qaqcchk(dat)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # get nutrient data, filter qaqc flags
@@
 > rem_reps(dat, FUN = func)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import data
@@
 > subset(dat, rem_rows = T, rem_cols = T)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import, qaqc removal
@@
 > subset(dat_nut, rem_rows = T, rem_cols = T)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # get nut, wq, and met data as separate objects
@@
 > comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # get data, keep all observations
@@
 ## [1] 16
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # import data, qaqc and subset
@@
 > lines(dat_smooth, col = 'red', lwd = 2)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 \end{figure}
-
-\end{knitrout}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # get data, qaqc and subset
@@
 > lines(dat)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # get data
@@
 > plot(dc_dat)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 > decomp_cj(dat, param = 'chla_n')
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 
-\end{knitrout}
-
 Detailed exploratory graphics are also useful for evaluating general trends in observed data.
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > ## import data
@@
 > plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > ## import water quality and weather data
@@
 > plot_metab(res)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 
-\end{knitrout}
-
 Finally, the \texttt{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 > # plot the stations at Jacques Cousteau reserve
 > map_reserve('jac')
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+
+\begin{figure}[!h]
 
@@
 % metabolism figure
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!h]
 
+\begin{figure}[!h]
 {\centering \includegraphics[width=0.00\textwidth]{figure/metab_plo-1}
@@
 
-\end{article}
-
-\end{document}
+\address{Marcus W Beck\\
+  ORISE Research Participation Program\\
+  USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division\\
+  1 Sabine Island Drive, Gulf Breeze, FL 32651\\
+  USA\\}
+\email{beck.marcus@epa.gov}

### 2015-09-02 swmpr_manu.Rnw
created plos reject branch for future ref
working version following R journal wrapper file

@@
-% Template for PLoS
-% Version 3.1 February 2015
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
+\documentclass[a4paper]{report}
 \usepackage[utf8]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% fixltx2e package for \textsubscript
-\usepackage{fixltx2e}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% rotating package for sideways tables
-\usepackage{rotating}
-
-% Remove comment for double spacing
-\usepackage{setspace} 
-\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-% for acronyms
-\usepackage{acronym}
-
-% packages I added
-\usepackage{cleveref}
-\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
-\usepackage[usenames,dvipsnames]{xcolor}
-\usepackage[acronym]{glossaries}
-\makeglossaries
-\include{glossary}
-
-%% Include all macros below
-
-% cleveref options
-\crefname{table}{Table}{Tables}
-\crefname{figure}{Fig.}{Figs.}
-\renewcommand{\figurename}{Fig.}
+\usepackage[T1]{fontenc}
+\usepackage{RJournal}
+\usepackage{amsmath,amssymb,array}
+\usepackage{booktabs}
+
+%% load any required packages
 
@@
 
-% acronyms from glossaries package
-\newacronym{CDMO}{CDMO}{Centralized Data Management Office}
-\newacronym{CRAN}{CRAN}{Comprehensive R Archive Network}
-\newacronym{DO}{DO}{dissolved oxygen}
-\newacronym{NERRS}{NERRS}{National Estuarine Research Reserve System}
-\newacronym{QAQC}{QAQC}{quality assurance/quality control}
-\newacronym{SWMP}{SWMP}{System Wide Monitoring Program}
-
-%% END MACROS SECTION
+\begin{document}
+
+%% do not edit, for illustration only
+\sectionhead{Contributed research article}
+\volume{XX}
+\volnumber{YY}
+\year{20ZZ}
+\month{AAAA}
+
+\begin{article}
 
@@
 options(replace.assign=TRUE,width=70,digits=1)
-@
-
-\begin{document}
-\vspace*{0.35in}
-\urlstyle{same}
-
-% Title must be 250 characters or less.
-% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
-\begin{flushleft}
-{\Large
-\textbf\newline{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Marcus William Beck\textsuperscript{1, *}
-
-\bigskip
-\bf{1} ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
-\\
-\bigskip
-
-* beck.marcus@epa.gov
-
-\end{flushleft}
-
-\section*{Abstract}
+knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
+@
+
+\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
+\author{by Marcus W Beck}
+
+\maketitle
+
+\abstract{
 Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \acrlong{NERRS}.
+One example in the United States is the System-Wide Monitoring Program (SWMP) that was implemented in 1995 by the federally-funded National Estuarine Research Reserve System.
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-\gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
+SWMP data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
 However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
-I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.
+I describe SWMPr, an open-source R package, for use with SWMP environmental data.
 The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.
@@
 Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
-\linenumbers
-\glsaddall
-\printglossaries
-
-\section*{Introduction}
+}
+
+\section{Introduction}
 
@@
   'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
-  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\gls{CDMO} using the zip downloads option.',
+  'Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.',
   'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
@@
   'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
-  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters.',
+  'Station identifier used by NERRS as a string with 7 or 8 characters.',
   "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
-  'Indicates if \\gls{QAQC} columns are present in the raw data.',
+  'Indicates if QAQC columns are present in the raw data.',
   'Start and end dates for the data.',
@@
 Only different data types can be combined.',
-  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
-  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each \\gls{QAQC} flag used by \\gls{CDMO}.  The output can be used to inform further processing.', 
+  'Remove QAQC columns and remove data based on QAQC flag values for a \\texttt{swmpr} object.  Only applies if QAQC columns are present. ', 
+  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
@@
   'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\texttt{ecometab} function.',
-  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes \\gls{QAQC} columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
-  'Parses HTML returned from \\gls{CDMO} data requests.  Used in the retrieval functions.',
+  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes QAQC columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
+  'Parses HTML returned from CDMO data requests.  Used in the retrieval functions.',
   'Creates a \\texttt{swmpr} object class.  Used in the data retrieval functions.',
@@
   caption = "Miscellaneous functions available from the SWMPr package.
-Most are used within the main functions above but may be useful for customized evaluations of \\gls{SWMP} data.
+Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
 Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
@@
 % metabolism figure
-<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\gls{NERRS} reserve.
+<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve.
 Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.
@@
 
-cap <- 'Trends in metabolism for two sites at each of the \\gls{NERRS} reserves.
+cap <- 'Trends in metabolism for two sites at each of the NERRS reserves.
 Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
@@
 
-\nolinenumbers
-
-\bibliography{swmpr_refs}
-
-\section*{Supporting Information}
-\label{supp_info}
-
-\subsection*{S1 Sample data files}
-\label{supp_zip}
-
-\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \gls{CDMO}.
-These data can be retrieved for use in R with the \texttt{import\_local} function.
-The same data are included with the SWMPr package as binary RData files.
-
-\subsection*{S2 \gls{SWMP} data as R binary files}
-
-\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \gls{SWMP} data from 1995 through 2014, one file per site.
-See the help file for \texttt{import\_remote} for more information.
-
-\subsection*{S3 Metabolism scripts}
-\label{met_scripts}
-
-\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data pre-processing script for all data requested from the CDMO using the zip downloads feature.
-\url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291} Metabolism script for all data from the previous script.
-
-\subsection*{S4 Source code for reproducing figures/tables}
-
-\href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R} Source code as an R script for reproducing figures and tables in the manuscript.
-
-\subsection*{S5 Data to reproduce figures/tables}
-
-\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the RData format.
+\bibliography{beck}
+
+\address{Marcus W Beck\\
+  ORISE Research Participation Program\\
+  USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division\\
+  1 Sabine Island Drive, Gulf Breeze, FL 32651\\
+  USA\\}
+\email{beck.marcus@epa.gov}
+
+\end{article}
 

### 2015-09-02 swmpr_manu.tex
working version following R journal wrapper file

@@
 \author{by Marcus W Beck}
-% ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
 
@@
 
+\address{Marcus W Beck\\
+  ORISE Research Participation Program\\
+  USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division\\
+  1 Sabine Island Drive, Gulf Breeze, FL 32651\\
+  USA\\}
+\email{beck.marcus@epa.gov}
+
 \end{article}

### 2015-09-02 swmpr_manu.Rnw
most of the formatting is done, ready to butcher text

@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.00\\textwidth')
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.8\\textwidth')
 options(replace.assign=TRUE,width=70,digits=1)
@@
 
-\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
+\title{\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
 \author{by Marcus W Beck}
@@
 \abstract{
-Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.
-One example in the United States is the System-Wide Monitoring Program (SWMP) that was implemented in 1995 by the federally-funded National Estuarine Research Reserve System.
+The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System.
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-SWMP data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.
-However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
-I describe SWMPr, an open-source R package, for use with SWMP environmental data.
-The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.
+Simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
+This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.
 Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.
-Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.
-Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
+The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and provides an effective approach to link quantitative information with analysis tools that will inform coastal protection and restoration.
 }
@@
 
-The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.
+The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \citep{Glasgow04,Fries08}.
 These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
-Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \cite{NDBC15}.
-Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.
+Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \citep{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \citep{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \citep{NDBC15}.
+Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \citep{Glasgow04}.
 However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information.
-A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \cite{Campbell13,Millie13}.
+A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \citep{Campbell13,Millie13}.
 
@@
 The SWMP network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
-These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).
+These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \citet{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \citet{Caffrey03,Caffrey04}, tidal characteristics \citet{Sanger02}, dissolved oxygen \citet{Wenner04}).
 However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current SWMP data.
-NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \cite{SWMP14}.
-
-This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R \cite{RDCT14}.
-SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.
-Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
+NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \citep{SWMP14}.
+
+This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R.
+\CRANpkg{SWMPr} (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.
+Functions provided by \pkg{SWMPr} address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
 Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.
@@
 Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
-As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.
+As such, the \pkg{SWMPr} package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 All reserve names, stations, and date ranges for each parameter type can be viewed on the CDMO website.
-Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.
-As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in SWMPr.
+Alternatively, the \code{site\_codes} (all sites) or \code{site\_codes\_ind} (single site) functions provided by the \pkg{SWMPr} package can be used to view the same information.
+As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in \pkg{SWMPr}.
 Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.
-The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on CDMO web services.
-For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.
-The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.
-
-\section{Structure of the SWMPr package}
+The data retrieval functions in \pkg{SWMPr} are simple calls to the existing retrieval functions on CDMO web services.
+For example, the \code{site\_codes} function in \pkg{SWMPr} uses the \code{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.
+The text below describes the data retrieval functions in more detail, including all other functions available in \pkg{SWMPr}.
+
+\section{Structure of the \pkg{SWMPr} package}
 
@@
 
-The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}.
-The SWMPr package can be downloaded from CRAN by executing the following commands at the R console prompt.
-The package is loaded in the workspace using the \texttt{library} command.
+The \pkg{SWMPr} package was developed for use with the R ($\geq$ v3.0.0) statistical programming language.
+The \pkg{SWMPr} package can be downloaded from CRAN by executing the following commands at the R console prompt.
+The package is loaded in the workspace using the \code{library} command.
 
@@
 
-The SWMPr package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
-Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.
-Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.
+The \pkg{SWMPr} package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
+Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.
+Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.
 An additional group of `miscellaneous' functions are included as helpers for the main functions.
-The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.
+The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.
 
@@
 
-Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
+Two basic approaches using \pkg{SWMPr} are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
 First, functions from the package can be used to import the data directly from the online server using CDMO web services.
 To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.
-The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
+The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
 
@@
 funcs <- c('all\\_params', 'all\\_params\\_dtrng', 'import\\_local', 'single\\_param', 'site\\_codes', 'site\\_codes\\_ind')
-funcs <- paste0('\\texttt{', funcs, '}')
+funcs <- paste0('\\code{', funcs, '}')
 
 descrips <- c(
-  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
-  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
+  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\code{exportAllParamsXMLNew} function on web services.',
+  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\code{exportAllParamsDateRangeXMLNew}.', 
   'Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.',
-  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
-  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
-  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
+  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\code{exportSingleParamXMLNew} function on web services.', 
+  'Metadata for all stations, wrapper to \\code{exportStationCodesXMLNew} function on web services.',
+  'Metadata for all stations at a single site, wrapper  to \\code{NERRFilterStationCodesXMLNew} function on web services.'
   )
@@
   file = '',
-  caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",
+  caption = "Retrieval functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 
-Data retrieval functions to import data directly into R from the CDMO include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.
+Data retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}: \code{all\_params} returns the most recent records of all parameters at a station, \code{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \code{single\_param} is identical to \code{all\_params} except that a single parameter is requested.
 Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.
-However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.
+However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.
 Download time can be excessive for longer time series.
@@
 
-The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from CDMO.
+The second approach for data retrieval is to use the \code{import\_local} function to import data into R that are locally available after downloading from CDMO.
 This approach is most appropriate for large, specific data requests.
-The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website.
+The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website.
 The zip downloads feature can be used to obtain data from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.
+The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.
 
@@
 These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.
-The \texttt{import\_local} function handles duplicate entries differently depending on the data type.
+The \code{import\_local} function handles duplicate entries differently depending on the data type.
 For water quality and nutrient data, duplicate time stamps are simply removed.
 Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
-Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.
+Nutrient data with replicates with unique time stamps are not removed but can be further processed using \code{rem\_reps}.
 Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
@@
 
-\subsection{The \texttt{swmpr} object class}
-
-All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.
-The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
+\subsection{The \code{"swmpr"} object class}
+
+All data retrieval functions return a \code{"swmpr"} object that includes relevant data and several attributes describing the dataset.
+The data include a \code{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
 Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.
-The following shows an example of the raw data imported using \texttt{all\_params}.
+The following shows an example of the raw data imported using \code{all\_params}.
 
@@
 
-The attributes for a \texttt{swmpr} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).
+The attributes for a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).
 These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
@@
 # view all attributes of dat
-atts <- paste0('\\texttt{', names(attributes(dat)), '}')
+atts <- paste0('\\code{', names(attributes(dat)), '}')
 atts <- gsub('_', '\\_', atts, fixed = T)
@@
 atts_desc <- c(
-  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
-  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
-  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
+  'Column names of the entire data set, inherited from the \\code{data.frame} object class.',
+  'Row names of the data set, inherited from the \\code{data.frame} object class.',
+  'Class of the data object indicating \\code{"swmpr"} and \\code{data.frame}.',
   'Station identifier used by NERRS as a string with 7 or 8 characters.',
-  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
+  "Character vector of column names for data parameters, e.g., \\code{`do\\_mgl'}, \\code{`turb'}, etc.",
   'Indicates if QAQC columns are present in the raw data.',
@@
 'Timezone of the station using the city/country format\\textsuperscript{a}.',
-  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
+  'Class of the \\code{datetimestamp} column, usually POSIXct unless data have been aggregated.'
 )
@@
 to_tab <- data.frame(Attributes = atts, Class = atts_class, Description = atts_desc, stringsAsFactors = F)
-foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\texttt{swmpr} objects and may not be cities in the United States.
+foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\code{"swmpr"} objects and may not be cities in the United States.
 For example, ``America/Jamaica" is used for Eastern Standard Time.')
@@
   file = '',
-  caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",
+  caption = 'Attributes of a \\code{"swmpr"} object that describe characteristics of the data.',
   rowlabel = 'Attributes',
@@
 
-The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.
-A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.
-Available methods for the \texttt{swmpr} class are described below and can also be viewed:
+The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.
+This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.
+A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.
+Available methods for the \code{"swmpr"} class are described below and can also be viewed:
 
@@
 This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
-These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.
-These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
-Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).
+These data are also included with the package as binary data files (RData) that can be loaded using the \code{data} function, as shown above.
+These include \code{"swmpr"} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
+Information for each file can be viewed in the help documentation (e.g., \code{?apacpnut}).
 
@@
 
-The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
-Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
+Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).
 Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.
-The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).
+The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
 Erroneous data entered as -99 are also removed.
-The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
+The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.
 
@@
 funcs <- c('comb', 'qaqc', 'qaqcchk', 'rem\\_reps', 'setstep', 'subset')
-funcs <- paste0('\\texttt{', funcs, '}')
+funcs <- paste0('\\code{', funcs, '}')
 
 descrips <- c(
-  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+  'Combines \\code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.',
-  'Remove QAQC columns and remove data based on QAQC flag values for a \\texttt{swmpr} object.  Only applies if QAQC columns are present. ', 
-  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
+  'Remove QAQC columns and remove data based on QAQC flag values for a \\code{"swmpr"} object.  Only applies if QAQC columns are present. ', 
+  'View a summary of the number of observations in a \\code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
-  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
-  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
+  'Format data from a \\code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \\code{comb}.',
+  'Subset by dates and/or columns for a \\code{"swmpr"} object.  This is a method passed to the generic \\code{subset} function provided in the base installation.'
   )
@@
   file = '',
-  caption = "Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",
+  caption = "Organizing functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 
-Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \texttt{qaqc}.
-The \texttt{qaqcchk} function can be used to view this information.
+Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \code{qaqc}.
+The \code{qaqcchk} function can be used to view this information.
 
@@
 Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
-The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
-Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.
-The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).
-The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.
+The \code{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
+The \code{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
+Other suggested functions include the \code{median}, \code{min}, or \code{max}.
+The entire function call, including treatment of \code{NA} values, should be passed to the \code{FUN} argument (see the examples).
+The function is meant to be used after \code{qaqc} processing, although it works with a warning if QAQC columns are present.
 
@@
 
-A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.
+A subset method added to the existing generic \code{subset} function in R is available for \code{"swmpr"} objects.
 This function is used to subset the data by date and/or a selected parameter.
 The date can be a single value or as two dates to select records within the range.
-The former case requires a binary operator as a character string passed to the \texttt{operator} argument, such as \texttt{`>'} or \texttt{`<='}.
+The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{`>'} or \code{`<='}.
 The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').
 Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
-\texttt{`2012-11-31 12:00'}.
+\code{`2012-11-31 12:00'}.
 Finally, the function can be used to remove rows and columns that do not contain data.
@@
 
-The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.
+The \code{setstep} function formats a \code{"swmpr"} object to a continuous time series at a given time step.
 This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.
-The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
-The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.
-Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.
-Likewise, the default value for \texttt{differ} is one half the time step.
-Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.
+The first argument of the function, \code{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
+The second argument, \code{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.
+Values for \code{differ} that are greater than one half the value of \code{timestep} are not allowed to prevent duplication of existing data.
+Likewise, the default value for \code{differ} is one half the time step.
+Time steps that do not match any existing data within the limits of the \code{differ} argument are not discarded, although a corresponding data value will not be assigned.
 
@@
 
-The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.
-The \texttt{setstep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.
-All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input to \texttt{comb}.
-The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.
-The time series for merging depends on the value passed to the \texttt{method} argument.
-Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
-Passing \texttt{`intersect'} to \texttt{method} will create a continuous time series from the set of dates that are shared between all input objects.
-Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
+The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.
+The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.
+All arguments must be called explicitly since an arbitrary number of \code{"swmpr"} objects can be used as input to \code{comb}.
+The function combines data by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.
+The time series for merging depends on the value passed to the \code{method} argument.
+Passing \code{`union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
+Passing \code{`intersect'} to \code{method} will create a continuous time series from the set of dates that are shared between all input objects.
+Finally, a seven or eight character station name passed to \code{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
 Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).
-The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
-These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
-Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
-Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.
-The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
-
-The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.
+The general purpose tools are \code{"swmpr"} methods that were developed for existing generic functions in the R base installation or relevant packages (\pkg{SWMPr} imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
+These functions include \code{"swmpr"} methods for \code{aggreswmp}, \code{filter}, and \code{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \code{plot}, \code{lines}, and \code{hist} methods.
+Decomposition functions, \code{decomp} and \code{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \code{ecometab} and \code{plot\_metab} functions.
+The analysis functions may or may not return a \code{"swmpr"} object depending on whether further processing with \code{"swmpr"} methods is possible from the output.
+
+The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
 Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.
-A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.
-Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
+A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for IDate objects created in the \CRANpkg{data.table} package \citep{Dowle14}.
+Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
 
@@
 funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
-funcs <- paste0('\\texttt{', funcs, '}')
+funcs <- paste0('\\code{', funcs, '}')
 
 descrips <- c(
-  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
-  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
+  'Aggregate \\code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
+  'Aggregate metabolism data from a \\code{"swmpr"} object.  This is primarly used within \\code{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
   'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
-  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \\texttt{decompose} \\cite{Kendall83}.
+  'Decompose a \\code{"swmpr"} time series into trend, seasonal, and residual components.
+This is a simple wrapper to \\code{decompose} \\citep{Kendall83}.
 Decomposition of monthly or daily trends is possible.',
-  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \\texttt{decompTs} in the wq package \\cite{Jassby14}.
+  'Decompose a \\code{"swmpr"} time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \\code{decompTs} in the \\CRANpkg{wq} package \\citep{Jassby14}.
 Only monthly decomposition is possible.',
-  'Plot a histogram for a \\texttt{swmpr} object.',
-  'Add lines to an existing plot created with \\texttt{plot}.',
-  'Create a map of all stations in a reserve using the ggmap package \\cite{Kahle13}.',
-  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
-  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
-  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a combined \\texttt{swmpr} object.',  
+  'Plot a histogram for a \\code{"swmpr"} object.',
+  'Add lines to an existing plot created with \\code{plot}.',
+  'Create a map of all stations in a reserve using the \\CRANpkg{ggmap} package \\citep{Kahle13}.',
+  'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.', 
+  'Plot a univariate  time series for a \\code{"swmpr"} object.  The parameter name must be specified.',
+  'Plot ecosystem metabolism estimates after running \\code{ecometab} on a combined \\code{"swmpr"} object.',  
   'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
-  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\texttt{filter}.'
+  'Smooth \\code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\code{filter}.'
   )
@@
   file = '',
-  caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",
+  caption = "Analysis functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",
   rowlabel = 'Function',
@@
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
-The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
-The \texttt{sides} argument specifies how the average is calculated for each observation.
-Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation.
-As before, the \texttt{params} argument specifies which parameters to smooth.
-
-<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\texttt{smoother} function.">>=
+The \code{smoother} function can be used to smooth parameters in a \code{"swmpr"} object using a specified window size.
+The \code{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
+The \code{sides} argument specifies how the average is calculated for each observation.
+Setting \code{sides = 1} will filter observations within the window that are previous to the current observation, whereas \code{sides = 2} will filter observations within the window centered at zero lag from the current observation.
+As before, the \code{params} argument specifies which parameters to smooth.
+
+<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\code{smoother} function.">>=
 # import data, qaqc and subset
@@
 In either case, an analyst should have a strong rationale for the chosen method.
-A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (Figure~\ref{fig:interp_ex}).
+A common approach implemented in the \pkg{SWMPr} package is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).
 A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
@@
 An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
-The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
-
-<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
+The \code{maxgap} argument passed to \code{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
+
+<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
 # get data, qaqc and subset
@@
 The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
-The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
+The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
 An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
-The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
-For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
+The \code{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
+For example, the \code{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
 The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
-For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+For simplicity, character strings of \code{`daily'} or \code{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
 A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
-For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).
-Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
-
-<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \\texttt{decomp} function.">>=
+For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} function for details).
+Use of the \code{setstep} function is also required to standardize the time step prior to decomposition.
+
+<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \\code{decomp} function.">>=
 # get data
@@
 
-An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.
-The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+An alternative approach to time series decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \CRANpkg{wq} package \citep{Cloern10,Jassby14}.
+The \code{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
 This approach works best for nutrient data that are typically obtained on a monthly cycle.
 The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.
-The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
+The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \code{decomp} function described above.
 For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
-Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
-Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-Figure~\ref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
-
-<<decomp_ex2, eval = T, fig.height = 6, warning = F, cache = T, fig.cap = "Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \\texttt{decomp\\_cj} function.">>=
+Additional arguments passed to \code{decompTs} can be used with \code{decomp\_cj}, such as \code{startyr}, \code{endyr}, and \code{type}.
+Values passed to \code{type} are \code{mult} (default) or \code{add}, referring to multiplicative or additive decomposition.
+Figure~\ref{fig:decomp_ex2} shows the results from the \code{decomp\_cj} function applied to a multi-year chlorophyll time series.
+
+<<decomp_ex2, eval = T, fig.height = 6, warning = F, cache = T, fig.cap = "Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \\code{decomp\\_cj} function.">>=
 # get data
@@
 Detailed exploratory graphics are also useful for evaluating general trends in observed data.
-Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (Figure~\ref{fig:summary_ex}).
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
 Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
-An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
-
-<<summary_ex, fig.height = 7, fig.width = 13, message = F, cache = T, fig.cap = "Summaries of a multi-year chlorophyll time series using the \\texttt{plot\\_summary} function.
+An interactive Shiny web application \citep{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
+
+<<summary_ex, fig.height = 7, fig.width = 13, message = F, cache = T, fig.cap = "Summaries of a multi-year chlorophyll time series using the \\code{plot\\_summary} function.
 Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)">>=
@@
 Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}.
+The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
 The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series.
@@
 
-The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.
+The \code{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \citep{Caffrey14}.
 Several assumptions must be met for a valid interpretation of the results.
@@
 Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
-Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object.
+Water column depth is calculated as the mean value of the depth variable across the time series in the \code{"swmpr"} object.
 Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
@@
 The metadata should be consulted for exact height.
-Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).
-
-The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.
-Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (Figure~\ref{fig:metab_ex}).
+Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \citet{Kemp12,Needoba12}).
+
+The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.
+Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).
 
@@
 
-Finally, the \texttt{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
+Finally, the \code{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the \CRANpkg{ggmap} package \citep{Kahle13}.
 This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
-The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
-
-<<map_ex, fig.height = 5, message = F, cache = T, fig.cap = "Locations of all sites at the Jacques Cousteau reserve using the \\texttt{map\\_reserve} function.">>=
+The current function is limited to Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for the \code{ggmap} function for more info on \code{zoom}.
+
+<<map_ex, fig.height = 5, message = F, cache = T, fig.cap = "Locations of all sites at the Jacques Cousteau reserve using the \\code{map\\_reserve} function.">>=
 # plot the stations at Jacques Cousteau reserve
@@
 funcs <- c('calckl', 'metab\\_day', 'param\\_names', 'parser', 'swmpr', 'time\\_vec')
-funcs <- paste0('\\texttt{', funcs, '}')
+funcs <- paste0('\\code{', funcs, '}')
 
 descrips <- c(
-  'Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \\texttt{ecometab} function.',
-  'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\texttt{ecometab} function.',
-  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes QAQC columns with \\texttt{f\\_} prefix. Used in the data retrieval functions.',
+  'Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \\code{ecometab} function.',
+  'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\code{ecometab} function.',
+  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes QAQC columns with \\code{f\\_} prefix. Used in the data retrieval functions.',
   'Parses HTML returned from CDMO data requests.  Used in the retrieval functions.',
-  'Creates a \\texttt{swmpr} object class.  Used in the data retrieval functions.',
-  'Converts time vectors to \\texttt{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.'
+  'Creates a \\code{"swmpr"} object class.  Used in the data retrieval functions.',
+  'Converts time vectors to \\code{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.'
   )
@@
   file = '',
-  caption = "Miscellaneous functions available from the SWMPr package.
+  caption = "Miscellaneous functions available from the \\pkg{SWMPr} package.
 Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
+Full documentation for each function is in the help file (e.g., execute \\code{?calckl} at the command line).",
   rowlabel = 'Function',
@@
 
-\section{Applications using the SWMPr package}
+\section{Applications using the \pkg{SWMPr} package}
 \label{swmp_apps}
 
-This section describes three examples using the SWMPr package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
+This section describes three examples using the \pkg{SWMPr} package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
 First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each NERRS reserve using all years of available data.
 The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent SWMP data.
-Caffrey \cite{Caffrey03} and Caffrey \cite{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.
-The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \texttt{ecometab} function provides a more accurate estimate by including weather data in the calculation (see Caffrey et al.
-\cite{Caffrey14} for details).
+\citet{Caffrey03} and \citet{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.
+The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \code{ecometab} function provides a more accurate estimate by including weather data in the calculation (see \citet{Caffrey14} for details).
 
 Water quality and weather observations from January 1995 to December 2014 for all NERRS sites were obtained through a bulk data request using the zip downloads feature of CDMO.
-All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files.
-This resulted in a single \texttt{swmpr} object for each parameter at each site.
+All csv files for each station were imported into R using the \code{import\_local} function, processed using the \code{setstep} and \code{qaqc} functions, then saved locally as binary RData files.
+This resulted in a single \code{"swmpr"} object for each parameter at each site.
 All files were then uploaded to a remote server for online access.
@@
 Mean annual metabolism values at each site, organized by region, are shown in Figure~\ref{fig:metab_plo}, whereas decadal comparisons are shown in Table~\ref{tab:metab_tab}.
-All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
+All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with \citet{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
 Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
-More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.
+More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the \pkg{SWMPr} package.
 
@@
 
-The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of SWMP data.
+The second and third examples are two interactive web applications \citep{Chang15} created using the \pkg{SWMPr} package that illustrate summaries and comparisons of SWMP data.
 The first web application evaluates trends in SWMP data within and between sites using an interactive map (Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
@@
 The trend over time is blue for decreasing and red for increasing.
-The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
+The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
 The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
@@
 \begin{center}
-\includegraphics[width = 0\textwidth]{figure/swmp_comp.pdf}
+\includegraphics[width = 0.8\textwidth]{figure/swmp_comp.pdf}
 \caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
@@
 This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
-Although the data provided by SWMP are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of NERRS researchers and staff to more effectively evaluate SWMP data could be greatly improved using the SWMPr package.
-
-The SWMPr package provides several functions to retrieve, organize, and analyze SWMP data to more effectively address common challenges working with large datasets.
+Although the data provided by SWMP are unique among coastal observing systems and have been used in a variety of applications \citep{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of NERRS researchers and staff to more effectively evaluate SWMP data could be greatly improved using the \pkg{SWMPr} package.
+
+The \pkg{SWMPr} package provides several functions to retrieve, organize, and analyze SWMP data to more effectively address common challenges working with large datasets.
 The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (Table~\ref{tab:retrieve}).
-Established QAQC methods and data processing techniques are also enhanced with SWMPr by functions that filter observations for different QAQC flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).
-Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize and combine time series.
+Established QAQC methods and data processing techniques are also enhanced with \pkg{SWMPr} by functions that filter observations for different QAQC flags (\code{qaqc}) and subset by selected dates or variables (\code{subset}).
+Additionally, cumbersome challenges comparing differents datasets are addressed by the \code{setstep} and \code{comb} functions that standardize and combine time series.
 Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.
-In particular, the \texttt{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \cite{Odum56,Caffrey14}.
+In particular, the \code{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \citep{Odum56,Caffrey14}.
 The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the NERRS estuaries using recent data to evaluate trends over time.
 Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
-Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by CDMO, SWMP, and NERRS.
+Further development of the \pkg{SWMPr} package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by CDMO, SWMP, and NERRS.
 

### 2015-09-02 swmpr_manu.tex
most of the formatting is done, ready to butcher text

@@
 
-Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
+Two basic approaches using \pkg{SWMPr} are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
 First, functions from the package can be used to import the data directly from the online server using CDMO web services.
 To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.
-The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
+The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
 
 % table for retrieval functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
-\caption{Retrieval functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} for individual functions or \texttt{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}}
+\caption{Retrieval functions available from the \pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}}
 \begin{center}
@@
 \hline
-\texttt{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
-\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
+\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
+\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
 Optional argument for a single parameter.
-Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
-\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
-\texttt{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
-\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
-\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
+Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
+\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
+\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
+\code{site\_codes}&Metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
+\code{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
 \hline
@@
 % attributes table
-%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
+%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
-\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.\label{tab:attributes}} 
+\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
 \begin{center}
@@
 \hline
-\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class.\tabularnewline
-\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class.\tabularnewline
-\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}.\tabularnewline
-\texttt{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
-\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}, \texttt{`turb'}, etc.\tabularnewline
-\texttt{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
-\texttt{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
-\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
-\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
+\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
+\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
+\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{data.frame}.\tabularnewline
+\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
+\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{`do\_mgl'}, \code{`turb'}, etc.\tabularnewline
+\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
+\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
+\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
+\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
 \hline
@@
 
-\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.
+\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.
 For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}
 
-The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.
-A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.
-Available methods for the \texttt{swmpr} class are described below and can also be viewed:
+The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.
+This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.
+A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.
+Available methods for the \code{"swmpr"} class are described below and can also be viewed:
 
@@
 
-The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
-Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).
+The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
+Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).
 Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.
-The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).
+The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).
 Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12.
 Erroneous data entered as -99 are also removed.
-The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.
+The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.
 
 % table for organize functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
-\caption{Organizing functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \texttt{?comb} for individual functions or \texttt{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}}
+\caption{Organizing functions available from the \pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}}
 \begin{center}
@@
 \hline
-\texttt{comb}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
+\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
 Only different data types can be combined.\tabularnewline
-\texttt{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present. \tabularnewline
-\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.
+\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.  Only applies if QAQC columns are present. \tabularnewline
+\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.
 The output can be used to inform further processing.\tabularnewline
-\texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
-\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \texttt{comb}.\tabularnewline
-\texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
+\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
+\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \code{comb}.\tabularnewline
+\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function provided in the base installation.\tabularnewline
 \hline
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).
-The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
-These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.
-Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
-Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.
-The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.
-
-The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.
+The general purpose tools are \code{"swmpr"} methods that were developed for existing generic functions in the R base installation or relevant packages (\pkg{SWMPr} imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
+These functions include \code{"swmpr"} methods for \code{aggreswmp}, \code{filter}, and \code{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \code{plot}, \code{lines}, and \code{hist} methods.
+Decomposition functions, \code{decomp} and \code{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
+Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \code{ecometab} and \code{plot\_metab} functions.
+The analysis functions may or may not return a \code{"swmpr"} object depending on whether further processing with \code{"swmpr"} methods is possible from the output.
+
+The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.
 This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
 Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.
-A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.
-Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
+A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for IDate objects created in the \CRANpkg{data.table} package \citep{Dowle14}.
+Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
 
 % table for analysis functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
-\caption{Analysis functions available from the SWMPr package.
-Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} for individual functions or \texttt{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}}
+\caption{Analysis functions available from the \pkg{SWMPr} package.
+Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}}
 \begin{center}
@@
 \hline
-\texttt{aggreswmp}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
-\texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.
-This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
-\texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
-\texttt{decomp}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.
+\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
+\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.
+This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
+\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
+\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.
+This is a simple wrapper to \code{decompose} \citep{Kendall83}.
 Decomposition of monthly or daily trends is possible.\tabularnewline
-\texttt{decomp\_cj}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.
+\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.
+This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.
 Only monthly decomposition is possible.\tabularnewline
-\texttt{hist}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
-\texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
-\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package \cite{Kahle13}.\tabularnewline
-\texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
-\texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
-\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a combined \texttt{swmpr} object.\tabularnewline
-\texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
-\texttt{smoother}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
+\code{hist}&Plot a histogram for a \code{"swmpr"} object.\tabularnewline
+\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
+\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
+\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
+\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
+\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
+\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
+\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \code{filter}.\tabularnewline
 \hline
@@
 Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.
-The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
-The \texttt{sides} argument specifies how the average is calculated for each observation.
-Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation.
-As before, the \texttt{params} argument specifies which parameters to smooth.
+The \code{smoother} function can be used to smooth parameters in a \code{"swmpr"} object using a specified window size.
+The \code{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
+The \code{sides} argument specifies how the average is calculated for each observation.
+Setting \code{sides = 1} will filter observations within the window that are previous to the current observation, whereas \code{sides = 2} will filter observations within the window centered at zero lag from the current observation.
+As before, the \code{params} argument specifies which parameters to smooth.
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/smooth_ex-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/smooth_ex-1}
 
@@
 
-\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.}\label{fig:smooth_ex}
+\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function.}\label{fig:smooth_ex}
 \end{figure}
@@
 An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
-The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
+The \code{maxgap} argument passed to \code{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/interp_ex-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/interp_ex-1}
 
@@
 
-\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
+\caption[Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
 \end{figure}
@@
 The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
-The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
+The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
 An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
-The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
-For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
+The \code{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
+For example, the \code{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
 The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
-For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+For simplicity, character strings of \code{`daily'} or \code{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
 A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
-For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).
-Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.
+For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} function for details).
+Use of the \code{setstep} function is also required to standardize the time step prior to decomposition.
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/decomp_ex1-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex1-1}
 
@@
 
-\caption[An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
+\caption[An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \code{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \code{decomp} function.}\label{fig:decomp_ex1}
 \end{figure}
@@
 
-An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.
-The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+An alternative approach to time series decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \CRANpkg{wq} package \citep{Cloern10,Jassby14}.
+The \code{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
 This approach works best for nutrient data that are typically obtained on a monthly cycle.
 The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.
-The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.
+The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \code{decomp} function described above.
 For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
-Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.
-Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.
-Figure~\ref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.
+Additional arguments passed to \code{decompTs} can be used with \code{decomp\_cj}, such as \code{startyr}, \code{endyr}, and \code{type}.
+Values passed to \code{type} are \code{mult} (default) or \code{add}, referring to multiplicative or additive decomposition.
+Figure~\ref{fig:decomp_ex2} shows the results from the \code{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/decomp_ex2-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex2-1}
 
@@
 
-\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \texttt{decomp\_cj} function.}\label{fig:decomp_ex2}
+\caption[Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \code{decomp\_cj} function]{Additive decomposition of a multi-year chlorophyll time series into the grandmean, annual, seasonal, and events components using the \code{decomp\_cj} function.}\label{fig:decomp_ex2}
 \end{figure}
@@
 Detailed exploratory graphics are also useful for evaluating general trends in observed data.
-Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (Figure~\ref{fig:summary_ex}).
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
 Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
-An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
+An interactive Shiny web application \citep{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/summary_ex-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/summary_ex-1}
 
@@
 
-\caption[Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function]{Summaries of a multi-year chlorophyll time series using the \texttt{plot\_summary} function.
+\caption[Summaries of a multi-year chlorophyll time series using the \code{plot\_summary} function]{Summaries of a multi-year chlorophyll time series using the \code{plot\_summary} function.
 Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)}\label{fig:summary_ex}
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/metab_ex-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/metab_ex-1}
 
@@
 
-Finally, the \texttt{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the ggmap package \cite{Kahle13}.
+Finally, the \code{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the \CRANpkg{ggmap} package \citep{Kahle13}.
 This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
-The current function is limited to Google maps of four types that can be set with the \texttt{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \texttt{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \texttt{ggmap} function for more info on \texttt{zoom}.
+The current function is limited to Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
+See the help documentation for the \code{ggmap} function for more info on \code{zoom}.
 
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/map_ex-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/map_ex-1}
 
@@
 
-\caption[Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \texttt{map\_reserve} function.}\label{fig:map_ex}
+\caption[Locations of all sites at the Jacques Cousteau reserve using the \code{map\_reserve} function]{Locations of all sites at the Jacques Cousteau reserve using the \code{map\_reserve} function.}\label{fig:map_ex}
 \end{figure}
@@
 % table for misc functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the SWMPr package.
+%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the \\pkg{SWMPr} package.
 Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
+Full documentation for each function is in the help file (e.g., execute \\code{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
 \begin{table}[!tbp]
-\caption{Miscellaneous functions available from the SWMPr package.
+\caption{Miscellaneous functions available from the \pkg{SWMPr} package.
 Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \texttt{?calckl} at the command line).\label{tab:misc}}
+Full documentation for each function is in the help file (e.g., execute \code{?calckl} at the command line).\label{tab:misc}}
 \begin{center}
@@
 \hline
-\texttt{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \texttt{ecometab} function.\tabularnewline
-\texttt{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \texttt{ecometab} function.\tabularnewline
-\texttt{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
-Includes QAQC columns with \texttt{f\_} prefix.
+\code{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \code{ecometab} function.\tabularnewline
+\code{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \code{ecometab} function.\tabularnewline
+\code{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
+Includes QAQC columns with \code{f\_} prefix.
 Used in the data retrieval functions.\tabularnewline
-\texttt{parser}&Parses HTML returned from CDMO data requests.  Used in the retrieval functions.\tabularnewline
-\texttt{swmpr}&Creates a \texttt{swmpr} object class.  Used in the data retrieval functions.\tabularnewline
-\texttt{time\_vec}&Converts time vectors to \texttt{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.\tabularnewline
+\code{parser}&Parses HTML returned from CDMO data requests.  Used in the retrieval functions.\tabularnewline
+\code{swmpr}&Creates a \code{"swmpr"} object class.  Used in the data retrieval functions.\tabularnewline
+\code{time\_vec}&Converts time vectors to \code{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.\tabularnewline
 \hline
@@
 
-{\centering \includegraphics[width=0.00\textwidth]{figure/metab_plo-1}
+{\centering \includegraphics[width=0.8\textwidth]{figure/metab_plo-1}
 

### 2015-09-02 swmpr_manu.Rnw
text butchered... need to seriously shorten

@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.8\\textwidth')
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '\\textwidth')
 options(replace.assign=TRUE,width=70,digits=1)
@@
 The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \citep{Glasgow04,Fries08}.
-These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.
-Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \citep{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \citep{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \citep{NDBC15}.
-Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \citep{Glasgow04}.
-However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information.
-A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \citep{Campbell13,Millie13}.
+An invaluable source of monitoring data for coastal regions in the United States is provided by the National Estuarine Research Reserve System (NERRS, \url{http://www.nerrs.noaa.gov/}).
+This network of 28 estuary reserves was created to address long-term research, monitoring, education, and stewardship goals in support of coastal management.
+The System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
+The SWMP network provides a continuous source of data collected at near real-time with the potential to evaluate causes of spatiotemporal variation in environmental condition and ecosystem function.
+Environmental researchers have expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \citep{SWMP14}.
 
-An invaluable source of monitoring data for coastal environments in the United States is provided by the National Estuarine Research Reserve System (NERRS, \url{http://www.nerrs.noaa.gov/}).
-This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.
-As part of this effort, the System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
-The SWMP network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.
-These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \citet{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \citet{Caffrey03,Caffrey04}, tidal characteristics \citet{Sanger02}, dissolved oxygen \citet{Wenner04}).
-However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current SWMP data.
-NERRS researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \citep{SWMP14}.
-
-This article describes a software package that was developed to address research needs of the NERRS program using the open-source statistical programming language R.
-\CRANpkg{SWMPr} (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System-Wide monitoring program.
-Functions provided by \pkg{SWMPr} address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
-Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.
-The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large datasets.
+This article describes the \CRANpkg{SWMPr} package that was developed for estuary monitoring data from the System-Wide Monitoring Program.
+Functions provided by \pkg{SWMPr} address many common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
+Additionally, web applications derived from \pkg{SWMPr} and \CRANpkg{shiny} illustrate potential applications using the functions in this package.
+The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large environmental datasets.
 
@@
 
-Four core data elements are collected through the SWMP monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.
-Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.
-As such, the \pkg{SWMPr} package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.
+The \pkg{SWMPr} package is developed for the continuous abiotic monitoring network that represents a majority of SWMP data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
@@
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are made accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.
+All data are accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.
 The final data include all timestamped observations including relevant QAQC flags with the appropriate qualifier.
 
-The CDMO web portal was established to support priority areas of SWMP that focus on the continuation and advancement of data management.
-As such, CDMO provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic QAQC procedures.
-Prior to any data request to the CDMO, the location, parameter type, and date ranges need to be identified based on the analysis needs.
-All stations in the SWMP network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
+At the time of writing, the CDMO web portal provides over 60 million water quality, weather, and nutrient records that have been authenticated through systematic QAQC procedures.
+Records for each station are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.
 For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').
-Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.
+Similarly, a suffix of `met' or `nut' specifies the weather (meteorological) or nutrient stations.
 All reserve names, stations, and date ranges for each parameter type can be viewed on the CDMO website.
-Alternatively, the \code{site\_codes} (all sites) or \code{site\_codes\_ind} (single site) functions provided by the \pkg{SWMPr} package can be used to view the same information.
-As noted below, the computer's IP address must be registered with CDMO before using the data retrieval functions in \pkg{SWMPr}.
+Alternatively, the \code{site\_codes} (all sites) or \code{site\_codes\_ind} (single site) functions provided by \pkg{SWMPr} can be used.
+As noted below, an IP address must be registered with CDMO before using the data retrieval functions in \pkg{SWMPr}.
 Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.
 The data retrieval functions in \pkg{SWMPr} are simple calls to the existing retrieval functions on CDMO web services.
-For example, the \code{site\_codes} function in \pkg{SWMPr} uses the \code{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the SWMP sites.
-The text below describes the data retrieval functions in more detail, including all other functions available in \pkg{SWMPr}.
 
@@
 
-\subsection{Installing the package}
-
-The \pkg{SWMPr} package was developed for use with the R ($\geq$ v3.0.0) statistical programming language.
-The \pkg{SWMPr} package can be downloaded from CRAN by executing the following commands at the R console prompt.
-The package is loaded in the workspace using the \code{library} command.
-
-<<eval = F, message = F>>=
-install.packages('SWMPr')
-library(SWMPr)
-@
 <<eval = T, echo = F, message = F, cache = F>>=
@@
 
-The \pkg{SWMPr} package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
+\pkg{SWMPr} functions are categorized by one of three steps in the data workflow: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
 Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.
 Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.
-An additional group of `miscellaneous' functions are included as helpers for the main functions.
 The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.
@@
 
-Two basic approaches using \pkg{SWMPr} are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}).
-First, functions from the package can be used to import the data directly from the online server using CDMO web services.
-To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.
+\pkg{SWMPr} can import data into R through direct download from the CDMO or by importing local data that was previously downloaded (Table~\ref{tab:retrieve}).
+The IP address for the computer making the request must be registered if the first approach is used (see CDMO \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{website}).
 The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
@@
 
-Data retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}: \code{all\_params} returns the most recent records of all parameters at a station, \code{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \code{single\_param} is identical to \code{all\_params} except that a single parameter is requested.
+Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}.
 Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.
@@
 Download time can be excessive for longer time series.
-
 <<eval = F>>=
@@
 
-The second approach for data retrieval is to use the \code{import\_local} function to import data into R that are locally available after downloading from CDMO.
+The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.
 This approach is most appropriate for large, specific data requests.
@@
 A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for IDate objects created in the \CRANpkg{data.table} package \citep{Dowle14}.
+Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for \code{"IDate"} objects created in the \CRANpkg{data.table} package \citep{Dowle14}.
 Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
@@
 
-funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
+funcs <- c('aggreswmp', 'aggremetab', 'ecometab', 'decomp', 'decomp\\_cj', 'hist', 'lines', 'map\\_reserve', 'na.approx', 'overplot', 'plot', 'plot\\_metab', 'plot\\_summary', 'smoother')
 funcs <- paste0('\\code{', funcs, '}')
@@
   'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.', 
+  'Plot multiple time series on the same y-axis.', 
   'Plot a univariate  time series for a \\code{"swmpr"} object.  The parameter name must be specified.',
@@
 Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
-In either case, an analyst should have a strong rationale for the chosen method.
-A common approach implemented in the \pkg{SWMPr} package is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).
+An approach implemented in \pkg{SWMPr} is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).
 A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
@@
 The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
-For simplicity, character strings of \code{`daily'} or \code{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+For simplicity, character strings of \code{'daily'} or \code{'annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
 A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
@@
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.
 An interactive Shiny web application \citep{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
@@
 
-Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+The \code{overplot} function provides an alternative approach to viewing observed data from the same station.
+This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
+
+<<overplot, fig.height = 4, fig.width = 9, message = F, cache = T, fig.cap = "The \\code{overplot} function plots multiple variables on the same y-axis.">>=
+## import data
+data(apacpwq)
+dat <- qaqc(apacpwq)
+
+## plot
+overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
+  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
+@
+
+Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
 The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
-The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series.
-The nighttime average DO flux is used to estimate respiration rates, while the daytime DO flux is used to estimate net primary production.
-To generate daily integrated rates, respiration rates are assumed constant such that hourly night time DO flux rates are multiplied by 24.
-Similarly, the daytime DO flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
-Respiration rates are subtracted from daily net production estimates to yield gross production rates.
-The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days
-
-The \code{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \citep{Caffrey14}.
-Several assumptions must be met for a valid interpretation of the results.
-First, the DO time series is assumed to be a sample of the same water mass over time.
-Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.
-The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
-Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
-Water column depth is calculated as the mean value of the depth variable across the time series in the \code{"swmpr"} object.
-Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
-Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
-The metadata should be consulted for exact height.
-Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \citet{Kemp12,Needoba12}).
-
+A detailed discussion of the method is beyond the scope of this article, although users are strongly encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
 The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.
@@
 
-\subsection{Miscellaneous functions}
-
-Several additional functions are provided that do not fit the above categories (Table~\ref{tab:misc}).
-These functions are generally used within the main functions but may be useful for more customized evaluation of SWMP data.
-
-% table for misc functions
-<<results = 'asis', echo = FALSE>>=
-
-funcs <- c('calckl', 'metab\\_day', 'param\\_names', 'parser', 'swmpr', 'time\\_vec')
-funcs <- paste0('\\code{', funcs, '}')
-
-descrips <- c(
-  'Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \\code{ecometab} function.',
-  'Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \\code{ecometab} function.',
-  'Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes QAQC columns with \\code{f\\_} prefix. Used in the data retrieval functions.',
-  'Parses HTML returned from CDMO data requests.  Used in the retrieval functions.',
-  'Creates a \\code{"swmpr"} object class.  Used in the data retrieval functions.',
-  'Converts time vectors to \\code{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.'
-  )
-
-to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
-
-library(Hmisc)
-
-latex(
-  to_tab[, 'Description', drop = F],
-  file = '',
-  caption = "Miscellaneous functions available from the \\pkg{SWMPr} package.
-Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \\code{?calckl} at the command line).",
-  rowlabel = 'Function',
-  colheads = 'Description',
-  rowname = to_tab$Functions,
-  caption.loc = 'top',
-  col.just=c("p{3.5in}"), 
-  label = 'tab:misc', 
-  table.env = FALSE, 
-  where = '!tbp'
-  )
-
-@
-
 \section{Applications using the \pkg{SWMPr} package}
@@
 
-This section describes three examples using the \pkg{SWMPr} package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
-First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each NERRS reserve using all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent SWMP data.
-\citet{Caffrey03} and \citet{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.
-The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \code{ecometab} function provides a more accurate estimate by including weather data in the calculation (see \citet{Caffrey14} for details).
-
-Water quality and weather observations from January 1995 to December 2014 for all NERRS sites were obtained through a bulk data request using the zip downloads feature of CDMO.
-All csv files for each station were imported into R using the \code{import\_local} function, processed using the \code{setstep} and \code{qaqc} functions, then saved locally as binary RData files.
-This resulted in a single \code{"swmpr"} object for each parameter at each site.
-All files were then uploaded to a remote server for online access.
-An additional R script was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
-Two water quality sites with the longest time series at each reserve were used.
-Mean annual metabolism values at each site, organized by region, are shown in Figure~\ref{fig:metab_plo}, whereas decadal comparisons are shown in Table~\ref{tab:metab_tab}.
-All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with \citet{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
-Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
-More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the \pkg{SWMPr} package.
-
-% metabolism figure
-<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve.
-Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.
-Two sites were chosen from each reserve that had the longest available time series.
-Sites were assigned to regions based on approximate geographic coordinates.', fig.height = 6, fig.width = 8>>=
-# # packages to use
-# library(SWMPr)
-# library(httr)
-# library(XML)
-# library(foreach)
-# library(doParallel)
-# 
-# # names of files on server
-# files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
-# files_s3 <- rawToChar(files_s3)
-# files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
-# files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
-# files_s3 <- gsub('\\.RData$', '', files_s3) 
-# 
-# # find only active water quality and weather sites
-# # IP address must be registed with CDMO 
-# meta <- site_codes()
-# sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
-# meta <- meta[sel, ]
-# 
-# # get wq sites from meta, then filter by those on server
-# wq_sites <- grep('wq$', meta$station_code, value = T)
-# wq_sites <- files_s3[gsub('\\.RData$', '', files_s3) %in% wq_sites]
-# 
-# # setup parallel backend for processing
-# cl <- makeCluster(8)
-# registerDoParallel(cl)
-# strt <- Sys.time()
-# 
-# # process
-# metabs <- foreach(wq_site = wq_sites) %dopar% {
-#  
-#   library(SWMPr)
-#   
-#   # progress
-#   sink('log.txt')
-#   cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
-#   print(Sys.time() - strt)
-#   sink()
-#   
-#   # find corresponding wx station
-#   met_site <- substr(wq_site, 1, 3)
-#   met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
-#   met_site <- gsub('\\.RData$', '', met_site)
-# 
-#   # continue if wx data found
-#   if(length(met_site) > 0){
-#     
-#     # if > 1 wx site, pick the one with more obs
-#     if(length(met_site) > 1){
-# 
-#       met_list <- vector('list', length(met_site))
-#       names(met_list) <- met_site
-#       for(met in met_site){
-#       
-#         met_tmp <- import_remote(met)
-#         met_list[[met]] <- met_tmp
-#       
-#       }
-#     
-#       # pick the one with most obs
-#       met_most <- which.max(unlist(lapply(met_list, nrow)))
-#       met <- met_list[met_most][[1]]
-# 
-#     # otherwise load only one
-#     } else {
-#     
-#       # met
-#       met <- import_remote(met_site)
-#     
-#     }
-#     
-#     ##
-#     # load the wq file
-#     
-#     # wq
-#     wq <- import_remote(wq_site)
-#     
-#     ## 
-#     # combine, estimate metabolism, reduce data volume
-#     dat <- comb(wq, met, method = attr(met, 'station'))
-#     dat <- ecometab(dat)
-#     dat <- dat[1, ]
-#     
-#     dat
-#     
-#   }
-#   
-# }
-# names(metabs) <- wq_sites
-# save(metabs, file = 'metabs.RData')
-
-# # take all metab estimates and summarize by year
-# # done for two stations at each reserve with the longest time series
-# data(metabs)
-# 
-# metabs <- lapply(metabs, attr, 'metabolism')
-# metabs <- do.call('rbind', metabs)
-# metabs$site <- gsub('\\.[0-9]*$', '', as.character(row.names(metabs)))
-# row.names(metabs) <- 1:nrow(metabs)
-# 
-# library(dplyr)
-# library(tidyr)
-# metabs <- select(metabs, date, Pg, Rt, NEM, site) %>% 
-#   mutate(date = as.numeric(strftime(date, '%Y'))) %>% 
-#   gather('metab', 'value', 2:4) %>% 
-#   group_by(date, site, metab) %>% 
-#   summarize(value = mean(value, na.rm = T)) %>% 
-#   group_by(site, metab) %>% 
-#   na.omit %>% 
-#   mutate(reserve = substr(site, 1, 3)) %>% 
-#   spread(metab, value)
-# 
-# # select two sites from each reserve, those with longest records
-# metabs <- split(metabs, metabs$reserve)
-# metabs <- lapply(metabs, 
-#   function(x){
-#     
-#     num_yrs <- aggregate(date ~ site, x, length)
-#     num_yrs <- num_yrs[order(num_yrs$date, decreasing = T), ]
-#     
-#     if(nrow(num_yrs) >1){
-#       site_sel <- num_yrs$site[c(1, 2)]
-#     } else {
-#       site_sel <- num_yrs$site[1]
-#     }
-#     
-#     out <- x[x$site %in% site_sel, ]
-#     return(out)
-#     
-#   }
-#   
-# )
-# metabs <- do.call('rbind', metabs)
-# 
-# metabs_sum <- metabs
-# save(metabs_sum, file = 'data/metabs_sum.RData')
-
-data(metabs_sum)
-
-library(dplyr)
-library(tidyr)
-
-# summarize for annual means, get conf ints
-metabs_plo <- gather(metabs_sum, 'metab', 'value', 4:6) %>% 
-  group_by(site, metab) %>% 
-  summarize(
-    margs = qt(1 - 0.05/2, length(value) - 1) * sd(value)/sqrt(length(value)),
-    value = mean(value)
-    ) %>% 
-  mutate(
-    value = 0.032 * value,
-    margs = 0.032 * margs,
-    upper = value + margs, 
-    lower = value - margs
-    )
-
-# get region
-data(stat_locs)
-names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
-metabs_plo$site <- gsub('wq$', '', metabs_plo$site)
-metabs_plo <- left_join(metabs_plo, stat_locs, by = 'site')
-region <- rep('Pacific', nrow(metabs_plo))
-region[metabs_plo$longitude > -110] <- 'Gulf'
-region[metabs_plo$latitude > 28 & metabs_plo$longitude > -84] <- 'Atlantic'
-
-# combine, select, sort
-metabs_plo <- select(metabs_plo, -c(latitude, longitude, gmt_off))
-metabs_plo$region <- region
-metabs_plo <- arrange(metabs_plo, region, site)
-
-# plot prep
-to_plo <- data.frame(metabs_plo)
-
-#reassign factors for ranking in plot
-sort_val <- to_plo[to_plo$metab == 'Pg', c('region', 'value', 'site')]
-sort_val <- order(sort_val$region, sort_val$value)
-to_plo$site <- factor(to_plo$site)
-to_plo$site <- factor(
-  to_plo$site,
-  levels = levels(to_plo$site)[sort_val],
-  labels = levels(to_plo$site)[sort_val]
-  )
-new_labs <- c('Mean net metabolism', 'Mean gross production', 'Mean total respiration')
-to_plo$metab <- factor(to_plo$metab, levels = c('NEM', 'Pg', 'Rt'), labels = new_labs)
-
-ylabs <- expression(paste('g ', O[2],' ', m^-2,' ', d^-1))
-library(ggplot2)
-p <- ggplot(to_plo, 
-    aes(x = site, y = value, group = metab, fill = metab)
-  ) + 
-  geom_bar(stat = 'identity', width = 1, colour = 'black') + 
-  scale_y_continuous(ylabs) + 
-  facet_grid(metab ~ region, scales = 'free') + 
-  scale_x_discrete(name = element_blank()) + 
-  theme_bw() + 
-  theme(
-    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, 
-      size = 7),
-    legend.position = 'none'
-    ) +
-  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0)
-print(p)
-
-@
-
-% table summary of trends
-<<eval = T, results = 'asis', cache = T, echo = F>>=
-
-data(metabs_sum)
-
-library(dplyr)
-library(tidyr)
-
-# summarize by time periods
-metabs_tab <- gather(metabs_sum, 'metab', 'value', 4:6)
-metabs_tab$yr_grp <- '1995-2004'
-metabs_tab$yr_grp[metabs_tab$date > 2004] <- '2005-2014'
-metabs_tab <- mutate(metabs_tab, value = 0.032 * value) %>% 
-  group_by(site, metab, yr_grp) %>% 
-  arrange(yr_grp, date, metab) %>% 
-  summarize(ave = round(mean(value, na.rm = TRUE), 2)) %>% 
-  spread(yr_grp, ave) 
-trends <- apply(metabs_tab, 1, function(x){
-
-  val <- as.numeric(x[['2005-2014']])
-  trnd <- val > as.numeric(x[['1995-2004']])
-
-  if(is.na(trnd)) return(val)
-  if(trnd) return(paste0('{\\bf ', val, '}'))
-  if(!trnd) return(paste0('{\\it ', val, '}'))
-
-})
-metabs_tab[, '2005-2014'] <- trends
-metabs_tab <- gather(metabs_tab, 'period', 'value', 3:4) %>% 
-  unite('var', metab, period, sep = ' ') %>% 
-  spread(var, value) %>% 
-  mutate(site = gsub('wq$', '', site))
-
-# get region
-data(stat_locs)
-names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
-metabs_tab <- left_join(metabs_tab, stat_locs, by = 'site')
-region <- rep('Pacific', nrow(metabs_tab))
-region[metabs_tab$longitude > -110] <- 'Gulf'
-region[metabs_tab$latitude > 28 & metabs_tab$longitude > -84] <- 'Atlantic'
-
-# combine, select, sort
-metabs_tab <- select(metabs_tab, -c(latitude, longitude, gmt_off)) %>% 
-  mutate(region = region) %>% 
-  arrange(region, site)
-
-nrgroups <- as.numeric(table(metabs_tab$region))
-# table
-
-cap <- 'Trends in metabolism for two sites at each of the NERRS reserves.
-Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
-Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
-Sites were assigned to regions based on approximate geographic coordinates.'
-foots <- '\\footnotesize{\\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
-
-library(Hmisc)
-latex(
-  metabs_tab[, !names(metabs_tab) %in% c('site', 'region')], 
-  file = '', 
-  caption = cap, 
-  cgroup = c('NEM\\textsuperscript{a}', 'Pg', 'Rt'), 
-  n.cgroup = c(2, 2, 2), 
-  colheads = rep(c('1995-2004', '2005-2014'), 3),
-  rowlabel = 'Site', 
-  rowname = metabs_tab$site,
-  digits = 2,
-  insert.bottom = foots,
-  rgroup = unique(metabs_tab$region),
-  n.rgroup = nrgroups,
-  label = 'tab:metab_tab',
-  size = 'footnotesize', 
-  where = '!tbp'
-  )
-
-@
-
-The second and third examples are two interactive web applications \citep{Chang15} created using the \pkg{SWMPr} package that illustrate summaries and comparisons of SWMP data.
-The first web application evaluates trends in SWMP data within and between sites using an interactive map (Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
+Two \CRANpkg{shiny} web applications hav been created that illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
+The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
-Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.
-Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.
-The trend over time is blue for decreasing and red for increasing.
+Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
 The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
-Both of the apps provide previously unavailable utilities to interact with SWMP data using an intuitive online interface.
-This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
-The applications have been used extensively with over 300 hours noted in a single month.
 
@@
 
-The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.
-Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.
-The System-Wide Monitoring Program has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the National Estuarine Research Reserve System.
-This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.
-Although the data provided by SWMP are unique among coastal observing systems and have been used in a variety of applications \citep{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of NERRS researchers and staff to more effectively evaluate SWMP data could be greatly improved using the \pkg{SWMPr} package.
-
-The \pkg{SWMPr} package provides several functions to retrieve, organize, and analyze SWMP data to more effectively address common challenges working with large datasets.
-The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (Table~\ref{tab:retrieve}).
+\pkg{SWMPr} was developed to augment existing data management programs (i.e., CDMO) by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (Table~\ref{tab:retrieve}).
 Established QAQC methods and data processing techniques are also enhanced with \pkg{SWMPr} by functions that filter observations for different QAQC flags (\code{qaqc}) and subset by selected dates or variables (\code{subset}).
-Additionally, cumbersome challenges comparing differents datasets are addressed by the \code{setstep} and \code{comb} functions that standardize and combine time series.
+Additionally, challenges comparing differents datasets are addressed by the \code{setstep} and \code{comb} functions that standardize and combine time series.
 Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.
-In particular, the \code{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \citep{Odum56,Caffrey14}.
-The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the NERRS estuaries using recent data to evaluate trends over time.
-Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism.
-Further development of the \pkg{SWMPr} package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by CDMO, SWMP, and NERRS.
+Further development of the package will include modifications and additional functions to better integrate data analysis with the quality of information provided by SWMP.
+For example, several functions include default methods to extend use beyond the \code{"swmpr"} object and additional development will continue to focus on modifying the package to handle arbitrary data structures.
 
@@
 Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.
-The views expressed in this article are those of the author and do not necessarily reflect the views or policies of the U.S.
-Environmental Protection Agency.
-The use of trade names or products does not constitute endorsement by the US Government.
 

### 2015-09-02 swmpr_manu.tex
text butchered... need to seriously shorten

@@
 
-\subsection{Installing the package}
-
-The \pkg{SWMPr} package was developed for use with the R ($\geq$ v3.0.0) statistical programming language.
-The \pkg{SWMPr} package can be downloaded from CRAN by executing the following commands at the R console prompt.
-The package is loaded in the workspace using the \code{library} command.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> install.packages('SWMPr')
-> library(SWMPr)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-The \pkg{SWMPr} package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
+\pkg{SWMPr} functions are categorized by one of three steps in the data workflow: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
 Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.
 Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.
-An additional group of `miscellaneous' functions are included as helpers for the main functions.
 The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.
@@
 Download time can be excessive for longer time series.
-
 \begin{knitrout}\small
@@
 \code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
+\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
 \code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/smooth_ex-1}
+{\centering \includegraphics[width=\textwidth]{figure/smooth_ex-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/interp_ex-1}
+{\centering \includegraphics[width=\textwidth]{figure/interp_ex-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex1-1}
+{\centering \includegraphics[width=\textwidth]{figure/decomp_ex1-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex2-1}
+{\centering \includegraphics[width=\textwidth]{figure/decomp_ex2-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/summary_ex-1}
+{\centering \includegraphics[width=\textwidth]{figure/summary_ex-1}
 
@@
 
-Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+The \code{overplot} function provides an alternative approach to viewing observed data from the same station.
+This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+> ## import data
+> data(apacpwq)
+> dat <- qaqc(apacpwq)
+> 
+> ## plot
+> overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
++   subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
+\end{verbatim}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=\textwidth]{figure/overplot-1}
+
+}
+
+\caption[The \code{overplot} function plots multiple variables on the same y-axis]{The \code{overplot} function plots multiple variables on the same y-axis.}\label{fig:overplot}
+\end{figure}
+
+\end{knitrout}
+
+Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
 The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
-The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series.
-The nighttime average DO flux is used to estimate respiration rates, while the daytime DO flux is used to estimate net primary production.
-To generate daily integrated rates, respiration rates are assumed constant such that hourly night time DO flux rates are multiplied by 24.
-Similarly, the daytime DO flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production.
-Respiration rates are subtracted from daily net production estimates to yield gross production rates.
-The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days
-
-The \code{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \citep{Caffrey14}.
-Several assumptions must be met for a valid interpretation of the results.
-First, the DO time series is assumed to be a sample of the same water mass over time.
-Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.
-The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.
-Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).
-Water column depth is calculated as the mean value of the depth variable across the time series in the \code{"swmpr"} object.
-Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.
-Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.
-The metadata should be consulted for exact height.
-Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \citet{Kemp12,Needoba12}).
-
+A detailed discussion of the method is beyond the scope of this article, although users are strongly encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
 The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/metab_ex-1}
+{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}
 
@@
 
-{\centering \includegraphics[width=0.8\textwidth]{figure/map_ex-1}
+{\centering \includegraphics[width=\textwidth]{figure/map_ex-1}
 
@@
 
-\subsection{Miscellaneous functions}
-
-Several additional functions are provided that do not fit the above categories (Table~\ref{tab:misc}).
-These functions are generally used within the main functions but may be useful for more customized evaluation of SWMP data.
-
-% table for misc functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the \\pkg{SWMPr} package.
-Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \\code{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
-\begin{table}[!tbp]
-\caption{Miscellaneous functions available from the \pkg{SWMPr} package.
-Most are used within the main functions above but may be useful for customized evaluations of SWMP data.
-Full documentation for each function is in the help file (e.g., execute \code{?calckl} at the command line).\label{tab:misc}}
-\begin{center}
-\begin{tabular}{lp{3.5in}}
-\hline\hline
-\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
-\code{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \code{ecometab} function.\tabularnewline
-\code{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \code{ecometab} function.\tabularnewline
-\code{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).
-Includes QAQC columns with \code{f\_} prefix.
-Used in the data retrieval functions.\tabularnewline
-\code{parser}&Parses HTML returned from CDMO data requests.  Used in the retrieval functions.\tabularnewline
-\code{swmpr}&Creates a \code{"swmpr"} object class.  Used in the data retrieval functions.\tabularnewline
-\code{time\_vec}&Converts time vectors to \code{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.\tabularnewline
-\hline
-\end{tabular}\end{center}
-
-\end{table}
-
 \section{Applications using the \pkg{SWMPr} package}
@@
 
-This section describes three examples using the \pkg{SWMPr} package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data.
-First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each NERRS reserve using all years of available data.
-The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent SWMP data.
-\citet{Caffrey03} and \citet{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.
-The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \code{ecometab} function provides a more accurate estimate by including weather data in the calculation (see \citet{Caffrey14} for details).
-
-Water quality and weather observations from January 1995 to December 2014 for all NERRS sites were obtained through a bulk data request using the zip downloads feature of CDMO.
-All csv files for each station were imported into R using the \code{import\_local} function, processed using the \code{setstep} and \code{qaqc} functions, then saved locally as binary RData files.
-This resulted in a single \code{"swmpr"} object for each parameter at each site.
-All files were then uploaded to a remote server for online access.
-An additional R script was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.
-Two water quality sites with the longest time series at each reserve were used.
-Mean annual metabolism values at each site, organized by region, are shown in Figure~\ref{fig:metab_plo}, whereas decadal comparisons are shown in Table~\ref{tab:metab_tab}.
-All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with \citet{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods.
-Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.
-More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the \pkg{SWMPr} package.
-
-% metabolism figure
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!h]
-
-{\centering \includegraphics[width=0.8\textwidth]{figure/metab_plo-1}
-
-}
-
-\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve.
-Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.
-Two sites were chosen from each reserve that had the longest available time series.
-Sites were assigned to regions based on approximate geographic coordinates.}\label{fig:metab_plo}
-\end{figure}
-
-\end{knitrout}
-
-% table summary of trends
-%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{a}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%
-\begin{table}[!tbp]
-{\footnotesize
-\caption{Trends in metabolism for two sites at each of the NERRS reserves.
-Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).
-Bold values indicate an increase from the first period, whereas italic values indicate a decrease.
-Sites were assigned to regions based on approximate geographic coordinates.\label{tab:metab_tab}}
-\begin{center}
-\begin{tabular}{lllcllcll}
-\hline\hline
-\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline
-\cline{2-3} \cline{5-6} \cline{8-9}
-\multicolumn{1}{l}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}\tabularnewline
-\hline
-{\bfseries Atlantic}&&&&&&&&\tabularnewline
-~~acebb&-0.04&{\bf -0.03}&&5.11&{\it 3.81}&&-5.15&{\bf -3.82}\tabularnewline
-~~acesp&-0.26&{\bf -0.1}&&4.68&{\bf 4.93}&&-4.94&{\it -5.03}\tabularnewline
-~~cbmip&-0.14&{\it -0.17}&&1.02&{\bf 1.91}&&-1.16&{\it -2.08}\tabularnewline
-~~cbmmc&-0.43&{\bf -0.24}&&6.06&{\bf 10.13}&&-6.48&{\it -10.38}\tabularnewline
-~~cbvgi&0.13&{\it 0}&&6.56&{\it 4.47}&&-6.43&{\bf -4.46}\tabularnewline
-~~cbvtc&-0.15&{\bf -0.07}&&5.58&{\bf 6.07}&&-5.73&{\it -6.14}\tabularnewline
-~~delbl&-0.59&{\bf -0.16}&&6.68&{\it 6.51}&&-7.27&{\bf -6.67}\tabularnewline
-~~delsl&-0.37&{\bf -0.33}&&2.84&{\bf 3.22}&&-3.21&{\it -3.55}\tabularnewline
-~~grblr&&-0.02&&&1.28&&&-1.31\tabularnewline
-~~grbor&&-0.04&&&3.16&&&-3.2\tabularnewline
-~~gtmfm&-0.09&{\bf -0.04}&&3.36&{\bf 4.19}&&-3.45&{\it -4.23}\tabularnewline
-~~gtmpc&-0.48&{\bf -0.45}&&1.95&{\bf 2.91}&&-2.43&{\it -3.37}\tabularnewline
-~~hudsk&0.02&{\it 0.02}&&-0.36&{\it -0.39}&&0.38&{\bf 0.41}\tabularnewline
-~~hudtn&-0.03&{\it -0.03}&&3.75&{\it 3.43}&&-3.78&{\bf -3.46}\tabularnewline
-~~jacb6&0.05&{\it 0.03}&&3.41&{\bf 3.93}&&-3.36&{\it -3.89}\tabularnewline
-~~jacb9&0.01&{\it -0.02}&&1.8&{\bf 3.1}&&-1.79&{\it -3.12}\tabularnewline
-~~narnc&-0.64&{\bf -0.53}&&12.64&{\it 11.33}&&-13.25&{\bf -11.86}\tabularnewline
-~~narpc&-0.14&{\bf -0.03}&&7.36&{\it 6.3}&&-7.5&{\bf -6.32}\tabularnewline
-~~niwcb&-0.46&{\bf -0.28}&&5.5&{\bf 6.1}&&-5.95&{\it -6.38}\tabularnewline
-~~niwdc&-0.13&{\it -0.2}&&7.28&{\bf 8.02}&&-7.41&{\it -8.23}\tabularnewline
-~~nocec&-0.05&{\it -0.06}&&4.86&{\bf 4.88}&&-4.92&{\it -4.94}\tabularnewline
-~~nocrc&-0.19&{\it -0.19}&&5.93&{\it 5.31}&&-6.12&{\bf -5.5}\tabularnewline
-~~owcbr&-0.17&{\bf -0.09}&&1.45&{\bf 1.9}&&-1.62&{\it -1.99}\tabularnewline
-~~owcwm&-0.36&{\bf -0.19}&&7.02&{\it 5.36}&&-7.38&{\bf -5.54}\tabularnewline
-~~saphd&-1.28&{\bf -1}&&1.89&{\bf 3.77}&&-3.17&{\it -4.77}\tabularnewline
-~~sapld&-0.16&{\it -0.43}&&2.32&{\bf 3.56}&&-2.49&{\it -3.99}\tabularnewline
-~~welht&-0.07&{\bf -0.04}&&0.36&{\bf 0.63}&&-0.43&{\it -0.67}\tabularnewline
-~~welin&-1.87&{\it -2.49}&&3.61&{\it 2.68}&&-5.48&{\bf -5.18}\tabularnewline
-~~wqbcr&-0.01&{\bf 0.01}&&8.13&{\it 7.41}&&-8.14&{\bf -7.4}\tabularnewline
-~~wqbmh&0.02&{\it 0}&&2.82&{\it 2.76}&&-2.79&{\bf -2.75}\tabularnewline
-\hline
-{\bfseries Gulf}&&&&&&&&\tabularnewline
-~~apaeb&-0.35&{\it -0.35}&&4.19&{\bf 4.31}&&-4.54&{\it -4.67}\tabularnewline
-~~apaes&-0.71&{\bf -0.67}&&3.35&{\it 3.05}&&-4.06&{\bf -3.73}\tabularnewline
-~~gndbc&-1.02&{\bf -0.88}&&3.5&{\bf 4.3}&&-4.51&{\it -5.18}\tabularnewline
-~~gndbh&-1.81&{\it -2}&&2.19&{\it 1.77}&&-4&{\bf -3.77}\tabularnewline
-~~job09&-0.34&{\bf -0.25}&&5.32&{\bf 5.95}&&-5.66&{\it -6.2}\tabularnewline
-~~job10&-0.15&{\it -0.15}&&3.58&{\bf 4.03}&&-3.72&{\it -4.2}\tabularnewline
-~~marab&&-0.02&&&3.38&&&-3.4\tabularnewline
-~~marce&&-0.01&&&3.62&&&-3.63\tabularnewline
-~~rkbfb&-0.28&{\it -0.37}&&5.28&{\bf 6.25}&&-5.56&{\it -6.62}\tabularnewline
-~~rkbfu&-0.33&{\it -0.41}&&2.95&{\bf 3.24}&&-3.28&{\it -3.64}\tabularnewline
-~~wkbfr&-0.29&{\bf -0.14}&&6.27&{\bf 6.35}&&-6.57&{\bf -6.49}\tabularnewline
-~~wkbwb&-0.29&{\bf -0.16}&&6.81&{\it 6.54}&&-7.1&{\bf -6.69}\tabularnewline
-\hline
-{\bfseries Pacific}&&&&&&&&\tabularnewline
-~~elkap&0.03&{\bf 0.11}&&14.7&{\bf 19.84}&&-14.67&{\it -19.74}\tabularnewline
-~~elknm&-0.24&{\bf -0.08}&&11.95&{\bf 16.37}&&-12.18&{\it -16.45}\tabularnewline
-~~kachd&0.2&{\it -0.66}&&7.31&{\it 3.2}&&-7.12&{\bf -3.86}\tabularnewline
-~~kacsd&-0.06&{\it -0.26}&&5.15&{\it 4.2}&&-5.21&{\bf -4.45}\tabularnewline
-~~pdbbp&0.01&{\bf 0.08}&&6.29&{\bf 7.31}&&-6.27&{\it -7.23}\tabularnewline
-~~pdbby&-0.07&{\bf -0.01}&&8.39&{\it 5.92}&&-8.47&{\bf -5.93}\tabularnewline
-~~sfbcc&&-0.24&&&3.56&&&-3.79\tabularnewline
-~~sfbgc&&-0.55&&&7.32&&&-7.87\tabularnewline
-~~sosva&-0.2&{\bf -0.05}&&6.98&{\bf 7.83}&&-7.18&{\it -7.87}\tabularnewline
-~~soswi&-0.28&{\bf -0.08}&&4.41&{\bf 4.74}&&-4.69&{\it -4.82}\tabularnewline
-~~tjrbr&-1.33&{\bf -0.17}&&7.72&{\it 7.06}&&-9.05&{\bf -7.22}\tabularnewline
-~~tjros&-0.26&{\it -0.34}&&10.18&{\bf 11.61}&&-10.44&{\it -11.96}\tabularnewline
-\hline
-\end{tabular}\end{center}}
-
-\footnotesize{\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}
-
-The second and third examples are two interactive web applications \citep{Chang15} created using the \pkg{SWMPr} package that illustrate summaries and comparisons of SWMP data.
-The first web application evaluates trends in SWMP data within and between sites using an interactive map (Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
+Two \CRANpkg{shiny} web applications hav been created that illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
+The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
-Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.
-Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.
-The trend over time is blue for decreasing and red for increasing.
+Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
 The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
-Both of the apps provide previously unavailable utilities to interact with SWMP data using an intuitive online interface.
-This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package.
-The applications have been used extensively with over 300 hours noted in a single month.
 

### 2015-09-03 swmpr_manu.Rnw
done butchering/shortening, ready to journalify

@@
 # set global chunk options
-opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '\\textwidth')
+opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = FALSE, size = 'small', out.width = '\\textwidth')
 options(replace.assign=TRUE,width=70,digits=1)
@@
 Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.
-The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and provides an effective approach to link quantitative information with analysis tools that will inform coastal protection and restoration.
+The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
 }
@@
 The System-Wide Monitoring Program (SWMP) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.
-The SWMP network provides a continuous source of data collected at near real-time with the potential to evaluate causes of spatiotemporal variation in environmental condition and ecosystem function.
 Environmental researchers have expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by SWMP \citep{SWMP14}.
@@
 This article describes the \CRANpkg{SWMPr} package that was developed for estuary monitoring data from the System-Wide Monitoring Program.
-Functions provided by \pkg{SWMPr} address many common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.
-Additionally, web applications derived from \pkg{SWMPr} and \CRANpkg{shiny} illustrate potential applications using the functions in this package.
+Functions provided by \CRANpkg{SWMPr} address many common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify parameters of interest.
+Additionally, web applications derived from \CRANpkg{SWMPr} and \CRANpkg{shiny} illustrate potential applications using the functions in this package.
 The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large environmental datasets.
 
-\section{SWMP overview and data retrieval}
+\section{Overview of the SWMP network}
 
-The \pkg{SWMPr} package is developed for the continuous abiotic monitoring network that represents a majority of SWMP data and, consequently, the most challenging to evaluate.
+The \pkg{SWMPr} package was developed for the continuous abiotic monitoring network that represents a majority of SWMP data and, consequently, the most challenging to evaluate.
 Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).
-Each reserve has no less than four water quality stations and one weather station at fixed locations.
+Each of the 28 estuary reserves has no less than four water quality stations and one weather station at fixed locations.
 Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.
-All data are accessible through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.
-The final data include all timestamped observations including relevant QAQC flags with the appropriate qualifier.
+Data are made available through the Centralized Data Management Office (CDMO) web portal (\url{http://cdmo.baruch.sc.edu/}), where quality assurance/quality control (QAQC) measures are used to screen the information for accuracy and reliability.
+The final data include timestamped observations with relevant QAQC flags.
 
@@
 Web services are provided by CDMO for direct access to SWMP data through http requests, in addition to standard graphical user interface options for selecting data.
-The data retrieval functions in \pkg{SWMPr} are simple calls to the existing retrieval functions on CDMO web services.
+The data retrieval functions in \pkg{SWMPr} are simple calls to the existing retrieval functions on CDMO web services, as explained below.
 
@@
 The IP address for the computer making the request must be registered if the first approach is used (see CDMO \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{website}).
-The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.
+The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view site metadata.
 
@@
   'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\code{exportSingleParamXMLNew} function on web services.', 
-  'Metadata for all stations, wrapper to \\code{exportStationCodesXMLNew} function on web services.',
-  'Metadata for all stations at a single site, wrapper  to \\code{NERRFilterStationCodesXMLNew} function on web services.'
+  'Get metadata for all stations, wrapper to \\code{exportStationCodesXMLNew} function on web services.',
+  'Get metadata for all stations at a single site, wrapper to \\code{NERRFilterStationCodesXMLNew} function on web services.'
   )
@@
 # get all parameters within a date range
-all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'))
 
 # get single parameter within a date range
-all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'), 
   param = 'do_mgl')
@@
 # single parameter for a station, most recent
-single_param('hudscwq', 'do_mgl')
+single_param('hudscwq', param = 'do_mgl')
 @
@@
 The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.
-This approach is most appropriate for large, specific data requests.
-The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website.
-The zip downloads feature can be used to obtain data from multiple stations in one request.
+This approach is most appropriate for large data requests.
+The \code{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO website.
+The zip downloads feature can be used to obtain a large number of records from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.
-
-Occasionally, non-unique observations are present in the raw data.
-These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.
-The \code{import\_local} function handles duplicate entries differently depending on the data type.
-For water quality and nutrient data, duplicate time stamps are simply removed.
-Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
-Nutrient data with replicates with unique time stamps are not removed but can be further processed using \code{rem\_reps}.
-Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
-Only duplicate values at 15 minutes are averaged for weather data during import.
-
-<<eval = F>>=
-# import local data for apaebmet
-
-# this is an example path with the decompressed csv files
-path <- 'C:/my_path'
-
-# this is an example path for zipped csv files
-path <- 'C:/my_path.zip'
-
-# import, do not include file extension
-import_local(path, 'apadbwq') 
-@
+The \code{import\_local} function can be used to import files directly from the zipped folder.
 
@@
 All data retrieval functions return a \code{"swmpr"} object that includes relevant data and several attributes describing the dataset.
-The data include a \code{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).
+The data include a \code{datetimestamp} column in the timezone for a station and additional parameters for the data type (weather, nutrients, or water quality).
 Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.
@@
 
-The attributes for a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).
+The attributes of a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).
 These act as metadata that are used internally by many of the package functions and are updated as the data are processed.
@@
   'Row names of the data set, inherited from the \\code{data.frame} object class.',
-  'Class of the data object indicating \\code{"swmpr"} and \\code{data.frame}.',
+  'Class of the data object indicating \\code{"swmpr"} and \\code{"data.frame"}.',
   'Station identifier used by NERRS as a string with 7 or 8 characters.',
-  "Character vector of column names for data parameters, e.g., \\code{`do\\_mgl'}, \\code{`turb'}, etc.",
+  "Character vector of column names for data parameters, e.g., \\code{'do\\_mgl'}, \\code{'turb'}, etc.",
   'Indicates if QAQC columns are present in the raw data.',
@@
 'Timezone of the station using the city/country format\\textsuperscript{a}.',
-  'Class of the \\code{datetimestamp} column, usually POSIXct unless data have been aggregated.'
+  'Class of the \\code{datetimestamp} column, usually \\code{"POSIXct"} unless data have been aggregated.'
 )
@@
 foot <- c('\\textsuperscript{a}\\footnotesize Time zones that do not observe daylight savings are used for \\code{"swmpr"} objects and may not be cities in the United States.
-For example, ``America/Jamaica" is used for Eastern Standard Time.')
+For example, \\code{America/Jamaica} is used for Eastern Standard Time.')
 
@@
 The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.
-A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.
+This uses the standard S3 object class system for R, such that specific methods for generic functions are developed for the object class.
+A \code{"swmpr"} object also secondarily inherits methods from the \code{"data.frame"} class.
 Available methods for the \code{"swmpr"} class are described below and can also be viewed:
@@
 
-A sample dataset can be downloaded for use with the examples below.
-This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.
-These data are also included with the package as binary data files (RData) that can be loaded using the \code{data} function, as shown above.
-These include \code{"swmpr"} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.
-Information for each file can be viewed in the help documentation (e.g., \code{?apacpnut}).
-
 \subsection{Data organizing}
 
-The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).
+The organize functions are used to `clean' or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).
 
 The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}).
-Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).
+Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{f\_do\_mgl} for \code{do\_mgl}).
 Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.
@@
 descrips <- c(
-  'Combines \\code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
-Only different data types can be combined.',
-  'Remove QAQC columns and remove data based on QAQC flag values for a \\code{"swmpr"} object.  Only applies if QAQC columns are present. ', 
-  'View a summary of the number of observations in a \\code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
+  'Combines \\code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.',
+  'Remove QAQC columns and remove data based on QAQC flag values for a \\code{"swmpr"} object.', 
+  'View a summary of the number of observations in a \\code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.', 
   'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
-  'Format data from a \\code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \\code{comb}.',
-  'Subset by dates and/or columns for a \\code{"swmpr"} object.  This is a method passed to the generic \\code{subset} function provided in the base installation.'
+  'Format data from a \\code{"swmpr"} object to a continuous time series at a given timestep.',
+  'Subset by dates and/or columns for a \\code{"swmpr"} object.  This is a method passed to the generic \\code{subset} function in the base installation.'
   )
@@
 
-Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
-The \code{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The \code{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
-Other suggested functions include the \code{median}, \code{min}, or \code{max}.
-The entire function call, including treatment of \code{NA} values, should be passed to the \code{FUN} argument (see the examples).
-The function is meant to be used after \code{qaqc} processing, although it works with a warning if QAQC columns are present.
-
-<<eval = F, cache = F>>=
-# get nutrient data, filter qaqc flags
-data(apacpnut)
-dat <- apacpnut
-dat <- qaqc(dat)
-
-# remove replicate nutrient data
-rem_reps(dat)
-
-# use different function to aggregate replicates
-func <- function(x) max(x, na.rm = T)
-rem_reps(dat, FUN = func)
-@
-
 A subset method added to the existing generic \code{subset} function in R is available for \code{"swmpr"} objects.
@@
 The date can be a single value or as two dates to select records within the range.
-The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{`>'} or \code{`<='}.
-The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').
-Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g.
-\code{`2012-11-31 12:00'}.
-Finally, the function can be used to remove rows and columns that do not contain data.
+The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{'>'} or \code{'<='}.
+The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., \code{'2007-01-01 06:30'}).
 
@@
 
-# subset records within a date range
-subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
-
 # subset records within a date range, select two parameters
@@
   select = c('atemp', 'totsorad'))
-
-# remove rows/columns that do not contain data
-subset(dat, rem_rows = T, rem_cols = T)
 @
@@
 The \code{setstep} function formats a \code{"swmpr"} object to a continuous time series at a given time step.
-This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.
+The function also has a default method making it useful for standardizing arbitrary time series to a given interval.
 The first argument of the function, \code{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.
@@
 Likewise, the default value for \code{differ} is one half the time step.
-Time steps that do not match any existing data within the limits of the \code{differ} argument are not discarded, although a corresponding data value will not be assigned.
 
@@
 setstep(dat, timestep = 120, differ = 30)
-
-# convert a nutrient time series to a continuous time series
-# then remove empty rows and columns
-data(apacpnut)
-dat_nut <- apacpnut
-dat_nut <- setstep(dat_nut, timestep = 60)
-subset(dat_nut, rem_rows = T, rem_cols = T)
 @
@@
 The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.
-All arguments must be called explicitly since an arbitrary number of \code{"swmpr"} objects can be used as input to \code{comb}.
-The function combines data by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.
+Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.
 The time series for merging depends on the value passed to the \code{method} argument.
-Passing \code{`union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects.
-Passing \code{`intersect'} to \code{method} will create a continuous time series from the set of dates that are shared between all input objects.
-Finally, a seven or eight character station name passed to \code{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.
-Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).
+Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.
+A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.
+As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.
 
@@
 The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).
-The general purpose tools are \code{"swmpr"} methods that were developed for existing generic functions in the R base installation or relevant packages (\pkg{SWMPr} imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}).
-These functions include \code{"swmpr"} methods for \code{aggreswmp}, \code{filter}, and \code{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \code{plot}, \code{lines}, and \code{hist} methods.
-Decomposition functions, \code{decomp} and \code{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components.
-Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \code{ecometab} and \code{plot\_metab} functions.
-The analysis functions may or may not return a \code{"swmpr"} object depending on whether further processing with \code{"swmpr"} methods is possible from the output.
-
-The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.
-This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.
-Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.
-A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.
-Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for \code{"IDate"} objects created in the \CRANpkg{data.table} package \citep{Dowle14}.
-Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).
+The general purpose tools are \code{"swmpr"} methods for existing S3 generics or are slight modifications to existing functions.
+These include \code{aggreswmp} to combine observations by set periods of time (e.g., weeks, months), \code{smoother} to average time series with a moving window, and \code{approx} to substitute missing data with interpolated values.
+For brevity, the general functions are not discussed.
+More specific functions for environmental time series include decomposition functions, \code{decomp} and \code{decomp\_cj}, and functions to estimate and plot ecosystem metabolism from combined water quality and weather data.
+Several plotting methods to facilitate analysis are also descibed below.
 
@@
   'Aggregate metabolism data from a \\code{"swmpr"} object.  This is primarly used within \\code{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
-  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
+  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \\citep{Odum56}.',
   'Decompose a \\code{"swmpr"} time series into trend, seasonal, and residual components.
@@
 Only monthly decomposition is possible.',
-  'Plot a histogram for a \\code{"swmpr"} object.',
+  'Plot a histogram for a single variable.',
   'Add lines to an existing plot created with \\code{plot}.',
   'Create a map of all stations in a reserve using the \\CRANpkg{ggmap} package \\citep{Kahle13}.',
-  'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.', 
+  'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object.', 
   'Plot multiple time series on the same y-axis.', 
-  'Plot a univariate  time series for a \\code{"swmpr"} object.  The parameter name must be specified.',
+  'Plot a univariate  time series for a \\code{"swmpr"} object.',
   'Plot ecosystem metabolism estimates after running \\code{ecometab} on a combined \\code{"swmpr"} object.',  
-  'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
-  'Smooth \\code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\code{filter}.'
+  'Create summary plots of seasonal/annual trends and anomalies for a single parameter.',
+  'Smooth \\code{"swmpr"} objects with a moving window average, passed to \\code{filter}.'
   )
@@
 
-<<eval = T, cache = F>>=
-# get data, keep all observations
-data(apacpnut)
-dat <- qaqc(apacpnut, qaqc_keep = NULL)
-
-# aggregate by quarters
-agg_dat <- aggreswmp(dat, by = 'quarters')
-nrow(agg_dat)
-
-# aggregate by quarters, remove rows with NA values
-# note the reduction in the number of rows
-agg_dat2 <- aggreswmp(dat, by = 'quarters', na.action = na.omit)
-nrow(agg_dat2)
-@
-
-Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).
-Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \code{smoother} function can be used to smooth parameters in a \code{"swmpr"} object using a specified window size.
-The \code{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
-The \code{sides} argument specifies how the average is calculated for each observation.
-Setting \code{sides = 1} will filter observations within the window that are previous to the current observation, whereas \code{sides = 2} will filter observations within the window centered at zero lag from the current observation.
-As before, the \code{params} argument specifies which parameters to smooth.
-
-<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\code{smoother} function.">>=
-# import data, qaqc and subset
-data(apadbwq)
-dat <- qaqc(apadbwq)
-dat <- subset(dat, select = 'do_mgl', 
-  subset = c('2012-07-09 00:00', '2012-07-24 00:00')
-  )
-
-# smooth using a window of 50 observations
-dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
-
-# plot raw and smoothed
-plot(dat)
-lines(dat_smooth, col = 'red', lwd = 2)
-@
-
-A common issue with any statistical analysis is the treatment of missing values.
-Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
-An approach implemented in \pkg{SWMPr} is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).
-A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
-However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.
-Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.
-An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
-The \code{maxgap} argument passed to \code{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
-
-<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
-# get data, qaqc and subset
-data(apadbwq)
-dat <- qaqc(apadbwq)
-dat <- subset(dat, select = 'do_mgl', 
-  subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
-
-# interpolate, maxgap of 10 records
-fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-
-# interpolate maxgap of 30 records
-fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
-
-# plot for comparison
-par(mfrow = c(3, 1))
-plot(dat, main = 'Raw')
-plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
-lines(dat)
-plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
-lines(dat)
-@
-
-The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
-The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
+The disaggregation of time series into additive or multiplicative components is a common application for trend analysis.
+The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
 An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
@@
 For example, the \code{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
-The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
 For simplicity, character strings of \code{'daily'} or \code{'annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
-A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
-For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} function for details).
-Use of the \code{setstep} function is also required to standardize the time step prior to decomposition.
+A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (see \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} for details).
 
@@
 
-An alternative approach to time series decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \CRANpkg{wq} package \citep{Cloern10,Jassby14}.
-The \code{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.
+An alternative approach for decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \CRANpkg{wq} package \citep{Cloern10,Jassby14}.
+The \code{decomp\_cj} function is a monthly decomposition for characterizing relatively long-term trends.
 This approach works best for nutrient data that are typically obtained on a monthly cycle.
-The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.
-The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \code{decomp} function described above.
-For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
+The time series is decomposed into the grandmean, annual, seasonal, and events components (Figure~\ref{fig:decomp_ex2}), as compared to trend, seasonal, and random components for the \code{decomp} function above.
+For both functions, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.
 Additional arguments passed to \code{decompTs} can be used with \code{decomp\_cj}, such as \code{startyr}, \code{endyr}, and \code{type}.
 Values passed to \code{type} are \code{mult} (default) or \code{add}, referring to multiplicative or additive decomposition.
-Figure~\ref{fig:decomp_ex2} shows the results from the \code{decomp\_cj} function applied to a multi-year chlorophyll time series.
 
@@
 
-Detailed exploratory graphics are also useful for evaluating general trends in observed data.
-Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
-The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-An interactive Shiny web application \citep{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
-
-<<summary_ex, fig.height = 7, fig.width = 13, message = F, cache = T, fig.cap = "Summaries of a multi-year chlorophyll time series using the \\code{plot\\_summary} function.
-Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)">>=
-## import data
-data(apacpnut)
-dat <- qaqc(apacpnut)
-
-## plot
-plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
-@
-
-The \code{overplot} function provides an alternative approach to viewing observed data from the same station.
-This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
-
-<<overplot, fig.height = 4, fig.width = 9, message = F, cache = T, fig.cap = "The \\code{overplot} function plots multiple variables on the same y-axis.">>=
-## import data
-data(apacpwq)
-dat <- qaqc(apacpwq)
-
-## plot
-overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
-  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
-@
-
-Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+Estimates of ecosystem metabolism provide a measure of system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
 The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
-A detailed discussion of the method is beyond the scope of this article, although users are strongly encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
-The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.
+A detailed discussion of the method is beyond the scope of this article, although users are encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
+Methods for estuaries have not previously been available in R, although the \CRANpkg{StreamMetabolism} package provides an approach for freshwater systems.
+The following is an example that shows use of \code{ecometab} with a combined water quality and weather data set.
 Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).
@@
 
-Finally, the \code{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the \CRANpkg{ggmap} package \citep{Kahle13}.
-This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
-The current function is limited to Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+Exploratory graphics are also useful for evaluating general trends in observed data.
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
+The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
+An interactive \CRANpkg{shiny} web application \citep{Chang15} that uses this function is available for viewing results for all SWMP sites (see the \nameref{swmp_apps} section).
+
+<<summary_ex, fig.height = 7, fig.width = 13, message = F, cache = T, fig.cap = "Summaries of a multi-year chlorophyll time series using the \\code{plot\\_summary} function.
+Summaries include monthly distributions (means on top left, quantiles on bottom left), monthly histograms (center), monthly means by year (top right), deviation from monthly means (middle right), and annual trends as deviations from the grand mean (bottom right)">>=
+## import data
+data(apacpnut)
+dat <- qaqc(apacpnut)
+
+## plot
+plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
+@
+
+Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.
+This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
+
+<<overplot, fig.height = 4, fig.width = 9, message = F, cache = T, fig.cap = "The \\code{overplot} function plots multiple variables on the same y-axis.">>=
+## import data
+data(apacpwq)
+dat <- qaqc(apacpwq)
+
+## plot
+overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
+  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
+@
+
+Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.
+The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
 The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \code{ggmap} function for more info on \code{zoom}.
 
@@
 
-Two \CRANpkg{shiny} web applications hav been created that illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
+Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
 The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
@@
 \begin{center}
-\includegraphics[width = 0.8\textwidth]{figure/swmp_comp.pdf}
+\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
 \caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
@@
 Further development of the package will include modifications and additional functions to better integrate data analysis with the quality of information provided by SWMP.
-For example, several functions include default methods to extend use beyond the \code{"swmpr"} object and additional development will continue to focus on modifying the package to handle arbitrary data structures.
+Several functions include default methods to extend use beyond the \code{"swmpr"} object and additional development will continue to focus on modifying the package to handle arbitrary data structures.
+These challenges are not unique to the SWMP database such that many of the functions will facilitate evaluations of more generic time series datasets.
 

### 2015-09-03 swmpr_manu.tex
done butchering/shortening, ready to journalify

@@
 \code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
-\code{site\_codes}&Metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
-\code{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
+\code{site\_codes}&Get metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
+\code{site\_codes\_ind}&Get metadata for all stations at a single site, wrapper to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
 \hline
@@
 \begin{verbatim}
-> # retrieve metadata for all sites
-> site_codes()
-> 
-> # retrieve metadata for a single site
-> site_codes_ind('apa')
+# retrieve metadata for all sites
+site_codes()
+
+# retrieve metadata for a single site
+site_codes_ind('apa')
 \end{verbatim}
@@
 \begin{verbatim}
-> # all parameters for a station, most recent
-> all_params('hudscwq')
-> 
-> # get all parameters within a date range
-> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
-> 
-> # get single parameter within a date range
-> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
-+   param = 'do_mgl')
-> 
-> # single parameter for a station, most recent
-> single_param('hudscwq', 'do_mgl')
+# all parameters for a station, most recent
+all_params('hudscwq')
+
+# get all parameters within a date range
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'))
+
+# get single parameter within a date range
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'), 
+  param = 'do_mgl')
+
+# single parameter for a station, most recent
+single_param('hudscwq', param = 'do_mgl')
 \end{verbatim}
@@
 The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.
-This approach is most appropriate for large, specific data requests.
-The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website.
-The zip downloads feature can be used to obtain data from multiple stations in one request.
+This approach is most appropriate for large data requests.
+The \code{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO website.
+The zip downloads feature can be used to obtain a large number of records from multiple stations in one request.
 The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).
-The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.
-
-Occasionally, non-unique observations are present in the raw data.
-These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.
-The \code{import\_local} function handles duplicate entries differently depending on the data type.
-For water quality and nutrient data, duplicate time stamps are simply removed.
-Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.
-Nutrient data with replicates with unique time stamps are not removed but can be further processed using \code{rem\_reps}.
-Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.
-Only duplicate values at 15 minutes are averaged for weather data during import.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # import local data for apaebmet
-> 
-> # this is an example path with the decompressed csv files
-> path <- 'C:/my_path'
-> 
-> # this is an example path for zipped csv files
-> path <- 'C:/my_path.zip'
-> 
-> # import, do not include file extension
-> import_local(path, 'apadbwq') 
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+The \code{import\_local} function can be used to import files directly from the zipped folder.
 
@@
 \begin{verbatim}
-> # import all paramaters for the station
-> # three most recent records
-> exdat <- all_params('apadbwq', Max = 3, trace = F)
-> exdat
+# import all paramaters for the station
+# three most recent records
+exdat <- all_params('apadbwq', Max = 3, trace = F)
+exdat
 ##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
@@
 \begin{verbatim}
-> # import sample data from package
-> data(apadbwq)
-> dat <- apadbwq
-> 
-> # view all attributes of dat
-> attributes(dat)
->  
-> # view a single attribute of dat
-> attr(dat, 'station')
+# import sample data from package
+data(apadbwq)
+dat <- apadbwq
+
+# view all attributes of dat
+attributes(dat)
+
+# view a single attribute of dat
+attr(dat, 'station')
 \end{verbatim}
@@
 \code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
-\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{data.frame}.\tabularnewline
+\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{"data.frame"}.\tabularnewline
 \code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
-\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{`do\_mgl'}, \code{`turb'}, etc.\tabularnewline
+\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{'do\_mgl'}, \code{'turb'}, etc.\tabularnewline
 \code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
@@
 \code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
-\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
+\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually \code{"POSIXct"} unless data have been aggregated.\tabularnewline
 \hline
@@
 \textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.
-For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}
+For example, \code{America/Jamaica} is used for Eastern Standard Time.\end{table}
 
 The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.
-This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.
-A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.
+This uses the standard S3 object class system for R, such that specific methods for generic functions are developed for the object class.
+A \code{"swmpr"} object also secondarily inherits methods from the \code{"data.frame"} class.
 Available methods for the \code{"swmpr"} class are described below and can also be viewed:
@@
 \begin{verbatim}
-> # view available methods for swmpr class
-> methods(class = 'swmpr')
+# view available methods for swmpr class
+methods(class = 'swmpr')
 \end{verbatim}
@@
 \hline
-\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.
-Only different data types can be combined.\tabularnewline
-\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.  Only applies if QAQC columns are present. \tabularnewline
-\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.
-The output can be used to inform further processing.\tabularnewline
+\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.\tabularnewline
+\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.\tabularnewline
+\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.\tabularnewline
 \code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
-\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \code{comb}.\tabularnewline
-\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function provided in the base installation.\tabularnewline
+\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.\tabularnewline
+\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function in the base installation.\tabularnewline
 \hline
@@
 \begin{verbatim}
-> # qaqc screen for a swmpr object, retain only '0'
-> qaqc(dat)
-> 
-> # retain all data regardless of flag
-> qaqc(dat, qaqc_keep = NULL)
-> 
-> # retain only '0' and '-1' flags
-> qaqc(dat, qaqc_keep = c(0, -1))
+# qaqc screen for a swmpr object, retain only '0'
+qaqc(dat)
+
+# retain all data regardless of flag
+qaqc(dat, qaqc_keep = NULL)
+
+# retain only '0' and '-1' flags
+qaqc(dat, qaqc_keep = c(0, -1))
 \end{verbatim}
@@
 \begin{verbatim}
-> # view the number of observations in each QAQC flag
-> qaqcchk(dat)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.
-The \code{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.
-The \code{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.
-Other suggested functions include the \code{median}, \code{min}, or \code{max}.
-The entire function call, including treatment of \code{NA} values, should be passed to the \code{FUN} argument (see the examples).
-The function is meant to be used after \code{qaqc} processing, although it works with a warning if QAQC columns are present.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get nutrient data, filter qaqc flags
-> data(apacpnut)
-> dat <- apacpnut
-> dat <- qaqc(dat)
-> 
-> # remove replicate nutrient data
-> rem_reps(dat)
-> 
-> # use different function to aggregate replicates
-> func <- function(x) max(x, na.rm = T)
-> rem_reps(dat, FUN = func)
+# view the number of observations in each QAQC flag
+qaqcchk(dat)
 \end{verbatim}
@@
 \begin{verbatim}
-> # import data
-> data(apaebmet)
-> dat <- apaebmet
-> 
-> # select two parameters from dat
-> subset(dat, select = c('rh', 'bp'))
-> 
-> # subset records greater than or equal to a date
-> subset(dat, subset = '2013-01-01 0:00', operator = '>=')
-> 
-> # subset records within a date range
-> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
-> 
-> # subset records within a date range, select two parameters
-> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
-+   select = c('atemp', 'totsorad'))
-> 
-> # remove rows/columns that do not contain data
-> subset(dat, rem_rows = T, rem_cols = T)
+# import data
+data(apaebmet)
+dat <- apaebmet
+
+# select two parameters from dat
+subset(dat, select = c('rh', 'bp'))
+
+# subset records greater than or equal to a date
+subset(dat, subset = '2013-01-01 0:00', operator = '>=')
+
+# subset records within a date range, select two parameters
+subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+  select = c('atemp', 'totsorad'))
 \end{verbatim}
@@
 \begin{verbatim}
-> # import, qaqc removal
-> data(apadbwq)
-> dat <- qaqc(apadbwq)
-> 
-> # convert time series to two hour invervals
-> # tolerance of +/- 30 minutes for matching existing data
-> setstep(dat, timestep = 120, differ = 30)
-> 
-> # convert a nutrient time series to a continuous time series
-> # then remove empty rows and columns
-> data(apacpnut)
-> dat_nut <- apacpnut
-> dat_nut <- setstep(dat_nut, timestep = 60)
-> subset(dat_nut, rem_rows = T, rem_cols = T)
+# import, qaqc removal
+data(apadbwq)
+dat <- qaqc(apadbwq)
+
+# convert time series to two hour invervals
+# tolerance of +/- 30 minutes for matching existing data
+setstep(dat, timestep = 120, differ = 30)
 \end{verbatim}
@@
 \begin{verbatim}
-> # get nut, wq, and met data as separate objects
-> data(apacpnut)
-> data(apacpwq)
-> data(apaebmet)
-> swmp1 <- apacpnut
-> swmp2 <- apacpwq
-> swmp3 <- apaebmet
-> 
-> # combine nut and wq data by union
-> comb(swmp1, swmp2, method = 'union')
-> 
-> # combine nut and wq data by intersect
-> comb(swmp1, swmp3, method = 'intersect')
-> 
-> # combine nut, wq, and met data by nut time series, two hour time step
-> comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
+# get nut, wq, and met data as separate objects
+data(apacpnut)
+data(apacpwq)
+data(apaebmet)
+swmp1 <- apacpnut
+swmp2 <- apacpwq
+swmp3 <- apaebmet
+
+# combine nut and wq data by union
+comb(swmp1, swmp2, method = 'union')
+
+# combine nut and wq data by intersect
+comb(swmp1, swmp3, method = 'intersect')
+
+# combine nut, wq, and met data by nut time series, two hour time step
+comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
 \end{verbatim}
@@
 This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
-\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
+\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \citep{Odum56}.\tabularnewline
 \code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.
@@
 Only monthly decomposition is possible.\tabularnewline
-\code{hist}&Plot a histogram for a \code{"swmpr"} object.\tabularnewline
+\code{hist}&Plot a histogram for a single variable.\tabularnewline
 \code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
 \code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
-\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
+\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
 \code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
-\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
+\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline
 \code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
-\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
-\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \code{filter}.\tabularnewline
+\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a single parameter.\tabularnewline
+\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
 \hline
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get data, keep all observations
-> data(apacpnut)
-> dat <- qaqc(apacpnut, qaqc_keep = NULL)
-> 
-> # aggregate by quarters
-> agg_dat <- aggreswmp(dat, by = 'quarters')
-> nrow(agg_dat)
-## [1] 47
-> # aggregate by quarters, remove rows with NA values
-> # note the reduction in the number of rows
-> agg_dat2 <- aggreswmp(dat, by = 'quarters', na.action = na.omit)
-> nrow(agg_dat2)
-## [1] 16
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).
-Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.
-The \code{smoother} function can be used to smooth parameters in a \code{"swmpr"} object using a specified window size.
-The \code{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.
-The \code{sides} argument specifies how the average is calculated for each observation.
-Setting \code{sides = 1} will filter observations within the window that are previous to the current observation, whereas \code{sides = 2} will filter observations within the window centered at zero lag from the current observation.
-As before, the \code{params} argument specifies which parameters to smooth.
+The disaggregation of time series into additive or multiplicative components is a common application for trend analysis.
+The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
+An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
+The \code{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
+For example, the \code{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
+For simplicity, character strings of \code{'daily'} or \code{'annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
+A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (see \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} for details).
 
@@
 \begin{verbatim}
-> # import data, qaqc and subset
-> data(apadbwq)
-> dat <- qaqc(apadbwq)
-> dat <- subset(dat, select = 'do_mgl', 
-+   subset = c('2012-07-09 00:00', '2012-07-24 00:00')
-+   )
-> 
-> # smooth using a window of 50 observations
-> dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
-> 
-> # plot raw and smoothed
-> plot(dat)
-> lines(dat_smooth, col = 'red', lwd = 2)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+# get data
+data(apadbwq)
+dat <- apadbwq
 
-{\centering \includegraphics[width=\textwidth]{figure/smooth_ex-1}
+# subset for daily decomposition
+dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
 
-}
-
-\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function.}\label{fig:smooth_ex}
-\end{figure}
-
-\end{knitrout}
-
-A common issue with any statistical analysis is the treatment of missing values.
-Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.
-An approach implemented in \pkg{SWMPr} is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).
-A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.
-However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.
-Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.
-An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.
-The \code{maxgap} argument passed to \code{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get data, qaqc and subset
-> data(apadbwq)
-> dat <- qaqc(apadbwq)
-> dat <- subset(dat, select = 'do_mgl', 
-+   subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
-> 
-> # interpolate, maxgap of 10 records
-> fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)
-> 
-> # interpolate maxgap of 30 records
-> fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
-> 
-> # plot for comparison
-> par(mfrow = c(3, 1))
-> plot(dat, main = 'Raw')
-> plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
-> lines(dat)
-> plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
-> lines(dat)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
-
-{\centering \includegraphics[width=\textwidth]{figure/interp_ex-1}
-
-}
-
-\caption[Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
-\end{figure}
-
-\end{knitrout}
-
-The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.
-The \code{decomp} function is a simple wrapper to \code{decompose} \citep{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).
-An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.
-The \code{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.
-For example, the \code{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).
-The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).
-For simplicity, character strings of \code{'daily'} or \code{'annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.
-A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.
-For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\code{ts}} function for details).
-Use of the \code{setstep} function is also required to standardize the time step prior to decomposition.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # get data
-> data(apadbwq)
-> dat <- apadbwq
-> 
-> # subset for daily decomposition
-> dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
-> 
-> # daily decomposition of DO and plot
-> dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
-> plot(dc_dat)
+# daily decomposition of DO and plot
+dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
+plot(dc_dat)
 \end{verbatim}
@@
 \begin{verbatim}
-> # get data
-> data(apacpnut)
-> dat <- apacpnut
-> dat <- qaqc(dat, qaqc_keep = NULL)
-> 
-> # decomposition of chl
-> decomp_cj(dat, param = 'chla_n')
+# get data
+data(apacpnut)
+dat <- apacpnut
+dat <- qaqc(dat, qaqc_keep = NULL)
+
+# decomposition of chl
+decomp_cj(dat, param = 'chla_n')
 \end{verbatim}
@@
 
-Detailed exploratory graphics are also useful for evaluating general trends in observed data.
-Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
-The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-An interactive Shiny web application \citep{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).
+Estimates of ecosystem metabolism provide a measure of system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
+The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
+A detailed discussion of the method is beyond the scope of this article, although users are encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
+Methods for estuaries have not previously been available in R, although the \CRANpkg{StreamMetabolism} package provides an approach for freshwater systems.
+The following is an example that shows use of \code{ecometab} with a combined water quality and weather data set.
+Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).
 
@@
 \begin{verbatim}
-> ## import data
-> data(apacpnut)
-> dat <- qaqc(apacpnut)
-> 
-> ## plot
-> plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
+## import water quality and weather data
+data(apadbwq)
+data(apaebmet)
+
+## qaqc, combine
+wq <- qaqc(apadbwq)
+met <- qaqc(apaebmet)
+dat <- comb(wq, met)
+
+## estimate metabolism
+res <- ecometab(dat, trace = FALSE)
+plot_metab(res)
+\end{verbatim}
+\end{kframe}\begin{figure}[!h]
+
+{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}
+
+}
+
+\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.}\label{fig:metab_ex}
+\end{figure}
+
+\end{knitrout}
+
+Exploratory graphics are also useful for evaluating general trends in observed data.
+Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \code{plot\_summary} function (Figure~\ref{fig:summary_ex}).
+The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
+An interactive \CRANpkg{shiny} web application \citep{Chang15} that uses this function is available for viewing results for all SWMP sites (see the \nameref{swmp_apps} section).
+
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## import data
+data(apacpnut)
+dat <- qaqc(apacpnut)
+
+## plot
+plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
 \end{verbatim}
@@
 
-The \code{overplot} function provides an alternative approach to viewing observed data from the same station.
+Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.
 This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
@@
 \begin{verbatim}
-> ## import data
-> data(apacpwq)
-> dat <- qaqc(apacpwq)
-> 
-> ## plot
-> overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
-+   subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
+## import data
+data(apacpwq)
+dat <- qaqc(apacpwq)
+
+## plot
+overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
+  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
 \end{verbatim}
@@
 
-Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.
-The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}.
-A detailed discussion of the method is beyond the scope of this article, although users are strongly encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).
-The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.
-Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).
+Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.
+The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
+The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
 
@@
 \begin{verbatim}
-> ## import water quality and weather data
-> data(apadbwq)
-> data(apaebmet)
-> 
-> ## qaqc, combine
-> wq <- qaqc(apadbwq)
-> met <- qaqc(apaebmet)
-> dat <- comb(wq, met)
-> 
-> ## estimate metabolism
-> res <- ecometab(dat, trace = FALSE)
-> plot_metab(res)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
-
-{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}
-
-}
-
-\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.}\label{fig:metab_ex}
-\end{figure}
-
-\end{knitrout}
-
-Finally, the \code{map\_reserve} function can be used to create a map with all stations at a reserve using functions in the \CRANpkg{ggmap} package \citep{Kahle13}.
-This map may be useful for aiding the interpretation of spatial trends in water quality parameters given the relative locations in a reserve.
-The current function is limited to Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
-The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.
-See the help documentation for the \code{ggmap} function for more info on \code{zoom}.
-
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-> # plot the stations at Jacques Cousteau reserve
-> map_reserve('jac')
+# plot the stations at Jacques Cousteau reserve
+map_reserve('jac')
 \end{verbatim}

### 2015-09-03 Rjournal/beck.tex
submitted to R Journal

--- 2015-09-03 swmpr_manu.tex
@@
-\documentclass[a4paper]{report}\usepackage[]{graphicx}\usepackage{xcolor}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{RJournal}
-\usepackage{amsmath,amssymb,array}
-\usepackage{booktabs}
-
-%% load any required packages
-
-% bold/italic text
-\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-
-%% do not edit, for illustration only
-\sectionhead{Contributed research article}
-\volume{XX}
-\volnumber{YY}
-\year{20ZZ}
-\month{AAAA}
-
-\begin{article}
-
-%knitr options
-
 \title{\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # retrieve metadata for all sites
@@
 site_codes_ind('apa')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 Download time can be excessive for longer time series.
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # all parameters for a station, most recent
@@
 single_param('hudscwq', param = 'do_mgl')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # import all paramaters for the station
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # import sample data from package
@@
 attr(dat, 'station')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # view available methods for swmpr class
 methods(class = 'swmpr')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # qaqc screen for a swmpr object, retain only '0'
@@
 qaqc(dat, qaqc_keep = c(0, -1))
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # view the number of observations in each QAQC flag
 qaqcchk(dat)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # import data
@@
   select = c('atemp', 'totsorad'))
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # import, qaqc removal
@@
 setstep(dat, timestep = 120, differ = 30)
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # get nut, wq, and met data as separate objects
@@
 comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\end{example}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # get data
@@
 plot(dc_dat)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/decomp_ex1-1}
+{\centering \includegraphics[width=\textwidth]{decomp_ex1-1}
 
@@
 decomp_cj(dat, param = 'chla_n')
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/decomp_ex2-1}
+{\centering \includegraphics[width=\textwidth]{decomp_ex2-1}
 
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 ## import water quality and weather data
@@
 plot_metab(res)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}
+{\centering \includegraphics[width=\textwidth]{metab_ex-1}
 
@@
 
-\end{knitrout}
-
 Exploratory graphics are also useful for evaluating general trends in observed data.
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 ## import data
@@
 plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/summary_ex-1}
+{\centering \includegraphics[width=\textwidth]{summary_ex-1}
 
@@
 
-\end{knitrout}
-
 Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.
@@
   subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/overplot-1}
+{\centering \includegraphics[width=\textwidth]{overplot-1}
 
@@
 
-\end{knitrout}
-
 Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.
@@
 
-\begin{knitrout}\small
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{example}
 # plot the stations at Jacques Cousteau reserve
 map_reserve('jac')
-\end{verbatim}
-\end{kframe}\begin{figure}[!h]
+\end{example}
+\begin{figure}[!h]
 
-{\centering \includegraphics[width=\textwidth]{figure/map_ex-1}
+{\centering \includegraphics[width=\textwidth]{map_ex-1}
 
@@
 \begin{center}
-\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
+\includegraphics[width = \textwidth]{swmp_comp.pdf}
 \caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
@@
 \email{beck.marcus@epa.gov}
-
-\end{article}
-
-\end{document}

### 2015-09-03 cover_letter.tex
submitted to R Journal

@@
 \emph{To:}\newline
-Dr. Damian Pattinson\newline
-Editorial Director\newline
-PLOS ONE
+Bettina Gr\"un\newline
+Editor in Chief\newline
+The R Journal
 
 \emph{From:}\newline
-Dr. Marcus W. Beck\newline
+Marcus W. Beck\newline
 ORISE Research Participation Program\newline
@@
 
-Enclosed please find my manuscript, entitled `SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries', to be considered as a research article in PLOS ONE.
+Enclosed please find my manuscript, entitled `SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries', to be considered as a research article in the R Journal.
+
 A continuing challenge in the environmental sciences is the effective use of continuous monitoring data.
 The increasing quantity of data from automated sensor networks suggests that more appropriate methods to synthesize and interpret this information are needed.
-This article describes the SWMPr (`swamper') software package that was developed to address research needs of the US National Estuarine Research Reserve System (NERRS) using the open-source statistical programming language R.
+This article describes the SWMPr (`swamper') software package for R that was developed in part to address research needs of the US National Estuarine Research Reserve System (NERRS).
 The System-Wide Monitoring Program (SWMP) was implemented by NERRS in 1995 to collect continuous water quality data at over 140 fixed stations at each of the estuaries in the reserve system.
-The SWMPr package address many common issues working with large datasets from automated sensor networks, such as data pre-processing, combining data from different sources, and exploratory analyses to identify parameters of interest.
-Additionally, a cross-reserve comparison of water quality trends and ecosystem metabolism estimates is provided to illustrate the utility of the package.
+The SWMPr package addresses many common issues working with large datasets from automated sensor networks, such as data pre-processing, combining data from different sources, and exploratory analyses to identify parameters of interest.
+As such, the package has value for the NERRS network as well as more generic applications to environmental time series data.
+Several functions provided in the package are novel contributions to CRAN.
 
-The manuscript has not been previously submitted to PLOS ONE and I have had no prior interactions with PLOS.
-Please note that Dr.
-Jeffrey Hollister (US EPA) has already provided a review of an earlier draft.
-I greatly appreciate the opportunity to publish my work in PLOS ONE.
-Below is a suggested list of appropriate editors to handle my manuscript.
-
-Zaid Abdo,
-Institution and Department: Agricultural Research Service,
-UNITED STATES
-
-Ruth H. Carmichael,
-Dauphin Island Sea Lab; University of South Alabama,
-UNITED STATES
-
-Juan Carlos Molinero,
-GEOMAR: Helmholtz Center for Ocean Research,
-GERMANY
-
-Tomoya Iwata,
-University of Yamanashi,
-JAPAN
+I greatly appreciate the opportunity to publish my work in the R Journal.  Please feel free to contact me directly should additional information be needed.
 

### 2015-09-03 Rjournal/Beck_code.R
added R code file for examples

@@ -1,0 +1,145 @@
+## load SWMPr
+## install.packages('SWMPr')
+library(SWMPr)
+
+## Most of the data retrieval functions will not work unless the IP address is registered
+## See http://cdmo.baruch.sc.edu/webservices.cfm
+
+# retrieve metadata for all sites
+site_codes()
+
+# retrieve metadata for a single site
+site_codes_ind('apa')
+
+# all parameters for a station, most recent
+all_params('hudscwq')
+
+# get all parameters within a date range
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'))
+
+# get single parameter within a date range
+all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'),
+  param = 'do_mgl')
+
+# single parameter for a station, most recent
+single_param('hudscwq', param = 'do_mgl')
+
+# import all paramaters for the station
+# three most recent records
+exdat <- all_params('apadbwq', Max = 3, trace = F)
+exdat
+
+# import sample data from package
+data(apadbwq)
+dat <- apadbwq
+
+# view all attributes of dat
+attributes(dat)
+
+# view a single attribute of dat
+attr(dat, 'station')
+
+# view available methods for swmpr class
+methods(class = 'swmpr')
+
+# qaqc screen for a swmpr object, retain only '0'
+qaqc(dat)
+
+# retain all data regardless of flag
+qaqc(dat, qaqc_keep = NULL)
+
+# retain only '0' and '-1' flags
+qaqc(dat, qaqc_keep = c(0, -1))
+
+# view the number of observations in each QAQC flag
+qaqcchk(dat)
+
+# import data
+data(apaebmet)
+dat <- apaebmet
+
+# select two parameters from dat
+subset(dat, select = c('rh', 'bp'))
+
+# subset records greater than or equal to a date
+subset(dat, subset = '2013-01-01 0:00', operator = '>=')
+
+# subset records within a date range, select two parameters
+subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+  select = c('atemp', 'totsorad'))
+
+# import, qaqc removal
+data(apadbwq)
+dat <- qaqc(apadbwq)
+
+# convert time series to two hour invervals
[…]

### 2015-10-18 swmpr_manu.Rnw
stupid abstract edits

@@
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-Simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
+However, the increasing quantity of data provided by the monitoring network have complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.
 This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.

### 2015-10-18 swmpr_manu.tex
stupid abstract edits

@@
 ##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
-## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
-## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
-## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
+## 1 2015-10-06 10:30:00   23      0     47        0  30     0     71
+## 2 2015-10-06 10:45:00   23      0     46        0  30     0     70
+## 3 2015-10-06 11:00:00   23      0     46        0  30     0     62
 ##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
-## 1        0      5        0     2       0  8    0   13      0       NA
-## 2        0      5        0     2       0  8    0    9      0       NA
-## 3        0      6        0     2       0  8    0   11      0       NA
+## 1        0      5        0     2       0  8    0    5      0       NA
+## 2        0      5        0     2       0  8    0  246      0       NA
+## 3        0      4        0     2       0  8    0    4      0       NA
 ##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
-## 1         -2    NA      -1     NA     NA       -1         
-## 2         -2    NA      -1     NA     NA       -1         
-## 3         -2    NA      -1     NA     NA       -1
+## 1         -2    NA      -1      2     NA        3         
+## 2         -2    NA      -1      2     NA        3         
+## 3         -2    NA      -1      2     NA        3
 \end{verbatim}

### 2015-11-16 review_response.Rnw
started revision from R journal

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}
\usepackage{amsmath}

%acronyms
% \acrodef{}{}

%knitr options
<<setup, cache = F, echo = F>>=
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\setlength{\parskip}{5mm}
\setlength{\parindent}{0in}

\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

\begin{document}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to reviewer comments ``SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries'', by M. W. Beck}

{\it The authors wish to thank the reviewers for providing thoughtful comments and suggestions on my manuscript.  My response to these comments are shown in italics.  Page and paragraph numbers refer to the original manuscript.}

\Bigtxt{Reviewer 1:}

*Main text*
I didn't find it obvious from the text whether the setstep and comb functions would handle missing data with interpolation or NA. (It's NA, as it turns out.) Perhaps it is because I don't do a lot of work in this domain, though.

*Package performance*

Some of the functions can spend a surprising amount of time parsing. For example, the call: all\_params\_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013')) took 19 seconds on my machine, and running under a profiler showed that about 12-3 of those seconds were spent parsing the data, not downloading it. The SWMPr::parser code calls htmlTreeParse, which is where all that time is being spent; I found that changing it to xmlTreeParse made almost all of that time go away. The only difference I noticed was that htmlTreeParse converts all of the tag names to lowercase, while xmlTreeParse does not. This can be easily fixed by having the parser() function change the `out` data frame's names to lowercase before returning. With these two modifications, the elapsed time drops to 6-7 seconds--the bulk of the time now being spent downloading the data (as far as I can tell).

*shiny\_comp app*

The Shiny app that uses Leaflet is using a deprecated API. The new API is documented at https://rstudio.github.io/leaflet/ and I think it would significantly simplify the code. I can lend a hand if the author wants.

The Leaflet map could use a color legend, which is possible with the new Leaflet API. (It could use a radius legend too, but that is not yet a feature of the Leaflet package.)

There is a feature in the new Leaflet API that lets you have a label <http://leaflet.github.io/Leaflet.label/> for each circle that is either always shown, or shown on mouseover. This is a nicer experience for the user than having to click. This feature has not yet made it to CRAN, but it is available in the GitHub master branch of Leaflet.

It would be interesting if the Leaflet app could let you select multiple stations and see the data plotted together.

*shiny\_summary app*

In the file https://github.com/fawda123/swmp\_summary/blob/master/server.R, the plotInput should not be a reactive(), but a regular function that takes no arguments. This is because reactives cache their values, and should not be executed for their side effects (in this case, plotting). As the app is written now, there doesn't appear to be an actual user-visible bug, but a change as simple as setting the plot width to "100\%" would introduce a bug (I'd expect it to fail to redraw the plot when the window was resized).

\Bigtxt{Reviewer 2:}

First, I did not see any information about how SWMPR treats censored data, i.e., measurements that can be reported only as less than or greater than an analysis or reporting limit. For example, measurements of nutrient parameters could include values that are below analytical detection limits. A considerable amount of research has shown that the assumptions and methods used to analyze data that include censored values can influence the conclusions of the analyses (for example, see Helsel 2012). Considerable attention is given in the manuscript, appropriately, to various quality assurance/quality control functions and flags, and I realize not every aspect of QA/QC needs to be covered explicitly in the manuscript. However, I believe that the topic of censored values is important enough that the paper would be improved by providing explicit information about how censored values may be coded, retrieved, and ultimately used in various data analyses carried out by SWMPr.

Second, in the "Applications using the SWMPr package" section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time. This particular statistical method may or may not be appropriate for a given data set. Alternative trend analysis methods exist, including several that treat censored values appropriately (again, Helsel 2012 is a useful reference). It is unclear from the manuscript what alternatives to SLR may be available in SWMPr (e.g., maximum likelihood estimation, Akritas-Theil-Sen nonparametric regression, others). Therefore, clarifying text that either identifies the availability of other trend analysis procedures in SWMPr or provides caveats to the use of SLR (e.g., used only for initial exploratory analysis) should be added to this section.

Reference:
Helsel, D.R. 2012. Statistics for Censored Environmental Data Using Minitab and R. 2nd edition, John Wiley \& Sons, Inc. Hoboken, New Jersey. 324pp.

\end{document}

### 2015-11-16 review_response.tex
started revision from R journal

--- 2015-11-16 review_response.Rnw
@@
 \newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}

### 2015-11-16 swmpr_manu.tex
started revision from R journal

@@
 ##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
-## 1 2015-10-06 10:30:00   23      0     47        0  30     0     71
-## 2 2015-10-06 10:45:00   23      0     46        0  30     0     70
-## 3 2015-10-06 11:00:00   23      0     46        0  30     0     62
+## 1 2015-11-03 11:15:00   26      0     45        0  29     0     78
+## 2 2015-11-03 11:30:00   26      0     46        0  30     0     76
+## 3 2015-11-03 11:45:00   26      0     46        0  30     0     75
 ##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
-## 1        0      5        0     2       0  8    0    5      0       NA
-## 2        0      5        0     2       0  8    0  246      0       NA
-## 3        0      4        0     2       0  8    0    4      0       NA
+## 1        0      5        0     2       0  8    0    2      0       NA
+## 2        0      5        0     2       0  8    0    5      0       NA
+## 3        0      5        0     2       0  8    0    5      0       NA
 ##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel

### 2015-11-16 review_response.Rnw
response to second reviewer done

@@
 
-{\it The authors wish to thank the reviewers for providing thoughtful comments and suggestions on my manuscript.
+{\it I thank the reviewers for providing thoughtful comments on my manuscript and the SWMPr package.
 My response to these comments are shown in italics.
@@
 *Main text*
+
 I didn't find it obvious from the text whether the setstep and comb functions would handle missing data with interpolation or NA.
@@
 
+{\it I fully agree that considering censored data is a necessary aspect of processing and analyzing water quality data.
+Censored data, either above or below the detection limit, are given unique QAQC flags during initial data processing through the Centralized Data Management Office (CDMO).
+Censored values in the imported data can be identified using functions in the SWMPr package.
+The default behavior of the qaqc function is to remove censored values, but the user can also retain these values by changing the function arguments.
+A data frame showing the number of observations for a given QAQC flag can also be retrieved using the qaqcchk function.
+I have revised the text to more clearly describe the presence of censored values in SWMP data and how SWMPr can be used to assess their distribution (page 4, paragraph 3):
+
+``SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.
+For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.
+The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series (Helsel, 2012).
+A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages (e.g., cents, McLeod et al., 2014).
+However, the functions in SWMPr can be used to identify censored data based on the appropriate QAQC flag for a given parameter.
+Viewing this information can be helpful for determining how to further process the data with the qaqc function or alternative methods outside of SWMPr.
+The qaqcchk function returns a data.frame of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.
+SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.''
+}
+
 Second, in the "Applications using the SWMPr package" section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time.
@@
 
+{\it I agree that the use of SLR is a simple method that may not be appropriate for all datasets provided by SWMP.
+The application is meant to be exploratory and I have revised the text to clarify that SLR is a simple and potentially inadequate method for trend analysis, particularly with censored data.
+The text description on the application was also revised.
+
+Page 12, paragraph one: `More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.'
+
+Added to app: `Please note that the use of simple regression to identify trends is for exploratory purposes only and may not be appropriate for all datasets.'
+}
+
 Reference:
@@
 
+{\it This reference was added to the manuscript.}
+
 \end{document}

### 2015-11-16 swmpr_manu.Rnw
response to second reviewer done

@@
 
-Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \code{qaqc}.
-The \code{qaqcchk} function can be used to view this information.
+SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.
+For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.
+The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series \citep{Helsel12}.
+A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.
+However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a given parameter.
+Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.
+The \code{qaqcchk} function returns a \code{data.frame} of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.
+SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.
 
@@
 Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.
+More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.

### 2015-11-16 review_response.Rnw
response to first reviewer done, need to proofread, recompile

@@
 
-{\it Response to reviewer comments ``SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries'', by M. W. Beck}
+{\it Response to reviewer comments ``SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries'', by M. W. Beck.}
 
@@
 
+{\it This was clarified in the revision, page 6 paragraph one:
+
+`Both functions treat missing data as NA values, either for observations that exceed the allowable tolerance for the differ argument of setstep or for time series that do not overlap given the method argument passed to comb.'
+}
+
 *Package performance*
@@
 Some of the functions can spend a surprising amount of time parsing.
-For example, the call: all\_params\_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013')) took 19 seconds on my machine, and running under a profiler showed that about 12-3 of those seconds were spent parsing the data, not downloading it.
+For example, the call: all\_params\_dtrng(`hudscwq', dtrng = c(`09/01/2013', `10/01/2013')) took 19 seconds on my machine, and running under a profiler showed that about 12-3 of those seconds were spent parsing the data, not downloading it.
 The SWMPr::parser code calls htmlTreeParse, which is where all that time is being spent; I found that changing it to xmlTreeParse made almost all of that time go away.
 The only difference I noticed was that htmlTreeParse converts all of the tag names to lowercase, while xmlTreeParse does not.
-This can be easily fixed by having the parser() function change the `out` data frame's names to lowercase before returning.
+This can be easily fixed by having the parser() function change the `out' data frame's names to lowercase before returning.
 With these two modifications, the elapsed time drops to 6-7 seconds--the bulk of the time now being spent downloading the data (as far as I can tell).
+
+{\it That is a fantastic suggestion and I've made these changes in the development repository of the package.
+I noticed an immediate increase in performance with the import functions.
+I will push the new version to CRAN if or when the paper is accepted for publication.
+}
 
@@
 
+{\it I would be interested in working together to improve the code.
+I made the app by borrowing an example from the rstudio GitHub repo (https://github.com/rstudio/leaflet/tree/master/inst/legacy/examples/population) and tweaking the code for my application.
+I admit that I have a fuzzy understanding of using leaflets maps reactively and I have always had trouble editing the code as a result.
+Simplification with the new API would be great not only for performance, but also to help me better understand reactive programming.}
+
 The Leaflet map could use a color legend, which is possible with the new Leaflet API. (It could use a radius legend too, but that is not yet a feature of the Leaflet package.)
 
-There is a feature in the new Leaflet API that lets you have a label <http://leaflet.github.io/Leaflet.label/> for each circle that is either always shown, or shown on mouseover.
+There is a feature in the new Leaflet API that lets you have a label http://leaflet.github.io/Leaflet.label/ for each circle that is either always shown, or shown on mouseover.
 This is a nicer experience for the user than having to click.
@@
 
+{\it Again, these are all great suggestions but I would probably need assistance to implement these changes with the new API.}
+
 *shiny\_summary app*
@@
 This is because reactives cache their values, and should not be executed for their side effects (in this case, plotting).
-As the app is written now, there doesn't appear to be an actual user-visible bug, but a change as simple as setting the plot width to "100\%" would introduce a bug (I'd expect it to fail to redraw the plot when the window was resized).
+As the app is written now, there doesn't appear to be an actual user-visible bug, but a change as simple as setting the plot width to ``100\%'' would introduce a bug (I'd expect it to fail to redraw the plot when the window was resized).
+
+{\it Yes, the plot fails to render on shinyapps when the window is resized.
+I had not noticed the bug so I appreciate the comment.
+I've changed the code in server.R to make plotInput a regular function.
+I've redeployed the app and the bug has been fixed.}
 
@@
 
-Second, in the "Applications using the SWMPr package" section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time.
+Second, in the ``Applications using the SWMPr package'' section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time.
 This particular statistical method may or may not be appropriate for a given data set.
@@
 
-{\it This reference was added to the manuscript.}
+{\it This reference was added for the text additions above.}
 

### 2015-11-16 swmpr_manu.tex
response to first reviewer done, need to proofread, recompile

@@
 
+\begin{knitrout}\small
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in loadNamespace(j <- i[[1L]], c(lib.loc, .libPaths()), versionCheck = vI[[j]]): there is no package called 'Rcpp'}}\end{kframe}
+\end{knitrout}
+
 \pkg{SWMPr} functions are categorized by one of three steps in the data workflow: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.
@@
 % table for retrieval functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
-\begin{table}[!tbp]
-\caption{Retrieval functions available from the \pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}}
-\begin{center}
-\begin{tabular}{lp{3.5in}}
-\hline\hline
-\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
-\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
-\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.
-Optional argument for a single parameter.
-Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
-\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
-\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
-\code{site\_codes}&Get metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
-\code{site\_codes\_ind}&Get metadata for all stations at a single site, wrapper to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}
 
@@
 exdat <- all_params('apadbwq', Max = 3, trace = F)
+\end{verbatim}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}all\_params"{}}}\begin{verbatim}
 exdat
-##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
-## 1 2015-11-03 11:15:00   26      0     45        0  29     0     78
-## 2 2015-11-03 11:30:00   26      0     46        0  30     0     76
-## 3 2015-11-03 11:45:00   26      0     46        0  30     0     75
-##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
-## 1        0      5        0     2       0  8    0    2      0       NA
-## 2        0      5        0     2       0  8    0    5      0       NA
-## 3        0      5        0     2       0  8    0    5      0       NA
-##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
-## 1         -2    NA      -1      2     NA        3         
-## 2         -2    NA      -1      2     NA        3         
-## 3         -2    NA      -1      2     NA        3
 \end{verbatim}
-\end{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'exdat' not found}}\end{kframe}
 \end{knitrout}
@@
 % attributes table
-%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
-\begin{table}[!tbp]
-\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
-\begin{center}
-\begin{tabular}{lp{0.75in}p{3.25in}}
-\hline\hline
-\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
-\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
-\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
-\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{"data.frame"}.\tabularnewline
-\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
-\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{'do\_mgl'}, \code{'turb'}, etc.\tabularnewline
-\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
-\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
-\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
-\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually \code{"POSIXct"} unless data have been aggregated.\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.
-For example, \code{America/Jamaica} is used for Eastern Standard Time.\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}
 
@@
 % table for organize functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
-\begin{table}[!tbp]
-\caption{Organizing functions available from the \pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}}
-\begin{center}
-\begin{tabular}{lp{3.5in}}
-\hline\hline
-\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
-\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.\tabularnewline
-\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.\tabularnewline
-\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.\tabularnewline
-\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
-\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.\tabularnewline
-\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function in the base installation.\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}
 
@@
 As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.
+Both functions treat missing data as \code{NA} values, either for observations that exceed the allowable tolerance for the \code{differ} argument of \code{setstep} or for time series that do not overlap given the \code{method} argument passed to \code{comb}.
 
@@
 % table for analysis functions
-%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
-\begin{table}[!tbp]
-\caption{Analysis functions available from the \pkg{SWMPr} package.
-Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}}
-\begin{center}
-\begin{tabular}{lp{3.5in}}
-\hline\hline
-\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
-\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
-\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.
-This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
-\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \citep{Odum56}.\tabularnewline
-\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.
-This is a simple wrapper to \code{decompose} \citep{Kendall83}.
-Decomposition of monthly or daily trends is possible.\tabularnewline
-\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.
-This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.
-Only monthly decomposition is possible.\tabularnewline
-\code{hist}&Plot a histogram for a single variable.\tabularnewline
-\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
-\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
-\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
-\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
-\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline
-\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
-\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a single parameter.\tabularnewline
-\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
-\hline
-\end{tabular}\end{center}
+\begin{kframe}
 
-\end{table}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error: package 'ggplot2' could not be loaded}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}latex"{}}}\end{kframe}
 

### 2015-11-17 review_response.Rnw
edits before resubmit

@@
 
-`Both functions treat missing data as NA values, either for observations that exceed the allowable tolerance for the differ argument of setstep or for time series that do not overlap given the method argument passed to comb.'
+`Both functions treat missing data as NA values, either for observations that exceed the allowable tolerance for the differ argument of setstep or for portions of time series that do not overlap given the method argument passed to comb.'
 }
@@
 
-{\it That is a fantastic suggestion and I've made these changes in the development repository of the package.
+{\it That is a fantastic suggestion and I've made these changes in the development repository of the package (see \href{https://github.com/fawda123/SWMPr/commit/3a90bddabbeace917b54942a68449ef9be93ad1d}{commit}).
 I noticed an immediate increase in performance with the import functions.
@@
 I had not noticed the bug so I appreciate the comment.
-I've changed the code in server.R to make plotInput a regular function.
+I've changed the code in server.R to make plotInput a regular function (see \href{https://github.com/fawda123/swmp_summary/commit/907abdd25b543cd5db372a8f30ecd9d7e49611e7}{commit}).
 I've redeployed the app and the bug has been fixed.}
@@
 {\it I fully agree that considering censored data is a necessary aspect of processing and analyzing water quality data.
-Censored data, either above or below the detection limit, are given unique QAQC flags during initial data processing through the Centralized Data Management Office (CDMO).
-Censored values in the imported data can be identified using functions in the SWMPr package.
-The default behavior of the qaqc function is to remove censored values, but the user can also retain these values by changing the function arguments.
-A data frame showing the number of observations for a given QAQC flag can also be retrieved using the qaqcchk function.
-I have revised the text to more clearly describe the presence of censored values in SWMP data and how SWMPr can be used to assess their distribution (page 4, paragraph 3):
+Censored data are given unique QAQC flags during initial data processing through the Centralized Data Management Office (CDMO).
+Censored values in the imported data can be identified using functions in the SWMPr package, although the previous draft did not sufficiently describe this information.
+I believe that the treatment of censored values is beyond the scope of the SWMPr package as there are already methods available in R for working with these observations.
+However, I agree that the issue should be stated more clearly in the manuscript and, more importantly, how SWMPr can be used to identy censored data.
+I have revised the text accordingly (page 4, paragraph 3):
 
@@
 For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.
-The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series (Helsel, 2012).
+The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series (Helsel, 2012).
 A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages (e.g., cents, McLeod et al., 2014).
-However, the functions in SWMPr can be used to identify censored data based on the appropriate QAQC flag for a given parameter.
+However, the functions in SWMPr can be used to identify censored data based on the appropriate QAQC flag for a parameter.
 Viewing this information can be helpful for determining how to further process the data with the qaqc function or alternative methods outside of SWMPr.
@@
 {\it I agree that the use of SLR is a simple method that may not be appropriate for all datasets provided by SWMP.
-The application is meant to be exploratory and I have revised the text to clarify that SLR is a simple and potentially inadequate method for trend analysis, particularly with censored data.
+The application is meant to be exploratory and I have revised the text to clarify that SLR may be an inadequate method for trend analysis, particularly with censored data.
 The text description on the application was also revised.
 
-Page 12, paragraph one: `More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.'
+Page 12, paragraph one: `More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.'
 

### 2015-11-17 swmpr_manu.Rnw
response to first reviewer done, need to proofread, recompile
edits before resubmit

@@
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-However, the increasing quantity of data provided by the monitoring network have complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.
+However, the increasing quantity of data provided by the monitoring network has complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.
 This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.
@@
 For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.
-The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series \citep{Helsel12}.
+The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series \citep{Helsel12}.
 A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.
-However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a given parameter.
+However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a parameter.
 Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.
@@
 As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.
+Both functions treat missing data as \code{NA} values, either for observations that exceed the allowable tolerance for the \code{differ} argument of \code{setstep} or for portions of time series that do not overlap given the \code{method} argument passed to \code{comb}.
 
@@
 
-# combine nut and wq data by intersect
+# combine nut and met data by intersect
 comb(swmp1, swmp3, method = 'intersect')
@@
   'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object.', 
-  'Plot multiple time series on the same y-axis.', 
+  'Plot multiple time series in a \\code{"swmpr"} object on the same y-axis.', 
   'Plot a univariate  time series for a \\code{"swmpr"} object.',
@@
 Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.
-This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.
+This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis (Figure~\ref{fig:overplot}).
 
@@
 
-Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.
+Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package (Figure~\ref{fig:map_ex}, \citet{Kahle13}).
 The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
@@
 Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.
-More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.
+More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
@@
 
-\begin{figure}
+\begin{figure}[!h]
 \begin{center}
@@
 Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions.
-Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.
+Thanks to Jeff Hollister for providing useful comments on an earlier draft.
 

### 2015-11-17 swmpr_manu.tex
edits before resubmit

--- 2015-11-16 swmpr_manu.tex
@@
 \code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
-\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
+\code{overplot}&Plot multiple time series in a \code{"swmpr"} object on the same y-axis.\tabularnewline
 \code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline

### 2015-11-17 Rjournal/beck.tex
journal resubmission files prepped

@@
 This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.
-Simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.
+However, the increasing quantity of data provided by the monitoring network has complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.
 This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.
@@
 ##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
-## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
-## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
-## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
+## 1 2015-11-03 11:15:00   26      0     45        0  29     0     78
+## 2 2015-11-03 11:30:00   26      0     46        0  30     0     76
+## 3 2015-11-03 11:45:00   26      0     46        0  30     0     75
 ##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
-## 1        0      5        0     2       0  8    0   13      0       NA
-## 2        0      5        0     2       0  8    0    9      0       NA
-## 3        0      6        0     2       0  8    0   11      0       NA
+## 1        0      5        0     2       0  8    0    2      0       NA
+## 2        0      5        0     2       0  8    0    5      0       NA
+## 3        0      5        0     2       0  8    0    5      0       NA
 ##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
-## 1         -2    NA      -1     NA     NA       -1         
-## 2         -2    NA      -1     NA     NA       -1         
-## 3         -2    NA      -1     NA     NA       -1
+## 1         -2    NA      -1      2     NA        3         
+## 2         -2    NA      -1      2     NA        3         
+## 3         -2    NA      -1      2     NA        3
 \end{example}
@@
 
-Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \code{qaqc}.
-The \code{qaqcchk} function can be used to view this information.
+SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.
+For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.
+The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series \citep{Helsel12}.
+A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.
+However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a parameter.
+Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.
+The \code{qaqcchk} function returns a \code{data.frame} of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.
+SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.
 
@@
 \end{example}
+
 \begin{figure}[!h]
@@
 Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.
+More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.
 The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.

### 2015-11-17 Rjournal/Beck_code.R

@@ -87,7 +87,7 @@ swmp3 <- apaebmet
 # combine nut and wq data by union
 comb(swmp1, swmp2, method = 'union')
 
-# combine nut and wq data by intersect
+# combine nut and met data by intersect
 comb(swmp1, swmp3, method = 'intersect')
 
 # combine nut, wq, and met data by nut time series, two hour time step

### 2016-01-17 Rjournal/beck.tex
minor edits for acceptance

@@
 Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.
-The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
+The \pkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
 }
@@
 This article describes the \CRANpkg{SWMPr} package that was developed for estuary monitoring data from the System-Wide Monitoring Program.
-Functions provided by \CRANpkg{SWMPr} address many common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify parameters of interest.
-Additionally, web applications derived from \CRANpkg{SWMPr} and \CRANpkg{shiny} illustrate potential applications using the functions in this package.
+Functions provided by \pkg{SWMPr} address many common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify parameters of interest.
+Additionally, web applications derived from \pkg{SWMPr} and \CRANpkg{shiny} illustrate potential applications using the functions in this package.
 The software is provided specifically for use with NERRS data, although many of the applications are relevant for addressing common challenges working with large environmental datasets.
@@
 \begin{tabular}{lp{3.5in}}
-\hline\hline
+\toprule
 \multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
+\midrule
 \code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
@@
 \code{site\_codes\_ind}&Get metadata for all stations at a single site, wrapper to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
-\hline
+\bottomrule
 \end{tabular}\end{center}
@@
 \begin{tabular}{lp{0.75in}p{3.25in}}
-\hline\hline
+\toprule
 \multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
+\midrule
 \code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
@@
 \code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually \code{"POSIXct"} unless data have been aggregated.\tabularnewline
-\hline
+\bottomrule
 \end{tabular}\end{center}
@@
 \begin{tabular}{lp{3.5in}}
-\hline\hline
+\toprule
 \multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
+\midrule
 \code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.\tabularnewline
@@
 \code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function in the base installation.\tabularnewline
-\hline
+\bottomrule
 \end{tabular}\end{center}
@@
 \begin{tabular}{lp{3.5in}}
-\hline\hline
+\toprule
 \multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
-\hline
+\midrule
 \code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
@@
 \code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
-\hline
+\bottomrule
 \end{tabular}\end{center}
@@
 
-An alternative approach for decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \CRANpkg{wq} package \citep{Cloern10,Jassby14}.
+An alternative approach for decomposition is provided by the \code{decomp\_cj} function, which is a simple wrapper to the \code{decompTs} function in the \pkg{wq} package \citep{Cloern10,Jassby14}.
 The \code{decomp\_cj} function is a monthly decomposition for characterizing relatively long-term trends.
@@
 The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.
-An interactive \CRANpkg{shiny} web application \citep{Chang15} that uses this function is available for viewing results for all SWMP sites (see the \nameref{swmp_apps} section).
+An interactive \pkg{shiny} web application \citep{Chang15} that uses this function is available for viewing results for all SWMP sites (see the \nameref{swmp_apps} section).
 
@@
 
-Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package (Figure~\ref{fig:map_ex}, \citet{Kahle13}).
+Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \pkg{ggmap} package (Figure~\ref{fig:map_ex}, \citet{Kahle13}).
 The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.
@@
 
-Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
-The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.
+Two \pkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.
+The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \url{https://beckmw.shinyapps.io/swmp_comp}.
 Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.
@@
 More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.
-The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.
+The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \url{https://beckmw.shinyapps.io/swmp_summary/}.
 The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.
@@
 \includegraphics[width = \textwidth]{swmp_comp.pdf}
-\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
+\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \url{https://beckmw.shinyapps.io/swmp_comp}}
 \label{fig:swmp_comp}
@@
 \address{Marcus W Beck\\
-  ORISE Research Participation Program\\
-  USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division\\
+  US Environmental Protection Agency\\
+  National Health and Environmental Effects Research Laboratory\\
+  Gulf Ecology Division\\
   1 Sabine Island Drive, Gulf Breeze, FL 32651\\
